---
title: "Duke pipeline"
subtitle: "Module 1: Import and QC"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

```{r set_tab_title, echo=FALSE, results='asis'}
cat('<script>document.title = "Duke | Module 1: Import and QC";</script>')
```

# Overview

**Module 1: Import and QC**

This module handles:

1. File discovery and import (FASTQ, FASTA, BAM)
2. Duplicate read name handling
3. Optional downsampling
4. Quality control metrics
5. Adapter trimming

**Outputs:**

- Trimmed sequences ready for alignment
- QC metrics and visualisations
- Adapter trimming statistics


```{r load_packages, message=FALSE, warning=FALSE, echo=FALSE}
# Required packages
required_packages <- c("ShortRead", "Biostrings", "Rsamtools", "pbapply", 
                      "dplyr", "tidyr", "data.table", "stringr", "ggplot2",
                      "scales", "DT", "openxlsx", "parallel",
                      "pbmcapply", "tibble")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r create_directories, echo=FALSE}
# Create output directories
temp_dir <- file.path(params$dir_out, "temp")
qc_dir    <- file.path(params$dir_out, "01_import_qc")
plots_dir <- file.path(params$dir_out, file.path("01_import_qc", "plots"))  # NEW: plots directory

if (!dir.exists(temp_dir)) dir.create(temp_dir, recursive = TRUE)
if (!dir.exists(qc_dir)) dir.create(qc_dir, recursive = TRUE)
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)  # NEW
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}
```

```{r timing_init, echo=FALSE}
# Initialise section timing
module_start_time <- proc.time()["elapsed"]
timings <- c()
```

---

# Sequencing data

Discover and import sequencing files from the data directory.

```{r file_paths, echo=FALSE}
# List files
file_paths <- list.files(params$dir_data,
                         full.names = TRUE,
                         pattern = paste(params$import_patterns, collapse = "|"),
                         recursive = params$import_recursive)

# Extract file names
file_names <- basename(file_paths)

# Extract file stems, first removing gz double extensions
file_stems <- gsub("\\.gz$", "", file_names)
file_stems <- tools::file_path_sans_ext(file_stems)

# Extract formats
file_formats <- ifelse(grepl("\\.gz$", file_names), 
                       paste0(tools::file_ext(gsub("\\.gz$", "", file_names)), ".gz"), 
                       tools::file_ext(file_names))

# Vector of file names and stems
file_name_vector <- setNames(file_names, file_names)
file_stem_vector <- setNames(file_stems, file_stems)

# Table of file details
file_details <- data.frame(file_name = file_names,
                           file_stem = file_stems,
                           file_format = file_formats,
                           file_path = file_paths) %>%
  dplyr::mutate(paired = replace_na(grepl(params$r1_pattern, file_name) | grepl(params$r2_pattern, file_name), FALSE),
                paired_read_direction = case_when(grepl(params$r1_pattern, file_name) ~ "R1",
                                                  grepl(params$r2_pattern, file_name) ~ "R2",
                                                  TRUE ~ NA),
                include = case_when(paired & paired_read_direction == params$select_one_of_pair ~ TRUE,
                                    paired ~ FALSE,
                                    TRUE ~ TRUE),
                sample_stem = ifelse(paired, gsub("_R[12]_.*", "", file_name), file_stem)) %>%
  dplyr::relocate(c(file_format, paired, paired_read_direction, include, sample_stem), 
                  .after = "file_name")
```

## File format summary

```{r file_format_table, echo=FALSE}
data.frame(table(file_formats)) %>%
  dplyr::rename(file_format = file_formats,
                count = Freq) %>%
  DT::datatable(
    caption = "File format",
    options = list(pageLength = 10, scrollX = TRUE, dom = 't'),
    rownames = FALSE,
    class = 'cell-border stripe compact'
  )
```

## Paired-end detection

```{r paired_end_summary, echo=FALSE}
file_details %>%
  distinct(sample_stem, paired) %>%
  group_by(paired) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%
  DT::datatable(
    caption = "Count single and paired end sequencing files",
    options = list(pageLength = 10, scrollX = TRUE, dom = 't'),
    rownames = FALSE,
    class = 'cell-border stripe compact'
  )
```

## File details

```{r file_details_table, echo=FALSE}
DT::datatable(
  file_details,
  caption = "File summary",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(file_details), fontSize = '85%')


# Export table
write_excel_function(named_list = list("file_details" = file_details),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

```{r select_one_of_pair, echo=FALSE}
if (!is.na(params$select_one_of_pair)) {
  # Files to keep
  file_stem_keep <- file_details %>%
    dplyr::filter(include)
  file_stem_keep <- file_stem_keep$file_stem
  
  # Filter file stem vector
  file_stem_vector <- file_stem_vector[file_stem_vector %in% file_stem_keep]
}
```

```{r t_import_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r sequencing_import, echo=FALSE}
# Import sequencing files
seq <- apply_fn(file_stem_vector, function(file_stem, file_details) {
  file_path <- file_details$file_path[match(file_stem, file_details$file_stem)]
  file_format <- file_details$file_format[match(file_stem, file_details$file_stem)]
  import_sequencing(file_path, file_format)
}, file_details = file_details)

# Extract results if using parallel processing
seq <- extract_apply_fn_result(seq)
```

```{r t_import_end, echo=FALSE}
timings["Import"] <- proc.time()["elapsed"] - .t
```

---

# Check for duplicate read names

Some sequencing technologies may contain duplicate read names. This section identifies duplicates and retains only the longest instance of each duplicated read.

```{r t_dedup_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r select_unique_readnames, echo=FALSE, message=FALSE, results='hide'}
if (params$check_duplicate_readnames) {
  # Find duplicate read names and select the longest
  my_path <- file.path(temp_dir, "01_unique_readname_result.RData")
  if (params$resume && safe_load_cache(my_path, environment())) {
    message("Loaded duplicate readname results from cache")
  } else {
    unique_readname_result <- apply_fn(seq, select_longest_duplicate_readname)
    unique_readname_result <- extract_apply_fn_result(unique_readname_result)
    save(unique_readname_result, file = my_path)
  }
  seq_unique <- lapply(unique_readname_result, "[[", "dna_stringset")
} else {
  message("⊗ Duplicate read name check skipped (check_duplicate_readnames = FALSE)")
  seq_unique <- seq
}
```

```{r t_dedup_end, echo=FALSE}
timings["Duplicate names"] <- proc.time()["elapsed"] - .t
```

```{r count_read_names, echo=FALSE, eval=params$check_duplicate_readnames}
# Read name counts
read_name_count <- lapply(file_stem_vector, function(file_stem, unique_readname_result) {
  unique_readname_result[[file_stem]][["read_name_count"]] %>%
    dplyr::mutate(file_stem = file_stem) %>%
    relocate(file_stem)
}, unique_readname_result = unique_readname_result)
read_name_count <- rbindlist(read_name_count)

# Summarise read name count
read_name_count_summary <- read_name_count %>%
  group_by(file_stem, count) %>%
  dplyr::summarise(n_readnames = n(), .groups = "drop")

# Total number of read names
total_readnames <- read_name_count %>%
  group_by(file_stem) %>%
  dplyr::summarise(total_readnames = n(), .groups = "drop")

# Add total read name count
read_name_count_summary <- read_name_count_summary %>%
  left_join(total_readnames, by = join_by(file_stem)) %>%
  dplyr::mutate(proportion = n_readnames / total_readnames)

# Flag: are any duplicates present across all samples?
has_any_duplicates <- any(read_name_count_summary$count > 1)
```

## Duplicate read names

```{r plot_read_name_count, echo=FALSE, results='asis', eval=params$check_duplicate_readnames}
n_samp       <- length(unique(read_name_count_summary$file_stem))
total_reads  <- scales::comma(sum(total_readnames$total_readnames))

if (!has_any_duplicates) {

  cat(sprintf(
    "\n\n✓ **No duplicate read names detected** across %d sample%s (%s total reads).\n\n",
    n_samp, ifelse(n_samp == 1, "", "s"), total_reads
  ))

} else {

  # Per-sample duplication rate
  dup_rates <- read_name_count_summary %>%
    dplyr::filter(count > 1) %>%
    group_by(file_stem) %>%
    dplyr::summarise(pct_duplicated = sum(proportion) * 100, .groups = "drop")

  all_samples <- data.frame(file_stem = unique(read_name_count_summary$file_stem))
  dup_rates <- all_samples %>%
    left_join(dup_rates, by = "file_stem") %>%
    dplyr::mutate(pct_duplicated = replace_na(pct_duplicated, 0))

  mean_dup <- mean(dup_rates$pct_duplicated)
  sd_dup   <- if (nrow(dup_rates) > 1) sd(dup_rates$pct_duplicated) else 0
  dup_rates <- dup_rates %>%
    dplyr::mutate(outlier = sd_dup > 0 & abs(pct_duplicated - mean_dup) > 2 * sd_dup)

  n_bins <- max(5, min(30, ceiling(sqrt(nrow(dup_rates)))))

  p_duplicates <- ggplot(dup_rates, aes(x = pct_duplicated)) +
    geom_histogram(bins = n_bins, fill = "#619CFF", colour = "white", alpha = 0.85) +
    geom_rug(aes(colour = outlier), sides = "b", linewidth = 0.8, alpha = 0.8) +
    scale_colour_manual(values = c("FALSE" = "#619CFF", "TRUE" = "red"), guide = "none") +
    scale_x_continuous(labels = function(x) paste0(x, "%")) +
    labs(title = "Duplicate read name rate across samples",
         subtitle = sprintf("%d sample%s  |  mean %.2f%%  |  range %.2f–%.2f%%",
                            n_samp, ifelse(n_samp == 1, "", "s"),
                            mean_dup, min(dup_rates$pct_duplicated),
                            max(dup_rates$pct_duplicated)),
         x = "% reads with duplicate names per sample",
         y = "Number of samples") +
    theme_bw()

  print(p_duplicates)

  # List outlier samples
  outlier_df <- dplyr::filter(dup_rates, outlier)
  if (nrow(outlier_df) > 0) {
    cat("\n\n**Outlier samples (>2 SD from mean duplication rate):**\n\n")
    for (i in seq_len(nrow(outlier_df))) {
      cat(sprintf("- %s: %.2f%% duplicated\n",
                  outlier_df$file_stem[i], outlier_df$pct_duplicated[i]))
    }
    cat("\n\n")
  }

  plot_path <- file.path(plots_dir, "read_name_duplicates.png")
  if (!file.exists(plot_path) || !params$resume) {
    ggsave(filename = plot_path, plot = p_duplicates,
           width = 10, height = 6, dpi = params$plot_dpi, bg = "white")
  }
}
```

---

```{r count_original_vs_unique_readnames, echo=FALSE}
# Compare read count (only meaningful when duplicate check ran)
if (params$check_duplicate_readnames) {
  readcount_unique_readnames <- 
    compare_read_count(dna_stringset1 = seq,
                       dna_stringset2 = seq_unique,
                       name1 = "original",
                       name2 = "unique_read_name")
}

# Remove previous sequencing file (always, regardless of duplicate check)
if (params$remove_intermediate) rm(seq)
```

## Read count comparison

```{r plot_original_vs_unique_readnames, echo=FALSE, results='asis', eval=params$check_duplicate_readnames}
n_samp <- nrow(readcount_unique_readnames)

# Build long format with numeric stage position (needed for consistent jitter)
comparison_long <- readcount_unique_readnames %>%
  pivot_longer(!file_stem, names_to = "stage", values_to = "read_count") %>%
  dplyr::mutate(
    stage     = factor(stage,
                       levels = c("original", "unique_read_name"),
                       labels = c("Before\ndeduplication", "After\ndeduplication")),
    stage_num = as.numeric(stage)
  )

# Pre-compute a single consistent horizontal jitter offset per sample
# so the connecting line between Before and After is perfectly vertical
set.seed(123)
jitter_df <- data.frame(
  file_stem = unique(comparison_long$file_stem),
  x_offset  = runif(n_samp, -0.12, 0.12)
)
comparison_long <- comparison_long %>%
  left_join(jitter_df, by = "file_stem") %>%
  dplyr::mutate(x_pos = stage_num + x_offset)

# Flag samples where reads were actually removed
change_lookup <- readcount_unique_readnames %>%
  dplyr::mutate(changed = unique_read_name < original) %>%
  dplyr::select(file_stem, changed)
comparison_long <- comparison_long %>%
  left_join(change_lookup, by = "file_stem")

# Scale point/line aesthetics to N
point_size  <- if (n_samp <= 50) 2.5 else 1.5
point_alpha <- if (n_samp <= 50) 0.8  else max(0.2, 1 - n_samp / 500)
line_alpha  <- if (n_samp <= 50) 0.5  else max(0.1, 0.5 - n_samp / 1000)

# Subtitle
if (has_any_duplicates) {
  reads_lost  <- sum(readcount_unique_readnames$original - readcount_unique_readnames$unique_read_name)
  pct_lost    <- reads_lost / sum(readcount_unique_readnames$original) * 100
  plot_subtitle <- sprintf("%d sample%s  |  %s reads removed (%.2f%% of total)",
                           n_samp, ifelse(n_samp == 1, "", "s"),
                           scales::comma(reads_lost), pct_lost)
} else {
  plot_subtitle <- sprintf("%d sample%s  |  no reads removed",
                           n_samp, ifelse(n_samp == 1, "", "s"))
}

stage_labels <- levels(comparison_long$stage)

p_dedup <- ggplot(comparison_long) +
  # Boxplot summarises the distribution at each stage
  geom_boxplot(aes(x = stage_num, y = read_count, group = stage_num),
               outlier.shape = NA, width = 0.3,
               fill = "grey92", colour = "grey40", linewidth = 0.6) +
  # Lines connect the same sample across stages — vertical when jitter is consistent
  geom_line(aes(x = x_pos, y = read_count, group = file_stem, colour = changed),
            linewidth = 0.4, alpha = line_alpha) +
  # Points overlaid on box
  geom_point(aes(x = x_pos, y = read_count, colour = changed),
             size = point_size, alpha = point_alpha) +
  # Mean diamond
  stat_summary(aes(x = stage_num, y = read_count, group = stage_num),
               fun = mean, geom = "point", shape = 18,
               colour = "red", size = 4, alpha = 0.6) +
  scale_colour_manual(
    values = c("FALSE" = "grey50", "TRUE" = "red"),
    labels = c("FALSE" = "Unchanged",  "TRUE" = "Reads removed"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 1:2, labels = stage_labels, limits = c(0.6, 2.4)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title    = "Read counts before and after duplicate read name removal",
       subtitle = plot_subtitle,
       x = NULL,
       y = "Read count per sample") +
  theme_bw()

# Suppress legend when all samples are unchanged (would only show one entry)
if (!has_any_duplicates) p_dedup <- p_dedup + guides(colour = "none")

print(p_dedup)

# Summary message / outlier list
if (!has_any_duplicates) {
  cat(sprintf(
    "\n\n✓ **No reads removed**: all read names unique across all %d sample%s.\n\n",
    n_samp, ifelse(n_samp == 1, "", "s")
  ))
} else {
  removed_df <- readcount_unique_readnames %>%
    dplyr::mutate(
      reads_removed = original - unique_read_name,
      pct_removed   = reads_removed / original * 100
    )
  mean_rem <- mean(removed_df$pct_removed)
  sd_rem   <- if (nrow(removed_df) > 1) sd(removed_df$pct_removed) else 0
  removed_df <- removed_df %>%
    dplyr::mutate(outlier = sd_rem > 0 & abs(pct_removed - mean_rem) > 2 * sd_rem)

  outlier_df <- dplyr::filter(removed_df, outlier)
  if (nrow(outlier_df) > 0) {
    cat("\n\n**Outlier samples (>2 SD from mean removal rate):**\n\n")
    for (i in seq_len(nrow(outlier_df))) {
      cat(sprintf("- %s: %.2f%% removed (%s reads)\n",
                  outlier_df$file_stem[i], outlier_df$pct_removed[i],
                  scales::comma(outlier_df$reads_removed[i])))
    }
    cat("\n\n")
  }
}

plot_path <- file.path(plots_dir, "unique_read_names.png")
if (!file.exists(plot_path) || !params$resume) {
  ggsave(filename = plot_path, plot = p_dedup,
         width = 8, height = 6, dpi = params$plot_dpi, bg = "white")
}
```

```{r table_original_vs_unique_readnames, echo=FALSE, eval=params$check_duplicate_readnames}
# Display table
DT::datatable(
  readcount_unique_readnames,
  caption = "Count unique read names",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(readcount_unique_readnames), fontSize = '85%')

# Export table
write_excel_function(named_list = list("readcount_unique_readnames" = readcount_unique_readnames),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

---

# Sample reads

Optionally downsample reads for faster processing and visualisation.

```{r t_downsample_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r downsample, echo=FALSE}
# Downsample
if (!is.null(params$downsample) && !is.na(params$downsample)) {
  seq_sample <- apply_fn(seq_unique, 
                         downsample, 
                         sample_size = params$downsample)
  seq_sample <- extract_apply_fn_result(seq_sample)
} else {
  seq_sample <- seq_unique
}
```

```{r t_downsample_end, echo=FALSE}
timings["Downsampling"] <- proc.time()["elapsed"] - .t
```

```{r downsample_read_counts, echo=FALSE}
# Compare read count
read_count_sample <- 
  compare_read_count(dna_stringset1 = seq_unique,
                     dna_stringset2 = seq_sample,
                     name1 = "original",
                     name2 = "sample")
```

## Downsampling reads

```{r plot_downsample_read_counts, echo=FALSE, results='asis'}
n_samp <- nrow(read_count_sample)

# Build long format with numeric stage position
counts_long <- read_count_sample %>%
  pivot_longer(!file_stem, names_to = "stage", values_to = "read_count") %>%
  dplyr::mutate(
    stage     = factor(stage,
                       levels = c("original", "sample"),
                       labels = c("Before\ndownsampling", "After\ndownsampling")),
    stage_num = as.numeric(stage)
  )

# Pre-compute a single consistent horizontal jitter offset per sample
set.seed(123)
jitter_df <- data.frame(
  file_stem = unique(counts_long$file_stem),
  x_offset  = runif(n_samp, -0.12, 0.12)
)
counts_long <- counts_long %>%
  left_join(jitter_df, by = "file_stem") %>%
  dplyr::mutate(x_pos = stage_num + x_offset)

# Flag samples where reads were actually removed by downsampling
change_lookup <- read_count_sample %>%
  dplyr::mutate(changed = sample < original) %>%
  dplyr::select(file_stem, changed)
counts_long <- counts_long %>%
  left_join(change_lookup, by = "file_stem")

# Downsampling info
downsample_applied <- !is.null(params$downsample) && !is.na(params$downsample)
n_downsampled      <- sum(change_lookup$changed)

# Summary stats (on original counts)
mean_count <- mean(read_count_sample$original)
sd_count   <- if (n_samp > 1) sd(read_count_sample$original) else 0

subtitle <- sprintf(
  "%d sample%s  |  mean %s  |  SD %s  |  range %s \u2013 %s",
  n_samp, ifelse(n_samp == 1, "", "s"),
  scales::comma(round(mean_count)),
  scales::comma(round(sd_count)),
  scales::comma(min(read_count_sample$original)),
  scales::comma(max(read_count_sample$original))
)
if (downsample_applied) {
  subtitle <- paste0(
    subtitle,
    sprintf("\nDownsample threshold: %s  \u2014  %d sample%s affected",
            scales::comma(params$downsample), n_downsampled,
            ifelse(n_downsampled == 1, "", "s"))
  )
} else {
  subtitle <- paste0(subtitle, "\nNo downsampling applied")
}

# Scale point/line aesthetics to N
point_size  <- if (n_samp <= 50) 2.5 else 1.5
point_alpha <- if (n_samp <= 50) 0.8  else max(0.2, 1 - n_samp / 500)
line_alpha  <- if (n_samp <= 50) 0.5  else max(0.1, 0.5 - n_samp / 1000)

stage_labels <- levels(counts_long$stage)

p_filtered_counts <- ggplot(counts_long) +
  geom_boxplot(aes(x = stage_num, y = read_count, group = stage_num),
               outlier.shape = NA, width = 0.3,
               fill = "grey92", colour = "grey40", linewidth = 0.6) +
  geom_line(aes(x = x_pos, y = read_count, group = file_stem, colour = changed),
            linewidth = 0.4, alpha = line_alpha) +
  geom_point(aes(x = x_pos, y = read_count, colour = changed),
             size = point_size, alpha = point_alpha) +
  stat_summary(aes(x = stage_num, y = read_count, group = stage_num),
               fun = mean, geom = "point", shape = 18,
               colour = "red", size = 4, alpha = 0.6) +
  scale_colour_manual(
    values = c("FALSE" = "grey50", "TRUE" = "red"),
    labels = c("FALSE" = "Unchanged", "TRUE" = "Downsampled"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 1:2, labels = stage_labels, limits = c(0.6, 2.4)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title    = "Read counts before and after downsampling",
       subtitle = subtitle,
       x = NULL,
       y = "Read count per sample") +
  theme_bw()

# Horizontal downsample threshold line
if (downsample_applied) {
  p_filtered_counts <- p_filtered_counts +
    geom_hline(yintercept = params$downsample, colour = "red", alpha = 0.7,
               linetype = "dashed", linewidth = 0.8) +
    annotate("text", x = 2.35, y = params$downsample,
             label = sprintf("Threshold\n(%s)", scales::comma(params$downsample)),
             vjust = -0.2, hjust = 1, colour = "red", size = 3, alpha = 0.7)
}

# Suppress legend when nothing was downsampled
if (n_downsampled == 0) p_filtered_counts <- p_filtered_counts + guides(colour = "none")

print(p_filtered_counts)

# List outlier samples (those downsampled) below plot
if (n_downsampled > 0) {
  downsampled_df <- change_lookup %>% dplyr::filter(changed) %>%
    left_join(read_count_sample, by = "file_stem")
  cat("\n\n**Samples affected by downsampling:**\n\n")
  for (i in seq_len(nrow(downsampled_df))) {
    cat(sprintf("- %s: %s \u2192 %s reads\n",
                downsampled_df$file_stem[i],
                scales::comma(downsampled_df$original[i]),
                scales::comma(downsampled_df$sample[i])))
  }
  cat("\n\n")
}

plot_path <- file.path(plots_dir, "read_counts_filtered.png")
if (!file.exists(plot_path) || !params$resume) {
  ggsave(filename = plot_path, plot = p_filtered_counts,
         width = 8, height = 6, dpi = params$plot_dpi, bg = "white")
}
```

```{r table_downsample_read_counts, echo=FALSE}
# Display table
DT::datatable(
  read_count_sample,
  caption = "Sampled read count",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(read_count_sample), fontSize = '85%')

# Export table
write_excel_function(named_list = list("read_count_sample" = read_count_sample),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)

# Remove previous sequencing file
if (params$remove_intermediate) rm(seq_unique)
```

---

# Quality

Calculate read quality metrics and length distributions.

```{r t_qc_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r qc, echo=FALSE}
# Quality and length of each read
sequencing_qc <- apply_fn(file_stem_vector, function(my_file_stem, seq_sample) {
  
  # Extract sequencing file
  sequencing_file <- seq_sample[[my_file_stem]]
  
  # Check if quality scores exist
  if(inherits(sequencing_file, "QualityScaledDNAStringSet")) {
  
    # Extract qualities
    quals <- quality(sequencing_file)
    
    # Sum -log10 pvalues for each read
    read_scores <- alphabetScore(quals)
    
  } else {
    
    # NA if no quality scores
    read_scores <- rep(NA, length(sequencing_file))
    
  }
  
  # Read lengths
  read_lengths <- width(sequencing_file)
  
  # Table
  qc_table <- data.frame(file_stem = rep(my_file_stem, length(sequencing_file)),
                         read_name = names(sequencing_file),
                         phred_sum = read_scores,
                         length = read_lengths) %>%
    dplyr::mutate(read_phred = phred_sum / length)
  
  # Output
  return(qc_table)
  
}, seq_sample = seq_sample)

# Extract results if using parallel processing
sequencing_qc <- extract_apply_fn_result(sequencing_qc)
  
# Summarise QC
sequencing_qc_summary <- 
  rbindlist(sequencing_qc) %>%
    group_by(file_stem) %>%
    dplyr::summarise(read_count = n(),
                     mean_length = mean(length, na.rm = TRUE),
                     median_length = median(length, na.rm = TRUE),
                     sd_length = sd(length, na.rm = TRUE),
                     mean_phred = mean(read_phred, na.rm = TRUE),
                     median_phred = median(read_phred, na.rm = TRUE),
                     sd_phred = sd(read_phred, na.rm = TRUE),
                     .groups = "drop")
```

```{r t_qc_end, echo=FALSE}
timings["QC metrics"] <- proc.time()["elapsed"] - .t
```

## QC metrics

```{r qc_plot, echo=FALSE, message=FALSE}
# Prepare data
filtered_read_summary <- sequencing_qc_summary %>%
  dplyr::mutate(mean_phred_plot = ifelse(is.na(mean_phred), 0, mean_phred),
                phred_available = ifelse(!is.na(mean_phred), "Phred available", "Phred unavailable"))

# Combined plot
p_qc_scatter <- filtered_read_summary %>%
  ggplot(aes(x = mean_length, y = mean_phred_plot)) +
  facet_wrap(vars(phred_available), scales = "free") +
  geom_point(aes(size = read_count), 
             fill = "skyblue", colour = "black", shape = 21, alpha = 0.75) +
  geom_smooth() +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(breaks = pretty_breaks()) +
  scale_fill_viridis_c(option = "D", direction = 1) +
  labs(x = "Mean read length (bp)",
       y = "Mean phred score",
       size = "Read count") +
  theme_minimal()

print(p_qc_scatter)

# Export combined plot
ggsave(
  filename = file.path(plots_dir, "quality_vs_length.png"),
  plot = p_qc_scatter,
  width = 12,
  height = 6,
  dpi = params$plot_dpi,
  bg = "white"
)
```

```{r qc_table, echo=FALSE}
# Display table
DT::datatable(
  sequencing_qc_summary,
  caption = "Sequencing QC",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(sequencing_qc_summary), fontSize = '85%')

# Export table
write_excel_function(named_list = list("sequencing_qc_summary" = sequencing_qc_summary),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

---

# Trim

Trim adapter sequences from reads.

```{r trim_function, echo=FALSE}
trim_reads <- function(dna_stringset, trim_patterns, max_mismatch, with_indels) {
  
  # Vector of trim sequence names
  trim_pattern_names <- setNames(names(trim_patterns), 
                                 names(trim_patterns))
  
  # Log locations of trim sequences
  trim_locations <- lapply(trim_pattern_names, function(trim_pattern_name,
                                                        trim_patterns,
                                                        dna_stringset,
                                                        max_mismatch,
                                                        with_indels) {
    # Extract trim sequence
    trim_pattern <- trim_patterns[[trim_pattern_name]]
    
    # Trim coordinates
    trim_location <- trimLRPatterns(Lpattern = trim_pattern,
                                    Rpattern = trim_pattern,
                                    subject = dna_stringset,
                                    max.Lmismatch = max_mismatch,
                                    max.Rmismatch = max_mismatch,
                                    with.Lindels = with_indels,
                                    with.Rindels = with_indels,
                                    ranges = TRUE)
    
    # Convert to dataframe
    trim_location <- as.data.frame(trim_location) %>%
      dplyr::mutate(trim_pattern_name = trim_pattern_name,
                    read_name = names(dna_stringset),
                    read_length = width(dna_stringset)) %>%
      relocate(c(trim_pattern_name, read_name, read_length))
    
    # Output
    return(trim_location)
    
  },
  trim_patterns = trim_patterns,
  dna_stringset = dna_stringset,
  max_mismatch = max_mismatch,
  with_indels = with_indels)
  
  # Rbind
  trim_locations <- rbindlist(trim_locations) %>%
    dplyr::mutate(trim_length = read_length - width,
                  trim = width < read_length)
  
  # Trim
  dna_stringset_trimmed <- dna_stringset
  for (trim_pattern in trim_patterns) {
    dna_stringset_trimmed <- trimLRPatterns(Lpattern = trim_pattern,
                                            Rpattern = trim_pattern,
                                            subject = dna_stringset_trimmed,
                                            max.Lmismatch = max_mismatch,
                                            max.Rmismatch = max_mismatch,
                                            with.Lindels = with_indels,
                                            with.Rindels = with_indels)
  }
  
  # Output
  return(list(trim_locations = trim_locations,
              dna_stringset_trimmed = dna_stringset_trimmed))
}
```

```{r trim_sequences, echo=FALSE}
# Import trim sequences (only if trimming enabled)
if (params$trim && !is.null(params$path_trim_patterns) && !is.na(params$path_trim_patterns)) {
  
  message("Loading adapter sequences for trimming...")
  
  trim_sequences <- read.csv(params$path_trim_patterns, fileEncoding = "UTF-8-BOM")
  trim_sequences <- setNames(trim_sequences$adapter_sequence,
                             trim_sequences$adapter_name)
  
  # Reverse complement the trim sequences
  trim_sequences_rc <- as.character(reverseComplement(DNAStringSet(trim_sequences)))
  names(trim_sequences_rc) <- paste0(names(trim_sequences_rc), "_rc")
  
  # Combine sense and reverse complement trim sequences
  combined_trim_sequences <- c(trim_sequences, trim_sequences_rc)
  
  # Display table
  trim_df <- enframe(combined_trim_sequences, name = "adapter_name", value = "adapter_sequence")
  DT::datatable(
    trim_df,
    caption = "Trim sequences",
    options = list(pageLength = 25, scrollX = TRUE, scrollY = "400px"),
    rownames = FALSE,
    filter = 'top',
    class = 'cell-border stripe compact'
  )
  
  # Export table
  write_excel_function(named_list = list("trim_sequences" = trim_df),
                       path = file.path(qc_dir, "qc.xlsx"),
                       overwrite_sheet = TRUE)
  
  cat("\n✓ Adapter trimming enabled:", length(trim_sequences), "adapters loaded\n")
  
} else {
  
  # Trimming disabled - create empty adapter list
  combined_trim_sequences <- character(0)
  
  if (!params$trim) {
    cat("\n⊗ Adapter trimming disabled (trim = FALSE)\n")
  } else {
    cat("\n⊗ Adapter trimming disabled (no path_trim_patterns provided)\n")
  }
  cat("  Sequences will not be trimmed for adapters\n")
  
}
```

```{r t_trim_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r trim, echo=FALSE, message=FALSE, results='hide'}
# Run trim function (only if adapters provided)
my_path <- file.path(temp_dir, "02_trim_result.RData")

if (length(combined_trim_sequences) > 0) {
  # Trimming enabled - process adapters
  
  if (params$resume && safe_load_cache(my_path, environment())) {
    message("Loading cached trim results...")
  } else {
    message("Trimming adapters from sequences...")
    
    trim_result <- apply_fn(seq_sample, trim_reads,
                            trim_patterns = combined_trim_sequences,
                            max_mismatch = params$trim_max_mismatch,
                            with_indels = params$trim_with_indels)
    trim_result <- extract_apply_fn_result(trim_result)
    save(trim_result, file = my_path)
  }
  
  # Extract trimmed sequencing files
  seq_trim <- lapply(trim_result, "[[", "dna_stringset_trimmed")
  
  cat("\nAdapter trimming complete\n")
  
} else {
  # Trimming disabled - pass through sequences unchanged
  message("Skipping adapter trimming (disabled or no adapters)")
  
  seq_trim <- seq_sample
  
  cat("\n⊗ No adapter trimming performed\n")
}

# Remove previous sequencing file
if (params$remove_intermediate) rm(seq_sample)
```

```{r t_trim_end, echo=FALSE}
timings["Adapter trimming"] <- proc.time()["elapsed"] - .t
```

## Proportion of reads trimmed

```{r trim_proportion, echo=FALSE}
if (length(combined_trim_sequences) > 0) {
# Extract trim locations
trim_locations <- lapply(file_stem_vector, function(my_file_stem, trim_result) {
  trim_result[[my_file_stem]]$trim_locations %>%
    dplyr::mutate(file_stem = my_file_stem) %>%
    relocate(file_stem)
}, trim_result = trim_result)

# Rbind
trim_locations <- rbindlist(trim_locations)

# Count trimmed reads by trim sequence for each file
count_trimmed <- trim_locations %>%
  group_by(file_stem, trim_pattern_name) %>%
  dplyr::summarise(read_count = sum(trim), .groups = "drop")

# Count untrimmed reads
count_untrimmed <- trim_locations %>%
  group_by(file_stem, read_name) %>%
  dplyr::summarise(all_untrimmed = all(!trim), .groups = "drop") %>%
  group_by(file_stem) %>%
  dplyr::summarise(read_count = sum(all_untrimmed), .groups = "drop")

# Merge trimmed and untrimmed read counts
count_trimmed <- plyr::rbind.fill(count_trimmed, count_untrimmed) %>%
  arrange(file_stem, trim_pattern_name) %>%
  dplyr::filter(read_count > 0) %>%
  dplyr::mutate(trim_pattern_name = ifelse(is.na(trim_pattern_name), "untrimmed", trim_pattern_name))

# Count total reads for each file
total_reads <- trim_locations %>%
  distinct(file_stem, read_name) %>%
  group_by(file_stem) %>%
  dplyr::summarise(total_reads = n(), .groups = "drop")

# Calculate proportion of reads trimmed
count_trimmed <- count_trimmed %>%
  left_join(total_reads, by = join_by(file_stem)) %>%
  dplyr::mutate(proportion = read_count / total_reads)

# Calculate mean proportion for each trim pattern to reorder the x-axis
mean_trim <- count_trimmed %>%
  group_by(trim_pattern_name) %>%
  dplyr::summarise(mean_proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(desc(mean_proportion))

# Reorder trim_pattern_name by mean count
count_trimmed <- count_trimmed %>%
  dplyr::mutate(trim_pattern_name = factor(trim_pattern_name, levels = mean_trim$trim_pattern_name))
} # end if (length(combined_trim_sequences) > 0)
```

```{r trim_proportion_plot, echo=FALSE}
if (length(combined_trim_sequences) > 0) {

p_trim_prop <- ggplot(count_trimmed, aes(x = trim_pattern_name, y = proportion)) +
  geom_boxplot(outlier.shape = NA, width = 0.4,
               fill = "grey92", colour = "grey40") +
  geom_jitter(height = 0, width = 0.1, alpha = 0.6, size = 2.5, colour = "grey40") +
  stat_summary(fun = mean, geom = "point", shape = 18,
               colour = "red", size = 4, alpha = 0.6) +
  stat_summary(fun = mean, geom = "text",
               aes(label = scales::percent(after_stat(y), accuracy = 0.1)),
               colour = "red", size = 3.5, vjust = -1.2, alpha = 0.7) +
  scale_x_discrete(guide = guide_axis(angle = 75)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     breaks = scales::pretty_breaks(n = 6),
                     expand = expansion(mult = c(0.1, 0.12))) +
  labs(x = "Trim sequence",
       y = "Proportion of reads trimmed") +
  theme_bw()

print(p_trim_prop)

# Export plot
plot_path <- file.path(plots_dir, "adapter_trimming_proportion.png")
if (!file.exists(plot_path) || !params$resume) {
  ggsave(
    filename = plot_path,
    plot = p_trim_prop,
    width = 10,
    height = 6,
    dpi = params$plot_dpi,
    bg = "white"
  )
}
} # end if (length(combined_trim_sequences) > 0)
```

## Trimming quality summary

Trimming occurs from sequence **ends** using `Biostrings::trimLRPatterns`:

- **left** = 5' end (start of read)
- **right** = 3' end (end of read)  
- **both** = adapters found at both ends

Compare trim length statistics to `adapter_length`:

- **Similar to adapter length** = good match (adapter fully present)
- **Shorter than adapter** = partial adapter or early termination
- **Longer than adapter** = adapter + extra sequence removed

### Summary by adapter

```{r trim_summary_by_adapter, echo=FALSE}
if (length(combined_trim_sequences) > 0) {
# Comprehensive summary by adapter
trim_by_adapter <- trim_locations %>%
  filter(trim == TRUE) %>%
  mutate(
    trim_side = case_when(
      start == 1 & end < read_length ~ "left",
      start > 1 & end == read_length ~ "right",
      start == 1 & end == read_length ~ "both",
      TRUE ~ "internal"
    )
  ) %>%
  group_by(trim_pattern_name, trim_side) %>%
  summarise(
    n_trimmed = n(),
    median_trim_length = median(trim_length, na.rm = TRUE),
    mean_trim_length = mean(trim_length, na.rm = TRUE),
    sd_trim_length = sd(trim_length, na.rm = TRUE),
    min_trim_length = min(trim_length, na.rm = TRUE),
    max_trim_length = max(trim_length, na.rm = TRUE),
    .groups = "drop"
  )

# Add adapter info and calculate proportion
total_reads <- length(unique(trim_locations$read_name))

trim_by_adapter <- trim_by_adapter %>%
  left_join(
    data.frame(
      trim_pattern_name = names(combined_trim_sequences),
      adapter_sequence = as.character(combined_trim_sequences),
      adapter_length = nchar(combined_trim_sequences)
    ),
    by = "trim_pattern_name"
  ) %>%
  mutate(
    proportion = n_trimmed / total_reads,
    total_reads = total_reads
  ) %>%
  relocate(adapter_sequence, adapter_length, trim_side, .after = trim_pattern_name) %>%
  relocate(n_trimmed, total_reads, proportion, .after = trim_side) %>%
  arrange(desc(n_trimmed))

# Display table
DT::datatable(
  trim_by_adapter,
  caption = "Trimming summary by adapter (across all samples)",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact'
) %>%
  DT::formatStyle(columns = colnames(trim_by_adapter), fontSize = '85%') %>%
  DT::formatRound(columns = c('median_trim_length', 'mean_trim_length', 'sd_trim_length', 'proportion'), digits = 2)
} # end if (length(combined_trim_sequences) > 0)
```

### Summary by sample

```{r trim_summary_by_sample, echo=FALSE}
if (length(combined_trim_sequences) > 0) {
# Comprehensive summary by sample
trim_by_sample <- trim_locations %>%
  filter(trim == TRUE) %>%
  mutate(
    trim_side = case_when(
      start == 1 & end < read_length ~ "left",
      start > 1 & end == read_length ~ "right",
      start == 1 & end == read_length ~ "both",
      TRUE ~ "internal"
    )
  ) %>%
  group_by(file_stem, trim_pattern_name, trim_side) %>%
  summarise(
    n_trimmed = n(),
    median_trim_length = median(trim_length, na.rm = TRUE),
    mean_trim_length = mean(trim_length, na.rm = TRUE),
    sd_trim_length = sd(trim_length, na.rm = TRUE),
    min_trim_length = min(trim_length, na.rm = TRUE),
    max_trim_length = max(trim_length, na.rm = TRUE),
    .groups = "drop"
  )

# Add adapter info and calculate proportion per sample
sample_totals <- trim_locations %>%
  group_by(file_stem) %>%
  summarise(total_reads = n_distinct(read_name), .groups = "drop")

trim_by_sample <- trim_by_sample %>%
  left_join(
    data.frame(
      trim_pattern_name = names(combined_trim_sequences),
      adapter_sequence = as.character(combined_trim_sequences),
      adapter_length = nchar(combined_trim_sequences)
    ),
    by = "trim_pattern_name"
  ) %>%
  left_join(sample_totals, by = "file_stem") %>%
  mutate(proportion = n_trimmed / total_reads) %>%
  relocate(adapter_sequence, adapter_length, trim_side, .after = trim_pattern_name) %>%
  relocate(n_trimmed, total_reads, proportion, .after = trim_side) %>%
  arrange(file_stem, desc(n_trimmed))

# Display table
DT::datatable(
  trim_by_sample,
  caption = "Trimming summary by sample (per adapter)",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(trim_by_sample), fontSize = '85%') %>%
  DT::formatRound(columns = c('median_trim_length', 'mean_trim_length', 
                               'sd_trim_length', 'proportion'), digits = 2)

# Export both tables
write_excel_function(
  named_list = list(
    "trim_by_adapter" = trim_by_adapter,
    "trim_by_sample" = trim_by_sample
  ),
  path = file.path(qc_dir, "qc.xlsx"),
  overwrite_sheet = TRUE
)
} # end if (length(combined_trim_sequences) > 0)
```

### Trim length distribution

```{r trim_length_plot, echo=FALSE}
if (length(combined_trim_sequences) > 0) {
# Use trim_by_sample for plotting
if (nrow(trim_by_sample) > 0) {
  # Calculate mean trim length for each adapter (for ordering x-axis)
  adapter_order <- trim_by_sample %>%
    group_by(trim_pattern_name) %>%
    summarise(mean_overall = mean(mean_trim_length, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_overall))
  
  # Prepare data for plotting
  plot_data <- trim_by_sample %>%
    mutate(trim_pattern_name = factor(trim_pattern_name, levels = adapter_order$trim_pattern_name))
  
  # Plot
  p_trim_length <- ggplot(plot_data, aes(x = trim_pattern_name, y = mean_trim_length)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.3) +
    geom_jitter(height = 0, width = 0.1, alpha = 0.5, size = 3) +
    stat_summary(fun = mean, geom = "point", shape = 18, colour = "red", size = 4, alpha = 0.6) +
    stat_summary(fun = mean, geom = "text", colour = "red", size = 3.5, alpha = 0.7,
                 angle = 45, vjust = -0.5, hjust = -0.25,
                 aes(label = after_stat(round(y, 1)))) +
    geom_point(aes(y = adapter_length), colour = "blue", shape = 3, size = 3) +
    scale_x_discrete(guide = guide_axis(angle = 75)) +
    scale_y_continuous(breaks = pretty_breaks(),
                       expand = expansion(mult = c(0.05, 0.2))) +
    labs(title = "Trim length distribution by adapter",
         subtitle = "Red diamond = mean, blue cross = expected adapter length, points = samples",
         x = "Adapter",
         y = "Trim length (bp)") +
    theme_minimal()
  
  print(p_trim_length)
  
  # Export plot
  plot_path <- file.path(plots_dir, "trim_length_distribution.png")
  if (!file.exists(plot_path) || !params$resume) {
    ggsave(
      filename = plot_path,
      plot = p_trim_length,
      width = 10,
      height = 6,
      dpi = params$plot_dpi,
      bg = "white"
    )
  }
}

message("All QC plots exported to: ", plots_dir)
} # end if (length(combined_trim_sequences) > 0)
```

---

# Summary

## Final read counts

```{r final_summary, echo=FALSE}
final_summary <- data.frame(
  sample = names(seq_trim),
  n_reads = sapply(seq_trim, length),
  mean_length = sapply(seq_trim, function(x) mean(width(x))),
  stringsAsFactors = FALSE
)

DT::datatable(
  final_summary,
  caption = "Final trimmed sequences",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(final_summary), fontSize = '85%') %>%
  DT::formatRound(columns = 'mean_length', digits = 1)

write_excel_function(named_list = list("final_summary" = final_summary),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

---

# Save output

```{r save_output, echo=FALSE}
# Compile module output
module1_output <- list(
  sequences        = seq_trim,
  file_details     = file_details,
  qc_summary       = sequencing_qc_summary,
  trim_summary     = if (exists("count_trimmed")) count_trimmed else NULL,
  trim_by_adapter  = if (exists("trim_by_adapter")) trim_by_adapter else NULL,
  trim_by_sample   = if (exists("trim_by_sample")) trim_by_sample else NULL,
  final_summary    = final_summary
)

# Create results directory
results_dir <- file.path(params$dir_out, "module_data")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

# Save complete module output
output_path <- file.path(results_dir, "01_import_qc.RData")
save(module1_output, file = output_path)
```

**Module 1 Complete!**

Results saved to: `r output_path`

- **`r length(seq_trim)`** samples processed
- **`r format(sum(final_summary$n_reads), big.mark = ",")`** total reads
- Mean reads per sample: **`r format(round(mean(final_summary$n_reads)), big.mark = ",")`**

---

# Module diagnostics

## Timing

```{r timing_table, echo=FALSE}
timing_df <- build_timing_table(timings)

DT::datatable(
  timing_df,
  caption = "Section timing — Module 1",
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Section", "Seconds", "Elapsed (mm:ss)", "Cumulative (mm:ss)")
) %>%
  DT::formatStyle(
    "section",
    target = "row",
    fontWeight = DT::styleEqual("Total", "bold")
  )
```

## Output objects

```{r manifest_table, echo=FALSE}
rdata_mb <- round(file.size(output_path) / 1024^2, 2)

manifest <- data.frame(
  object = names(module1_output),
  description = c(
    "Trimmed sequences ready for alignment (named list of XStringSet, one per sample)",
    "Input file metadata: names, formats, paths, paired-end status",
    "Per-sample QC metrics: read count, mean/median length, Phred quality",
    "Read counts by adapter and sample (NULL if trimming disabled)",
    "Trimming statistics per adapter across all samples (NULL if trimming disabled)",
    "Trimming statistics per adapter per sample (NULL if trimming disabled)",
    "Final read counts and mean lengths after all processing steps"
  ),
  class = sapply(module1_output, function(x) if (is.null(x)) "NULL" else class(x)[1]),
  size = sapply(module1_output, function(x) {
    format_size(if (is.null(x)) 0 else as.numeric(object.size(x)))
  }),
  bytes = sapply(module1_output, function(x) {
    if (is.null(x)) 0L else as.numeric(object.size(x))
  }),
  stringsAsFactors = FALSE
)

DT::datatable(
  manifest[, c("object", "description", "class", "size")],
  caption = sprintf("Objects saved to module1_output — RData file: %s on disk (%s)",
                    format_size(file.size(output_path)), output_path),
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Object", "Description", "Class", "In-memory size")
)
```

```{r diagnostics_export, echo=FALSE}
write_excel_function(
  named_list = list(
    "timing"   = timing_df,
    "manifest" = manifest
  ),
  path = file.path(qc_dir, "qc.xlsx"),
  overwrite_sheet = TRUE
)

message(sprintf("Module 1 complete — total time: %s  |  RData: %s",
                format_elapsed(sum(timings)), format_size(file.size(output_path))))
```

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```

