---
title: "Duke Pipeline - Module 2: Alignment"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

# Overview

**Module 2: Alignment and Processing**

This module performs comprehensive alignment processing:

1. Load trimmed sequences from Module 1
2. Parse reference sequence (split at NNNNN into left/right flanks)
3. Align reads to reference flanks using minimap2
4. **Filter alignments** for reads spanning both flanks
5. **Analyse strand orientation**
6. **Assess alignment quality**
7. **Visualise alignments** (optional)
8. **Correct strand orientation** (reverse complement antisense reads)
9. **Filter for commonest reference order** (removes spurious alignments)
10. **Allow manual read exclusion** (optional)
11. **Split sequences** into pre/left/mid/right/post segments

**Key improvements over original:**

- Standard genomic coordinates (ref_start < ref_end always)
- Proper MAPQ tie-breaking (eliminates duplicate alignments)
- Commonest order filtering after strand correction (catches spurious alignments)
- Better error handling and progress reporting

**Why no pre-correction order validation:**

Query coordinates from minimap2 CIGAR strings are unreliable for antisense reads before strand correction. The original duke.Rmd approach validates order AFTER correction when coordinates are consistent.

**Outputs:**

- Filtered alignment tables for each sample
- QC metrics and visualisations
- Strand-corrected sequences split into functional segments
- Sequences ready for repeat detection (Module 3)

```{r load_packages, message=FALSE, warning=FALSE, echo=FALSE}
# Set working directory
# if (interactive()) {
#   setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
# } else {
#   rmd_path <- knitr::current_input(dir = TRUE)
#   if (!is.null(rmd_path) && rmd_path != "") {
#     setwd(rmd_path)
#   }
# }

# Required packages
required_packages <- c("Biostrings", "Rsamtools", "GenomicAlignments",
                      "pbapply", "dplyr", "tidyr", "data.table", "stringr", 
                      "ggplot2", "scales", "DT", "openxlsx",
                      "parallel", "pbmcapply", "tibble", "stringi", "forcats")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}
```

```{r create_directories, echo=FALSE}
# Create output directories
temp_dir <- file.path(params$dir_out, "temp")
alignment_dir <- file.path(params$dir_out, "02_alignment", "alignment")  # Consolidated alignment outputs
results_dir <- file.path(params$dir_out, "module_data")
plots_dir <- file.path(params$dir_out, file.path("02_alignment", "plots"))

if (!dir.exists(temp_dir)) dir.create(temp_dir, recursive = TRUE)
if (!dir.exists(alignment_dir)) dir.create(alignment_dir, recursive = TRUE)
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)
```

---

# Load Module 1 results

Load the trimmed sequences from Module 1.

```{r load_module1, echo=FALSE, results='asis'}
# Load Module 1 results
module1_path <- file.path(params$dir_out, "module_data", "01_import_qc_results.RData")

if (!file.exists(module1_path)) {
  stop("Module 1 results not found. Please run Module 1 first: ", module1_path)
}

load(module1_path)

# Extract sequences and metadata
sequences <- module1_output$sequences
file_details <- module1_output$file_details

# Create named vectors (matching Module 1 pattern)
file_stems <- names(sequences)
file_stem_vector <- setNames(file_stems, file_stems)

message("Loaded ", length(sequences), " samples from Module 1")

# Format sample list for HTML
n_samples <- length(file_stem_vector)
if (n_samples <= 10) {
  cat("\n\n**", n_samples, " samples:**\n\n", 
      paste(names(file_stem_vector), collapse = ", "), "\n\n", sep="")
} else {
  cat("\n\n**", n_samples, " samples:**\n\n", 
      paste(head(names(file_stem_vector), 10), collapse = ", "), 
      ", ... and ", n_samples - 10, " more\n\n", sep="")
}

message("Total reads: ", format(sum(sapply(sequences, length)), big.mark = ","))
```

---

# Reference sequences

Parse the reference sequence file, which should contain a single sequence with the repeat region marked by at least 5 consecutive Ns (NNNNN). The sequence will be split into left and right flanking regions.

```{r parse_reference, echo=FALSE}
# Parse reference sequence
ref_data <- parse_reference_sequence(params$path_ref)
ref_sequences <- ref_data$sequences
ref_summary <- ref_data$summary

# Also save original full reference for downstream variant calling
ref_original <- readDNAStringSet(params$path_ref)
ref_full_sequence <- as.character(ref_original[[1]])

# Display reference summary
DT::datatable(
  ref_summary,
  caption = "Reference sequences",
  options = list(pageLength = 10, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact'
)
```

**Reference structure:**

- Left flank: `r ref_summary$ref_length[1]` bp
- Right flank: `r ref_summary$ref_length[2]` bp
- Repeat region marked by Ns in original reference

---

# Alignment

Align reads to left and right reference flanks separately using minimap2. This approach ensures that reads spanning the repeat region (which may vary in length) can align to both flanks independently.

```{r run_alignment, echo=FALSE, message=TRUE, results='asis'}
# Check for cached alignments
alignment_cache_path <- file.path(temp_dir, "02_alignments.RData")

if (params$resume && file.exists(alignment_cache_path)) {
  message("Loading cached alignments...")
  load(alignment_cache_path)
  
  # Verify loaded data has proper names
  if (is.null(names(alignments))) {
    warning("Cached alignments have no names - reconstructing from file_stem_vector")
    names(alignments) <- file_stem_vector
  }
  
} else {
  message("Running minimap2 alignment...")
  
  # Run alignment for each sample using named vector
  alignments <- apply_fn(
    file_stem_vector,  # Named vector ensures names are preserved
    minimap2_align_sample,
    sequences = sequences,
    ref_sequences = ref_sequences,
    out_dir = temp_dir,
    minimap2_args = params$minimap2_args,
    resume = params$resume
  )
  
  # Extract results
  alignments <- extract_apply_fn_result(alignments)
  
  # Verify names were preserved (defensive check)
  if (is.null(names(alignments))) {
    warning("Names were lost during processing - restoring from file_stem_vector")
    names(alignments) <- file_stem_vector
  }
  
  # Save cache
  save(alignments, file = alignment_cache_path)
}

# Final verification before proceeding
if (is.null(names(alignments)) || length(names(alignments)) != length(alignments)) {
  stop("ERROR: Alignment list is not properly named. Cannot proceed safely.")
}

# Format sample list for HTML display
alignment_samples <- names(alignments)
n_samples <- length(alignment_samples)

cat("\n\n**Alignments retrieved for ", n_samples, " samples:**\n\n", sep="")
if (n_samples <= 10) {
  cat(paste(alignment_samples, collapse = ", "), "\n\n", sep="")
} else {
  cat(paste(head(alignment_samples, 10), collapse = ", "), 
      ", ... and ", n_samples - 10, " more\n\n", sep="")
}

# Add file_stem column to each alignment table
alignments <- lapply(file_stem_vector, function(my_file_stem, alignments) {
  alignment_info <- alignments[[my_file_stem]]
  if (!is.null(alignment_info) && nrow(alignment_info) > 0) {
    alignment_info <- alignment_info %>%
      dplyr::mutate(file_stem = my_file_stem) %>%
      relocate(file_stem)
  }
  return(alignment_info)
}, alignments = alignments)

# Restore names after lapply (lapply returns unnamed list)
names(alignments) <- file_stem_vector

# Merge into single table for summary statistics
alignment_info_merge <- data.table::rbindlist(alignments, fill = TRUE)

message("Alignment complete!")
message("Total alignments: ", format(nrow(alignment_info_merge), big.mark = ","))
```

## Alignment summary

```{r alignment_summary, echo=FALSE}
# Summarise alignments per sample
alignment_summary <- alignment_info_merge %>%
  group_by(file_stem) %>%
  dplyr::summarise(
    total_alignments = n(),
    unique_reads = n_distinct(qname),
    mean_mapq = mean(mapq, na.rm = TRUE),
    mean_mismatches = mean(mismatches, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  alignment_summary,
  caption = "Alignment summary per sample",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(alignment_summary), fontSize = '85%') %>%
  DT::formatRound(columns = c('mean_mapq', 'mean_mismatches'), digits = 1)
```

---

# Filter alignments

Filter reads to retain only those that align to **both left and right flanks** of the same reference. This ensures the repeat region is bounded on both sides.

**Filtering criteria:**

1. Read must align to both left AND right flank
2. Both alignments must be on the same strand
3. Keep only the best alignment to each flank (highest MAPQ, then longest)

**Note:** Proper alignment order validation occurs AFTER strand correction, when query coordinates are reliable for all reads.

```{r filter_alignments, echo=FALSE}
# Add ref_id and ref_flank information
alignment_info_merge <- alignment_info_merge %>%
  left_join(ref_summary %>% dplyr::select(rname, ref_id, ref_flank),
            by = "rname") %>%
  relocate(ref_id, ref_flank, .after = "rname")

# Check for cached filtered alignments
cache_file <- file.path(temp_dir, "02_alignment_filtered.RData")

if (params$resume && file.exists(cache_file)) {
  message("Loading cached filtered alignments from: ", cache_file)
  load(cache_file)
} else {
  # Apply filtering
  message("Filtering alignments for dual-flank reads...")
  alignment_filtered <- filter_dual_flank_alignments(alignment_info_merge, ref_summary)
  
  # Cache the result
  save(alignment_filtered, file = cache_file)
  message("Cached filtered alignments to: ", cache_file)
}

message("Dual-flank filtering complete!")
message("Reads before filtering: ", format(n_distinct(alignment_info_merge$qname), big.mark = ","))
message("Reads after filtering: ", format(n_distinct(alignment_filtered$qname), big.mark = ","))
message("Retention rate: ", 
        round(100 * n_distinct(alignment_filtered$qname) / n_distinct(alignment_info_merge$qname), 1), "%")
```

## Filtering summary

```{r filtering_summary, echo=FALSE}
# Count reads before and after filtering
read_count_before <- alignment_info_merge %>%
  group_by(file_stem) %>%
  dplyr::summarise(before_filtering = n_distinct(qname), .groups = "drop")

read_count_after <- alignment_filtered %>%
  group_by(file_stem) %>%
  dplyr::summarise(after_filtering = n_distinct(qname), .groups = "drop")

filtering_summary <- read_count_before %>%
  left_join(read_count_after, by = "file_stem") %>%
  dplyr::mutate(
    retained = after_filtering / before_filtering,
    lost = before_filtering - after_filtering
  )

DT::datatable(
  filtering_summary,
  caption = "Reads retained after dual-flank filtering",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(filtering_summary), fontSize = '85%') %>%
  DT::formatRound(columns = 'retained', digits = 3)
```

```{r plot_filtering, echo=FALSE, fig.height=6, fig.width=10}
# Prepare data for plotting
filtering_plot_data <- filtering_summary %>%
  pivot_longer(cols = c(before_filtering, after_filtering),
               names_to = "stage", values_to = "read_count") %>%
  dplyr::mutate(stage = factor(stage, levels = c("before_filtering", "after_filtering"),
                               labels = c("Before", "After")))

# Plot
p_filtering <- ggplot(filtering_plot_data, aes(x = stage, y = read_count)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.5, size = 3) +
  geom_line(aes(group = file_stem), alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 18, 
               colour = "red", size = 4) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Read counts before and after dual-flank filtering",
       x = "Filtering stage",
       y = "Read count (log10 scale)") +
  theme_minimal()

print(p_filtering)

# Export plot
ggsave(
  filename = file.path(plots_dir, "filtering_stages.png"),
  plot = p_filtering,
  width = 8,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

```{r save_filtering_summary, echo=FALSE}
# Save to Excel
write_excel_function(
  named_list = list("filtering_summary" = filtering_summary),
  path = file.path(alignment_dir, "alignment_qc.xlsx"),
  overwrite_sheet = TRUE
)
```

```{r cleanup_alignment_merge, echo=FALSE}
# Clean up large objects if requested
if (params$remove_intermediate) {
  rm(alignment_info_merge)
}
```

---

# Correct strand orientation

Reverse complement reads that aligned on the reverse strand (-) so that all reads are oriented 5' to 3' relative to the reference. After correction, these reads are marked with strand "+c" (corrected).

**What happens:**

1. Sequences are reverse complemented
2. Query coordinates are flipped to match new orientation
3. Strand is changed from "-" to "+c"

```{r correct_strand, echo=FALSE}
# Check for cached strand-corrected data
strand_cache_path <- file.path(temp_dir, "02_strand_corrected.RData")

if (params$resume && file.exists(strand_cache_path)) {
  message("Loading cached strand-corrected alignments...")
  load(strand_cache_path)
} else {
  message("Correcting strand orientation...")
  
  # Split alignments back into list by sample
  alignment_list <- split(alignment_filtered, alignment_filtered$file_stem)
  names(alignment_list) <- file_stem_vector[names(alignment_list)]
  
  # Apply strand correction using named vector
  alignment_corrected <- apply_fn(
    file_stem_vector,
    correct_strand_orientation,
    alignment_list = alignment_list,
    out_dir = temp_dir,
    resume = params$resume
  )
  
  # Extract results
  alignment_corrected <- extract_apply_fn_result(alignment_corrected)
  
  # Verify names were preserved
  if (is.null(names(alignment_corrected))) {
    warning("Names were lost during strand correction - restoring")
    names(alignment_corrected) <- file_stem_vector
  }
  
  # Save cache
  save(alignment_corrected, file = strand_cache_path)
}

message("Strand correction complete!")
```

```{r cleanup_alignments, echo=FALSE}
# Clean up if requested
if (params$remove_intermediate) {
  rm(alignment_list, alignment_filtered)
}
```

---

# Validate reference order

**Critical quality control:** After strand correction, all reads should show left→right reference order. This step validates the biologically expected order and removes reads with backwards or unusual patterns (spurious alignments).

```{r validate_correct_order, echo=FALSE}
# Validate correct reference order (left→right)
order_validation_result <- validate_correct_reference_order(alignment_corrected)

# Extract results
alignment_validated <- order_validation_result$alignments
validation_summary <- order_validation_result$validation_summary

# Merge back into single table
alignment_final <- data.table::rbindlist(alignment_validated, fill = TRUE)

message("Reference order validation complete!")

# Count before and after
reads_before_validation <- sum(sapply(alignment_corrected, function(x) n_distinct(x$qname)))
reads_after_validation <- n_distinct(alignment_final$qname)

message("Reads before validation: ", format(reads_before_validation, big.mark = ","))
message("Reads after validation: ", format(reads_after_validation, big.mark = ","))
message("Reads with incorrect order removed: ", 
        format(reads_before_validation - reads_after_validation, big.mark = ","))
message("Retention rate: ", 
        round(100 * reads_after_validation / reads_before_validation, 1), "%")
```

## Validation summary

```{r validation_summary_table, echo=FALSE}
DT::datatable(
  validation_summary,
  caption = "Reference order validation by sample",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(validation_summary), fontSize = '85%') %>%
  DT::formatRound(columns = 'retention_rate', digits = 3)
```

```{r plot_validation, echo=FALSE, fig.height=6, fig.width=10}
# Plot retention after order validation
p_retention <- ggplot(validation_summary, aes(x = file_stem, y = retention_rate)) +
  geom_bar(stat = "identity", fill = "#619CFF") +
  geom_text(aes(label = scales::percent(retention_rate, accuracy = 0.1)),
            vjust = -0.5, size = 3) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(title = "Retention rate after reference order validation",
       subtitle = "Removes reads with backwards or unusual flank order",
       x = "Sample",
       y = "Retention rate") +
  theme_minimal()

print(p_retention)

# Export plot
# Calculate width (capped at 50 inches)
n_samples <- length(unique(validation_summary$file_stem))
plot_width <- min(50, max(8, n_samples * 0.8))

ggsave(
  filename = file.path(plots_dir, "retention_rates.png"),
  plot = p_retention,
  width = plot_width,
  height = 6,
  dpi = 300,
  bg = "white",
  limitsize = FALSE
)
```

```{r cleanup_corrected, echo=FALSE}
# Clean up if requested
if (params$remove_intermediate) {
  rm(alignment_corrected, alignment_validated)
}
```

---

# Strand analysis

Analyse the strand orientation of **validated** alignments. After strand correction and order validation, reads are marked as "+" (originally forward) or "+c" (corrected from reverse).

```{r strand_analysis, echo=FALSE}
# Summarise strand distribution
strand_summary <- alignment_final %>%
  group_by(file_stem, strand) %>%
  dplyr::summarise(read_count = n_distinct(qname), .groups = "drop") %>%
  group_by(file_stem) %>%
  dplyr::mutate(proportion = read_count / sum(read_count)) %>%
  ungroup() %>%
  # Fix strand label: "." should be "+" for display
  mutate(strand = ifelse(strand == ".", "+", as.character(strand)))

DT::datatable(
  strand_summary,
  caption = "Strand distribution (validated reads only)",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(strand_summary), fontSize = '85%') %>%
  DT::formatRound(columns = 'proportion', digits = 3)
```

```{r plot_strand, echo=FALSE, fig.height=6, fig.width=10}
p_strand <- ggplot(strand_summary, aes(x = file_stem, y = proportion, fill = strand)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = scales::percent(proportion, accuracy = 1)),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c("+" = "#619CFF", "+c" = "#00BA38")) +
  scale_y_continuous(labels = scales::percent) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(title = "Strand distribution per sample (validated reads only)",
       subtitle = "After strand correction and order validation",
       x = "Sample",
       y = "Proportion of reads",
       fill = "Strand") +
  theme_minimal()

print(p_strand)

# Export plot
# Calculate width (capped at 50 inches)
n_samples <- length(unique(strand_summary$file_stem))
plot_width <- min(50, max(8, n_samples * 0.8))

ggsave(
  filename = file.path(plots_dir, "strand_distribution.png"),
  plot = p_strand,
  width = plot_width,
  height = 6,
  dpi = 300,
  bg = "white",
  limitsize = FALSE
)
```

---

# Alignment quality

Assess alignment quality metrics for **validated** alignments only. These statistics reflect only high-quality reads with correct flank order.

```{r alignment_quality, echo=FALSE}
# Calculate quality metrics per sample and flank
quality_summary <- alignment_final %>%
  group_by(file_stem, ref_flank) %>%
  dplyr::summarise(
    n_alignments = n(),
    mean_alignment_length = mean(alignment_length_on_ref, na.rm = TRUE),
    median_alignment_length = median(alignment_length_on_ref, na.rm = TRUE),
    mean_mapq = mean(mapq, na.rm = TRUE),
    mean_mismatches = mean(mismatches, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  quality_summary,
  caption = "Alignment quality metrics (validated reads only)",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(quality_summary), fontSize = '85%') %>%
  DT::formatRound(columns = c('mean_mapq', 'mean_alignment_length'), digits = 1)
```

```{r plot_alignment_length, echo=FALSE, fig.height=6, fig.width=10}
# Determine if we should generate combined plot
n_samples <- length(unique(alignment_final$file_stem))

if (n_samples <= 30) {
  # Plot alignment length with jitter + boxplot overlay
  p_alignment_length <- ggplot(alignment_final, aes(x = ref_flank, y = alignment_length_on_ref)) +
    geom_jitter(height = 0, width = 0.2, alpha = 0.2, size = 0.8, color = "#619CFF") +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, width = 0.4, fill = "#619CFF") +
    facet_wrap(~ file_stem, scales = "free_x") +
    labs(title = "Alignment length distribution by flank (validated reads only)",
         subtitle = "After removing reads with incorrect reference order",
         x = "Reference flank",
         y = "Alignment length (bp)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p_alignment_length)
  
  # Export combined plot
  plot_width <- min(30, max(10, n_samples * 3))
  
  ggsave(
    filename = file.path(plots_dir, "alignment_length_by_flank.png"),
    plot = p_alignment_length,
    width = plot_width,
    height = 6,
    dpi = 300,
    bg = "white",
    limitsize = FALSE
  )
} else {
  message("Skipping combined alignment length plot (", n_samples, " samples > 30)")
  message("  Individual per-sample plots will be generated below")
}

# Create subdirectory for per-sample plots
alignment_length_by_sample_dir <- file.path(plots_dir, "alignment_length_by_flank_by_sample")
dir.create(alignment_length_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

# Individual per-sample plots
for (sample in unique(alignment_final$file_stem)) {
  sample_data <- alignment_final %>% filter(file_stem == sample)
  
  p_alignment_sample <- ggplot(sample_data, aes(x = ref_flank, y = alignment_length_on_ref)) +
    geom_jitter(height = 0, width = 0.2, alpha = 0.3, size = 1.2, color = "#619CFF") +
    geom_boxplot(alpha = 0.3, outlier.shape = NA, width = 0.4, fill = "#619CFF") +
    labs(title = paste("Alignment length by flank -", sample),
         subtitle = "Validated reads only",
         x = "Reference flank",
         y = "Alignment length (bp)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave(
    filename = file.path(alignment_length_by_sample_dir, paste0(sample, "-alignment_length_by_flank.png")),
    plot = p_alignment_sample,
    width = 8,
    height = 6,
    dpi = 300,
    bg = "white"
  )
}
```

```{r plot_mapq, echo=FALSE, fig.height=6, fig.width=10}
# Determine if we should generate combined plot
n_samples <- length(unique(alignment_final$file_stem))

if (n_samples <= 30) {
  p_mapq <- ggplot(alignment_final, aes(x = ref_flank, y = mapq)) +
    geom_violin(alpha = 0.6, fill = "#619CFF") +
    geom_jitter(height = 0, width = 0.2, alpha = 0.2, size = 0.8, color = "#619CFF") +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.3) +
    facet_wrap(~ file_stem, scales = "free_x") +
    labs(title = "MAPQ distribution by flank (validated reads only)",
         subtitle = "After strand correction and order validation",
         x = "Reference flank",
         y = "Mapping quality (MAPQ)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p_mapq)
  
  # Export combined plot
  plot_width <- min(30, max(10, n_samples * 3))
  
  ggsave(
    filename = file.path(plots_dir, "mapq_distribution.png"),
    plot = p_mapq,
    width = plot_width,
    height = 6,
    dpi = 300,
    bg = "white",
    limitsize = FALSE
  )
} else {
  message("Skipping combined MAPQ plot (", n_samples, " samples > 30)")
  message("  Individual per-sample plots will be generated below")
}

# Create subdirectory for per-sample plots
mapq_by_sample_dir <- file.path(plots_dir, "mapq_distribution_by_sample")
dir.create(mapq_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

# Individual per-sample plots
for (sample in unique(alignment_final$file_stem)) {
  sample_data <- alignment_final %>% filter(file_stem == sample)
  
  p_mapq_sample <- ggplot(sample_data, aes(x = ref_flank, y = mapq)) +
    geom_violin(alpha = 0.6, fill = "#619CFF") +
    geom_jitter(height = 0, width = 0.2, alpha = 0.3, size = 1.2, color = "#619CFF") +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.3) +
    labs(title = paste("MAPQ distribution by flank -", sample),
         subtitle = "Validated reads only",
         x = "Reference flank",
         y = "Mapping quality (MAPQ)") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  ggsave(
    filename = file.path(mapq_by_sample_dir, paste0(sample, "-mapq_distribution.png")),
    plot = p_mapq_sample,
    width = 8,
    height = 6,
    dpi = 300,
    bg = "white"
  )
}
```

---

# Visualise alignments (optional)

Visualise alignment positions on the reference sequence. **Only validated reads are shown** - spurious and backwards alignments have been removed.

```{r alignment_vis_function, echo=FALSE}
# Function to visualise alignments for one sample
visualise_sample_alignments <- function(sample_data, sample_name, 
                                       max_reads = 100) {
  
  # Exit if NULL or empty
  if (is.null(sample_data) || nrow(sample_data) == 0) {
    message("No data for ", sample_name)
    return(NULL)
  }
  
  # Get unique reads
  unique_reads <- unique(sample_data$qname)
  
  # Check if unique_reads is valid
  if (is.null(unique_reads) || length(unique_reads) == 0) {
    message("No unique reads found for ", sample_name)
    return(NULL)
  }
  
  # Ensure max_reads has a default value
  if (is.null(max_reads) || is.na(max_reads)) {
    max_reads <- 100
  }
  
  # Sample reads if too many
  if (length(unique_reads) > max_reads) {
    sampled_reads <- sample(unique_reads, max_reads)
    sample_data <- sample_data %>%
      dplyr::filter(qname %in% sampled_reads)
    message("Sampled ", max_reads, " reads from ", sample_name, 
            " (out of ", length(unique_reads), " total)")
  }
  
  # Create plot
  p <- ggplot(sample_data, aes(x = ref_start, xend = ref_end,
                                y = qname, yend = qname,
                                colour = strand)) +
    geom_segment(linewidth = 2, alpha = 0.6) +
    facet_wrap(~ rname, scales = "free_x") +
    scale_colour_manual(values = c("+" = "#619CFF", "+c" = "#00BA38")) +
    labs(title = paste("Alignments:", sample_name, "(validated reads only)"),
         subtitle = "After strand correction and order validation",
         x = "Reference position (bp)",
         y = "Read",
         colour = "Strand") +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
  
  return(p)
}
```

```{r visualise_alignments, echo=FALSE, message=FALSE, fig.height=8, fig.width=12}
if (params$visualise_alignment) {
  message("Generating alignment visualisations...")
  
  # Create subdirectory for coverage plots
  coverage_by_sample_dir <- file.path(plots_dir, "coverage_by_sample")
  dir.create(coverage_by_sample_dir, showWarnings = FALSE, recursive = TRUE)
  
  # Split alignments by sample
  alignment_by_sample <- split(alignment_final, alignment_final$file_stem)
  
  # Generate plots
  for (sample_name in names(alignment_by_sample)) {
    sample_data <- alignment_by_sample[[sample_name]]
    
    p <- visualise_sample_alignments(
      sample_data, 
      sample_name,
      max_reads = params$visualise_alignment_downsample
    )
    
    if (!is.null(p)) {
      # Don't print to HTML (would make file huge with many samples)
      # Just save to file
      
      # Export coverage plot
      ggsave(
        filename = file.path(coverage_by_sample_dir, paste0(sample_name, "-coverage.png")),
        plot = p,
        width = 12,
        height = 8,
        dpi = 300,
        bg = "white"
      )
    }
  }
  
  message("Coverage plots exported to: ", coverage_by_sample_dir)
}
```

---

# Manual read exclusion (optional)

Apply manual read exclusions if specified in the parameters.

```{r manual_exclusions, echo=FALSE}
# Check if manual exclusion path is provided and valid
has_exclusions <- !is.null(params$path_manual_exclusions) && 
                  !is.na(params$path_manual_exclusions) && 
                  params$path_manual_exclusions != ""

if (has_exclusions) {
  message("Loading manual read exclusions...")
  
  # Load exclusion list
  exclusion_list <- load_manual_read_exclusions(params$path_manual_exclusions)
  
  if (!is.null(exclusion_list)) {
    # Count before exclusion
    reads_before_exclusion <- nrow(alignment_final)
    
    # Apply exclusions to each sample
    for (sample_name in names(exclusion_list)) {
      alignment_final <- apply_manual_exclusions(
        alignment_final,
        exclusion_list,
        sample_name = sample_name
      )
    }
    
    # Count after exclusion
    reads_after_exclusion <- nrow(alignment_final)
    
    message("Manual exclusion complete!")
    message("Alignments excluded: ", 
            format(reads_before_exclusion - reads_after_exclusion, big.mark = ","))
  }
} else {
  message("No manual exclusions specified")
}
```

---

# Split sequences

Split each read into five segments based on alignment coordinates:

- **pre**: sequence before left flank alignment
- **left**: sequence aligned to left flank
- **mid**: sequence between flanks (contains repeat region)
- **right**: sequence aligned to right flank
- **post**: sequence after right flank alignment

```{r split_sequences, echo=FALSE}
message("Splitting sequences into segments...")

# Split sequences
alignment_split <- split_sequences_enhanced(alignment_final)

message("Sequence splitting complete!")
message("Reads with all segments: ", sum(!is.na(alignment_split$mid)))
```

## Segment length summary

```{r segment_summary, echo=FALSE}
# Calculate segment lengths
segment_summary <- alignment_split %>%
  dplyr::mutate(
    pre_length = nchar(pre),
    left_length = nchar(left),
    mid_length = nchar(mid),
    right_length = nchar(right),
    post_length = nchar(post)
  ) %>%
  group_by(file_stem) %>%
  dplyr::summarise(
    n_reads = n(),
    mean_pre = mean(pre_length, na.rm = TRUE),
    mean_left = mean(left_length, na.rm = TRUE),
    mean_mid = mean(mid_length, na.rm = TRUE),
    mean_right = mean(right_length, na.rm = TRUE),
    mean_post = mean(post_length, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  segment_summary,
  caption = "Mean segment lengths (bp)",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(segment_summary), fontSize = '85%') %>%
  DT::formatRound(columns = c('mean_pre', 'mean_left', 'mean_mid', 'mean_right', 'mean_post'), digits = 1)
```

```{r plot_mid_length, echo=FALSE, fig.height=6, fig.width=10}
# Plot mid (repeat) region length distribution
p_mid_length <- ggplot(alignment_split, aes(x = file_stem, y = nchar(mid))) +
  geom_violin(fill = "lightblue", alpha = 0.6) +
  geom_jitter(height = 0, width = 0.2, alpha = 0.2, size = 0.8, color = "#619CFF") +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.3) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(title = "Distribution of repeat region length (mid segment)",
       x = "Sample",
       y = "Mid segment length (bp)") +
  theme_minimal()

print(p_mid_length)

# Export combined plot
# Calculate width (capped at 50 inches)
n_samples <- length(unique(alignment_split$file_stem))
plot_width <- min(50, max(8, n_samples * 0.8))

ggsave(
  filename = file.path(plots_dir, "repeat_segment_length.png"),
  plot = p_mid_length,
  width = plot_width,
  height = 6,
  dpi = 300,
  bg = "white",
  limitsize = FALSE
)

# Create subdirectory for per-sample plots
repeat_segment_by_sample_dir <- file.path(plots_dir, "repeat_segment_length_by_sample")
dir.create(repeat_segment_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

# Individual per-sample plots (horizontal orientation to match Module 7)
for (sample in unique(alignment_split$file_stem)) {
  sample_data <- alignment_split %>% filter(file_stem == sample)
  
  p_mid_sample <- ggplot(sample_data, aes(x = nchar(mid), y = 1)) +
    geom_violin(fill = "lightblue", alpha = 0.6, width = 0.8) +
    geom_jitter(width = 0, alpha = 0.3, size = 1.2, color = "#619CFF") +
    geom_boxplot(height = 0.15, outlier.shape = NA, alpha = 0.3, fill = "white") +
    labs(title = paste("Repeat region length -", sample),
         subtitle = "Mid segment distribution",
         x = "Mid segment length (bp)",
         y = "") +
    theme_minimal() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
  
  ggsave(
    filename = file.path(repeat_segment_by_sample_dir, paste0(sample, "-repeat_segment_length.png")),
    plot = p_mid_sample,
    width = 8,
    height = 6,
    dpi = 300,
    bg = "white"
  )
}

message("All alignment plots exported to: ", plots_dir)
```

---

# Save outputs

Save the processed alignment data for use in Module 3.

```{r save_outputs, echo=FALSE}
# Compile module output
module2_output <- list(
  alignment_split = alignment_split,
  ref_summary = ref_summary,
  ref_sequences = ref_sequences,
  ref_full_sequence = ref_full_sequence,  # Full original for variant calling
  segment_summary = segment_summary,
  filtering_summary = filtering_summary,
  strand_summary = strand_summary,
  quality_summary = quality_summary,
  validation_summary = validation_summary
)

# Save complete module output
output_path <- file.path(results_dir, "02_alignment_results.RData")
save(module2_output, file = output_path)

message("Module 2 outputs saved to: ", output_path)
```

**Module 2 Complete!**

Results saved to: `r output_path`

- **`r nrow(alignment_split)`** reads processed
- **`r length(unique(alignment_split$file_stem))`** samples
- Mean repeat region length: **`r round(mean(nchar(alignment_split$mid), na.rm = TRUE))` bp**

Ready for Module 3: Repeat Detection

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```
