---
title: "Duke pipeline"
subtitle: "Module 3: Repeat Detection"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```
```{r set_tab_title, echo=FALSE, results='asis'}
cat('<script>document.title = "Duke | Module 3: Repeat Detection";</script>')
```


```{r load_packages, message=FALSE, warning=FALSE, echo=FALSE}
# Required packages
required_packages <- c("Biostrings", "pbapply", "pbmcapply", "dplyr", "tidyr", 
                       "data.table", "stringr", "ggplot2", "DT", 
                       "openxlsx", "ggridges", "scales")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}
```

```{r timing_init, echo=FALSE}
module_start_time <- proc.time()["elapsed"]
timings <- c()
```

```{r create_directories, echo=FALSE}
# Create output directories (derived from params$dir_out)
temp_dir <- file.path(params$dir_out, "temp")
results_dir <- file.path(params$dir_out, "module_data")
repeat_dir <- file.path(params$dir_out, "03_repeat_detection")
plots_dir <- file.path(params$dir_out, file.path("03_repeat_detection", "plots"))  # NEW: plots directory

if (!dir.exists(temp_dir)) dir.create(temp_dir, recursive = TRUE)
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
if (!dir.exists(repeat_dir)) dir.create(repeat_dir, recursive = TRUE)
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)  # NEW
```

```{r set_param_defaults, echo=FALSE}
# Set default values for optional parameters
if (is.null(params$repeat_count_round_digits)) params$repeat_count_round_digits <- 0
if (is.null(params$histogram_binwidth)) params$histogram_binwidth <- 1
if (is.null(params$plot_detection_rate)) params$plot_detection_rate <- TRUE
if (is.null(params$plot_histograms)) params$plot_histograms <- TRUE
if (is.null(params$repeat_count_method)) params$repeat_count_method <- "repeat_count_full"
if (is.null(params$na_repeat_handling)) params$na_repeat_handling <- "convert_to_zero"
if (is.null(params$rm_flank_length_outliers)) params$rm_flank_length_outliers <- FALSE
if (is.null(params$flank_iqr_multiplier)) params$flank_iqr_multiplier <- 1.5
if (is.null(params$check_duplicate_readnames)) params$check_duplicate_readnames <- TRUE
if (is.null(params$plot_dpi)) params$plot_dpi <- 150
if (is.null(params$plot_per_sample)) params$plot_per_sample <- TRUE
if (is.null(params$export_read_counts)) params$export_read_counts <- TRUE
```

# Overview

This module identifies and counts repeat tracts in the mid sequence, performs flank length QC, and optionally filters reads with outlier combined flank lengths.

**Purpose:**

1. Find repeat tracts matching the target pattern (e.g., CAG)
2. Validate repeat detection rates
3. Count repeats using robust length-based method
4. **Flank length QC** — assess combined (left + right) flank length distribution per sample; optionally filter outlier reads before downstream analysis
5. Count method comparison and repeat length distribution visualisation

**Key parameters:** `rm_flank_length_outliers` (default `FALSE`), `flank_iqr_multiplier` (default `1.5`, standard IQR rule; only active when `rm_flank_length_outliers = TRUE`)

**Input:** `02_alignment.RData` from Module 2  
**Output:** `03_repeat_detection.RData`, `repeat_detection.xlsx`, `flank/` plots

---

# Load previous results

```{r load_results, echo=FALSE}
# Load Module 2 results
module2_path <- file.path(results_dir, "02_alignment.RData")
safe_load_module(module2_path, module_num = 2, envir = environment())

# Extract alignment_split
alignment_split <- module2_output$alignment_split

message("Loaded Module 2 results:")
message("  Reads: ", format(nrow(alignment_split), big.mark = ","))
message("  Samples: ", n_distinct(alignment_split$file_stem))
```

---

# Find repeat tracts

Search for repeat tracts in the mid sequence (between left and right flanks).

**Parameters:**

| Parameter | Value |
|-----------|-------|
| Pattern | `r params$rpt_pattern` |
| Min repeats | `r params$rpt_min_repeats` |
| Max mismatch | `r params$rpt_max_mismatch` |
| Perfect at start | `r params$rpt_start_perfect_repeats` |
| Perfect at end | `r params$rpt_end_perfect_repeats` |
| Max gap between repeats | `r params$rpt_max_gap` |
| Max gap between tracts | `r params$rpt_max_tract_gap` |
| Return option | `r params$rpt_return_option` |

```{r t_detect_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r repeat_detection, echo=FALSE}
message("Processing per sample: find tracts → reposition → count")
message("")

# Check for cached alignment_counts (combined from all samples)
alignment_counts_cache <- file.path(temp_dir, "03_alignment_counts.RData")

if (params$resume && safe_load_cache(alignment_counts_cache, environment())) {
  message("✓ Loading cached alignment counts...")
  message("  Loaded ", format(nrow(alignment_counts), big.mark = ","), " reads from ",
          length(unique(alignment_counts$file_stem)), " samples")

} else {
  message("Processing all samples with full parallelisation...")
  
  # Get sample names
  samples <- unique(alignment_split$file_stem)
  n_samples <- length(samples)
  message("  Samples: ", n_samples)
  message("  Threads: ", params$threads)
  message("")
  
  # Process each sample through ALL three steps (parallelised)
  # - Find repeat tracts
  # - Reposition segments
  # - Count repeats
  alignment_counts_list <- apply_fn(
    samples,
    process_sample_repeat_detection,
    alignment_split = alignment_split,
    pattern = params$rpt_pattern,
    min_repeats = params$rpt_min_repeats,
    max_mismatch = params$rpt_max_mismatch,
    start_perfect_repeats = params$rpt_start_perfect_repeats,
    end_perfect_repeats = params$rpt_end_perfect_repeats,
    max_gap = params$rpt_max_gap,
    max_tract_gap = params$rpt_max_tract_gap,
    return_option = params$rpt_return_option,
    repeat_count_method = params$repeat_count_method,
    round_digits = params$repeat_count_round_digits,
    out_dir = temp_dir,
    resume = params$resume
  )
  
  # Extract results
  alignment_counts_list <- extract_apply_fn_result(alignment_counts_list)
  
  # Combine all samples
  message("Combining results from all samples...")
  alignment_counts <- bind_rows(alignment_counts_list)
  
  # Save cache for alignment_counts
  save(alignment_counts, file = alignment_counts_cache)
  message("✓ Saved cache: ", basename(alignment_counts_cache))
}

message("")
message("Processing complete. Generating summaries and visualisations...")
message("")

# Handle reads where no repeat tract was found
n_na <- sum(is.na(alignment_counts$repeat_count))

if (n_na > 0) {
  na_rate <- n_na / nrow(alignment_counts)
  message("  Reads without detectable repeats: ", format(n_na, big.mark = ","), 
          " (", round(100 * na_rate, 1), "%)")
  
  # na_repeat_handling controls what happens to reads where no repeat tract
  # was detected within the mid sequence.
  #
  # "convert_to_zero" (default): assign repeat_count = 0.
  #   Appropriate for long-read data where every read spans the full locus —
  #   Module 2 already requires alignment to both flanks, so a missing tract
  #   genuinely means no (or sub-threshold) repeat sequence is present.
  #
  # "filter": remove these reads entirely.
  #   Use when NA reads represent a distinct technical failure class you want
  #   excluded from all downstream analysis.
  #
  # NA: leave repeat_count as NA and propagate as-is.
  #   Reads remain in alignment_counts but will be excluded from any
  #   calculation that ignores NAs. Useful for inspecting the NA reads
  #   before deciding how to handle them.
  
  if (!is.null(params$na_repeat_handling) && !is.na(params$na_repeat_handling)) {
    if (params$na_repeat_handling == "convert_to_zero") {
      alignment_counts <- alignment_counts %>%
        mutate(
          repeat_count = tidyr::replace_na(repeat_count, 0),
          repeat_count_full = tidyr::replace_na(repeat_count_full, 0),
          repeat_count_match = tidyr::replace_na(repeat_count_match, 0)
        )
      message("  → Converted NA to 0 (na_repeat_handling = 'convert_to_zero')")
    } else if (params$na_repeat_handling == "filter") {
      n_before <- nrow(alignment_counts)
      alignment_counts <- alignment_counts %>%
        filter(!is.na(repeat_count))
      n_after <- nrow(alignment_counts)
      message("  → Filtered out ", format(n_before - n_after, big.mark = ","), 
              " reads (na_repeat_handling = 'filter')")
    } else {
      warning("  Unrecognised na_repeat_handling value: '", params$na_repeat_handling,
              "'. Expected 'convert_to_zero', 'filter', or NA. Reads left as NA.")
    }
  } else {
    message("  → NAs retained (na_repeat_handling = NA)")
  }
}
```

```{r t_detect_end, echo=FALSE}
timings["Repeat detection"] <- proc.time()["elapsed"] - .t
```

## Detection summary

```{r repeat_detection_summary, echo=FALSE}
# Summary statistics by sample
tract_summary <- alignment_counts %>%
  group_by(file_stem) %>%
  summarise(
    total_reads = n(),
    reads_with_repeat = sum(!is.na(repeat_count) & repeat_count > 0),
    reads_without_repeat = sum(is.na(repeat_count) | repeat_count == 0),
    detection_rate = mean(!is.na(repeat_count) & repeat_count > 0),
    mean_repeat = mean(repeat_count, na.rm = TRUE),
    median_repeat = median(repeat_count, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  tract_summary,
  caption = "Repeat detection summary by sample",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(tract_summary), fontSize = '85%') %>%
  DT::formatRound(columns = c('detection_rate', 'mean_repeat', 'median_repeat'), digits = 2)
```

```{r plot_detection_rate, echo=FALSE, fig.height=6, fig.width=10, eval=params$plot_detection_rate}
if (params$plot_detection_rate) {
  n_det_samples <- nrow(tract_summary)

  # Shared jitter position with seed so text labels align with points
  jitter_pos <- position_jitter(width = 0.08, height = 0, seed = 42)

  p_detection <- ggplot(tract_summary, aes(x = "", y = detection_rate)) +
    geom_boxplot(outlier.shape = NA, width = 0.35,
                 fill = "grey92", colour = "grey40") +
    geom_point(position = jitter_pos, size = 3,
               colour = "#619CFF", alpha = 0.8) +
    geom_text(aes(label = format(reads_without_repeat, big.mark = ",")),
              position = jitter_pos, hjust = -0.25, size = 3, colour = "grey30") +
    stat_summary(fun = mean, geom = "point", shape = 18,
                 colour = "red", size = 4, alpha = 0.6) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.001),
                       expand = expansion(mult = c(0.05, 0.05))) +
    labs(title    = "Repeat tract detection rate",
         subtitle = sprintf("Pattern: %s  |  n = %d\nPoint label = reads without repeat",
                            params$rpt_pattern, n_det_samples),
         x        = NULL,
         y        = sprintf("Detection rate\n(%s tract detected)", params$rpt_pattern)) +
    theme_bw() +
    theme(axis.ticks.x = element_blank())

  print(p_detection)

  ggsave(
    filename = file.path(plots_dir, "repeat_detection_rate.png"),
    plot     = p_detection,
    width    = 5,
    height   = 5,
    dpi      = params$plot_dpi,
    bg       = "white"
  )
}
```

---

# Counting methods

Before counting, this module **refines the `mid` segment** inherited from Module 2. Module 2 defines `mid` as everything between the left and right flank alignments — this may include non-repeat sequence at either end (e.g. the terminal CAA+CAG interruption in HTT). Module 3 runs repeat tract detection within `mid` and repositions the boundaries:

- Any sequence **before** the detected tract start is appended to `left`
- Any sequence **after** the detected tract end is prepended to `right`  
- `mid` is updated to contain **only the detected repeat tract**

The repositioning is controlled by your parameters — in particular, `rpt_end_perfect_repeats = `r params$rpt_end_perfect_repeats`` requires the tract to end with `r params$rpt_end_perfect_repeats` perfect `r params$rpt_pattern` unit(s), so terminal interruptions (e.g. CAACAG in HTT) are correctly excluded from `mid` and moved into `right`. All three counting methods below operate on this repositioned `mid`.

The pipeline calculates three counting methods for all reads:

**1. repeat_count_full** (Default - Most robust)
- Formula: Tract length ÷ pattern length (rounded to nearest integer)
- Includes all bases in the repositioned tract — gap bases from within-tract interruptions (allowed by `rpt_max_gap = `r params$rpt_max_gap``) contribute to the length
- Best for: Oxford Nanopore data with sequencing errors
- Example: 63 bp CAG tract ÷ 3 bp = 21 repeats

**2. repeat_count_match** (Stringent)
- Formula: Count only perfect pattern matches within the repositioned tract
- Excludes gap bases — reads with within-tract interruptions will give a lower count than repeat_count_full
- Best for: High-accuracy data (PacBio HiFi, Illumina)
- Example: Requires exact CAG CAG CAG... (no interruptions)

**3. repeat_count_tracts** (Structural analysis)
- Formula: Reports tract structure as "3,2,2" (sizes separated by gaps)
- Use: QC visualisation and interruption analysis
- Note: Returns string, not used as primary count

**Current method:** `r params$repeat_count_method`

All three methods are calculated and available in the output data. Discordance between `repeat_count_full` and `repeat_count_match` indicates reads with within-tract interruptions (gaps permitted by `rpt_max_gap`).

---

## Count method comparison

```{r count_method_comparison, echo=FALSE, fig.height=5, fig.width=12}
# compute_cmc defined here — used in this chunk and in flank_qc_post below
compute_cmc <- function(df) {
  df %>%
    group_by(file_stem) %>%
    dplyr::summarise(
      n_reads                        = n(),
      mean_repeat_count_full         = mean(repeat_count_full,  na.rm = TRUE),
      mean_repeat_count_match        = mean(repeat_count_match, na.rm = TRUE),
      median_repeat_count_full       = median(repeat_count_full,  na.rm = TRUE),
      median_repeat_count_match      = median(repeat_count_match, na.rm = TRUE),
      mean_diff_full_minus_match     = mean(repeat_count_full - repeat_count_match, na.rm = TRUE),
      concordance_pct_reads_agreeing = mean(round(repeat_count_full) == repeat_count_match,
                                            na.rm = TRUE),
      .groups = "drop"
    )
}

# Baseline comparison on unfiltered alignment_counts
count_method_comparison <- compute_cmc(alignment_counts)
n_cmc <- nrow(count_method_comparison)

lm_cmc  <- lm(mean_repeat_count_match ~ mean_repeat_count_full, data = count_method_comparison)
lm_coef <- coef(lm_cmc)
lm_r2   <- summary(lm_cmc)$r.squared
eq_label <- sprintf("y = %.3fx + %.2f\nR\u00b2 = %.4f", lm_coef[2], lm_coef[1], lm_r2)

ax_min <- min(c(count_method_comparison$mean_repeat_count_full,
                count_method_comparison$mean_repeat_count_match), na.rm = TRUE)
ax_max <- max(c(count_method_comparison$mean_repeat_count_full,
                count_method_comparison$mean_repeat_count_match), na.rm = TRUE)
ax_pad <- (ax_max - ax_min) * 0.05

p_method_scatter <- ggplot(count_method_comparison,
                            aes(x = mean_repeat_count_full, y = mean_repeat_count_match)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_smooth(method = "lm", se = TRUE, colour = "#619CFF", fill = "#619CFF",
              alpha = 0.15, formula = y ~ x) +
  geom_point(size = 3, alpha = 0.8, colour = "grey30") +
  annotate("text", x = ax_min + ax_pad, y = ax_max - ax_pad,
           label = eq_label, hjust = 0, vjust = 1, size = 3.5) +
  scale_x_continuous(limits = c(ax_min - ax_pad, ax_max + ax_pad)) +
  scale_y_continuous(limits = c(ax_min - ax_pad, ax_max + ax_pad)) +
  labs(title    = "Count method comparison: full vs match",
       subtitle = sprintf(
         "repeat_count_full (length / %d bp) vs repeat_count_match (exact)\nDashed = identity | Blue = regression | n = %d samples",
         nchar(params$rpt_pattern), n_cmc),
       x = "Mean repeat count (repeat_count_full — length-based)",
       y = "Mean repeat count (repeat_count_match — exact)") +
  theme_bw() +
  coord_fixed()

p_method_conc <- ggplot(count_method_comparison,
                         aes(x = "", y = concordance_pct_reads_agreeing)) +
  geom_boxplot(outlier.shape = NA, width = 0.35, fill = "grey92", colour = "grey40") +
  geom_jitter(width = 0.08, height = 0, size = 3, alpha = 0.8, colour = "grey30") +
  stat_summary(fun = mean, geom = "point", shape = 18, colour = "red", size = 4, alpha = 0.6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0.02, 0.02))) +
  labs(title    = "Method concordance per sample",
       subtitle = sprintf("round(full) == match | Red = mean | n = %d", n_cmc),
       x = NULL, y = "Concordance") +
  theme_bw() +
  theme(axis.ticks.x = element_blank())

if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  p_method_combined <- p_method_scatter + p_method_conc + plot_layout(widths = c(2, 1))
  print(p_method_combined)
  ggsave(file.path(plots_dir, "count_method_comparison.png"), p_method_combined,
         width = 12, height = 5, dpi = params$plot_dpi, bg = "white")
} else {
  print(p_method_scatter); print(p_method_conc)
}

mean_conc <- mean(count_method_comparison$concordance_pct_reads_agreeing, na.rm = TRUE)
if (mean_conc >= 0.95) {
  message(sprintf("High concordance (mean = %.1f%%): repeat_count_match appropriate.",
                  100 * mean_conc))
} else if (mean_conc >= 0.85) {
  message(sprintf("Moderate concordance (mean = %.1f%%): consider repeat_count_full.",
                  100 * mean_conc))
} else {
  message(sprintf("Low concordance (mean = %.1f%%): use repeat_count_full.",
                  100 * mean_conc))
}
```

---

# Flank length QC

The left and right flank lengths reflect the Module 3 repeat tract repositioning. Total combined flank length (left + right) should be stable per allele — the repeat is the primary source of read-to-read variation. Variation in total flank length indicates chimeras, truncated reads, or alignment boundary shifts.

**Flank length filtering:** controlled by `rm_flank_length_outliers` (default `FALSE`) and `flank_iqr_multiplier`. If enabled, reads with outlier combined flank lengths are removed from `alignment_counts` here and the filtered table propagates to all downstream modules (4-7). Review the plots and sweep table below before enabling — the effect on the count method comparison immediately below shows whether outlier reads cluster at particular repeat lengths.

```{r t_flank_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r flank_qc, echo=FALSE, fig.height=4, fig.width=6}
flank_dir <- file.path(params$dir_out, "03_repeat_detection", "plots", "flank")
dir.create(flank_dir, showWarnings = FALSE, recursive = TRUE)

# Capture read count BEFORE any flank filtering — used in post-filter read count plot
n_reads_before_flank_filter <- nrow(alignment_counts)

# IQR multiplier — read from params, fall back to 1.5 if not set
iqr_mult <- if (!is.null(params$flank_iqr_multiplier)) params$flank_iqr_multiplier else 1.5

# Combined flank length + repeat count per read (before any filtering)
flank_qc_data <- alignment_counts %>%
  dplyr::filter(!is.na(mid), mid != "") %>%
  dplyr::mutate(
    left_length  = nchar(left),
    right_length = nchar(right),
    flank_total  = left_length + right_length,
    repeat_count = .data[[params$repeat_count_method]]
  )

# Per-sample summary of combined flank length
flank_summary <- flank_qc_data %>%
  dplyr::group_by(file_stem) %>%
  dplyr::summarise(
    n_reads              = n(),
    flank_mean_bp        = round(mean(flank_total,  na.rm = TRUE), 0),
    flank_sd_bp          = round(sd(flank_total,    na.rm = TRUE), 1),
    flank_iqr_bp         = round(IQR(flank_total,   na.rm = TRUE), 1),
    flank_modal_bp       = as.numeric(names(which.max(table(flank_total)))),
    flank_pct_at_modal   = round(100 * sum(flank_total == flank_modal_bp) / n(), 1),
    flank_min_bp         = min(flank_total, na.rm = TRUE),
    flank_max_bp         = max(flank_total, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  flank_summary,
  caption  = "Combined flank length (left + right) per sample — flank_pct_at_modal = % of reads at most common combined length",
  options  = list(pageLength = 25, scrollX = TRUE, dom = 'ft'),
  rownames = FALSE,
  class    = 'cell-border stripe compact'
) %>%
  DT::formatStyle(columns = colnames(flank_summary), fontSize = '85%')

# Cohort-level IQR plot — one point per sample, for Module 5 calibration
p_flank_iqr <- ggplot(flank_summary, aes(x = "", y = flank_iqr_bp)) +
  geom_boxplot(outlier.shape = NA, width = 0.35, fill = "#619CFF", alpha = 0.6) +
  geom_jitter(width = 0.08, height = 0, size = 2.5, alpha = 0.8, colour = "grey30") +
  stat_summary(fun = mean, geom = "point", shape = 18, colour = "red", size = 4, alpha = 0.6) +
  labs(title    = "Per-sample combined flank IQR",
       subtitle = sprintf("n = %d | red = mean | informs Module 5 filter",
                          nrow(flank_summary)),
       x = NULL,
       y = "Within-sample IQR of (left + right) bp") +
  theme_bw() +
  theme(axis.ticks.x = element_blank())

print(p_flank_iqr)

ggsave(
  filename = file.path(flank_dir, "flank_length_iqr.png"),
  plot     = p_flank_iqr,
  width    = 4,
  height   = 4,
  dpi      = params$plot_dpi,
  bg       = "white"
)

# Per-sample box+whisker of actual combined flank lengths with downsampled points
# Cohort-level combined flank distribution: one point per sample (modal flank length)
# with the current IQR filter threshold shown as dotted lines.
# Reads outside the dotted lines would be removed by the filter.

# Per-sample modal and filter thresholds
# Always compute 1.5x IQR (standard reference) plus the user-set multiplier
filter_active <- isTRUE(params$rm_flank_length_outliers)
show_custom   <- filter_active && (abs(iqr_mult - 1.5) > 0.01)  # show custom line only if different from 1.5

flank_thresholds <- flank_qc_data %>%
  dplyr::group_by(file_stem) %>%
  dplyr::summarise(
    modal_flank    = as.numeric(names(which.max(table(flank_total)))),
    modal_repeat   = as.numeric(names(which.max(table(repeat_count)))),
    fq1            = quantile(flank_total, 0.25, na.rm = TRUE),
    fq3            = quantile(flank_total, 0.75, na.rm = TRUE),
    fiqr           = IQR(flank_total, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    ref_thresh_lo    = fq1 - 1.5    * fiqr,
    ref_thresh_hi    = fq3 + 1.5    * fiqr,
    custom_thresh_lo = fq1 - iqr_mult * fiqr,
    custom_thresh_hi = fq3 + iqr_mult * fiqr
  )

mean_ref_lo    <- mean(flank_thresholds$ref_thresh_lo,    na.rm = TRUE)
mean_ref_hi    <- mean(flank_thresholds$ref_thresh_hi,    na.rm = TRUE)
mean_custom_lo <- mean(flank_thresholds$custom_thresh_lo, na.rm = TRUE)
mean_custom_hi <- mean(flank_thresholds$custom_thresh_hi, na.rm = TRUE)

# Subtitle describes what lines are shown
filter_label <- if (!filter_active) {
  "filter disabled — lines shown for reference only"
} else if (show_custom) {
  sprintf("grey dashed = 1.5 x IQR | blue dashed = user setting (%.1f x IQR)", iqr_mult)
} else {
  sprintf("blue dashed = %.1f x IQR (active filter)", iqr_mult)
}

# Per-sample % within each threshold (computed on all reads, not downsampled)
flank_pct <- flank_qc_data %>%
  dplyr::left_join(flank_thresholds %>%
                     dplyr::select(file_stem, ref_thresh_lo, ref_thresh_hi,
                                   custom_thresh_lo, custom_thresh_hi),
                   by = "file_stem") %>%
  dplyr::group_by(file_stem) %>%
  dplyr::summarise(
    n_reads        = n(),
    pct_within_ref = round(100 * mean(flank_total >= ref_thresh_lo &
                                       flank_total <= ref_thresh_hi), 1),
    pct_within_custom = round(100 * mean(flank_total >= custom_thresh_lo &
                                          flank_total <= custom_thresh_hi), 1),
    .groups = "drop"
  )

# Join onto flank_thresholds for the cohort plot
flank_thresholds <- flank_thresholds %>%
  dplyr::left_join(flank_pct %>%
                     dplyr::select(file_stem, pct_within_ref, pct_within_custom),
                   by = "file_stem")

# Cohort repeat vs flank scatter: one point per sample (modal repeat, modal flank)
# % within 1.5x IQR labelled next to each point.
# The horizontal dashed lines show the MEAN of the per-sample IQR fences across all
# samples. They are visual guides only — the actual filter (when enabled) applies
# per-sample fences computed independently for each sample.
p_flank_box <- ggplot(flank_thresholds, aes(x = modal_repeat, y = modal_flank)) +
  geom_hline(yintercept = mean_ref_lo, linetype = "dashed",
             colour = "grey50", linewidth = 0.7, alpha = 0.7) +
  geom_hline(yintercept = mean_ref_hi, linetype = "dashed",
             colour = "grey50", linewidth = 0.7, alpha = 0.7) +
  { if (show_custom) list(
      geom_hline(yintercept = mean_custom_lo, linetype = "dashed",
                 colour = "steelblue", linewidth = 0.8, alpha = 0.7),
      geom_hline(yintercept = mean_custom_hi, linetype = "dashed",
                 colour = "steelblue", linewidth = 0.8, alpha = 0.7),
      annotate("text", x = -Inf, y = mean_custom_hi,
               label = sprintf("%.1f x IQR", iqr_mult),
               vjust = -0.4, hjust = -0.1, size = 3, colour = "steelblue"),
      annotate("text", x = -Inf, y = mean_custom_lo,
               label = sprintf("%.1f x IQR", iqr_mult),
               vjust = 1.3, hjust = -0.1, size = 3, colour = "steelblue")
  )} +
  annotate("text", x = -Inf, y = mean_ref_hi,
           label = "1.5 x IQR", vjust = -0.4, hjust = -0.1,
           size = 3, colour = "grey50") +
  annotate("text", x = -Inf, y = mean_ref_lo,
           label = "1.5 x IQR", vjust = 1.3, hjust = -0.1,
           size = 3, colour = "grey50") +
  geom_point(aes(colour = pct_within_ref), size = 3, alpha = 0.9) +
  scale_colour_gradient(low = "#E41A1C", high = "#4DAF4A",
                        name = "% within\n1.5 x IQR",
                        limits = c(0, 100),
                        breaks = c(0, 25, 50, 75, 100),
                        labels = function(x) paste0(x, "%")) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +
  labs(title    = "Modal repeat vs modal combined flank length",
       subtitle = sprintf("n = %d samples | colour = %% within 1.5 x IQR | %s",
                          nrow(flank_thresholds), filter_label),
       x        = sprintf("Modal repeat length (%s, CAG units)", params$repeat_count_method),
       y        = "Modal combined flank length (left + right, bp)") +
  theme_bw()

print(p_flank_box)

ggsave(
  filename = file.path(flank_dir, "flank_length_distribution.png"),
  plot     = p_flank_box,
  width    = 6,
  height   = 5,
  dpi      = params$plot_dpi,
  bg       = "white"
)

# Per-sample repeat vs flank scatter — saved to directory, gated by plot_per_sample
flank_box_dir <- file.path(flank_dir, "flank_boxplot_by_sample")
dir.create(flank_box_dir, showWarnings = FALSE, recursive = TRUE)

if (params$plot_per_sample) {
  max_pts <- 5000
  for (samp in unique(flank_qc_data$file_stem)) {
    samp_data <- flank_qc_data %>%
      dplyr::filter(file_stem == samp)
    n_samp    <- nrow(samp_data)

    samp_thresh <- flank_thresholds %>% dplyr::filter(file_stem == samp)

    # Flag reads by zone — computed on all reads for accurate percentages
    samp_data <- samp_data %>%
      dplyr::mutate(
        outlier_ref = flank_total < samp_thresh$ref_thresh_lo |
                      flank_total > samp_thresh$ref_thresh_hi,
        outlier_custom = if (show_custom) {
          flank_total < samp_thresh$custom_thresh_lo |
          flank_total > samp_thresh$custom_thresh_hi
        } else outlier_ref,
        point_cat = dplyr::case_when(
          !outlier_ref    ~ "within 1.5 x IQR",
          !outlier_custom ~ sprintf("%.1f-1.5 x IQR", iqr_mult),
          TRUE            ~ "outlier"
        )
      )

    # Count per zone for annotation (all reads, not downsampled)
    zone_counts <- samp_data %>%
      dplyr::count(point_cat) %>%
      dplyr::mutate(pct = round(100 * n / sum(n), 1),
                    label = sprintf("%s: %s (%.1f%%)", point_cat,
                                    format(n, big.mark = ","), pct))

    corner_label <- paste(zone_counts$label, collapse = "\n")

    samp_plot <- samp_data %>%
      dplyr::slice_sample(n = min(max_pts, n_samp))

    pt_colours <- c("within 1.5 x IQR" = "grey60", "outlier" = "#E41A1C")
    if (show_custom)
      pt_colours[sprintf("%.1f-1.5 x IQR", iqr_mult)] <- "#FF7F00"

    p_samp <- ggplot(samp_plot, aes(x = repeat_count, y = flank_total, colour = point_cat)) +
      geom_hline(yintercept = samp_thresh$ref_thresh_lo, linetype = "dashed",
                 colour = "grey50", linewidth = 0.7, alpha = 0.7) +
      geom_hline(yintercept = samp_thresh$ref_thresh_hi, linetype = "dashed",
                 colour = "grey50", linewidth = 0.7, alpha = 0.7) +
      { if (show_custom) list(
          geom_hline(yintercept = samp_thresh$custom_thresh_lo, linetype = "dashed",
                     colour = "steelblue", linewidth = 0.7, alpha = 0.7),
          geom_hline(yintercept = samp_thresh$custom_thresh_hi, linetype = "dashed",
                     colour = "steelblue", linewidth = 0.7, alpha = 0.7)
      )} +
      geom_point(size = 0.6, alpha = 0.4) +
      annotate("text", x = Inf, y = Inf, label = corner_label,
               hjust = 1.05, vjust = 1.3, size = 2.8, colour = "grey20") +
      scale_colour_manual(values = pt_colours, name = NULL) +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +
      labs(title    = paste("Repeat vs flank length —", samp),
           subtitle = sprintf("n = %s | IQR = %.1f bp | %s",
                              format(n_samp, big.mark = ","),
                              samp_thresh$fiqr, filter_label),
           x = sprintf("Repeat length (%s, CAG units)", params$repeat_count_method),
           y = "Combined flank length (left + right, bp)") +
      theme_bw() +
      theme(legend.position = "top")

    ggsave(
      filename = file.path(flank_box_dir, paste0(samp, "-flank_scatter.png")),
      plot     = p_samp,
      width    = 6, height   = 5,
      dpi      = params$plot_dpi,
      bg       = "white"
    )
  }
  message("Per-sample repeat vs flank scatter plots saved to: ", flank_box_dir)
}

# Deviation category table — per sample, how far do reads fall from the modal flank length?
# Helps diagnose whether variability is biological (small, e.g. polyproline variation)
# or technical (large, e.g. chimeras, truncated reads, boundary shifts)
flank_deviation <- flank_qc_data %>%
  dplyr::group_by(file_stem) %>%
  dplyr::mutate(modal = as.numeric(names(which.max(table(flank_total)))),
                dev   = abs(flank_total - modal)) %>%
  dplyr::summarise(
    n_reads       = n(),
    modal_bp      = modal[1],
    at_modal      = sum(dev == 0),
    within_10bp   = sum(dev > 0  & dev <= 10),
    within_50bp   = sum(dev > 10 & dev <= 50),
    over_50bp     = sum(dev > 50),
    .groups = "drop"
  ) %>%
  dplyr::mutate(dplyr::across(c(at_modal, within_10bp, within_50bp, over_50bp),
    ~ sprintf("%s (%.1f%%)", format(., big.mark = ","), 100 * . / n_reads),
    .names = "{.col}"))

DT::datatable(
  flank_deviation,
  caption  = "Combined flank length deviation from per-sample modal — large deviations suggest chimeras or alignment issues",
  options  = list(pageLength = 25, scrollX = TRUE, dom = 'ft'),
  rownames = FALSE,
  class    = 'cell-border stripe compact',
  colnames = c("Sample", "Reads", "Modal (bp)",
               "At modal", "1-10 bp off", "11-50 bp off", ">50 bp off")
) %>%
  DT::formatStyle(columns = colnames(flank_deviation), fontSize = '85%')
multipliers <- c(1.0, 1.5, 2.0, 3.0, 5.0, 10.0)
iqr_mult    <- if (!is.null(params$flank_iqr_multiplier)) params$flank_iqr_multiplier else 1.5

sweep_long <- dplyr::bind_rows(lapply(multipliers, function(mult) {
  flank_qc_data %>%
    dplyr::group_by(file_stem) %>%
    dplyr::mutate(
      fq1     = quantile(flank_total, 0.25),
      fq3     = quantile(flank_total, 0.75),
      outlier = flank_total < fq1 - mult * (fq3 - fq1) |
                flank_total > fq3 + mult * (fq3 - fq1)
    ) %>%
    dplyr::summarise(pct_removed = round(mean(outlier) * 100, 1), .groups = "drop") %>%
    dplyr::mutate(flank_iqr_multiplier = mult)
}))

# Wide table for export
sweep_df <- sweep_long %>%
  tidyr::pivot_wider(names_from  = flank_iqr_multiplier,
                     values_from = pct_removed,
                     names_prefix = "x") %>%
  dplyr::rename_with(~ paste0(.x, "_pct_removed"), -file_stem)

DT::datatable(
  sweep_df,
  caption  = "% reads removed (combined flank) per sample at each IQR multiplier — higher multiplier removes fewer reads",
  options  = list(pageLength = 25, dom = 'ft'),
  rownames = FALSE,
  class    = 'cell-border stripe compact'
) %>%
  DT::formatStyle(columns = colnames(sweep_df), fontSize = '85%')

# Sweep line plot
sweep_mean <- sweep_long %>%
  dplyr::group_by(flank_iqr_multiplier) %>%
  dplyr::summarise(mean_pct = mean(pct_removed), .groups = "drop")

mean_flank_iqr <- mean(flank_summary$flank_iqr_bp, na.rm = TRUE)

show_custom_sweep <- abs(iqr_mult - 1.5) > 0.01

p_sweep <- ggplot(sweep_long,
                  aes(x = flank_iqr_multiplier, y = pct_removed, group = file_stem)) +
  # Always: 1.5x reference vline (blue, matching CMC overlay)
  geom_vline(xintercept = 1.5, linetype = "dashed", colour = "#619CFF",
             linewidth = 0.8, alpha = 0.8) +
  # User multiplier vline (orange) only if different from 1.5
  { if (show_custom_sweep)
      geom_vline(xintercept = iqr_mult, linetype = "dashed", colour = "#E07B39",
                 linewidth = 0.8, alpha = 0.8) } +
  geom_line(colour = "grey60", linewidth = 0.6, alpha = 0.7) +
  geom_point(colour = "grey60", size = 1.5, alpha = 0.7) +
  geom_line(data = sweep_mean, aes(x = flank_iqr_multiplier, y = mean_pct, group = 1),
            colour = "red", linewidth = 1.2, alpha = 0.6, inherit.aes = FALSE) +
  geom_point(data = sweep_mean, aes(x = flank_iqr_multiplier, y = mean_pct),
             colour = "red", size = 3, shape = 18, alpha = 0.6, inherit.aes = FALSE) +
  annotate("text", x = max(multipliers), y = Inf,
           label = sprintf("Mean IQR: %.1f bp", mean_flank_iqr),
           hjust = 1, vjust = 1.3, size = 3, colour = "grey30") +
  # Labels on vlines
  annotate("text", x = 1.5, y = Inf,
           label = "1.5 x", hjust = -0.15, vjust = 1.3, size = 3, colour = "#619CFF") +
  { if (show_custom_sweep)
      annotate("text", x = iqr_mult, y = Inf,
               label = sprintf("%.1f x", iqr_mult),
               hjust = -0.15, vjust = 1.3, size = 3, colour = "#E07B39") } +
  scale_x_continuous(breaks = multipliers) +
  scale_y_continuous(limits = c(0, NA),
                     labels = scales::label_number(accuracy = 1)) +
  labs(title    = "Flank outlier filter: % reads removed vs IQR multiplier",
       subtitle = sprintf(
         "Grey = samples (n=%d) | Red = mean | Blue = 1.5 x IQR%s | flat = IQR \u2248 0",
         nrow(flank_summary),
         if (show_custom_sweep) sprintf(" | Orange = %.1f x IQR (user setting)", iqr_mult) else ""),
       x = "IQR multiplier (higher = removes fewer reads)",
       y = "% reads removed") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_sweep)

ggsave(
  filename = file.path(flank_dir, "flank_iqr_sweep.png"),
  plot     = p_sweep,
  width    = 6,
  height   = 4,
  dpi      = params$plot_dpi,
  bg       = "white"
)

# Interpretation note
mean_flank_iqr <- mean(flank_summary$flank_iqr_bp, na.rm = TRUE)
message(sprintf("Combined flank: mean IQR = %.1f bp, mean length = %.0f bp",
                mean_flank_iqr, mean(flank_summary$flank_mean_bp, na.rm = TRUE)))

if (mean_flank_iqr < 2) {
  message(sprintf(
    "Note: Very low combined flank IQR (%.1f bp). ", mean_flank_iqr),
    "The standard 1.5 x IQR filter will remove all non-modal reads regardless of multiplier. ",
    "See the sweep plot above — flat curves confirm this. ",
    "Consider leaving rm_flank_length_outliers = FALSE for this data.")
}

# Apply flank length filter to alignment_counts if enabled
# This filtered table feeds into clustering and all downstream modules
if (isTRUE(params$rm_flank_length_outliers)) {
  n_before <- nrow(alignment_counts)

  flank_fences <- flank_qc_data %>%
    dplyr::group_by(file_stem) %>%
    dplyr::summarise(
      fq1  = quantile(flank_total, 0.25, na.rm = TRUE),
      fq3  = quantile(flank_total, 0.75, na.rm = TRUE),
      fiqr = IQR(flank_total, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      fence_lo = fq1 - iqr_mult * fiqr,
      fence_hi = fq3 + iqr_mult * fiqr
    )

  alignment_counts <- alignment_counts %>%
    dplyr::mutate(
      left_length  = nchar(left),
      right_length = nchar(right),
      flank_total  = left_length + right_length
    ) %>%
    dplyr::left_join(flank_fences %>% dplyr::select(file_stem, fence_lo, fence_hi,
                                                     fiqr),
                     by = "file_stem") %>%
    dplyr::filter(
      fiqr == 0 |  # skip filter if IQR = 0 (fence would be at modal only)
      (flank_total >= fence_lo & flank_total <= fence_hi)
    ) %>%
    dplyr::select(-left_length, -right_length, -flank_total,
                  -fence_lo, -fence_hi, -fiqr)

  n_after   <- nrow(alignment_counts)
  n_removed <- n_before - n_after
  message(sprintf(
    "Flank filter applied (%s x IQR): removed %s reads (%.1f%%) from alignment_counts.",
    iqr_mult, format(n_removed, big.mark = ","),
    100 * n_removed / n_before))
  message("Filtered data propagates to count method comparison below and to Modules 4-7.")
} else {
  message("rm_flank_length_outliers = FALSE: all reads retained. Review the plots above before enabling.")
}
```

```{r t_flank_end, echo=FALSE}
timings["Flank QC"] <- proc.time()["elapsed"] - .t
```

## Read counts and count method comparison after flank QC

```{r flank_qc_post, echo=FALSE, fig.height=5, fig.width=12}
# Read count before / after flank filter — same style as Module 2 filtering_stages.png
n_before_flank <- n_reads_before_flank_filter   # captured before filter was applied
n_after_flank  <- nrow(alignment_counts)
n_removed_flank <- n_before_flank - n_after_flank
flank_filter_summary <- dplyr::tibble(
  file_stem        = unique(flank_qc_data$file_stem)
) %>%
  dplyr::left_join(
    flank_qc_data %>% dplyr::count(file_stem, name = "before_filter"),
    by = "file_stem"
  ) %>%
  dplyr::left_join(
    alignment_counts %>% dplyr::count(file_stem, name = "after_filter"),
    by = "file_stem"
  ) %>%
  dplyr::mutate(after_filter = dplyr::coalesce(after_filter, 0L),
                reads_removed = before_filter - after_filter,
                pct_removed   = round(100 * reads_removed / before_filter, 1))

DT::datatable(
  flank_filter_summary,
  caption  = "Read counts before and after flank length filter",
  options  = list(pageLength = 25, scrollX = TRUE, dom = 'ft'),
  rownames = FALSE,
  class    = 'cell-border stripe compact'
) %>%
  DT::formatStyle(columns = colnames(flank_filter_summary), fontSize = '85%')

# Connected dot plot
n_samp_flank <- nrow(flank_filter_summary)
point_size   <- if (n_samp_flank <= 50) 2.5 else 1.5
point_alpha  <- if (n_samp_flank <= 50) 0.8 else max(0.2, 1 - n_samp_flank / 500)
line_alpha   <- if (n_samp_flank <= 50) 0.5 else max(0.1, 0.5 - n_samp_flank / 1000)

flank_filter_long <- flank_filter_summary %>%
  tidyr::pivot_longer(cols = c(before_filter, after_filter),
                      names_to = "stage", values_to = "read_count") %>%
  dplyr::mutate(
    stage     = factor(stage,
                       levels = c("before_filter", "after_filter"),
                       labels = c("Before\nflank filter", "After\nflank filter")),
    stage_num = as.numeric(stage),
    changed   = reads_removed > 0
  )

set.seed(123)
jitter_flank <- data.frame(
  file_stem = unique(flank_filter_long$file_stem),
  x_offset  = runif(n_samp_flank, -0.12, 0.12)
)
flank_filter_long <- flank_filter_long %>%
  dplyr::left_join(jitter_flank, by = "file_stem") %>%
  dplyr::mutate(x_pos = stage_num + x_offset)

p_flank_filter <- ggplot(flank_filter_long) +
  geom_boxplot(aes(x = stage_num, y = read_count, group = stage_num),
               outlier.shape = NA, width = 0.3, fill = "grey90", colour = "grey40") +
  geom_line(aes(x = x_pos, y = read_count, group = file_stem, colour = changed),
            linewidth = 0.4, alpha = line_alpha) +
  geom_point(aes(x = x_pos, y = read_count, colour = changed),
             size = point_size, alpha = point_alpha) +
  stat_summary(aes(x = stage_num, y = read_count, group = stage_num),
               fun = mean, geom = "point", shape = 18,
               colour = "red", size = 4, alpha = 0.6) +
  scale_colour_manual(
    values = c("FALSE" = "grey50", "TRUE" = "#E07B39"),
    labels = c("FALSE" = "Unchanged", "TRUE" = "Reads removed"),
    name   = NULL
  ) +
  scale_x_continuous(breaks = 1:2,
                     labels = c("Before\nflank filter", "After\nflank filter"),
                     limits = c(0.6, 2.4)) +
  scale_y_log10(labels = scales::comma) +
  labs(title    = "Read counts before and after flank length filter",
       subtitle = if (n_removed_flank == 0) {
         "No reads removed — flank filter not applied or IQR = 0 for all samples"
       } else {
         sprintf("%s reads removed cohort-wide (%.1f%%) using %.1f x IQR threshold",
                 format(n_removed_flank, big.mark = ","),
                 100 * n_removed_flank / n_before_flank,
                 iqr_mult)
       },
       x = NULL, y = "Read count (log10 scale)") +
  theme_bw()

if (!any(flank_filter_summary$reads_removed > 0))
  p_flank_filter <- p_flank_filter + guides(colour = "none")

print(p_flank_filter)
ggsave(file.path(flank_dir, "flank_filter_reads.png"), p_flank_filter,
       width = 6, height = 5, dpi = params$plot_dpi, bg = "white")

# Count method comparison with IQR filter series overlaid
# Always computed for all three series regardless of whether filter is applied.
# If no filter applied, the series will overlap — that itself is informative.
cmc_all <- compute_cmc(alignment_counts)   # current (post-filter if filter on, else all reads)

# Compute 1.5x and custom filtered versions from the unfiltered flank_qc_data
compute_cmc_filtered <- function(df, mult) {
  df %>%
    dplyr::group_by(file_stem) %>%
    dplyr::mutate(
      fq1  = quantile(flank_total, 0.25, na.rm = TRUE),
      fq3  = quantile(flank_total, 0.75, na.rm = TRUE),
      fiqr = fq3 - fq1,
      keep = fiqr == 0 | (flank_total >= fq1 - mult * fiqr & flank_total <= fq3 + mult * fiqr)
    ) %>%
    dplyr::filter(keep) %>%
    dplyr::ungroup() %>%
    compute_cmc()
}

cmc_ref    <- compute_cmc_filtered(flank_qc_data, 1.5)
cmc_custom <- if (abs(iqr_mult - 1.5) > 0.01) compute_cmc_filtered(flank_qc_data, iqr_mult) else NULL

# Build labelled long table for overlaid plot
series_list <- list(
  dplyr::mutate(compute_cmc(flank_qc_data), series = "All reads"),
  dplyr::mutate(cmc_ref,                    series = "1.5 x IQR filtered")
)
if (!is.null(cmc_custom))
  series_list[[3]] <- dplyr::mutate(cmc_custom,
                                    series = sprintf("%.1f x IQR filtered", iqr_mult))

cmc_overlay <- dplyr::bind_rows(series_list) %>%
  dplyr::mutate(series = factor(series, levels = sapply(series_list, function(x) x$series[1])))

series_colours <- c("All reads"          = "grey40",
                    "1.5 x IQR filtered" = "#619CFF")
if (!is.null(cmc_custom))
  series_colours[sprintf("%.1f x IQR filtered", iqr_mult)] <- "#E07B39"

# Axis limits across all series
ax_all_min <- min(c(cmc_overlay$mean_repeat_count_full,
                    cmc_overlay$mean_repeat_count_match), na.rm = TRUE)
ax_all_max <- max(c(cmc_overlay$mean_repeat_count_full,
                    cmc_overlay$mean_repeat_count_match), na.rm = TRUE)
ax_all_pad <- (ax_all_max - ax_all_min) * 0.05

p_cmc_overlay <- ggplot(cmc_overlay,
                         aes(x = mean_repeat_count_full, y = mean_repeat_count_match,
                             colour = series, fill = series)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.1, linewidth = 0.8, formula = y ~ x) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_colour_manual(values = series_colours, name = NULL) +
  scale_fill_manual(values  = series_colours, name = NULL) +
  scale_x_continuous(limits = c(ax_all_min - ax_all_pad, ax_all_max + ax_all_pad)) +
  scale_y_continuous(limits = c(ax_all_min - ax_all_pad, ax_all_max + ax_all_pad)) +
  labs(title    = "Count method comparison: effect of flank IQR filter",
       subtitle = sprintf(
         "Dashed = identity | Shaded = 95%% CI | n = %d samples%s",
         nrow(cmc_all),
         if (is.null(cmc_custom)) "" else
           sprintf(" | Orange = %.1f x IQR (user setting)", iqr_mult)),
       x = "Mean repeat count (repeat_count_full)",
       y = "Mean repeat count (repeat_count_match)") +
  theme_bw() +
  theme(legend.position = "top") +
  coord_fixed()

print(p_cmc_overlay)
ggsave(file.path(flank_dir, "count_method_comparison_post_filter.png"), p_cmc_overlay,
       width = 7, height = 7, dpi = params$plot_dpi, bg = "white")

if (n_removed_flank == 0) {
  message("Flank filter not applied — all three series overlap (this is expected).")
} else {
  message(sprintf("Overlay shows effect of filtering on count method agreement (%s reads removed).",
                  format(n_removed_flank, big.mark = ",")))
}

# Update count_method_comparison to post-filter version for Excel export
count_method_comparison <- cmc_all
```


---

# Visualise distributions

## Repeat count histograms

Per-sample histograms are saved to `plots/repeat_histogram_by_sample/` (one file per sample). They are not shown inline as they do not scale to large cohorts.

```{r plot_histograms, echo=FALSE, eval=params$plot_histograms}
if (params$plot_histograms) {
  n_samples <- length(unique(alignment_counts$file_stem))

  # Per-sample histograms (gated by plot_per_sample)
  if (params$plot_per_sample) {
    histogram_by_sample_dir <- file.path(plots_dir, "repeat_histogram_by_sample")
    dir.create(histogram_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

    message("Generating individual histograms for ", n_samples, " samples...")
    for (sample in unique(alignment_counts$file_stem)) {
      plot_path <- file.path(histogram_by_sample_dir, paste0(sample, "-repeat_histogram.png"))
      if (file.exists(plot_path) && params$resume) next

      sample_data <- alignment_counts %>% filter(file_stem == sample)

      p_histogram_sample <- ggplot(sample_data, aes(x = repeat_count)) +
        geom_histogram(binwidth = params$histogram_binwidth,
                       fill = "#619CFF", colour = "black", alpha = 0.7) +
        scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
        labs(title = paste("Repeat count distribution -", sample),
             x     = "Repeat count",
             y     = "Number of reads") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

      ggsave(
        filename = plot_path,
        plot     = p_histogram_sample,
        width    = 10,
        height   = 6,
        dpi      = params$plot_dpi,
        bg       = "white"
      )
    }

    message("Per-sample histograms saved to: ", histogram_by_sample_dir)
  }
}
```

## Distribution visualisation

**Important:** Samples are typically **bimodal** (two alleles with different repeat counts). Mean ± SD is misleading for bimodal distributions — it falls between the peaks and doesn't represent either allele. Use the heatmap and ridge plots below to visually identify alleles. Formal allele calling is in Module 4.

```{r plot_distributions, echo=FALSE, fig.height=6, fig.width=12}
n_samples <- length(unique(alignment_counts$file_stem))

# Build ordered sample index (used by heatmap and lookup table)
repeat_summary_scatter <- alignment_counts %>%
  group_by(file_stem) %>%
  dplyr::summarise(
    median_rc = median(repeat_count, na.rm = TRUE),
    q25_rc    = quantile(repeat_count, 0.25, na.rm = TRUE),
    q75_rc    = quantile(repeat_count, 0.75, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  dplyr::arrange(median_rc) %>%
  dplyr::mutate(sample_idx = seq_len(n()))

# --- Heatmap: distribution across all samples ---
# Samples on x (indexed by median order — no crowded labels)
# Fill = % of reads in each repeat count bin, normalised per sample
heatmap_bin_width <- max(1, params$histogram_binwidth)

heatmap_data <- alignment_counts %>%
  dplyr::mutate(rc_bin = floor(repeat_count / heatmap_bin_width) * heatmap_bin_width) %>%
  dplyr::count(file_stem, rc_bin) %>%
  dplyr::group_by(file_stem) %>%
  dplyr::mutate(frac = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(
    repeat_summary_scatter %>% dplyr::select(file_stem, sample_idx),
    by = "file_stem"
  )
# Note: tiles are only drawn where reads exist — white space = no reads at that
# repeat count for that sample, which is informative and should remain white.

p_heatmap <- ggplot(heatmap_data, aes(x = sample_idx, y = rc_bin, fill = frac)) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma", name = "% of reads",
                       labels = scales::percent_format(accuracy = 0.1)) +
  scale_x_continuous(
    breaks = if (n_samples <= 20) seq_len(n_samples) else
               scales::pretty_breaks(n = 5)(seq_len(n_samples)),
    expand = c(0, 0)
  ) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10), expand = c(0, 0)) +
  labs(title    = "Repeat count distribution heatmap",
       subtitle = sprintf(
         "Samples ordered by median repeat count  |  Fill = %% of reads  |  n = %d samples  |  See heatmap_sample_index in repeat_detection.xlsx",
         n_samples),
       x        = "Sample index (ordered by median repeat count)",
       y        = "Repeat count") +
  theme_bw() +
  theme(panel.grid  = element_blank(),
        axis.text.x = element_text(size = 8))

print(p_heatmap)

ggsave(
  filename  = file.path(plots_dir, "repeat_distribution_heatmap.png"),
  plot      = p_heatmap,
  width     = min(50, max(10, n_samples * 0.12 + 6)),
  height    = 6,
  dpi       = params$plot_dpi,
  bg        = "white",
  limitsize = FALSE
)

# Per-sample violin plots (gated by plot_per_sample)
if (params$plot_per_sample) {
  violin_by_sample_dir <- file.path(plots_dir, "repeat_violin_by_sample")
  dir.create(violin_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

  for (sample in unique(alignment_counts$file_stem)) {
    plot_path <- file.path(violin_by_sample_dir, paste0(sample, "-repeat_violin.png"))
    if (file.exists(plot_path) && params$resume) next

    sample_data <- alignment_counts %>% filter(file_stem == sample)

    p_violin_sample <- ggplot(sample_data, aes(x = 1, y = repeat_count)) +
      geom_violin(fill = "#619CFF", alpha = 0.6, scale = "width", width = 0.8) +
      geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8, fill = "white") +
      stat_summary(fun = median, geom = "point", size = 4, colour = "red", alpha = 0.6) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 15)) +
      labs(title    = paste("Repeat count distribution -", sample),
           subtitle = "Violin = distribution shape | Boxplot = quartiles | Red dot = median",
           x        = "",
           y        = "Repeat count") +
      theme_minimal() +
      theme(axis.text.x  = element_blank(),
            axis.ticks.x = element_blank())

    ggsave(
      filename = plot_path,
      plot     = p_violin_sample,
      width    = 6,
      height   = 8,
      dpi      = params$plot_dpi,
      bg       = "white"
    )
  }
  message("Per-sample violin plots saved to: ", violin_by_sample_dir)
}
```

```{r plot_ridges, echo=FALSE, fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
# Ridge plot - excellent for comparing distributions across samples
if (requireNamespace("ggridges", quietly = TRUE)) {
  library(ggridges)

  n_samples <- length(unique(alignment_counts$file_stem))

  # Parameters for point overlay (match analysis.Rmd ridge convention)
  ridge_show_points        <- if (!is.null(params$ridge_show_points))        params$ridge_show_points        else TRUE
  ridge_point_size         <- if (!is.null(params$ridge_point_size))         params$ridge_point_size         else 0.5
  ridge_point_alpha        <- if (!is.null(params$ridge_point_alpha))        params$ridge_point_alpha        else 0.15
  ridge_point_jitter_height <- if (!is.null(params$ridge_point_jitter_height)) params$ridge_point_jitter_height else 0.1

  p_ridge_all <- ggplot(alignment_counts,
                        aes(x = repeat_count, y = file_stem, fill = file_stem))

  if (isTRUE(ridge_show_points)) {
    p_ridge_all <- p_ridge_all +
      stat_density_ridges(
        geom             = "density_ridges",
        alpha            = 0.7, scale = 2,
        jittered_points  = TRUE,
        point_size       = ridge_point_size,
        point_alpha      = ridge_point_alpha,
        position         = position_points_jitter(height = ridge_point_jitter_height)
      )
  } else {
    p_ridge_all <- p_ridge_all +
      geom_density_ridges(alpha = 0.7, scale = 2)
  }

  p_ridge_all <- p_ridge_all +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
    labs(title    = "Repeat count distributions across samples",
         subtitle = sprintf("Each ridge = full distribution; peaks indicate alleles  |  n = %d samples",
                            n_samples),
         x        = "Repeat count",
         y        = "Sample") +
    theme_minimal() +
    theme(legend.position = "none")

  print(p_ridge_all)

  plot_height <- min(50, max(8, n_samples * 0.6))

  ggsave(
    filename = file.path(plots_dir, "repeat_density.png"),
    plot = p_ridge_all,
    width = 12,
    height = plot_height,
    dpi = params$plot_dpi,
    bg = "white",
    limitsize = FALSE
  )

  # Per-sample density plots (gated by plot_per_sample)
  if (params$plot_per_sample) {
    density_by_sample_dir <- file.path(plots_dir, "repeat_density_by_sample")
    dir.create(density_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

    for (sample in unique(alignment_counts$file_stem)) {
      plot_path <- file.path(density_by_sample_dir, paste0(sample, "-repeat_density.png"))
      if (file.exists(plot_path) && params$resume) next

      sample_data <- alignment_counts %>% filter(file_stem == sample)

      p_density_sample <- ggplot(sample_data, aes(x = repeat_count)) +
        geom_density(fill = "#619CFF", alpha = 0.7, colour = "black", linewidth = 1) +
        scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
        labs(title = paste("Repeat count density -", sample),
             subtitle = "Peaks indicate alleles",
             x = "Repeat count",
             y = "Density") +
        theme_minimal()

      ggsave(
        filename = plot_path,
        plot = p_density_sample,
        width = 10,
        height = 6,
        dpi = params$plot_dpi,
        bg = "white"
      )
    }
  }

  message("Repeat detection plots exported to: ", plots_dir)

} else {
  message("Install 'ggridges' package for ridge plot: install.packages('ggridges')")
}
```

```{r summary_table_clean, echo=FALSE}
# Simple summary statistics (leave allele detection for Module 4 clustering)
clean_summary <- alignment_counts %>%
  dplyr::group_by(file_stem) %>%
  dplyr::summarise(
    n_reads = n(),
    median_repeats = median(repeat_count, na.rm = TRUE),
    mean_repeats = mean(repeat_count, na.rm = TRUE),
    mode_repeats = as.numeric(names(sort(table(repeat_count), decreasing = TRUE)[1])),
    min_repeats = min(repeat_count, na.rm = TRUE),
    max_repeats = max(repeat_count, na.rm = TRUE),
    range_repeats = max_repeats - min_repeats,
    sd_repeats = sd(repeat_count, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  clean_summary,
  caption = "Repeat count summary by sample",
  colnames = c("Sample", "N reads", "Median", "Mean", "Mode", "Min", "Max", "Range", "SD"),
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(clean_summary), fontSize = '85%') %>%
  DT::formatRound(columns = c('median_repeats', 'mean_repeats', 'mode_repeats', 
                               'min_repeats', 'max_repeats', 'range_repeats', 'sd_repeats'), 
                  digits = 1)
```

---

# Save outputs

Save processed data for downstream modules.

```{r save_outputs, echo=FALSE}
# Slim counts table: join keys + the three columns repositioned by Module 3
alignment_counts_slim <- alignment_counts %>%
  dplyr::select(file_stem, qname,
                left, mid, right,
                repeat_count, repeat_count_full,
                repeat_count_match, repeat_count_tracts)

# Compile module output
module3_output <- list(
  alignment_counts_slim = alignment_counts_slim,
  count_summary         = clean_summary,
  tract_summary         = tract_summary,
  parameters            = params
)

# Export summaries to Excel
repeat_excel_path <- file.path(repeat_dir, "repeat_detection.xlsx")

heatmap_sample_index <- clean_summary %>%
  dplyr::select(file_stem, median_repeats) %>%
  dplyr::arrange(median_repeats) %>%
  dplyr::mutate(sample_index = row_number()) %>%
  dplyr::select(sample_index, file_stem, median_repeats)

write_excel_function(
  named_list = list(
    "count_summary"           = clean_summary,
    "tract_detection"         = tract_summary,
    "count_method_comparison" = count_method_comparison,
    "heatmap_sample_index"    = heatmap_sample_index,
    "flank_summary"           = if (exists("flank_summary"))   flank_summary   else NULL,
    "flank_iqr_sweep"         = if (exists("sweep_df"))        sweep_df        else NULL
  ) %>% { Filter(Negate(is.null), .) },
  path = repeat_excel_path
)
message("Repeat summaries exported to: ", repeat_excel_path)

# Save complete module output
output_path <- file.path(results_dir, "03_repeat_detection.RData")
save(module3_output, file = output_path)
message("Module 3 outputs saved to: ", output_path)
```

```{r export_read_counts, echo=FALSE}
if (isTRUE(params$export_read_counts)) {

  rc_dir <- file.path(params$dir_out, "03_repeat_detection", "read_counts")
  if (!dir.exists(rc_dir)) dir.create(rc_dir, recursive = TRUE)

  samples <- unique(alignment_counts$file_stem)

  invisible(lapply(samples, function(s) {
    out <- alignment_counts %>%
      dplyr::filter(file_stem == s) %>%
      dplyr::select(
        read_id             = qname,
        repeat_count_full,
        repeat_count_match,
        repeat_count_tracts
      )
    out_path <- file.path(rc_dir, paste0(s, ".tsv.gz"))
    data.table::fwrite(out, file = out_path, sep = "\t", compress = "gzip")
  }))

  message(sprintf(
    "Per-read repeat counts exported: %d samples → %s",
    length(samples), rc_dir
  ))

} else {
  message("export_read_counts = FALSE: per-read count files not written.")
}
```

```{r cleanup, echo=FALSE}
if (params$remove_intermediate) {
  rm(alignment_split)
  gc()
}
```

**Module 3 Complete!**

Results saved to: `r output_path`

- **`r format(nrow(alignment_counts), big.mark = ",")`** reads with repeat counts
- **`r n_distinct(alignment_counts$file_stem)`** samples analysed
- Mean repeat count across samples: **`r round(mean(clean_summary$mean_repeats), 1)`**

---

# Module diagnostics

## Timing

```{r timing_table, echo=FALSE}
timing_df <- build_timing_table(timings)
DT::datatable(
  timing_df,
  caption = "Section timing — Module 3",
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Section", "Seconds", "Elapsed (mm:ss)", "Cumulative (mm:ss)")
) %>%
  DT::formatStyle("section", target = "row",
                  fontWeight = DT::styleEqual("Total", "bold"))
```

## Output objects

```{r manifest_table, echo=FALSE}
manifest <- data.frame(
  object = names(module3_output),
  description = c(
    "Per-read repeat counts with repositioned left/mid/right sequences (slim — join with Module 2 for full data)",
    "Per-sample repeat count summary statistics (mean, median, SD, detection rate)",
    "Per-sample repeat tract detection rates and tract structure statistics",
    "Pipeline parameters used in this module run"
  ),
  class = sapply(module3_output, function(x) if (is.null(x)) "NULL" else class(x)[1]),
  size = sapply(module3_output, function(x) {
    format_size(if (is.null(x)) 0 else as.numeric(object.size(x)))
  }),
  bytes = sapply(module3_output, function(x) {
    if (is.null(x)) 0L else as.numeric(object.size(x))
  }),
  stringsAsFactors = FALSE
)

DT::datatable(
  manifest[, c("object", "description", "class", "size")],
  caption = sprintf("Objects saved to module3_output — RData file: %s on disk (%s)",
                    format_size(file.size(output_path)), output_path),
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Object", "Description", "Class", "In-memory size")
)
```

```{r diagnostics_export, echo=FALSE}
write_excel_function(
  named_list = list("timing" = timing_df, "manifest" = manifest),
  path = repeat_excel_path,
  overwrite_sheet = TRUE
)
message(sprintf("Module 3 complete — total time: %s  |  RData: %s",
                format_elapsed(sum(timings)), format_size(file.size(output_path))))
```

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```
