---
title: "Duke pipeline"
subtitle: "Module 4: Allele Calling"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

```{r set_tab_title, echo=FALSE, results='asis'}
cat('<script>document.title = "Duke | Module 4: Allele Calling";</script>')
```

```{r load_packages, message=FALSE, warning=FALSE, echo=FALSE}
# Required packages
required_packages <- c("Biostrings", "DECIPHER", "mclust", "cluster", 
                      "pbapply", "dplyr", "tidyr", "data.table", 
                      "stringr", "ggplot2", "openxlsx", "DT",
                      "pbmcapply")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}
```

```{r timing_init, echo=FALSE}
module_start_time <- proc.time()["elapsed"]
timings <- c()
```

# Overview

This module performs **clustering** to assign reads to distinct alleles, generates **consensus sequences**, and **calls variants** against the reference.

**Input:** Repeat count data from Module 3

**Processing:**

1. Cluster reads by repeat length, haplotype, or both
2. Assign cluster numbers ordered by abundance (cluster 1 = most reads)
3. Generate consensus sequences for each cluster
4. Call variants in consensus sequences (if reference provided)
5. Calculate cluster-specific summary statistics

**Outputs:**

- `alignment_clustered_slim` — cluster assignments per read (join keys + cluster number; join with Module 2/3 for full data)
- Cluster summary statistics table (`cluster_summary`)
- Cluster count per sample plot (`cluster_count_per_sample.png`)
- Per-sample violin and scatter plots (`violin_by_sample/`, `scatter_by_sample/`)
- Consensus sequences — Excel (`consensus.xlsx`) and FASTA (full, left, mid, right per cluster)
- Variant calls — Excel (`variant.xlsx`), CSV, and VCF files (per sample and combined)
- `04_allele_calling.RData` for downstream modules

**Key Parameters:**

- `cluster_by`: Clustering strategy — `"repeat"` (by repeat count), `"haplotype"` (by flanking sequence variants), or `"both"` (two-step)
- `repeat_cluster_max`: Maximum number of repeat-length clusters to detect per sample
- `haplotype_cluster_max`: Maximum number of haplotype clusters to detect
- `cluster_consensus`: Generate consensus sequences (`TRUE`/`FALSE`)
- `consensus_threshold`: Minimum agreement for consensus base call (0.5 = majority rule)
- `consensus_downsample`: Maximum reads used per cluster for consensus alignment (performance vs accuracy tradeoff)
- `call_variants`: Call variants in consensus sequences against the reference (`TRUE`/`FALSE`)

---

```{r setup_directories, echo=FALSE}
# Derive directory paths from params$dir_out
results_dir <- file.path(params$dir_out, "module_data")
temp_dir <- file.path(params$dir_out, "temp")

# Module 4 output directories
consensus_dir <- file.path(params$dir_out, file.path("04_allele_calling", "consensus"))
variants_dir <- file.path(params$dir_out, file.path("04_allele_calling", "variants"))

# Create directories if needed
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(temp_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(consensus_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(variants_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r load_data, echo=FALSE, results='asis'}
# Load Module 2 results (alignment_split with sequences)
module2_path <- file.path(results_dir, "02_alignment.RData")
safe_load_module(module2_path, module_num = 2, envir = environment())
alignment_split <- module2_output$alignment_split
message("Loaded ", format(nrow(alignment_split), big.mark = ","),
        " reads from Module 2 (alignment_split)")

# Load Module 3 results (slim repeat counts)
module3_path <- file.path(results_dir, "03_repeat_detection.RData")
safe_load_module(module3_path, module_num = 3, envir = environment())
counts_slim <- module3_output$alignment_counts_slim
message("Loaded ", format(nrow(counts_slim), big.mark = ","),
        " rows from Module 3 (alignment_counts_slim)")

# Join: start from alignment_split, replace left/mid/right with Module 3
# repositioned versions, and add repeat count columns.
# alignment_split$left/mid/right reflect Module 2 alignment boundaries;
# counts_slim$left/mid/right have been trimmed to the detected repeat tract.
alignment_counts <- alignment_split %>%
  dplyr::select(-left, -mid, -right) %>%
  dplyr::left_join(counts_slim, by = c("file_stem", "qname"))

n_unmatched <- sum(is.na(alignment_counts$repeat_count))
if (n_unmatched > 0) {
  message("Note: ", format(n_unmatched, big.mark = ","),
          " reads have no repeat count (NA after join)")
}
message("Combined: ", format(nrow(alignment_counts), big.mark = ","), " reads ready for clustering")

# Format sample list for HTML
samples   <- unique(alignment_counts$file_stem)
n_samples <- length(samples)
if (n_samples <= 10) {
  cat("\n\n**", n_samples, " samples:**\n\n",
      paste(samples, collapse = ", "), "\n\n", sep = "")
} else {
  cat("\n\n**", n_samples, " samples:**\n\n",
      paste(head(samples, 10), collapse = ", "),
      ", ... and ", n_samples - 10, " more\n\n", sep = "")
}

# Load reference sequence for variant calling (already loaded Module 2 above)
reference_seq <- NULL
if (params$call_variants) {
  if (!is.null(module2_output$ref_full_sequence)) {
    reference_seq <- module2_output$ref_full_sequence
    message("Reference sequence: ", nchar(reference_seq), " bp")
  } else {
    message("Warning: Module 2 output does not contain full reference sequence — variant calling will be skipped")
  }
}
```

---

# Clustering

Assign reads to clusters (alleles) using the strategy: **`r params$cluster_by`**

```{r t_cluster_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r clustering, echo=FALSE}
# Check resume
cache_path <- file.path(temp_dir, "04_clustering.RData")

if (params$resume && safe_load_cache(cache_path, environment())) {
  message("Loading cached clustering results...")
  message("Clustering loaded from cache")

} else {
  
  if (params$cluster) {
    
    message("Performing clustering...")
    message("Strategy: ", params$cluster_by)
    if (params$cluster_by %in% c("haplotype", "both")) {
      message("  Haplotype region: ", params$haplotype_region)
      message("  Haplotype method: ", params$haplotype_method)
    }
    
    # Split by sample for per-sample clustering
    alignment_list <- split(alignment_counts, alignment_counts$file_stem)
    n_samples <- length(alignment_list)
    
    message("\nClustering ", n_samples, " samples...")
    
    # Apply clustering to each sample using apply_fn (parallelised if threads > 1)
    alignment_clustered_list <- apply_fn(seq_along(alignment_list), function(i) {
      sample_data <- alignment_list[[i]]
      sample_name <- unique(sample_data$file_stem)
      
      # Check for cached result
      sample_cache_file <- file.path(temp_dir, paste0("cluster_", sample_name, ".RData"))
      if (params$resume && safe_load_cache(sample_cache_file, environment())) {
        return(sample_clustered)
      }
      
      # Apply clustering using the wrapper function
      cluster_result <- apply_clustering(
        alignment_data = sample_data,
        cluster_by = params$cluster_by,
        repeat_column = params$repeat_count_method,
        sequence_column = "query_sequence",
        left_column = "left",
        right_column = "right",
        haplotype_cluster_max = params$haplotype_cluster_max,
        repeat_cluster_max = params$repeat_cluster_max,
        repeat_cluster_downsample = params$repeat_cluster_downsample,
        haplotype_cluster_downsample = params$haplotype_cluster_downsample,
        haplotype_method = params$haplotype_method,
        haplotype_region = params$haplotype_region,
        haplotype_trim_length = params$haplotype_trim_length
      )
      
      # Handle result - can be vector or list with metadata
      if (is.list(cluster_result) && !is.null(cluster_result$cluster_number)) {
        # "both" strategy returns metadata
        sample_data$cluster_number <- cluster_result$cluster_number
        sample_data$repeat_cluster <- cluster_result$repeat_cluster
        sample_data$haplotype_cluster <- cluster_result$haplotype_cluster
      } else {
        # Simple strategies return vector
        sample_data$cluster_number <- cluster_result
        sample_data$repeat_cluster <- NA_character_
        sample_data$haplotype_cluster <- NA_character_
      }
      
      # Save per-sample cache
      sample_clustered <- sample_data
      save(sample_clustered, file = sample_cache_file)
      
      return(sample_clustered)
    })
    
    # Extract results (follows pattern from modules 1-3)
    alignment_clustered_list <- extract_apply_fn_result(alignment_clustered_list)
    
    # Recombine into single data frame
    alignment_clustered <- bind_rows(alignment_clustered_list)
    
    # Relocate cluster_number column
    alignment_clustered <- alignment_clustered %>%
      relocate(cluster_number, .after = all_of(params$repeat_count_method))
    
    # Summary statistics
    total_clusters <- alignment_clustered %>%
      group_by(file_stem) %>%
      summarise(n_clusters = length(unique(cluster_number[!is.na(cluster_number)])),
                .groups = "drop")
    
    message("Clustering complete!")
    message("  Total samples: ", n_samples)
    message("  Total clusters: ", sum(total_clusters$n_clusters))
    message("  Clusters per sample: ", 
            sprintf("%.1f ± %.1f (range: %d-%d)",
                    mean(total_clusters$n_clusters),
                    sd(total_clusters$n_clusters),
                    min(total_clusters$n_clusters),
                    max(total_clusters$n_clusters)))
    
  } else {
    
    # No clustering - assign all reads to cluster 1
    message("Clustering disabled - assigning all reads to cluster 1")
    alignment_clustered <- alignment_counts %>%
      mutate(cluster_number = 1,
             repeat_cluster = NA_character_,
             haplotype_cluster = NA_character_) %>%
      relocate(cluster_number, .after = all_of(params$repeat_count_method))
    
  }
  
  # Cache results
  save(alignment_clustered, file = cache_path)
  message("Clustering results cached to: ", cache_path)
}
```

```{r t_cluster_end, echo=FALSE}
timings["Clustering"] <- proc.time()["elapsed"] - .t
```

**Clustering Strategy Explanation:**

```{r clustering_explanation, echo=FALSE, results='asis'}
# Handle both vector and single value inputs
if (length(params$cluster_by) == 2) {
  # Two-step clustering
  cat("**Combined Clustering:**\n\n")
  cat("- Clustering order: ", paste(params$cluster_by, collapse = " → "), "\n")
  cat("- Step 1: Group by ", params$cluster_by[1], "\n")
  cat("- Step 2: Within each group, cluster by ", params$cluster_by[2], "\n")
  cat("- Captures both sequence variants AND repeat length differences\n")
  if (identical(params$cluster_by, c("repeat", "haplotype"))) {
    cat("- **Repeat-first**: More robust to sequencing errors (recommended for long reads)\n")
  } else {
    cat("- **Haplotype-first**: Traditional approach\n")
  }
} else if (length(params$cluster_by) == 1) {
  # Single method clustering
  if (params$cluster_by == "repeat") {
    cat("**Repeat Length Clustering:**\n\n")
    cat("- Reads grouped solely by repeat count\n")
    cat("- Standard approach for detecting alleles by size\n")
    cat("- Best when alleles differ primarily in repeat length\n")
  } else if (params$cluster_by == "haplotype") {
    cat("**Haplotype Clustering:**\n\n")
    cat("- Reads grouped by variants in flanking regions (", params$haplotype_region, ")\n")
    cat("- Uses ", params$haplotype_method, " distance to measure similarity\n")
    cat("- Best for detecting alleles that differ by SNPs/indels, not just repeat length\n")
  } else if (params$cluster_by == "none") {
    cat("**No Clustering:**\n\n")
    cat("- All reads combined into single consensus\n")
    cat("- Use for homozygous samples or when clustering not needed\n")
  }
}
```

---

## Cluster summary

Statistics for each cluster per sample.

```{r cluster_summary, echo=FALSE}
# Calculate cluster summary statistics
cluster_summary <- alignment_clustered %>%
  group_by(file_stem, cluster_number) %>%
  summarise(
    read_count = n(),
    repeat_cluster = dplyr::first(repeat_cluster),      # Use first() explicitly
    haplotype_cluster = dplyr::first(haplotype_cluster), # Use first() explicitly
    repeat_count_mode = {
      # Mode: most common value (best for discrete integer data)
      counts <- !!sym(params$repeat_count_method)
      counts <- counts[!is.na(counts)]
      if (length(counts) == 0) {
        NA_real_
      } else {
        ux <- unique(counts)
        tab <- tabulate(match(counts, ux))
        ux[which.max(tab)]
      }
    },
    repeat_count_median = median(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_mean = mean(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_sd = sd(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_min = min(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_max = max(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_iqr = IQR(!!sym(params$repeat_count_method), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(file_stem, cluster_number)

# Renumber clusters by read count within each sample (cluster 1 = most abundant)
cluster_summary <- cluster_summary %>%
  group_by(file_stem) %>%
  arrange(file_stem, desc(read_count)) %>%
  mutate(
    cluster_number_original = cluster_number,  # Keep original for reference
    cluster_number = row_number(),             # Renumber: 1 = most abundant
    read_proportion = read_count / sum(read_count)
  ) %>%
  ungroup() %>%
  relocate(cluster_number_original, .after = cluster_number) %>%
  relocate(read_proportion, .after = read_count)

# Also renumber in alignment_clustered to match
# Note: rename() avoided here due to S4 dispatch conflict with Bioconductor packages;
# column renaming done inside select() instead.
cluster_mapping <- cluster_summary %>%
  dplyr::select(file_stem,
                cluster_number_old = cluster_number_original,
                cluster_number_new = cluster_number)

alignment_clustered <- alignment_clustered %>%
  dplyr::left_join(cluster_mapping,
                   by = c("file_stem", "cluster_number" = "cluster_number_old")) %>%
  dplyr::select(-cluster_number) %>%
  dplyr::rename(cluster_number = cluster_number_new) %>%
  dplyr::relocate(cluster_number, .after = all_of(params$repeat_count_method))

message("Clusters renumbered by abundance (cluster 1 = most reads)")
message("Updated cluster numbers in both cluster_summary and alignment_clustered")

# Display table with DT::datatable for better handling of large tables
datatable(cluster_summary,
          options = list(
            pageLength = 15,
            autoWidth = TRUE,
            scrollX = TRUE,
            initComplete = JS("function(settings, json) {",
                              "$(this.api().table().header()).css({'font-size': '10px', 'font-family': 'Arial'});",
                              "}")
          ),
          rownames = FALSE,
          class = 'cell-border stripe',
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 12px; font-weight: bold;',
            "Cluster Summary (cluster 1 = most abundant, R = repeat cluster, H = haplotype cluster)"
          ),
          colnames = c("Sample", "Cluster", "Original", "Reads", "Proportion",
                       "R", "H",
                       "Mode", "Median", "Mean", "SD", "Min", "Max", "IQR")
) %>%
  formatStyle(columns = names(cluster_summary), fontSize = '10px', fontFamily = 'Arial') %>%
  formatStyle(columns = 0, fontSize = '10px', fontFamily = 'Arial') %>%
  formatRound(columns = c('read_proportion', 'repeat_count_median', 'repeat_count_mean', 
                          'repeat_count_sd'), digits = 3) %>%
  formatRound(columns = c('repeat_count_mode', 'repeat_count_min', 'repeat_count_max', 
                          'repeat_count_iqr'), digits = 1)
```

**Interpretation:**

- **Cluster numbers renumbered** by abundance (cluster 1 = most reads)

- **Original cluster number** kept for reference

- **R (Repeat cluster)** and **H (Haplotype cluster)** show sub-clustering:
  - Same R value = same repeat length cluster
  - Same H value = same haplotype cluster
  - Example: R1-H1 and R1-H2 have same repeat but different haplotypes
  - Useful for diagnosing why many clusters were detected

- **Proportion** shows percentage of total reads per sample

- **Mode** is the most common repeat count (robust to outliers)

- For **heterozygous diploid** samples, expect 2 clusters with ~0.5 proportion each

- For **homozygous** samples, expect 1 dominant cluster with ~1.0 proportion

- **Read count imbalance** may indicate allele dropout or amplification bias

---

## Cluster visualisation

The **cluster count plot** (shown inline) summarises how many clusters were detected per sample — a quick check on whether clustering is behaving as expected. For diploid samples with two alleles, expect 2 dominant clusters.

**Per-sample violin and scatter plots** are saved to `cluster/violin_by_sample/` and `cluster/scatter_by_sample/`. Each shows all clusters for one sample: violin plots show the full repeat count distribution per cluster; scatter plots show mean ± SD with point size proportional to read count.

```{r cluster_plots, echo=FALSE, fig.height=8, fig.width=12, warning=FALSE}
# Generate colours for any number of clusters
n_clusters <- length(unique(alignment_clustered$cluster_number[!is.na(alignment_clustered$cluster_number)]))
if (n_clusters <= 12) {
  cluster_colors <- RColorBrewer::brewer.pal(max(3, n_clusters), "Set3")
} else {
  # Use viridis for many clusters
  cluster_colors <- viridisLite::viridis(n_clusters, option = "turbo")
}

# Ensure cluster directory is available
cluster_dir <- file.path(params$dir_out, "04_allele_calling", "cluster")
dir.create(cluster_dir, showWarnings = FALSE, recursive = TRUE)

# Determine dynamic plot dimensions based on number of samples
n_samples <- length(unique(alignment_clustered$file_stem))
n_clusters_max <- max(table(cluster_summary$file_stem))

# Define plotting strategy based on sample count
# Small (≤30): All plots
# Medium (31-100): Combined plot limited, all individual plots
# Large (101-200): Skip combined plot, generate individual plots with message
# Very large (>200): Skip combined plot, generate only summary statistics
if (n_samples <= 30) {
  plot_strategy <- "all"
  message("Generating all plots (", n_samples, " samples)")
} else if (n_samples <= 100) {
  plot_strategy <- "limited_combined"
  message("Note: ", n_samples, " samples detected.")
  message("Note: ", n_samples, " samples detected. Per-sample plots will be generated.")
  message("  Individual plots per sample will be clearer for detailed analysis.")
} else if (n_samples <= 200) {
  plot_strategy <- "individual_only"
  message("Note: ", n_samples, " samples detected (>100).")
  message("  Skipping combined violin plot (would be unreadable).")
  message("  Generating individual plots per sample for detailed analysis.")
} else {
  plot_strategy <- "summary_only"
  message("WARNING: ", n_samples, " samples detected (>200).")
  message("  Skipping individual violin plots (would generate >200 files).")
  message("  Generating summary plots only (scatter, ridge).")
  message("  For detailed per-sample plots, filter to a subset of samples.")
}

# Calculate violin plot dimensions (3 columns of facets)
violin_ncol <- 3
violin_nrow <- ceiling(n_samples / violin_ncol)
# Cap height at 30 inches (enough for ~10 rows) to avoid graphics device limits
violin_height <- min(30, max(6, violin_nrow * 3))  # Between 6-30 inches
violin_width <- 12

# Calculate scatter plot dimensions
# Cap width at 40 inches to avoid graphics device limits
scatter_width <- min(40, max(8, n_samples * 0.8))  # Between 8-40 inches
scatter_height <- 6

# Only cluster_counts and y_indicator are needed for per-sample violin plots
cluster_counts <- alignment_clustered %>%
  group_by(file_stem, cluster_number) %>%
  summarise(n = n(), .groups = "drop")

y_min <- min(alignment_clustered[[params$repeat_count_method]], na.rm = TRUE)
y_max <- max(alignment_clustered[[params$repeat_count_method]], na.rm = TRUE)
y_indicator <- y_min - (y_max - y_min) * 0.1

alignment_with_counts <- alignment_clustered %>%
  left_join(cluster_counts, by = c("file_stem", "cluster_number"))

# Per-sample violin plots (gated by plot_per_sample and plot_strategy)
# No colour by cluster — avoids large legends when many clusters exist.
# Point size at baseline encodes read count.
violin_by_sample_dir <- file.path(cluster_dir, "violin_by_sample")

if (params$plot_per_sample && plot_strategy != "summary_only") {
  dir.create(violin_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

  for (sample in unique(alignment_clustered$file_stem)) {
    sample_data   <- alignment_with_counts %>% filter(file_stem == sample)
    sample_counts <- cluster_counts        %>% filter(file_stem == sample)

    p_violin_sample <- ggplot(sample_data, aes(x = factor(cluster_number),
                                               y = !!sym(params$repeat_count_method))) +
      geom_violin(fill = "#619CFF", alpha = 0.6, scale = "width") +
      geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9, fill = "white") +
      geom_point(data = sample_counts,
                 aes(x = factor(cluster_number), y = y_indicator, size = n),
                 shape = 21, fill = "grey30", color = "black", alpha = 0.8,
                 inherit.aes = FALSE) +
      geom_text(data = sample_counts,
                aes(x = factor(cluster_number), y = y_indicator, label = n),
                size = 3, color = "black", fontface = "bold",
                angle = 90, vjust = 0.5, inherit.aes = FALSE) +
      scale_size_continuous(name = "Read count", range = c(4, 15)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 15)) +
      labs(title = paste("Repeat count distribution -", sample),
           x     = "Cluster",
           y     = paste0("Repeat count (", params$repeat_count_method, ")")) +
      theme_minimal() +
      theme(axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
            legend.position = "right")

    ggsave(
      filename = file.path(violin_by_sample_dir, paste0(sample, "-violin.png")),
      plot     = p_violin_sample,
      width    = 10,
      height   = 6,
      dpi      = params$plot_dpi,
      bg       = "white"
    )
  }
  message("Per-sample violin plots saved to: ", violin_by_sample_dir)

} else if (plot_strategy == "summary_only") {
  message("Skipped individual violin plots (>200 samples)")
} else {
  message("Per-sample violin plots skipped (plot_per_sample = FALSE)")
}
```

```{r cluster_scatter, echo=FALSE, fig.height=4, fig.width=6, warning=FALSE}

# --- 1. Cluster count per sample (always shown — scales to any n) ---
cluster_count_per_sample <- cluster_summary %>%
  dplyr::group_by(file_stem) %>%
  dplyr::summarise(n_clusters = dplyr::n(), .groups = "drop")

p_cluster_count <- ggplot(cluster_count_per_sample, aes(x = "", y = n_clusters)) +
  geom_boxplot(outlier.shape = NA, width = 0.35, fill = "grey92", colour = "grey40") +
  geom_jitter(width = 0.08, height = 0, size = 3, colour = "#619CFF", alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 18, colour = "red", size = 4, alpha = 0.6) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +
  labs(title    = "Number of clusters per sample",
       subtitle = sprintf("Each point = one sample  |  n = %d\nRed diamond = mean",
                          n_samples),
       x        = NULL,
       y        = "Number of clusters") +
  theme_bw() +
  theme(axis.ticks.x = element_blank())

print(p_cluster_count)

ggsave(
  filename = file.path(cluster_dir, "cluster_count_per_sample.png"),
  plot     = p_cluster_count,
  width    = 4,
  height   = 4,
  dpi      = params$plot_dpi,
  bg       = "white"
)

# --- 2. Per-sample scatter plots (gated by plot_per_sample) ---
# Coloured by cluster number is avoided to prevent large legends when many clusters exist.
# Point size encodes read count; cluster number is labelled directly on each point.
scatter_by_sample_dir <- file.path(cluster_dir, "scatter_by_sample")

if (params$plot_per_sample) {
  dir.create(scatter_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

  for (sample in unique(cluster_summary$file_stem)) {
    sample_data <- cluster_summary %>% filter(file_stem == sample)

    p_scatter_sample <- ggplot(sample_data,
                               aes(x = factor(cluster_number), y = repeat_count_mean,
                                   size = read_count)) +
      geom_errorbar(aes(ymin = repeat_count_mean - repeat_count_sd,
                        ymax = repeat_count_mean + repeat_count_sd),
                    width = 0.3, alpha = 0.5, linewidth = 0.7, colour = "grey50") +
      geom_point(alpha = 0.8, shape = 21, fill = "#619CFF", colour = "black", stroke = 0.8) +
      geom_text(aes(label = read_count), size = 3, colour = "black", fontface = "bold") +
      scale_size_continuous(name = "Read count", range = c(6, 20)) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
      labs(title = paste("Cluster centres -", sample),
           x     = "Cluster",
           y     = paste0("Mean repeat count (", params$repeat_count_method, ")")) +
      theme_minimal() +
      theme(axis.text.x = element_text(size = 11), legend.position = "right")

    ggsave(
      filename = file.path(scatter_by_sample_dir, paste0(sample, "-scatter.png")),
      plot     = p_scatter_sample,
      width    = max(8, nrow(sample_data) * 0.6),
      height   = 6,
      dpi      = params$plot_dpi,
      bg       = "white"
    )
  }
  message("Per-sample scatter plots saved to: ", scatter_by_sample_dir)
}
```

---

# Consensus generation

For each cluster, reads are aligned using `DECIPHER::AlignSeqs` and a consensus sequence is called at each position with `DECIPHER::ConsensusSequence`. Consensus is generated independently for four segments:

- **Full sequence** — the complete read, used for variant calling
- **Left flank** — sequence upstream of the repeat tract
- **Mid (repeat region)** — the repositioned repeat tract only
- **Right flank** — sequence downstream of the repeat tract

**Downsampling:** Up to `consensus_downsample = `r params$consensus_downsample`` reads are used per cluster (`reads_for_consensus` in `consensus.xlsx`). For large clusters this may be a small fraction of all reads (`reads_in_cluster`). Increasing this improves consensus accuracy but slows alignment considerably for long sequences.

**Consensus threshold:** `consensus_threshold = `r params$consensus_threshold`` means a base is called if ≥`r params$consensus_threshold * 100`% of reads agree at that position; otherwise `N` is recorded. Ambiguity characters are not used.

Results are exported to `consensus/consensus.xlsx` (lengths and sequences in one table) and four FASTA files (`consensus_sequences.fasta`, `consensus_left_sequences.fasta`, `consensus_mid_sequences.fasta`, `consensus_right_sequences.fasta`). FASTA headers encode sample, cluster number, read count, modal repeat count, and cluster proportion.

```{r t_consensus_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r consensus_generation, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Suppress DECIPHER progress bars
options(verbose = FALSE)
old_verbose <- getOption("verbose")
on.exit(options(verbose = old_verbose))

# Check resume
consensus_cache_path <- file.path(temp_dir, "04_consensus.RData")

if (params$resume && safe_load_cache(consensus_cache_path, environment())) {
  message("Loading cached consensus results...")
  message("Consensus sequences loaded from cache")

} else {
  
  if (params$cluster_consensus) {
    
    message("Generating consensus sequences...")
    
    # Create composite key for splitting (more efficient than nest())
    alignment_clustered$cluster_key <- paste0(alignment_clustered$file_stem, 
                                              "_cluster", 
                                              alignment_clustered$cluster_number)
    
    # Split by cluster key (much faster than nest() for large datasets)
    cluster_list <- split(alignment_clustered, alignment_clustered$cluster_key)
    
    # Create lookup for sample names and cluster numbers
    cluster_info <- alignment_clustered %>%
      select(cluster_key, file_stem, cluster_number) %>%
      distinct()
    
    message("\nGenerating consensus sequences for ", length(cluster_list), " clusters...")
    
    # Generate consensus for each cluster using apply_fn
    consensus_results_list <- apply_fn(seq_along(cluster_list), function(i) {
      
      cluster_data <- cluster_list[[i]]
      cluster_key <- names(cluster_list)[i]
      
      # Extract sample name and cluster number from first row
      sample_name <- cluster_data$file_stem[1]
      cluster_num <- cluster_data$cluster_number[1]
      
      # Check for cached result
      cluster_cache_file <- file.path(temp_dir,
                                      paste0("consensus_", sample_name, "_cluster", cluster_num, ".RData"))
      if (params$resume && safe_load_cache(cluster_cache_file, environment())) {
        return(cluster_consensus)
      }
      
      # Get sequences for this cluster
      sequences <- cluster_data$query_sequence
      
      # Skip if no sequences
      if (length(sequences) == 0 || all(is.na(sequences))) {
        cluster_consensus <- list(
          file_stem = sample_name,
          cluster_number = cluster_num,
          consensus_full = NA_character_,
          consensus_left = NA_character_,
          consensus_mid = NA_character_,
          consensus_right = NA_character_,
          read_count = 0,
          n_seqs_consensus = 0L
        )
        save(cluster_consensus, file = cluster_cache_file)
        return(cluster_consensus)
      }
      
      # Create DNAStringSet
      valid_seqs <- sequences[!is.na(sequences) & sequences != ""]
      if (length(valid_seqs) == 0) {
        cluster_consensus <- list(
          file_stem = sample_name,
          cluster_number = cluster_num,
          consensus_full = NA_character_,
          consensus_left = NA_character_,
          consensus_mid = NA_character_,
          consensus_right = NA_character_,
          read_count = 0,
          n_seqs_consensus = 0L
        )
        save(cluster_consensus, file = cluster_cache_file)
        return(cluster_consensus)
      }
      
      dna_stringset <- DNAStringSet(valid_seqs)
      
      # Generate consensus (suppress DECIPHER progress bars)
      # IMPORTANT: Use threads = 1 inside parallel workers to avoid nested parallelisation
      # The outer apply_fn() already parallelizes across clusters
      sink(tempfile())  # Redirect output
      consensus_result <- tryCatch({
        consensus_function(
          dna_stringset = dna_stringset,
          threads = 1,  # Force single-threaded (outer loop is parallelised)
          consensus_threshold = params$consensus_threshold,
          sample_size = params$consensus_downsample
        )
      }, finally = {
        sink()  # Restore output
      })
      
      # Also generate consensus for segments (suppress DECIPHER output)
      sink(tempfile())
      consensus_left <- tryCatch({
        if (!all(is.na(cluster_data$left))) {
          left_seqs <- cluster_data$left[!is.na(cluster_data$left) & cluster_data$left != ""]
          if (length(left_seqs) > 0) {
            consensus_function(DNAStringSet(left_seqs), 
                             threads = 1,  # Single-threaded (inside parallel worker)
                             consensus_threshold = params$consensus_threshold,
                             sample_size = params$consensus_downsample)$consensus
          } else NA_character_
        } else NA_character_
      }, finally = {})
      
      consensus_mid <- tryCatch({
        if (!all(is.na(cluster_data$mid))) {
          mid_seqs <- cluster_data$mid[!is.na(cluster_data$mid) & cluster_data$mid != ""]
          if (length(mid_seqs) > 0) {
            consensus_function(DNAStringSet(mid_seqs),
                             threads = 1,  # Single-threaded (inside parallel worker)
                             consensus_threshold = params$consensus_threshold,
                             sample_size = params$consensus_downsample)$consensus
          } else NA_character_
        } else NA_character_
      }, finally = {})
      
      consensus_right <- tryCatch({
        if (!all(is.na(cluster_data$right))) {
          right_seqs <- cluster_data$right[!is.na(cluster_data$right) & cluster_data$right != ""]
          if (length(right_seqs) > 0) {
            consensus_function(DNAStringSet(right_seqs),
                             threads = 1,  # Single-threaded (inside parallel worker)
                             consensus_threshold = params$consensus_threshold,
                             sample_size = params$consensus_downsample)$consensus
          } else NA_character_
        } else NA_character_
      }, finally = {
        sink()  # Restore output after all segments done
      })
      
      # Save per-cluster cache
      cluster_consensus <- list(
        file_stem = sample_name,
        cluster_number = cluster_num,
        consensus_full = consensus_result$consensus,
        consensus_left = consensus_left,
        consensus_mid = consensus_mid,
        consensus_right = consensus_right,
        read_count = length(valid_seqs),
        n_seqs_consensus = length(consensus_result$input_seqs)
      )
      save(cluster_consensus, file = cluster_cache_file)
      
      return(cluster_consensus)
    })
    
    # Extract results (follows pattern from modules 1-3)
    consensus_results_list <- extract_apply_fn_result(consensus_results_list)
    
    # Convert to data frame
    consensus_summary <- bind_rows(consensus_results_list)
    
    # Merge with cluster_summary to get repeat count statistics
    consensus_summary <- consensus_summary %>%
      left_join(cluster_summary %>% select(file_stem, cluster_number, repeat_count_mode, repeat_count_median, repeat_count_mean),
                by = c("file_stem", "cluster_number"))
    
    message("Consensus generation complete!")
    message("Generated consensus sequences for ", nrow(consensus_summary), " clusters")
    
  } else {
    
    message("Consensus generation disabled")
    consensus_summary <- NULL
    
  }
  
  # Cache results
  save(consensus_summary, file = consensus_cache_path)
  message("Consensus results cached to: ", consensus_cache_path)
}
```

```{r t_consensus_end, echo=FALSE}
timings["Consensus generation"] <- proc.time()["elapsed"] - .t
```

```{r consensus_table, echo=FALSE}
# Display consensus summary
if (exists("consensus_summary") && !is.null(consensus_summary)) {
  
  # Single combined table: lengths then sequences, full/left/mid/right order
  # reads_in_cluster = total reads assigned to cluster (from clustering)
  # reads_for_consensus = reads actually used after downsampling (params$consensus_downsample)
  consensus_export <- consensus_summary %>%
    dplyr::left_join(
      cluster_summary %>% dplyr::select(file_stem, cluster_number,
                                        reads_in_cluster = read_count),
      by = c("file_stem", "cluster_number")
    ) %>%
    dplyr::rename(reads_for_consensus = n_seqs_consensus) %>%
    dplyr::mutate(
      full_length_bp  = nchar(consensus_full),
      left_length_bp  = nchar(consensus_left),
      mid_length_bp   = nchar(consensus_mid),
      right_length_bp = nchar(consensus_right)
    ) %>%
    dplyr::select(file_stem, cluster_number,
                  reads_in_cluster, reads_for_consensus,
                  full_length_bp, left_length_bp, mid_length_bp, right_length_bp,
                  consensus_full, consensus_left, consensus_mid, consensus_right)

  # Use lengths-only version for the inline DT table
  consensus_display <- consensus_export %>%
    dplyr::select(file_stem, cluster_number,
                  reads_in_cluster, reads_for_consensus,
                  full_length_bp, left_length_bp, mid_length_bp, right_length_bp)

  # Delete existing file first so stale sheets don't persist
  consensus_excel_path <- file.path(consensus_dir, "consensus.xlsx")
  if (file.exists(consensus_excel_path)) file.remove(consensus_excel_path)
  write_excel_function(
    named_list = list("consensus" = consensus_export),
    path = consensus_excel_path
  )
  message("Consensus exported to: ", consensus_excel_path)
  
  # Display consensus summary table
  datatable(consensus_display,
            options = list(
              pageLength = 15,
              autoWidth = TRUE,
              scrollX = TRUE,
              initComplete = JS("function(settings, json) {",
                                "$(this.api().table().header()).css({'font-size': '10px', 'font-family': 'Arial'});",
                                "}")
            ),
            rownames = FALSE,
            class = 'cell-border stripe',
            caption = htmltools::tags$caption(
              style = 'caption-side: top; text-align: left; font-size: 12px; font-weight: bold;',
              "Consensus Sequence Summary"
            ),
            colnames = c("Sample", "Cluster",
                         "Reads in cluster", "Reads for consensus",
                         "Full (bp)", "Left (bp)", "Mid (bp)", "Right (bp)")
  ) %>%
    formatStyle(columns = names(consensus_display), fontSize = '10px', fontFamily = 'Arial') %>%
    formatStyle(columns = 0, fontSize = '10px', fontFamily = 'Arial')}
```

---

# Variant calling

Each cluster's **full consensus sequence** is aligned to the reference using global pairwise alignment (`pairwiseAlignment`, gap open = 5, gap extend = 2). Positions that differ between consensus and reference are recorded as mismatches, insertions, or deletions.

**Reference repeat region:** The reference contains NNNNN in place of the CAG repeat. Any variant at a position where the reference base is N is automatically excluded from the VCF — this cleanly removes alignment noise from the repeat region without manual coordinate filtering.

**Output files:**

- `variant.xlsx` — summary of variant counts by type per cluster (`summary` tab) and all individual variants with reference coordinates (`all_variants` tab). Both tabs include `reads_in_cluster` and `reads_for_consensus` so the consensus provenance is traceable.
- `variants.csv` — all variants in flat CSV format
- `vcf_by_sample/*.vcf` — per-sample VCF with one column per cluster, for loading in IGV alongside the reference
- `all_samples_variants.vcf` — combined multi-sample VCF with one column per sample×cluster combination

**Note:** Raw variant counts (shown in the completion summary) include any remaining alignment artefacts from the repeat region boundaries. The VCF positions after N-filtering represent genuine flank variants.

```{r t_variant_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r variant_calling, echo=FALSE}
# Cache variant calling results
variant_cache_path <- file.path(temp_dir, "04_variants.RData")

if (params$resume && safe_load_cache(variant_cache_path, environment())) {
  message("Loading cached variant calling results...")
  message("Variant results loaded from cache")

} else {
  
  if (params$call_variants && exists("reference_seq") && !is.null(reference_seq) && 
      exists("consensus_summary") && !is.null(consensus_summary)) {
    
    message("Calling variants in consensus sequences...")
    
    # Call variants for all consensus sequences
    variant_results <- call_variants_for_clusters(consensus_summary, reference_seq)
    
    if (!is.null(variant_results) && nrow(variant_results) > 0) {
      
      message("Found ", nrow(variant_results), " variants across all clusters")
      
      # Summary of variants
      variant_summary_table <- summarise_variants(variant_results)
      
    } else {
      message("No variants found (all consensus sequences match reference)")
      variant_results <- NULL
      variant_summary_table <- NULL
    }
    
  } else {
    if (!params$call_variants) {
      message("Variant calling disabled")
    } else if (is.null(reference_seq)) {
      message("Variant calling skipped (no reference sequence provided)")
    }
    variant_results <- NULL
    variant_summary_table <- NULL
  }
  
  # Cache results
  save(variant_results, variant_summary_table, file = variant_cache_path)
  message("Variant results cached to: ", variant_cache_path)
}
```

```{r t_variant_end, echo=FALSE}
timings["Variant calling"] <- proc.time()["elapsed"] - .t
```

```{r variant_table, echo=FALSE}
if (exists("variant_results") && !is.null(variant_results) && nrow(variant_results) > 0) {
  
  # Export summaries to Excel — join read counts from consensus_summary
  variant_counts <- if (exists("consensus_summary") && !is.null(consensus_summary)) {
    consensus_summary %>%
      dplyr::left_join(
        cluster_summary %>% dplyr::select(file_stem, cluster_number,
                                          reads_in_cluster = read_count),
        by = c("file_stem", "cluster_number")
      ) %>%
      dplyr::rename(reads_for_consensus = n_seqs_consensus) %>%
      dplyr::select(file_stem, cluster_number, reads_in_cluster, reads_for_consensus)
  } else {
    NULL
  }

  add_counts <- function(df) {
    if (!is.null(variant_counts)) {
      dplyr::left_join(df, variant_counts, by = c("file_stem", "cluster_number"))
    } else df
  }

  variant_excel_path <- file.path(variants_dir, "variant.xlsx")
  if (file.exists(variant_excel_path)) file.remove(variant_excel_path)
  write_excel_function(
      named_list = list(
        "summary"      = add_counts(variant_summary_table),
        "all_variants" = add_counts(variant_results)
      ),
      path = variant_excel_path
    )
    message("Variant summaries exported to: ", variant_excel_path)
    
    # Display variant summary
    if (!is.null(variant_summary_table)) {
      # Get actual column names
      col_names <- names(variant_summary_table)
      
      # Create display names (capitalize first letter)
      display_names <- sapply(col_names, function(x) {
        if (x == "file_stem") return("Sample")
        if (x == "cluster_number") return("Cluster")
        # Capitalize first letter for variant types
        paste0(toupper(substring(x, 1, 1)), substring(x, 2))
      }, USE.NAMES = FALSE)
      
      datatable(variant_summary_table,
                options = list(
                  pageLength = 15,
                  autoWidth = TRUE,
                  scrollX = TRUE,
                  initComplete = JS("function(settings, json) {",
                                    "$(this.api().table().header()).css({'font-size': '10px', 'font-family': 'Arial'});",
                                    "}")
                ),
                rownames = FALSE,
                class = 'cell-border stripe',
                caption = htmltools::tags$caption(
                  style = 'caption-side: top; text-align: left; font-size: 12px; font-weight: bold;',
                  "Variant Summary by Cluster"
                ),
                colnames = display_names
      ) %>%
        formatStyle(columns = col_names, fontSize = '10px', fontFamily = 'Arial') %>%
        formatStyle(columns = 0, fontSize = '10px', fontFamily = 'Arial')
    }
}
```

---

# Export files

Exports consensus sequences as FASTA files. Each FASTA header encodes key metadata in the format:

`{sample}_cluster{NN}_n{read_count}_mode{repeat}_prop{proportion}_{segment}`

For example: `sample1_cluster01_n31237_mode17_prop0.257_left`

Four files are written — full sequence, left flank, mid (repeat region only), and right flank — allowing downstream tools (e.g. BLAST, IGV, alignment viewers) to work with each segment independently.

```{r export_consensus, echo=FALSE}
# Check if consensus was generated
if (exists("consensus_summary") && !is.null(consensus_summary) && nrow(consensus_summary) > 0) {
  
  # Export full consensus sequences
  consensus_fasta_path <- file.path(consensus_dir, "consensus_sequences.fasta")
  
  # Calculate proportion of reads per cluster (within each sample)
  consensus_with_props <- consensus_summary %>%
    group_by(file_stem) %>%
    mutate(
      total_reads = sum(read_count),
      proportion = read_count / total_reads
    ) %>%
    ungroup()
  
  # Calculate dynamic padding for cluster numbers
  # Ensures proper alphabetical sorting even with >999 clusters
  max_cluster <- max(consensus_with_props$cluster_number, na.rm = TRUE)
  n_digits <- nchar(as.character(max_cluster))
  format_string <- paste0("%0", n_digits, "d")
  
  # Create informative FASTA names with read counts, mode repeat, and proportions
  # Format: {sample}_cluster{N}_n{count}_mode{repeat}_prop{proportion}
  # Mode is best for discrete integer data (most common value, robust to outliers)
  # Cluster number padding adjusts to max clusters (e.g., 01-99, 001-999, 0001-9999)
  consensus_names <- paste0(
    consensus_with_props$file_stem, 
    "_cluster", sprintf(format_string, consensus_with_props$cluster_number),
    "_n", consensus_with_props$read_count,
    "_mode", round(consensus_with_props$repeat_count_mode, 0),
    "_prop", sprintf("%.3f", consensus_with_props$proportion)
  )
  consensus_seqs <- setNames(consensus_with_props$consensus_full, consensus_names)
  
  # Remove NA sequences
  consensus_seqs <- consensus_seqs[!is.na(consensus_seqs)]
  
  if (length(consensus_seqs) > 0) {
    write_fasta(consensus_seqs, sequence_type = "DNA", output_path = consensus_fasta_path)
    message("Consensus sequences exported to: ", consensus_fasta_path)
    message("  Cluster number format: ", n_digits, " digits (max cluster: ", max_cluster, ")")
  }
  
  # Also export mid sequences (repeat regions only)
  mid_fasta_path <- file.path(consensus_dir, "consensus_mid_sequences.fasta")
  mid_names <- paste0(
    consensus_with_props$file_stem,
    "_cluster", sprintf(format_string, consensus_with_props$cluster_number),
    "_n", consensus_with_props$read_count,
    "_mode", round(consensus_with_props$repeat_count_mode, 0),
    "_prop", sprintf("%.3f", consensus_with_props$proportion),
    "_mid"
  )
  mid_seqs <- setNames(consensus_with_props$consensus_mid, mid_names)
  mid_seqs <- mid_seqs[!is.na(mid_seqs)]
  
  if (length(mid_seqs) > 0) {
    write_fasta(mid_seqs, sequence_type = "DNA", output_path = mid_fasta_path)
    message("Mid consensus sequences exported to: ", mid_fasta_path)
  }
  
  # Export left flank sequences (independently generated consensus)
  left_fasta_path <- file.path(consensus_dir, "consensus_left_sequences.fasta")
  left_names <- paste0(
    consensus_with_props$file_stem, 
    "_cluster", sprintf(format_string, consensus_with_props$cluster_number),
    "_n", consensus_with_props$read_count,
    "_mode", round(consensus_with_props$repeat_count_mode, 0),
    "_prop", sprintf("%.3f", consensus_with_props$proportion),
    "_left"
  )
  left_seqs <- setNames(consensus_with_props$consensus_left, left_names)
  left_seqs <- left_seqs[!is.na(left_seqs)]
  
  if (length(left_seqs) > 0) {
    write_fasta(left_seqs, sequence_type = "DNA", output_path = left_fasta_path)
    message("Left consensus sequences exported to: ", left_fasta_path)
  }
  
  # Export right flank sequences (independently generated consensus)
  right_fasta_path <- file.path(consensus_dir, "consensus_right_sequences.fasta")
  right_names <- paste0(
    consensus_with_props$file_stem, 
    "_cluster", sprintf(format_string, consensus_with_props$cluster_number),
    "_n", consensus_with_props$read_count,
    "_mode", round(consensus_with_props$repeat_count_mode, 0),
    "_prop", sprintf("%.3f", consensus_with_props$proportion),
    "_right"
  )
  right_seqs <- setNames(consensus_with_props$consensus_right, right_names)
  right_seqs <- right_seqs[!is.na(right_seqs)]
  
  if (length(right_seqs) > 0) {
    write_fasta(right_seqs, sequence_type = "DNA", output_path = right_fasta_path)
    message("Right consensus sequences exported to: ", right_fasta_path)
  }
}
```

```{r export_variants, echo=FALSE}
if (exists("variant_results") && !is.null(variant_results) && nrow(variant_results) > 0) {
  
  # Prepare variant export — rename position column in select to avoid S4 dispatch conflict
  if ("position" %in% names(variant_results)) {
    variant_export <- variant_results %>%
      dplyr::select(file_stem, cluster_number,
                    position_on_reference = position,
                    ref, query, variant_type)
  } else {
    variant_export <- variant_results %>%
      dplyr::select(file_stem, cluster_number,
                    position_on_reference = position_on_reference,
                    ref, query, variant_type)
  }
  
  # Export variants to CSV
  variants_csv_path <- file.path(variants_dir, "variants.csv")
  write.csv(variant_export, variants_csv_path, row.names = FALSE)
  message("Variants exported to: ", variants_csv_path)
  
  # Export variants to VCF format (for IGV, etc.)
  # Filter out variants with N bases (ambiguous) and variants at position 1 (no anchor)
  # Create one VCF per sample
  samples <- unique(variant_export$file_stem)
  
  # Read reference sequence for anchor bases
  ref_seq <- as.character(reference_seq)
  ref_name <- sub("^>", "", readLines(params$path_ref, n = 1))
  
  # Filter variants for VCF: remove N bases and position 1
  vcf_valid_variants <- variant_export %>%
    filter(position_on_reference > 1) %>%
    filter(ref != "N" & query != "N" & ref != "-" & query != "-")
  
  # Individual VCFs: one column per cluster
  message("Exporting individual VCF files...")
  
  # Create subdirectory for individual sample VCF files
  vcf_by_sample_dir <- file.path(variants_dir, "vcf_by_sample")
  dir.create(vcf_by_sample_dir, showWarnings = FALSE, recursive = TRUE)
  
  vcf_export_count <- 0
  vcf_total_variants <- 0
  
  for (sample in samples) {
    sample_variants <- vcf_valid_variants %>%
      filter(file_stem == sample)
    
    if (nrow(sample_variants) > 0) {
      vcf_path <- file.path(vcf_by_sample_dir, paste0(sample, "_variants.vcf"))
      
      # Get all clusters for this sample
      clusters <- sort(unique(sample_variants$cluster_number))
      cluster_names <- paste0("cluster", clusters)
      
      # VCF header
      vcf_header <- c(
        "##fileformat=VCFv4.2",
        paste0("##fileDate=", format(Sys.Date(), "%Y%m%d")),
        "##source=Duke_Pipeline_Module4",
        paste0("##reference=", basename(params$path_ref)),
        paste0("##contig=<ID=", ref_name, ",length=", nchar(ref_seq), ">"),
        "##INFO=<ID=TYPE,Number=1,Type=String,Description=\"Variant type: SNP, INS, or DEL\">",
        "##FORMAT=<ID=GT,Number=1,Type=String,Description=\"Genotype\">",
        paste0("#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO\tFORMAT\t", paste(cluster_names, collapse = "\t"))
      )
      
      # Group variants by position and create merged records
      position_groups <- sample_variants %>%
        group_by(position_on_reference) %>%
        group_split()
      
      vcf_records <- lapply(position_groups, function(pos_variants) {
        # Get position info from first variant at this position
        pos <- unique(pos_variants$position_on_reference)[1]
        
        # Determine VCF position and alleles
        # For each unique REF/ALT at this position, create one VCF line
        unique_variants <- pos_variants %>%
          distinct(ref, query, variant_type, .keep_all = TRUE)
        
        # Process each unique variant at this position
        lapply(1:nrow(unique_variants), function(i) {
          var <- unique_variants[i, ]
          ref_base <- as.character(var$ref)
          query_base <- as.character(var$query)
          var_type <- as.character(var$variant_type)
          
          # Convert to VCF format
          ref_vcf <- NULL
          alt_vcf <- NULL
          pos_vcf <- NULL
          
          if (var_type %in% c("SNP", "mismatch")) {
            ref_vcf <- ref_base
            alt_vcf <- query_base
            pos_vcf <- pos
          } else if (var_type == "insertion" && query_base != "-") {
            if (pos > 1) {
              anchor <- substr(ref_seq, pos - 1, pos - 1)
              ref_vcf <- anchor
              alt_vcf <- paste0(anchor, query_base)
              pos_vcf <- pos - 1
            }
          } else if (var_type == "deletion" && ref_base != "-") {
            if (pos > 1) {
              anchor <- substr(ref_seq, pos - 1, pos - 1)
              ref_vcf <- paste0(anchor, ref_base)
              alt_vcf <- anchor
              pos_vcf <- pos - 1
            }
          }
          
          if (is.null(ref_vcf) || is.null(alt_vcf) || is.null(pos_vcf)) {
            return(NULL)
          }
          
          # Which clusters have this exact variant?
          clusters_with_variant <- pos_variants %>%
            filter(ref == ref_base, query == query_base) %>%
            pull(cluster_number)
          
          # Generate genotypes for all clusters
          genotypes <- sapply(clusters, function(c) {
            if (c %in% clusters_with_variant) "0/1" else "0/0"
          })
          
          chrom <- ref_name
          id <- "."
          qual <- "."
          filter <- "PASS"
          info <- paste0("TYPE=", var_type)
          format <- "GT"
          
          paste(chrom, pos_vcf, id, ref_vcf, alt_vcf, qual, filter, info, format,
                paste(genotypes, collapse = "\t"), sep = "\t")
        })
      })
      
      # Flatten and remove NULLs
      vcf_records <- unlist(unlist(vcf_records, recursive = FALSE))
      vcf_records <- vcf_records[!sapply(vcf_records, is.null)]
      
      if (length(vcf_records) > 0) {
        writeLines(c(vcf_header, vcf_records), vcf_path)
        vcf_export_count <- vcf_export_count + 1
        vcf_total_variants <- vcf_total_variants + length(vcf_records)
      }
    }
  }
  
  # Summary message instead of per-sample messages
  message("VCF files exported: ", vcf_export_count, " samples")
  message("  Location: ", variants_dir)
  message("  Total variant positions: ", vcf_total_variants)
  
  # Create combined multi-sample VCF with sample_cluster columns
  combined_vcf_path <- file.path(variants_dir, "all_samples_variants.vcf")
  
  if (nrow(vcf_valid_variants) > 0) {
    # Get all sample_cluster combinations
    sample_clusters <- vcf_valid_variants %>%
      distinct(file_stem, cluster_number) %>%
      arrange(file_stem, cluster_number) %>%
      mutate(sample_cluster = paste0(file_stem, "_cluster", cluster_number))
    
    all_sample_cluster_names <- sample_clusters$sample_cluster
    
    # VCF header
    vcf_header_combined <- c(
      "##fileformat=VCFv4.2",
      paste0("##fileDate=", format(Sys.Date(), "%Y%m%d")),
      "##source=Duke_Pipeline_Module4",
      paste0("##reference=", basename(params$path_ref)),
      paste0("##contig=<ID=", ref_name, ",length=", nchar(ref_seq), ">"),
      "##INFO=<ID=TYPE,Number=1,Type=String,Description=\"Variant type: SNP, INS, or DEL\">",
      "##FORMAT=<ID=GT,Number=1,Type=String,Description=\"Genotype\">",
      paste0("#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO\tFORMAT\t", 
             paste(all_sample_cluster_names, collapse = "\t"))
    )
    
    # Group all variants by position
    position_groups <- vcf_valid_variants %>%
      group_by(position_on_reference) %>%
      group_split()
    
    vcf_records_combined <- lapply(position_groups, function(pos_variants) {
      pos <- unique(pos_variants$position_on_reference)[1]
      
      # Get unique variants at this position
      unique_variants <- pos_variants %>%
        distinct(ref, query, variant_type, .keep_all = TRUE)
      
      # Process each unique variant
      lapply(1:nrow(unique_variants), function(i) {
        var <- unique_variants[i, ]
        ref_base <- as.character(var$ref)
        query_base <- as.character(var$query)
        var_type <- as.character(var$variant_type)
        
        # Convert to VCF format
        ref_vcf <- NULL
        alt_vcf <- NULL
        pos_vcf <- NULL
        
        if (var_type %in% c("SNP", "mismatch")) {
          ref_vcf <- ref_base
          alt_vcf <- query_base
          pos_vcf <- pos
        } else if (var_type == "insertion" && query_base != "-") {
          if (pos > 1) {
            anchor <- substr(ref_seq, pos - 1, pos - 1)
            ref_vcf <- anchor
            alt_vcf <- paste0(anchor, query_base)
            pos_vcf <- pos - 1
          }
        } else if (var_type == "deletion" && ref_base != "-") {
          if (pos > 1) {
            anchor <- substr(ref_seq, pos - 1, pos - 1)
            ref_vcf <- paste0(anchor, ref_base)
            alt_vcf <- anchor
            pos_vcf <- pos - 1
          }
        }
        
        if (is.null(ref_vcf) || is.null(alt_vcf) || is.null(pos_vcf)) {
          return(NULL)
        }
        
        # Which sample_clusters have this variant?
        sample_clusters_with_variant <- pos_variants %>%
          filter(ref == ref_base, query == query_base) %>%
          mutate(sample_cluster = paste0(file_stem, "_cluster", cluster_number)) %>%
          pull(sample_cluster)
        
        # Generate genotypes for all sample_clusters
        genotypes <- sapply(all_sample_cluster_names, function(sc) {
          if (sc %in% sample_clusters_with_variant) "0/1" else "0/0"
        })
        
        chrom <- ref_name
        id <- "."
        qual <- "."
        filter <- "PASS"
        info <- paste0("TYPE=", var_type)
        format <- "GT"
        
        paste(chrom, pos_vcf, id, ref_vcf, alt_vcf, qual, filter, info, format,
              paste(genotypes, collapse = "\t"), sep = "\t")
      })
    })
    
    # Flatten and remove NULLs
    vcf_records_combined <- unlist(unlist(vcf_records_combined, recursive = FALSE))
    vcf_records_combined <- vcf_records_combined[!sapply(vcf_records_combined, is.null)]
    
    if (length(vcf_records_combined) > 0) {
      writeLines(c(vcf_header_combined, vcf_records_combined), combined_vcf_path)
      message("Combined VCF exported: ", combined_vcf_path,
              " (", length(vcf_records_combined), " variant positions, ",
              length(all_sample_cluster_names), " sample_cluster combinations)")
    }
  }
  
  message("Note: VCF files can be viewed in IGV. Load reference first, then VCF files.")
}
```

---

# Save results

Saves a slim cluster assignment table (`alignment_clustered_slim`) containing only the columns added by Module 4 — `file_stem`, `qname`, `cluster_number`, `repeat_cluster`, `haplotype_cluster`. Full read-level data (sequences, alignments, repeat counts) remains in the Module 2 and 3 RData files and should be joined on `file_stem` + `qname` by downstream modules.

```{r save_outputs, echo=FALSE}
# Slim cluster assignments: only the new columns added by Module 4.
alignment_clustered_slim <- alignment_clustered %>%
  dplyr::select(file_stem, qname,
                cluster_number, repeat_cluster, haplotype_cluster)

# Compile module output
module4_output <- list(
  alignment_clustered_slim = alignment_clustered_slim,
  cluster_summary          = cluster_summary,
  consensus_summary        = consensus_summary,
  variant_results          = if (exists("variant_results")) variant_results else NULL,
  parameters               = params
)

# Save module output
output_path <- file.path(results_dir, "04_allele_calling.RData")
save(module4_output, file = output_path)
message("Module 4 outputs saved to: ", output_path)

# Export key tables to Excel
allele_excel_path <- file.path(params$dir_out, "04_allele_calling", "allele_calling.xlsx")
dir.create(dirname(allele_excel_path), showWarnings = FALSE, recursive = TRUE)
if (file.exists(allele_excel_path)) invisible(file.remove(allele_excel_path))

excel_tabs <- list("cluster_summary" = cluster_summary)
write_excel_function(named_list = excel_tabs, path = allele_excel_path)
message("Allele calling tables exported to: ", allele_excel_path)
```

```{r cleanup, echo=FALSE}
if (params$remove_intermediate) {
  rm(alignment_counts)
  gc()
}
```

**Module 4 Complete!**

Results saved to: `r output_path`

- **`r format(nrow(alignment_clustered), big.mark = ",")`** reads with cluster assignments
- **`r n_distinct(alignment_clustered$file_stem)`** samples analysed
- **`r if(!is.null(consensus_summary)) nrow(consensus_summary) else 0`** consensus sequences generated
- **`r if(exists("variant_results") && !is.null(variant_results)) nrow(variant_results) else 0`** variants called

---

# Module diagnostics

## Timing

```{r timing_table, echo=FALSE}
timing_df <- build_timing_table(timings)
DT::datatable(
  timing_df,
  caption = "Section timing — Module 4",
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Section", "Seconds", "Elapsed (mm:ss)", "Cumulative (mm:ss)")
) %>%
  DT::formatStyle("section", target = "row",
                  fontWeight = DT::styleEqual("Total", "bold"))
```

## Output objects

```{r manifest_table, echo=FALSE}
manifest <- data.frame(
  object = names(module4_output),
  description = c(
    "Per-read cluster assignments (slim — join with Modules 2/3 on file_stem + qname for full data)",
    "Per-sample/cluster summary: read counts, modal repeat, proportion, cluster boundaries",
    "Consensus sequences and quality metrics per cluster (NULL if cluster_consensus = FALSE)",
    "Variant calls vs reference for each cluster (NULL if call_variants = FALSE)",
    "Pipeline parameters used in this module run"
  ),
  class = sapply(module4_output, function(x) if (is.null(x)) "NULL" else class(x)[1]),
  size = sapply(module4_output, function(x) {
    format_size(if (is.null(x)) 0 else as.numeric(object.size(x)))
  }),
  bytes = sapply(module4_output, function(x) {
    if (is.null(x)) 0L else as.numeric(object.size(x))
  }),
  stringsAsFactors = FALSE
)

DT::datatable(
  manifest[, c("object", "description", "class", "size")],
  caption = sprintf("Objects saved to module4_output — RData file: %s on disk (%s)",
                    format_size(file.size(output_path)), output_path),
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Object", "Description", "Class", "In-memory size")
)
```

```{r diagnostics_export, echo=FALSE}
write_excel_function(
  named_list = list("timing" = timing_df, "manifest" = manifest),
  path = allele_excel_path,
  overwrite_sheet = TRUE
)
message(sprintf("Module 4 complete — total time: %s  |  RData: %s",
                format_elapsed(sum(timings)), format_size(file.size(output_path))))
```

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```
