---
title: "Duke Pipeline - Module 5: Waterfall Plots"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

```{r load_packages, echo=FALSE, message=FALSE}
# Required packages
required_packages <- c("dplyr", "tidyr", "ggplot2", "data.table", "stringr", 
                      "knitr", "DT", "scales", "pbapply")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r create_directories, echo=FALSE}
# Create output directory for plots
plots_dir <- file.path(params$dir_out, file.path("05_waterfall", "plots"))

if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}
```

---

# Module 5: Waterfall Plots

Visualise read-level structure showing left flank, repeat region, and right flank for each read.

**Key features:**

- Colour-coded by base (A/C/T/G)
- Ordered by repeat length
- Per-cluster visualisation
- Outlier removal optional
- Downsampling for performance

---

# Load data

```{r load_data, echo=FALSE}
# Load Module 4 output
results_dir <- file.path(params$dir_out, "module_data")
load(file.path(results_dir, "04_allele_calling_results.RData"))

# Extract clustered data
alignment_clustered <- module4_output$alignment_clustered

# Check data
message("Loaded clustered alignment data:")
message("  Samples: ", length(unique(alignment_clustered$file_stem)))
message("  Total reads: ", nrow(alignment_clustered))
message("  Clusters: ", length(unique(alignment_clustered$cluster_number)))
```

---

# Prepare data for waterfall

```{r prepare_waterfall_data, echo=FALSE}
# Track total reads per sample before filtering
total_reads_per_sample <- alignment_clustered %>%
  group_by(file_stem) %>%
  summarise(total_reads = n(), .groups = "drop")

# Prepare data for waterfall plots using library function
waterfall_data <- prepare_waterfall_data(
  alignment_data = alignment_clustered,
  repeat_count_method = params$repeat_count_method,
  remove_outliers = params$waterfall_rm_flank_length_outliers,
  group_vars = c("file_stem", "cluster_number")
)

# Summary with total reads
waterfall_summary <- waterfall_data %>%
  group_by(file_stem) %>%
  summarise(
    n_reads = n(),
    n_clusters = n_distinct(cluster_number),
    mean_left_length = mean(left_length, na.rm = TRUE),
    mean_mid_length = mean(mid_length, na.rm = TRUE),
    mean_right_length = mean(right_length, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(total_reads_per_sample, by = "file_stem") %>%
  relocate(total_reads, .after = n_reads)

DT::datatable(
  waterfall_summary,
  caption = "Waterfall data summary",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(waterfall_summary), fontSize = '85%') %>%
  DT::formatRound(columns = c('mean_left_length', 'mean_mid_length', 'mean_right_length'), digits = 1)
```

---

# Waterfall plots

Generate waterfall plots showing read structure per sample and cluster.

## Plotting function

Waterfall plots are generated using the `create_waterfall_plot()` function from `lib/waterfall.R`.

## Generate plots

```{r generate_waterfall_plots, echo=FALSE, results='asis', fig.width=14, fig.height=10}
# Generate waterfall plot for each sample
for (sample in unique(waterfall_data$file_stem)) {
  
  cat("\n\n###", sample, "\n\n")
  
  sample_data <- waterfall_data %>% filter(file_stem == sample)
  
  # Get total reads for this sample
  sample_total <- total_reads_per_sample %>% 
    filter(file_stem == sample) %>% 
    pull(total_reads)
  
  message("Generating waterfall for ", sample, " (", nrow(sample_data), " reads)")
  
  # Create plot using library function
  p <- create_waterfall_plot(
    data = sample_data,
    sample_name = sample,
    downsample = params$waterfall_downsample,
    dna_colours = params$dna_colours,
    repeat_count_col = "repeat_length",
    y_axis_labels = params$waterfall_y_axis_labels,
    total_reads = sample_total
  )
  
  # Don't print to HTML (would make file huge with many samples)
  # Just save to file and show message
  message("  → Saved waterfall plot")
  
  # Export
  ggsave(
    filename = file.path(plots_dir, paste0("waterfall_", sample, ".png")),
    plot = p,
    width = 14,
    height = 10,
    dpi = 300,
    bg = "white"
  )
  
  cat("\n\n")
}
```

## Per-cluster plots

Generate separate waterfall plots for each cluster within each sample (optional).

```{r generate_cluster_waterfall_plots, echo=FALSE, results='asis', fig.width=12, fig.height=8}
# Only generate if enabled
if (params$waterfall_per_cluster) {
  
  # Create subdirectory for cluster plots
  waterfall_by_cluster_dir <- file.path(plots_dir, "waterfall_by_cluster")
  dir.create(waterfall_by_cluster_dir, showWarnings = FALSE, recursive = TRUE)
  
  # Generate per-cluster waterfall plots
  for (sample in unique(waterfall_data$file_stem)) {
    
    cat("\n\n###", sample, "- by cluster\n\n")
    
    sample_data <- waterfall_data %>% filter(file_stem == sample)
    
    # Get total reads for this sample
    sample_total <- total_reads_per_sample %>% 
      filter(file_stem == sample) %>% 
      pull(total_reads)
    
    # Get clusters for this sample
    clusters <- sort(unique(sample_data$cluster_number))
    
    for (clust in clusters) {
      
      cluster_data <- sample_data %>% filter(cluster_number == clust)
      
      if (nrow(cluster_data) < 5) next  # Skip small clusters
      
      cat("\n\n####", paste("Cluster", clust), "\n\n")
      
      # Calculate actual plotted count (after downsampling)
      cluster_total <- nrow(cluster_data)
      cluster_plotted <- if (!is.na(params$waterfall_downsample) && 
                             cluster_total > params$waterfall_downsample) {
        params$waterfall_downsample
      } else {
        cluster_total
      }
      
      message("  Cluster ", clust, ": ", cluster_plotted, " reads (", cluster_total, " total)")
      
      # Create plot using library function
      p_cluster <- create_waterfall_plot(
        data = cluster_data,
        sample_name = paste(sample, "- Cluster", clust),
        downsample = params$waterfall_downsample,
        dna_colours = params$dna_colours,
        repeat_count_col = "repeat_length",
        y_axis_labels = params$waterfall_y_axis_labels,
        total_reads = sample_total
      ) +
        labs(subtitle = paste(cluster_plotted, "reads plotted from", 
                             cluster_total, "in cluster", clust, 
                             paste0("(", sample_total, " total in sample)")))
      
      # Don't print to HTML (would make file huge)
      # Just save to file
      message("    → Saved cluster waterfall plot")
      
      # Export
      ggsave(
        filename = file.path(waterfall_by_cluster_dir, paste0(sample, "_cluster", clust, "-waterfall.png")),
        plot = p_cluster,
        width = 12,
        height = 8,
        dpi = 300,
        bg = "white"
      )
      
      cat("\n\n")
    }
  }
  
  message("All waterfall plots exported to: ", plots_dir)
  
} else {
  cat("\n\nPer-cluster waterfall plots skipped (waterfall_per_cluster = FALSE)\n\n")
  message("Per-cluster plots skipped. Set waterfall_per_cluster = TRUE to enable.")
}
```

---

# Summary

**Waterfall plots generated:**

- Per-sample plots showing all reads
- Per-cluster plots for detailed inspection
- Color-coded by base composition
- Ordered by repeat length

**Outputs saved to:**
- `05_plots/` - PNG files
- `05_waterfall/` - Analysis files (if any)

---

# Save outputs

```{r save_outputs, echo=FALSE}
# Compile module output
module5_output <- list(
  waterfall_data = waterfall_data,
  waterfall_summary = waterfall_summary
)

# Save complete module output
results_dir <- file.path(params$dir_out, "module_data")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

output_path <- file.path(results_dir, "05_waterfall_results.RData")
save(module5_output, file = output_path)

message("Module 5 outputs saved to: ", output_path)
```

**Module 5 Complete!**

Results saved to: `r output_path`

- **`r length(unique(waterfall_data$file_stem))`** samples processed
- **`r nrow(waterfall_data)`** reads visualised
- **`r sum(waterfall_summary$n_clusters)`** total clusters across all samples

---
