---
title: "Duke pipeline"
subtitle: "Module 5: Waterfall Plots"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

```{r set_tab_title, echo=FALSE, results='asis'}
cat('<script>document.title = "Duke | Module 5: Waterfall Plots";</script>')
```

```{r load_packages, echo=FALSE, message=FALSE}
# Required packages
required_packages <- c("dplyr", "tidyr", "ggplot2", "data.table", "stringr",
                      "knitr", "DT", "scales", "pbapply", "openxlsx")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r create_directories, echo=FALSE}
# Create output directory for plots
plots_dir <- file.path(params$dir_out, file.path("05_waterfall", "plots"))

if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}
```

```{r timing_init, echo=FALSE}
module_start_time <- proc.time()["elapsed"]
timings <- c()
```

---

# Overview

**Module 5: Waterfall Plots**

Generates read-level waterfall plots showing the base composition of each read across the left flank, repeat region (mid), and right flank. Reads are stacked vertically and ordered by repeat length, making allele structure and somatic instability immediately visible.

**Input:** Cluster assignments from Module 4 (joined with repeat counts and segment sequences from Module 3)

**Output:** PNG waterfall plots per sample (`waterfall_{sample}.png`) and optionally per cluster (`waterfall_by_cluster/`)

**Key parameters:**

- `waterfall_downsample`: Maximum reads plotted per sample (default 1,000) — keeps file sizes manageable
- `waterfall_per_cluster`: Generate separate plots per cluster within each sample
- `dna_colours`: Base colour scheme (A/C/T/G/gap/N)

---

# Load data

```{r load_data, echo=FALSE}
results_dir <- file.path(params$dir_out, "module_data")

# Load Module 3 slim (repositioned left/mid/right + repeat counts)
module3_path <- file.path(results_dir, "03_repeat_detection_results.RData")
if (!file.exists(module3_path)) stop("Module 3 output not found: ", module3_path)
load(module3_path)
counts_slim <- module3_output$alignment_counts_slim

# Load Module 4 slim (cluster assignments)
module4_path <- file.path(results_dir, "04_allele_calling_results.RData")
if (!file.exists(module4_path)) stop("Module 4 output not found: ", module4_path)
load(module4_path)
clusters_slim <- module4_output$alignment_clustered_slim

# Join: slim repeat counts + cluster assignments
# left/mid/right are the repositioned (Module 3 trimmed) versions
alignment_clustered <- counts_slim %>%
  dplyr::left_join(clusters_slim, by = c("file_stem", "qname"))

message("Loaded clustered alignment data:")
message("  Samples: ",  length(unique(alignment_clustered$file_stem)))
message("  Total reads: ", format(nrow(alignment_clustered), big.mark = ","))
message("  Clusters: ", length(unique(alignment_clustered$cluster_number)))
```

---

# Prepare data for waterfall

```{r t_prepare_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r prepare_waterfall_data, echo=FALSE, message=FALSE}
# Track total reads per sample
total_reads_per_sample <- alignment_clustered %>%
  group_by(file_stem) %>%
  summarise(total_reads = n(), .groups = "drop")

# Prepare data for waterfall — no outlier filtering here (done in Module 4 if enabled)
waterfall_data <- suppressMessages(prepare_waterfall_data(
  alignment_data      = alignment_clustered,
  repeat_count_method = params$repeat_count_method,
  group_vars          = c("file_stem", "cluster_number")
))

# Per-sample read counts for subtitle
n_per_sample <- waterfall_data %>%
  group_by(file_stem) %>%
  summarise(n_reads = n(), .groups = "drop")

# Summary table
waterfall_summary <- n_per_sample %>%
  left_join(total_reads_per_sample, by = "file_stem") %>%
  left_join(
    waterfall_data %>%
      group_by(file_stem) %>%
      summarise(n_clusters  = n_distinct(cluster_number),
                mean_mid_bp = round(mean(mid_length, na.rm = TRUE), 1),
                .groups = "drop"),
    by = "file_stem")
```

```{r t_prepare_end, echo=FALSE}
timings["Prepare data"] <- proc.time()["elapsed"] - .t
```

---

# Waterfall plots

```{r t_plots_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r generate_waterfall_plots, echo=FALSE, message=FALSE}
# Generate waterfall plot for each sample
for (sample in unique(waterfall_data$file_stem)) {

  sample_data <- waterfall_data %>% filter(file_stem == sample)

  # Get total reads for this sample
  sample_total <- total_reads_per_sample %>%
    filter(file_stem == sample) %>%
    pull(total_reads)

  # Get reads for this sample
  n_available <- n_per_sample %>% filter(file_stem == sample) %>% pull(n_reads)
  n_plotted   <- min(n_available, params$waterfall_downsample, na.rm = TRUE)

  # Create plot using library function
  p <- create_waterfall_plot(
    data = sample_data,
    sample_name = sample,
    downsample = params$waterfall_downsample,
    dna_colours = params$dna_colours,
    repeat_count_col = "repeat_length",
    y_axis_labels = params$waterfall_y_axis_labels,
    total_reads = sample_total
  )

  # Override subtitle
  p <- p + labs(subtitle = sprintf(
    "%s reads plotted  |  %s total",
    format(n_plotted, big.mark = ","),
    format(sample_total, big.mark = ",")))

  # Export
  ggsave(
    filename = file.path(plots_dir, paste0("waterfall_", sample, ".png")),
    plot = p,
    width = 14,
    height = 10,
    dpi = params$plot_dpi,
    bg = "white"
  )
}
```

```{r generate_cluster_waterfall_plots, echo=FALSE, message=FALSE}
# Only generate if enabled
if (params$waterfall_per_cluster) {

  # Create subdirectory for cluster plots
  waterfall_by_cluster_dir <- file.path(plots_dir, "waterfall_by_cluster")
  dir.create(waterfall_by_cluster_dir, showWarnings = FALSE, recursive = TRUE)

  # Generate per-cluster waterfall plots
  for (sample in unique(waterfall_data$file_stem)) {

    sample_data <- waterfall_data %>% filter(file_stem == sample)

    # Get total reads for this sample
    sample_total <- total_reads_per_sample %>%
      filter(file_stem == sample) %>%
      pull(total_reads)

    # Get clusters for this sample
    clusters <- sort(unique(sample_data$cluster_number))

    for (clust in clusters) {

      cluster_data <- sample_data %>% filter(cluster_number == clust)

      if (nrow(cluster_data) < 5) next  # Skip small clusters

      # Calculate actual plotted count (after downsampling)
      cluster_total <- nrow(cluster_data)
      cluster_plotted <- if (!is.na(params$waterfall_downsample) &&
                             cluster_total > params$waterfall_downsample) {
        params$waterfall_downsample
      } else {
        cluster_total
      }

      # Create plot using library function
      p_cluster <- create_waterfall_plot(
        data = cluster_data,
        sample_name = paste(sample, "- Cluster", clust),
        downsample = params$waterfall_downsample,
        dna_colours = params$dna_colours,
        repeat_count_col = "repeat_length",
        y_axis_labels = params$waterfall_y_axis_labels,
        total_reads = sample_total
      ) +
        labs(subtitle = paste(cluster_plotted, "reads plotted from",
                             cluster_total, "in cluster", clust,
                             paste0("(", sample_total, " total in sample)")))

      # Export
      ggsave(
        filename = file.path(waterfall_by_cluster_dir, paste0(sample, "_cluster", clust, "-waterfall.png")),
        plot = p_cluster,
        width = 12,
        height = 8,
        dpi = params$plot_dpi,
        bg = "white"
      )
    }
  }

} else {
  message("Per-cluster plots skipped (waterfall_per_cluster = FALSE)")
}
```

```{r t_plots_end, echo=FALSE}
timings["Generate plots"] <- proc.time()["elapsed"] - .t
```

---

# Save outputs

```{r save_outputs, echo=FALSE}
message("Module 5 complete. Waterfall plots saved to: ", plots_dir)
```

**Module 5 Complete!**

- **`r length(unique(waterfall_data$file_stem))`** samples processed
- **`r nrow(waterfall_data)`** reads in waterfall data
- Plots saved to `05_waterfall/plots/`

---

# Module diagnostics

## Timing

```{r timing_table, echo=FALSE}
timing_df <- build_timing_table(timings)
DT::datatable(
  timing_df,
  caption = "Section timing — Module 5",
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Section", "Seconds", "Elapsed (mm:ss)", "Cumulative (mm:ss)")
) %>%
  DT::formatStyle("section", target = "row",
                  fontWeight = DT::styleEqual("Total", "bold"))
```

```{r diagnostics_export, echo=FALSE}
waterfall_excel_path <- file.path(params$dir_out, "05_waterfall", "waterfall.xlsx")
dir.create(dirname(waterfall_excel_path), showWarnings = FALSE, recursive = TRUE)
write_excel_function(
  named_list = list("timing" = timing_df),
  path = waterfall_excel_path,
  overwrite_sheet = TRUE
)
message(sprintf("Module 5 complete — total time: %s", format_elapsed(sum(timings))))
```

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```
