---
title: "Duke pipeline"
subtitle: "Module 6: Range analysis"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
# Suppress pbapply/pbmcapply progress bars globally
pbapply::pboptions(type = "none")
options(pbmcapply.use_pb = FALSE)
```

```{r set_tab_title, echo=FALSE, results='asis'}
cat('<script>document.title = "Duke | Module 6: Range analysis";</script>')
```

```{r load_packages, echo=FALSE, message=FALSE}
# Required packages
required_packages <- c("dplyr", "tidyr", "ggplot2", "ggridges", "stringr", "readxl",
                      "knitr", "DT", "pbapply", "moments",
                      "openxlsx", "pracma", "scales", "pbmcapply")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r create_directories, echo=FALSE}
# Create output directory
range_dir <- file.path(params$dir_out, "06_range_analysis")

if (!dir.exists(range_dir)) dir.create(range_dir, recursive = TRUE)
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}
```

```{r timing_init, echo=FALSE}
module_start_time <- proc.time()["elapsed"]
timings <- c()
```

---


# Overview

**Module 6: Range analysis & instability metrics**

Analyses repeat length distributions within user-defined ranges and calculates instability metrics relative to control samples.

**Input:** Repeat counts and cluster assignments (Modules 3 and 4), settings Excel file defining analysis ranges

**Key features:**

- Settings-based range definitions per sample
- Peak detection per range
- Distribution summaries (~30 metrics per range)
- Group control definition and instability metrics
- Excel exports

---

# Load data

```{r load_data, echo=FALSE}
results_dir <- file.path(params$dir_out, "module_data")

# Load Module 3 slim (repeat counts)
module3_path <- file.path(results_dir, "03_repeat_detection.RData")
safe_load_module(module3_path, module_num = 3, envir = environment())
counts_slim <- module3_output$alignment_counts_slim

# Load Module 4 slim (cluster assignments)
module4_path <- file.path(results_dir, "04_allele_calling.RData")
safe_load_module(module4_path, module_num = 4, envir = environment())
clusters_slim <- module4_output$alignment_clustered_slim

# Join: repeat counts + cluster assignments
alignment_clustered <- counts_slim %>%
  dplyr::left_join(clusters_slim, by = c("file_stem", "qname"))

message("Loaded clustered alignment data:")
message("  Samples: ",  length(unique(alignment_clustered$file_stem)))
message("  Total reads: ", format(nrow(alignment_clustered), big.mark = ","))
message("  Clusters: ", length(unique(alignment_clustered$cluster_number)))
```

---

# Load settings

Settings Excel defines analysis ranges, group assignments, and control samples.

**Expected columns:**

- `file_name`: Sample filename
- `analysis_ranges`: Range definitions (e.g., `[0-35][36-1000]`)
- `floor`: Minimum frequency threshold per range (e.g., `[3][3]`)
- `max_peaks`: Maximum peaks to detect per range (e.g., `[2][1]`)
- `manual_control_repeat_length`: Optional manual control values (e.g., `[14][100]`)
- `group`: Group assignment for comparisons
- `group_control_sample`: Flag control samples (1 = control, NA = experimental)
- `time`: Timepoint or order
- `exclude`: Flag samples to exclude from analysis

```{r load_settings, echo=FALSE}
# Check if settings file exists
if (is.na(params$path_settings) || !file.exists(params$path_settings)) {
  stop("Settings file not found. Please provide settings_path parameter.")
}

# Load settings (supports CSV and Excel)
file_ext <- tolower(tools::file_ext(params$path_settings))

if (file_ext == "csv") {
  # Read CSV
  settings <- read.csv(params$path_settings, header = TRUE, stringsAsFactors = FALSE)
} else if (file_ext %in% c("xlsx", "xls")) {
  # Read Excel
  settings <- read_excel(params$path_settings, sheet = 1)
} else {
  stop("Unsupported settings file format: ", file_ext, 
       ". Please use .csv, .xlsx, or .xls")
}

# Convert file_name to file_stem (remove extensions)
settings <- settings %>%
  mutate(file_stem = str_remove(file_name, "\\.(fastq|fasta|fa|bam)(\\.gz)?$"))

message("Settings loaded from: ", params$path_settings)
message("  Format: ", toupper(file_ext))
message("  Total samples in settings: ", nrow(settings))

# Filter settings to only include samples present in the actual data
samples_in_data <- unique(alignment_clustered$file_stem)
settings_all <- settings  # Keep original for reference
settings <- settings %>%
  filter(file_stem %in% samples_in_data)

# Report filtering
samples_filtered <- setdiff(settings_all$file_stem, settings$file_stem)
if (length(samples_filtered) > 0) {
  message("\n=== Settings Filtering ===")
  message("  Samples in settings file: ", nrow(settings_all))
  message("  Samples in dataset: ", length(samples_in_data))
  message("  Samples matched: ", nrow(settings))
  message("  Samples excluded (not in data): ", length(samples_filtered))
  if (length(samples_filtered) <= 10) {
    message("  Excluded samples: ", paste(samples_filtered, collapse = ", "))
  } else {
    message("  Excluded samples (first 10): ", 
            paste(head(samples_filtered, 10), collapse = ", "), " ...")
  }
}

# Check if any data samples are missing from settings
samples_missing_settings <- setdiff(samples_in_data, settings$file_stem)
if (length(samples_missing_settings) > 0) {
  warning("\n=== Missing Settings ===")
  warning("  Samples in data without settings: ", length(samples_missing_settings))
  if (length(samples_missing_settings) <= 10) {
    warning("  Samples: ", paste(samples_missing_settings, collapse = ", "))
  } else {
    warning("  Samples (first 10): ", 
            paste(head(samples_missing_settings, 10), collapse = ", "), " ...")
  }
  warning("  These samples will be skipped in Module 6 analysis")
}

# Display filtered settings table
datatable(settings,
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Settings file (filtered to samples in dataset)",
          rownames = FALSE) %>%
  formatStyle(columns = colnames(settings), fontSize = '10px')
```

---

# Parse analysis ranges

Parse range definitions from settings for each sample.

```{r parse_ranges, echo=FALSE}
# Parse ranges for all samples
settings_parsed <- settings %>%
  filter(!is.na(analysis_ranges), !is.na(floor), !is.na(max_peaks)) %>%
  rowwise() %>%
  mutate(
    # Parse ranges, floor, and max_peaks
    ranges_list = list(parse_analysis_ranges(analysis_ranges)),
    floor_list = list(parse_bracketed_param(floor)),
    max_peaks_list = list(parse_bracketed_param(max_peaks)),
    n_ranges = length(ranges_list)
  ) %>%
  ungroup()

# Show summary
range_summary <- settings_parsed %>%
  select(file_stem, analysis_ranges, n_ranges)

DT::datatable(
  range_summary,
  caption = "Analysis ranges per sample",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(range_summary), fontSize = '85%')

# ── Pad range names at source ───────────────────────────────────────────────
# Ranges are named range1, range2, etc. by parse_analysis_ranges().
# If any sample has >= 10 ranges, pad with leading zeros so they sort correctly
# (range01, range02, ...) without needing any display-time workarounds.
max_n_ranges <- max(settings_parsed$n_ranges, na.rm = TRUE)

if (max_n_ranges >= 10) {
  pad_width <- nchar(as.character(max_n_ranges))
  
  rename_range_keys <- function(lst) {
    if (is.null(names(lst))) return(lst)
    nums <- suppressWarnings(as.integer(gsub("^range", "", names(lst))))
    if (anyNA(nums)) return(lst)  # leave non-standard names unchanged
    setNames(lst, paste0("range", formatC(nums, width = pad_width, flag = "0")))
  }
  
  settings_parsed <- settings_parsed %>%
    rowwise() %>%
    mutate(
      ranges_list    = list(rename_range_keys(ranges_list)),
      floor_list     = list(rename_range_keys(floor_list)),
      max_peaks_list = list(rename_range_keys(max_peaks_list))
    ) %>%
    ungroup()
  
  message(sprintf("Range names padded to width %d (max ranges in any sample: %d)",
                  pad_width, max_n_ranges))
} else {
  message(sprintf("Range names unpadded (max ranges in any sample: %d, <10)",
                  max_n_ranges))
}

# Example for first sample (shows final padded names)
message("\nExample range parsing for: ", settings_parsed$file_stem[1])
for (rng in names(settings_parsed$ranges_list[[1]])) {
  r <- settings_parsed$ranges_list[[1]][[rng]]
  f <- settings_parsed$floor_list[[1]][[rng]]
  m <- settings_parsed$max_peaks_list[[1]][[rng]]
  message("  ", rng, ": [", r[1], "-", r[2], "] floor=", f, " max_peaks=", m)
}
```

---

# Experiment overview

```{r plot_experiment_map, echo=FALSE, fig.width=10, fig.height=5, message=FALSE}
# Build sample map: group × time, bubble size = number of samples
exp_map_data <- settings %>%
  select(file_stem, group, time) %>%
  mutate(time = suppressWarnings(as.numeric(time))) %>%
  group_by(group, time) %>%
  summarise(n_samples = n(), .groups = "drop")

has_time <- any(!is.na(exp_map_data$time))

if (has_time) {
  p_exp_map <- ggplot(exp_map_data,
                      aes(x = time, y = group, size = n_samples)) +
    geom_point(colour = "steelblue", alpha = 0.7) +
    geom_text(aes(label = n_samples), size = 3.5, colour = "white",
              fontface = "bold") +
    scale_size_continuous(range = c(6, 18), name = "Samples") +
    labs(title = "Experiment map",
         subtitle = "Each bubble = one group × time combination. Size and label = number of samples.",
         x = "Time",
         y = "Group") +
    theme_bw() +
    theme(panel.grid.minor = element_blank())
} else {
  # No time data — just show sample counts per group as horizontal bars
  p_exp_map <- ggplot(exp_map_data %>% group_by(group) %>%
                        summarise(n_samples = sum(n_samples), .groups = "drop"),
                      aes(x = n_samples, y = group)) +
    geom_col(fill = "steelblue", alpha = 0.7, width = 0.6) +
    geom_text(aes(label = n_samples), hjust = -0.3, size = 4) +
    labs(title = "Experiment overview",
         subtitle = "Number of samples per group (no time data provided).",
         x = "Number of samples",
         y = "Group") +
    theme_bw()
}

print(p_exp_map)

plots_dir <- file.path(params$dir_out, file.path("06_range_analysis", "plots"))
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)
ggsave(file.path(plots_dir, "01_experiment_map.png"),
       p_exp_map, width = 10, height = 5, dpi = params$plot_dpi)

message("Groups: ", paste(unique(settings$group), collapse = ", "))
message("Total samples: ", nrow(settings))
```

---

# Ranges per sample

```{r plot_n_ranges, echo=FALSE, fig.width=8, fig.height=5}
# Join n_ranges to group information from settings
n_ranges_data <- settings_parsed %>%
  select(file_stem, n_ranges)

p_n_ranges <- ggplot(n_ranges_data,
                     aes(x = factor(1), y = n_ranges)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey50",
               width = 0.4) +
  geom_jitter(colour = "grey40", width = 0.15, height = 0,
              alpha = 0.7, size = 2.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5,
               colour = "red", alpha = 0.6) +
  scale_x_discrete(labels = NULL, breaks = NULL) +
  labs(title = "Number of analysis ranges per sample",
       subtitle = "Box = IQR; points = samples; red diamond = mean.",
       x = NULL,
       y = "Number of ranges") +
  theme_bw() +
  theme(axis.ticks.x = element_blank())

print(p_n_ranges)

plots_dir <- file.path(params$dir_out, file.path("06_range_analysis", "plots"))
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)
ggsave(file.path(plots_dir, "02_n_ranges_per_sample.png"),
       p_n_ranges, width = 8, height = 5, dpi = params$plot_dpi)

message("Ranges per sample:")
message("  Min: ",  min(n_ranges_data$n_ranges))
message("  Max: ",  max(n_ranges_data$n_ranges))
message("  Mean: ", round(mean(n_ranges_data$n_ranges), 1))
```

---

# Range summary

Build a summary table showing all range parameters for each sample.

```{r t_range_summary_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r comprehensive_range_summary, echo=FALSE, message=FALSE}
message("\n=== Building Comprehensive Range Summary ===")

# Build comprehensive range summary for each sample and range
comprehensive_range_summary_list <- lapply(1:nrow(settings_parsed), function(i) {
  
  my_settings <- settings_parsed[i, ]
  my_file_stem <- my_settings$file_stem
  
  # Get ranges and parameters
  ranges <- my_settings$ranges_list[[1]]
  floor_vals <- my_settings$floor_list[[1]]
  max_peaks_vals <- my_settings$max_peaks_list[[1]]
  
  # Get manual controls if they exist
  manual_controls <- list()
  if (!is.null(my_settings$manual_control_list) && 
      length(my_settings$manual_control_list) > 0) {
    manual_controls <- my_settings$manual_control_list[[1]]
  }
  
  # Get sample data to calculate actual boundaries
  sample_data <- alignment_clustered %>%
    filter(file_stem == my_file_stem)
  
  # This should not happen since we pre-filtered settings, but check anyway
  if (nrow(sample_data) == 0) {
    warning("UNEXPECTED: No data for sample: ", my_file_stem, 
            " (should have been filtered out earlier)")
    return(NULL)
  }
  
  repeat_counts <- sample_data[[params$repeat_count_method]]
  
  # Build summary for each range
  range_summaries <- lapply(names(ranges), function(range_name) {
    
    my_range <- ranges[[range_name]]
    my_floor <- floor_vals[[range_name]]
    my_max_peaks <- max_peaks_vals[[range_name]]
    
    # Manual control if provided
    manual_control <- if (length(manual_controls) > 0 && 
                          range_name %in% names(manual_controls)) {
      manual_controls[[range_name]]
    } else {
      NA_real_
    }
    
    # Store original boundaries from settings
    settings_min <- my_range[1]
    settings_max <- my_range[2]
    
    # Replace NA with actual data boundaries for analysis
    if (is.na(my_range[1])) my_range[1] <- min(repeat_counts, na.rm = TRUE)
    if (is.na(my_range[2])) my_range[2] <- max(repeat_counts, na.rm = TRUE)
    
    # Filter to range to get observed boundaries
    counts_in_range <- repeat_counts[repeat_counts >= my_range[1] & 
                                     repeat_counts <= my_range[2]]
    
    # Observed boundaries (what's actually in the data)
    observed_min <- if (length(counts_in_range) > 0) {
      min(counts_in_range, na.rm = TRUE)
    } else {
      NA_real_
    }
    
    observed_max <- if (length(counts_in_range) > 0) {
      max(counts_in_range, na.rm = TRUE)
    } else {
      NA_real_
    }
    
    data.frame(
      file_stem = my_file_stem,
      analysis_range_name = range_name,
      settings_min = settings_min,
      settings_max = settings_max,
      analysis_min = my_range[1],
      analysis_max = my_range[2],
      observed_min = observed_min,
      observed_max = observed_max,
      floor_threshold = my_floor,
      max_peaks_allowed = my_max_peaks,
      manual_control_repeat_length = manual_control,
      stringsAsFactors = FALSE
    )
  })
  
  bind_rows(range_summaries)
})

# Combine all summaries
comprehensive_range_summary <- bind_rows(comprehensive_range_summary_list)

# Add group and time information
comprehensive_range_summary <- comprehensive_range_summary %>%
  left_join(
    select(settings, file_stem, group, group_control_sample, time),
    by = "file_stem"
  ) %>%
  relocate(file_stem, group, time, analysis_range_name)

message("  Rows created: ", nrow(comprehensive_range_summary))
message("  Samples: ", length(unique(comprehensive_range_summary$file_stem)))
message("  Ranges: ", length(unique(comprehensive_range_summary$analysis_range_name)))
```

```{r t_range_summary_end, echo=FALSE}
timings["Range summary"] <- proc.time()["elapsed"] - .t
```

**Column descriptions:**

- **settings_min/max**: Original range boundaries from settings file (may be NA for open ranges)
- **analysis_min/max**: Actual boundaries used for analysis (NA replaced with data min/max)
- **observed_min/max**: Minimum and maximum repeat lengths actually observed in the data within this range
- **floor_threshold**: Minimum read frequency threshold applied
- **max_peaks_allowed**: Maximum number of peaks allowed to detect
- **manual_control_repeat_length**: Manually specified control value (highlighted in yellow if present)

---

# Peak detection

Detect modal peaks within each analysis range for each sample.

```{r t_peaks_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r peak_detection, echo=FALSE, message=FALSE}
# Get unique file stems that have settings
file_stems_with_settings <- settings_parsed$file_stem

# Peak detection per sample and range
suppressMessages({
modal_peaks_list <- apply_fn(file_stems_with_settings, function(my_file_stem) {
  
  # Get sample data
  sample_data <- alignment_clustered %>%
    filter(file_stem == my_file_stem)
  
  # This should not happen since we pre-filtered settings
  if (nrow(sample_data) == 0) {
    warning("UNEXPECTED: No data for sample: ", my_file_stem, 
            " (should have been filtered out earlier)")
    return(NULL)
  }
  
  # Get settings for this sample
  my_settings <- settings_parsed %>%
    filter(file_stem == my_file_stem) %>%
    slice(1)
  
  # This should not happen since we pre-filtered settings
  if (nrow(my_settings) == 0) {
    warning("UNEXPECTED: No settings for sample: ", my_file_stem, 
            " (should have been filtered out earlier)")
    return(NULL)
  }
  
  # Get repeat counts
  repeat_counts <- sample_data[[params$repeat_count_method]]
  
  # Detect peaks for each range
  ranges <- my_settings$ranges_list[[1]]
  floor_vals <- my_settings$floor_list[[1]]
  max_peaks_vals <- my_settings$max_peaks_list[[1]]
  
  range_results <- lapply(names(ranges), function(range_name) {
    
    my_range <- ranges[[range_name]]
    my_floor <- floor_vals[[range_name]]
    my_max_peaks <- max_peaks_vals[[range_name]]
    
    # Store original window (before NA replacement)
    original_window_min <- my_range[1]
    original_window_max <- my_range[2]
    
    # Replace NA in range with data min/max
    if (is.na(my_range[1])) {
      my_range[1] <- min(repeat_counts, na.rm = TRUE)
    }
    if (is.na(my_range[2])) {
      my_range[2] <- max(repeat_counts, na.rm = TRUE)
    }
    
    # Find peaks
    peaks <- find_modal_peaks(
      repeat_counts = repeat_counts,
      range_min = my_range[1],
      range_max = my_range[2],
      floor = my_floor,
      max_peaks = my_max_peaks,
      span = params$range_peak_span
    )
    
    # Add metadata
    peaks %>%
      mutate(
        file_stem = my_file_stem,
        analysis_range_name = range_name,
        window_min = original_window_min,  # Settings boundary (may be NA)
        window_max = original_window_max,  # Settings boundary (may be NA)
        range_min = my_range[1],           # Actual boundary used
        range_max = my_range[2]            # Actual boundary used
      )
  })
  
  bind_rows(range_results)
})

# Combine all results
modal_peaks_list <- extract_apply_fn_result(modal_peaks_list)
})

modal_peaks <- bind_rows(modal_peaks_list) %>%
  relocate(file_stem, analysis_range_name)

message("\nPeak detection complete:")
message("  Samples: ", length(unique(modal_peaks$file_stem)))
message("  Total peaks detected: ", sum(!is.na(modal_peaks$modal_length)))

# Display peak table
datatable(modal_peaks,
          options = list(pageLength = 15, scrollX = TRUE),
          caption = "Modal peaks detected per sample and range",
          rownames = FALSE) %>%
  formatRound(columns = c('modal_length'), digits = 1) %>%
  formatStyle(columns = colnames(modal_peaks), fontSize = '10px')
```

```{r t_peaks_end, echo=FALSE}
timings["Peak detection"] <- proc.time()["elapsed"] - .t
```

```{r add_peaks_to_comprehensive_summary, echo=FALSE, message=FALSE}
# Add detected peaks to comprehensive summary
message("\n=== Adding Peaks to Comprehensive Summary ===")

# Get peaks in wide format for joining
message("  Converting peaks to wide format...")

# First, find the maximum number of peaks detected across all samples/ranges
max_peaks_found <- modal_peaks %>%
  filter(!is.na(modal_length)) %>%
  group_by(file_stem, analysis_range_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  pull(n) %>%
  max()

message("  Maximum peaks detected in any sample/range: ", max_peaks_found)

# Dynamically create peak columns based on what exists
peaks_wide <- modal_peaks %>%
  filter(!is.na(modal_length)) %>%
  arrange(file_stem, analysis_range_name, peak_rank) %>%
  group_by(file_stem, analysis_range_name) %>%
  summarise(
    n_peaks_detected = n(),
    .groups = "drop"
  )

# Add peak columns dynamically
for (i in 1:max_peaks_found) {
  # Extract peaks for this rank
  peak_data <- modal_peaks %>%
    filter(!is.na(modal_length), peak_rank == i) %>%
    select(file_stem, analysis_range_name, modal_length, modal_frequency) %>%
    rename(!!paste0("peak", i, "_length") := modal_length,
           !!paste0("peak", i, "_frequency") := modal_frequency)
  
  # Join to peaks_wide
  peaks_wide <- peaks_wide %>%
    left_join(peak_data, by = c("file_stem", "analysis_range_name"))
}

message("  Created ", max_peaks_found, " peak column pairs (length + frequency)")

# Join peaks to comprehensive summary
comprehensive_range_summary <- comprehensive_range_summary %>%
  left_join(peaks_wide, by = c("file_stem", "analysis_range_name"))

message("  Peaks added successfully")
if ("peak1_length" %in% colnames(comprehensive_range_summary)) {
  message("  Total peak1 detected: ", sum(!is.na(comprehensive_range_summary$peak1_length)))
}

# Dynamically identify peak columns for formatting
peak_cols <- grep("^peak[0-9]+_(length|frequency)$", 
                  colnames(comprehensive_range_summary), 
                  value = TRUE)
peak_length_cols <- grep("^peak[0-9]+_length$", peak_cols, value = TRUE)
peak_freq_cols <- grep("^peak[0-9]+_frequency$", peak_cols, value = TRUE)

message("  Peak columns in table: ", paste(peak_cols, collapse = ", "))

```

```{r plot_peaks, echo=FALSE, fig.width=12, fig.height=7, message=FALSE}
# Create plots directory
plots_dir <- file.path(params$dir_out, file.path("06_range_analysis", "plots"))
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

# Cache range boundaries for efficient per-sample lookups
range_boundaries_cache <- modal_peaks %>%
  select(file_stem, analysis_range_name, range_min, range_max) %>%
  distinct()

message(sprintf("Cached %d unique sample-range combinations", nrow(range_boundaries_cache)))


# Cohort plot: box+jitter per range, one point per sample, red mean diamond
# Note: range names (range1, range2, etc.) may have different boundaries in
# different samples — this plot shows the cohort distribution of peak values
# detected within each user-defined range window.
peak_cohort_data <- modal_peaks %>%
  filter(!is.na(modal_length), peak_rank == 1) %>%
  left_join(select(settings, file_stem, group), by = "file_stem")

p_peaks <- ggplot(peak_cohort_data,
                  aes(x = analysis_range_name, y = modal_length)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey50",
               width = 0.6) +
  geom_jitter(colour = "grey40", width = 0.2, height = 0,
              alpha = 0.7, size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4,
               colour = "red", alpha = 0.6) +
  labs(title = "Modal peak length by analysis range",
       subtitle = "Box = IQR; points = samples; red diamond = mean.\nNote: the same range name may span different boundaries in different samples.",
       x = "Analysis range",
       y = "Modal length (repeats)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_peaks)
ggsave(file.path(plots_dir, "03_modal_peaks.png"),
       p_peaks, width = 12, height = 7, dpi = params$plot_dpi)

# Generate individual peak plots by sample
message("\nGenerating individual peak plots by sample...")

modal_peaks_by_sample_dir <- file.path(plots_dir, "03_modal_peaks_by_sample")
if (!dir.exists(modal_peaks_by_sample_dir)) {
  dir.create(modal_peaks_by_sample_dir, recursive = TRUE)
}

for (my_sample in unique(modal_peaks$file_stem)) {
  sample_peaks <- modal_peaks %>%
    filter(file_stem == my_sample)
  
  if (nrow(sample_peaks) == 0) next
  
  # Get range boxes for this sample
  sample_boxes <- sample_peaks %>%
    select(analysis_range_name, window_min, window_max, range_min, range_max) %>%
    distinct() %>%
    mutate(y_position = seq_along(analysis_range_name))
  
  # Create peak plot for this sample
  p_sample_peaks <- ggplot() +
    # Range windows as shaded rectangles — uniform grey fill, no colour legend
    geom_rect(data = sample_boxes,
              aes(xmin = range_min, xmax = range_max,
                  ymin = y_position - 0.4, ymax = y_position + 0.4),
              fill = "grey80", alpha = 0.4) +
    # Detected peaks as points
    geom_point(data = sample_peaks %>% filter(!is.na(modal_length)),
               aes(x = modal_length, 
                   y = match(analysis_range_name, sample_boxes$analysis_range_name),
                   size = modal_frequency,
                   shape = factor(peak_rank)),
               colour = "black", alpha = 0.8) +
    # Labels for peaks
    geom_text(data = sample_peaks %>% filter(!is.na(modal_length)),
              aes(x = modal_length,
                  y = match(analysis_range_name, sample_boxes$analysis_range_name),
                  label = paste0(modal_length, " (", modal_frequency, ")")),
              vjust = -1.2, hjust = -0.1, angle = 45, size = 3) +
    scale_y_reverse(breaks = sample_boxes$y_position,
                    labels = sample_boxes$analysis_range_name) +
    scale_shape_manual(values = c("1" = 16, "2" = 17, "3" = 15),
                      name = "Peak rank
(1 = main)") +
    scale_size_continuous(range = c(3, 8), name = "Read count") +
    labs(title = paste0("Modal peaks - ", my_sample),
         subtitle = "Shaded regions = analysis range windows; Points = detected peaks; Labels = length (frequency)",
         x = "Repeat length",
         y = "Analysis range") +
    theme_bw() +
    theme(legend.position = "right",
          panel.grid.major.y = element_blank())
  
  ggsave(file.path(modal_peaks_by_sample_dir, 
                  paste0(my_sample, "-modal_peaks.png")),
         p_sample_peaks, width = 12, height = max(4, nrow(sample_boxes) * 1.5), 
         dpi = params$plot_dpi)
}

message("  Individual peak plots saved to: ", modal_peaks_by_sample_dir)
```

---

# Distribution summary

Calculate distribution statistics for each range and sample.

**Range boundary columns:**

- **`window_min/max`**: Original boundaries from settings (e.g., `[0-35]` → 0, 35; may be NA for open-ended)
- **`range_min/max`**: Actual boundaries used (NA values replaced with data min/max)
- **`observed_min/max`**: Actual minimum and maximum repeat lengths observed in the data within this window

**Floor filtering:**

Low-frequency repeat lengths can be filtered before calculating statistics to reduce noise:

- **`floor_threshold`**: The minimum frequency threshold from settings (can be absolute count or proportion)
  - If 0-1: Proportion of modal frequency (e.g., 0.05 = 5% of peak height)
  - If ≥1: Absolute count (e.g., 3 = must appear at least 3 times)
- **`n_reads_total`**: Total reads in the range window (before filtering)
- **`n_reads_used`**: Reads remaining after floor filtering (used for all statistics)

**Example:** If `floor_threshold = 3`, any repeat length appearing <3 times is excluded. If you have 5000 reads total but 200 are low-frequency singletons/doubletons, then `n_reads_used = 4800`.

```{r t_dist_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r distribution_summaries, echo=FALSE, message=FALSE}
# Calculate distribution summaries
suppressMessages({
distribution_summary_list <- apply_fn(file_stems_with_settings, function(my_file_stem) {
  
  # Get sample data
  sample_data <- alignment_clustered %>%
    filter(file_stem == my_file_stem)
  
  if (nrow(sample_data) == 0) return(NULL)
  
  # Get settings
  my_settings <- settings_parsed %>%
    filter(file_stem == my_file_stem) %>%
    slice(1)
  
  if (nrow(my_settings) == 0) return(NULL)
  
  # Get repeat counts
  repeat_counts <- sample_data[[params$repeat_count_method]]
  
  # Get ranges
  ranges <- my_settings$ranges_list[[1]]
  floor_vals <- my_settings$floor_list[[1]]
  
  # Summary for each range
  range_summaries <- lapply(names(ranges), function(range_name) {
    
    my_range <- ranges[[range_name]]
    my_floor <- floor_vals[[range_name]]
    
    # Store original window (before NA replacement)
    original_window_min <- my_range[1]
    original_window_max <- my_range[2]
    
    # Replace NA with data min/max
    if (is.na(my_range[1])) my_range[1] <- min(repeat_counts, na.rm = TRUE)
    if (is.na(my_range[2])) my_range[2] <- max(repeat_counts, na.rm = TRUE)
    
    # Filter to range
    counts_in_range <- repeat_counts[repeat_counts >= my_range[1] & 
                                     repeat_counts <= my_range[2]]
    
    # Calculate observed min/max within this window
    observed_min <- if (length(counts_in_range) > 0) min(counts_in_range, na.rm = TRUE) else NA_real_
    observed_max <- if (length(counts_in_range) > 0) max(counts_in_range, na.rm = TRUE) else NA_real_
    
    # Calculate summary
    summary_stats <- distribution_summary_function(
      repeat_counts = counts_in_range,
      floor = my_floor
    )
    
    # Convert to data frame
    as.data.frame(summary_stats) %>%
      mutate(
        file_stem = my_file_stem,
        analysis_range_name = range_name,
        window_min = original_window_min,  # Settings boundary (may be NA)
        window_max = original_window_max,  # Settings boundary (may be NA)
        range_min = my_range[1],           # Actual boundary used (NA replaced)
        range_max = my_range[2],           # Actual boundary used (NA replaced)
        observed_min = observed_min,       # Observed minimum in data
        observed_max = observed_max        # Observed maximum in data
      )
  })
  
  bind_rows(range_summaries)
})

# Combine all summaries
distribution_summary_list <- extract_apply_fn_result(distribution_summary_list)
})

distribution_summary <- bind_rows(distribution_summary_list) %>%
  relocate(file_stem, analysis_range_name)

message("\nDistribution summaries calculated:")
message("  Total summaries: ", nrow(distribution_summary))

# Join with settings for group info
distribution_summary <- distribution_summary %>%
  left_join(
    select(settings, file_stem, group, group_control_sample, time, exclude),
    by = "file_stem"
  )

# Display summary table (key columns only)
key_cols <- c("file_stem", "analysis_range_name", 
              "window_min", "window_max", "range_min", "range_max",
              "observed_min", "observed_max",
              "modal_length", "modal_frequency", "mean_length", "median_length", 
              "sd", "floor_threshold", "n_reads_total", "n_reads_used")

datatable(distribution_summary[, key_cols],
          options = list(pageLength = 15, scrollX = TRUE),
          caption = "Distribution summary statistics (key metrics)",
          rownames = FALSE) %>%
  formatRound(columns = c('window_min', 'window_max', 'range_min', 'range_max',
                         'observed_min', 'observed_max',
                         'modal_length', 'mean_length', 'median_length', 'sd'), 
              digits = 2) %>%
  formatStyle(columns = key_cols, fontSize = '10px')
```

```{r t_dist_end, echo=FALSE}
timings["Distribution summaries"] <- proc.time()["elapsed"] - .t
```

```{r plot_distributions, echo=FALSE, fig.width=10, fig.height=8, message=FALSE}
# --- Clean non-finite values (NaN, Inf, -Inf) before plotting ---
# Edge-case samples (e.g. all reads at modal → tail_balance = 0/0 = NaN,
# mean ≈ 0 → CV = Inf) cause ggplot to fail when computing axis breaks.
distribution_summary <- distribution_summary %>%
  mutate(across(where(is.numeric), ~ifelse(is.finite(.) | is.na(.), ., NA_real_)))

# ── Plot 02: Modal vs Mean (faceted by range, each point = one sample) ─────────
# Compute per-range lm stats for annotation
modal_mean_lm <- distribution_summary %>%
  group_by(analysis_range_name) %>%
  summarise(
    slope     = tryCatch({coef(lm(mean_length ~ modal_length, data = cur_data()))[2]}, error = function(e) NA_real_),
    intercept = tryCatch({coef(lm(mean_length ~ modal_length, data = cur_data()))[1]}, error = function(e) NA_real_),
    r2        = tryCatch({summary(lm(mean_length ~ modal_length, data = cur_data()))$r.squared}, error = function(e) NA_real_),
    x_pos = max(modal_length, na.rm = TRUE),
    y_pos = min(mean_length,  na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    sign_str = ifelse(!is.na(intercept), ifelse(intercept >= 0, " + ", " - "), ""),
    abs_icpt = ifelse(!is.na(intercept), abs(round(intercept, 2)), NA_real_),
    label    = ifelse(!is.na(slope),
                      paste0("y = ", round(slope, 2), "x", sign_str, abs_icpt,
                             "\nR² = ", round(r2, 3)),
                      NA_character_)
  )

p_modal_mean <- ggplot(distribution_summary,
                       aes(x = modal_length, y = mean_length)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, colour = "#4393C3") +
  geom_point(size = 3, alpha = 0.75, colour = "grey30") +
  geom_text(data = modal_mean_lm %>% filter(!is.na(label)),
            aes(x = x_pos, y = y_pos, label = label),
            colour = "grey20", size = 2.8, hjust = 1, vjust = 0, inherit.aes = FALSE) +
  facet_wrap(~ analysis_range_name, scales = "free") +
  labs(title = "Modal vs mean repeat lengths by range",
       subtitle = paste0("Each point = one sample. Blue line = lm fit across all samples; dashed = identity.",
                         "\nNote: the same range name may span different boundaries in different samples."),
       x = "Modal length (repeats)",
       y = "Mean length (repeats)") +
  theme_bw() +
  theme(strip.text = element_text(size = 8))
print(p_modal_mean)
ggsave(file.path(plots_dir, "04_modal_vs_mean.png"),
       p_modal_mean, width = 16, height = 12, dpi = params$plot_dpi)

# ── Plot 03: Distribution spread (SD) by range ──────────────────────────────────
p_spread <- ggplot(distribution_summary,
                   aes(x = analysis_range_name, y = sd)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey50", width = 0.6) +
  geom_jitter(colour = "grey40", width = 0.2, height = 0, alpha = 0.7, size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4,
               colour = "red", alpha = 0.6) +
  labs(title = "Distribution spread by analysis range",
       subtitle = "Standard deviation per range. Box = IQR; points = samples; red diamond = mean.",
       x = "Analysis range",
       y = "Standard deviation") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_spread)
ggsave(file.path(plots_dir, "05_spread_by_range.png"),
       p_spread, width = 12, height = 6, dpi = params$plot_dpi)

# ── Plot 04: Floor filtering effect by range ─────────────────────────────────────
filtering_data <- distribution_summary %>%
  filter(n_reads_total > 0) %>%
  mutate(pct_filtered = ((n_reads_total - n_reads_used) / n_reads_total) * 100)

p_filtering <- ggplot(filtering_data,
                      aes(x = analysis_range_name, y = pct_filtered)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey50", width = 0.6) +
  geom_jitter(colour = "grey40", width = 0.2, height = 0, alpha = 0.7, size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4,
               colour = "red", alpha = 0.6) +
  labs(title = "Floor filtering effect by analysis range",
       subtitle = "Percentage of reads removed by floor threshold. Floor values in distribution_summary Excel tab. Box = IQR; points = samples; red diamond = mean.",
       x = "Analysis range",
       y = "% reads filtered") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_filtering)
ggsave(file.path(plots_dir, "06_filtering_effect_by_range.png"),
       p_filtering, width = 12, height = 6, dpi = params$plot_dpi)

# ── Plot 05: Tail balance by range ───────────────────────────────────────────────
# Grouped box+jitter: contraction / at modal / expansion proportions per range
tail_balance_long <- distribution_summary %>%
  mutate( at_modal_proportion = 1 - contraction_proportion - expansion_proportion) %>%
  select(file_stem, analysis_range_name, group,
         contraction_proportion, at_modal_proportion, expansion_proportion) %>%
  pivot_longer(
    cols = c(contraction_proportion, at_modal_proportion, expansion_proportion),
    names_to = "direction", values_to = "proportion"
  ) %>%
  mutate(direction = case_when(
    direction == "contraction_proportion" ~ "Contraction",
    direction == "at_modal_proportion"    ~ "At modal",
    direction == "expansion_proportion"   ~ "Expansion"
  ),
  direction = factor(direction, levels = c("Contraction", "At modal", "Expansion")))

p_tail_balance <- ggplot(tail_balance_long,
                         aes(x = analysis_range_name, y = proportion, fill = direction)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.7,
               position = position_dodge(0.8)) +
  geom_jitter(aes(colour = direction),
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.6, size = 1.5) +
  scale_fill_manual(
    values = c("Contraction" = "#2166AC", "At modal" = "#F7F7F7", "Expansion" = "#D6604D")
  ) +
  scale_colour_manual(
    values = c("Contraction" = "#2166AC", "At modal" = "#AAAAAA",  "Expansion" = "#D6604D")
  ) +
  labs(title = "Tail balance by analysis range",
       subtitle = "Proportion of reads in contraction, at modal, and expansion tails. Box = IQR; points = samples.",
       x = "Analysis range",
       y = "Proportion of reads",
       fill = "Direction", colour = "Direction") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
print(p_tail_balance)
ggsave(file.path(plots_dir, "07_tail_balance_by_range.png"),
       p_tail_balance, width = 14, height = 7, dpi = params$plot_dpi)

# ── Plot 06: Coefficient of Variation by range ───────────────────────────────────
p_cv <- ggplot(distribution_summary,
               aes(x = analysis_range_name, y = coefficient_of_variation)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey50", width = 0.6) +
  geom_jitter(colour = "grey40", width = 0.2, height = 0, alpha = 0.7, size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4,
               colour = "red", alpha = 0.6) +
  labs(title = "Coefficient of variation by analysis range",
       subtitle = "Relative variability (SD / mean) per range. Box = IQR; points = samples; red diamond = mean.",
       x = "Analysis range",
       y = "Coefficient of variation (CV)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p_cv)
ggsave(file.path(plots_dir, "08_cv_by_range.png"),
       p_cv, width = 12, height = 6, dpi = params$plot_dpi)

# ── Plot 08: Skewness vs Kurtosis (faceted by range) ─────────────────────────────
shape_data <- distribution_summary %>%
  filter(!is.na(skewness), !is.na(kurtosis))

if (nrow(shape_data) > 0) {
  # Compute per-range lm stats for annotation
    shape_lm_stats <- shape_data %>%
    group_by(analysis_range_name) %>%
    summarise(
      slope     = tryCatch({coef(lm(kurtosis ~ skewness, data = cur_data()))[2]}, error = function(e) NA_real_),
      intercept = tryCatch({coef(lm(kurtosis ~ skewness, data = cur_data()))[1]}, error = function(e) NA_real_),
      r2        = tryCatch({summary(lm(kurtosis ~ skewness, data = cur_data()))$r.squared}, error = function(e) NA_real_),
      .groups = "drop"
    ) %>%
    mutate(
      sign_str = ifelse(!is.na(intercept), ifelse(intercept >= 0, " + ", " - "), ""),
      abs_icpt = ifelse(!is.na(intercept), abs(round(intercept, 2)), NA_real_),
      label = ifelse(!is.na(slope),
                     paste0("y = ", round(slope, 2), "x", sign_str, abs_icpt,
                            "\nR\u00b2 = ", round(r2, 3)),
                     NA_character_),
      # Inf/-Inf places label at bottom-right corner regardless of per-panel axis range
      x_pos = Inf, y_pos = -Inf)

  p_shape <- ggplot(shape_data,
                    aes(x = skewness, y = kurtosis)) +
    geom_hline(yintercept = 3, linetype = "dashed", colour = "grey50", alpha = 0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
    geom_point(size = 3, alpha = 0.7) +
    geom_text(data = shape_lm_stats %>% filter(!is.na(label)),
              aes(x = x_pos, y = y_pos, label = label),
              colour = "grey30", size = 2.8, hjust = 1, vjust = 0, inherit.aes = FALSE) +
    facet_wrap(~ analysis_range_name, scales = "free") +
    labs(title = "Distribution shape: skewness vs kurtosis by range",
         subtitle = "Each point = one sample. Reference lines: skewness=0 (symmetric), kurtosis=3 (normal). Line = lm fit.",
         x = "Skewness",
         y = "Kurtosis",
  ) +
    theme_bw() +
    theme(legend.position = "bottom",
          strip.text = element_text(size = 8))

  # Save to disk only — many panels would be heavy inline
  ggsave(file.path(plots_dir, "09_distribution_shape.png"),
         p_shape, width = 14, height = 10, dpi = params$plot_dpi)
  message("Distribution shape plot saved to: 09_distribution_shape.png")
}

# Plot 21: Density plots by group and time (full range + per analysis range)
message("\nGenerating density plots by group and time...")

# Create directory for density plots
density_by_group_and_time_dir <- file.path(plots_dir, "10_density_by_group_and_time")
if (!dir.exists(density_by_group_and_time_dir)) {
  dir.create(density_by_group_and_time_dir, recursive = TRUE)
}

# Prepare base data: join repeat counts with group/time metadata
density_data_full <- alignment_clustered %>%
  left_join(
    select(settings, file_stem, group, time),
    by = "file_stem"
  ) %>%
  filter(!is.na(group), !is.na(time),
         !is.na(.data[[params$repeat_count_method]]),
         is.finite(.data[[params$repeat_count_method]])) %>%
  arrange(group, time) %>%
  mutate(
    sample_label = paste0(file_stem, " (t=", time, ")"),
    # Order samples by time within each group (earliest = lowest number)
    time_order = as.numeric(time)
  )

if (nrow(density_data_full) > 0) {
  
  # Get unique analysis ranges from settings
  all_ranges <- unique(unlist(lapply(settings_parsed$ranges_list, names)))
  
  # Create a list of range definitions: "Full" + each analysis range
  range_list <- c(
    list(Full = c(-Inf, Inf)),  # Full range (all data)
    setNames(
      lapply(all_ranges, function(rng) {
        # Get the most common range boundaries for this range name
        range_bounds <- sapply(settings_parsed$ranges_list, function(x) {
          if (rng %in% names(x)) x[[rng]] else c(NA, NA)
        })
        # Take the first non-NA range
        valid_bounds <- range_bounds[, !is.na(range_bounds[1,]), drop = FALSE]
        if (ncol(valid_bounds) > 0) {
          bounds <- valid_bounds[, 1]
          # Replace NA with -Inf/Inf for filtering
          c(ifelse(is.na(bounds[1]), -Inf, bounds[1]),
            ifelse(is.na(bounds[2]), Inf, bounds[2]))
        } else {
          c(-Inf, Inf)
        }
      }),
      all_ranges
    )
  )
  
  # Generate plots for each range
  for (range_name in names(range_list)) {
    
    message("\n  Processing range: ", range_name)
    
    # Filter data by range
    range_bounds <- range_list[[range_name]]
    density_data_range <- density_data_full %>%
      filter(.data[[params$repeat_count_method]] >= range_bounds[1] &
               .data[[params$repeat_count_method]] <= range_bounds[2])
    
    if (nrow(density_data_range) == 0) {
      message("    No data in this range, skipping")
      next
    }
    
    # Create subdirectory for this range
    density_range_dir <- file.path(density_by_group_and_time_dir, range_name)
    if (!dir.exists(density_range_dir)) {
      dir.create(density_range_dir, recursive = TRUE)
    }
    
    # COMBINED PLOT: All groups in columns for this range
    p_density_combined <- ggplot(density_data_range, 
                              aes(x = .data[[params$repeat_count_method]], 
                                  y = reorder(sample_label, -time_order),  # Earliest at top
                                  fill = group)) +
      geom_density_ridges(alpha = 0.7, scale = 2, rel_min_height = 0.01,
                          jittered_points = TRUE,
                          position = position_points_jitter(width = 0.5, height = 0.15),
                          point_shape = 16, point_size = 0.4, point_alpha = 0.2) +
      facet_grid(~ group, scales = "free_y") +  # Groups as COLUMNS
      scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
      labs(
        title = paste0("Repeat length distribution by group and time - ", range_name),
        subtitle = "Samples ordered by time within each group (earliest at top)",
        x = "Repeat count",
        y = "Sample (Time)",
        fill = "Group"
      ) +
      theme_minimal() +
      theme(
        legend.position = "none",
        strip.text = element_text(face = "bold", size = 12),
        axis.text.y = element_text(size = 8)
      )
    
    # Not printed inline — saved to disk to keep HTML compact
    # (17 ridge plots × multiple groups = significant file size)
    
    # Calculate dimensions
    n_samples_range <- length(unique(density_data_range$file_stem))
    n_groups <- length(unique(density_data_range$group))
    plot_height <- max(8, min(50, n_samples_range * 0.4))
    plot_width <- max(10, min(40, n_groups * 8))
    
    # Determine filename based on range
    combined_filename <- if (range_name == "Full") {
      "density.png"
    } else {
      paste0("density-", range_name, ".png")
    }
    
    ggsave(
      file.path(density_range_dir, combined_filename),
      p_density_combined,
      width = plot_width,
      height = plot_height,
      dpi = params$plot_dpi,
      bg = "white",
      limitsize = FALSE
    )
    
    message("    Combined plot saved: ", combined_filename)
    
    # INDIVIDUAL PLOTS: One per group for this range
    for (my_group in unique(density_data_range$group)) {
      group_data <- density_data_range %>%
        filter(group == my_group) %>%
        arrange(time)
      
      if (nrow(group_data) == 0) next
      
      # Create plot for this group
      p_density_group <- ggplot(group_data, 
                             aes(x = .data[[params$repeat_count_method]], 
                                 y = reorder(sample_label, -time_order),  # Earliest at top
                                 fill = group)) +
        geom_density_ridges(alpha = 0.7, scale = 2, rel_min_height = 0.01,
                            jittered_points = TRUE,
                            position = position_points_jitter(width = 0.5, height = 0.15),
                            point_shape = 16, point_size = 0.4, point_alpha = 0.2) +
        scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
        labs(
          title = paste0("Repeat length distribution - ", my_group, " (", range_name, ")"),
          subtitle = "Samples ordered by time (earliest at top)",
          x = "Repeat count",
          y = "Sample (Time)",
          fill = "Group"
        ) +
        theme_minimal() +
        theme(
          legend.position = "none",
          axis.text.y = element_text(size = 9)
        )
      
      # Calculate height for this group
      n_samples_group <- length(unique(group_data$file_stem))
      group_height <- max(6, min(40, n_samples_group * 0.5))
      
      # Determine filename based on range
      group_filename <- if (range_name == "Full") {
        paste0("density-", my_group, ".png")
      } else {
        paste0("density-", my_group, "-", range_name, ".png")
      }
      
      ggsave(
        file.path(density_range_dir, group_filename),
        p_density_group,
        width = 12,
        height = group_height,
        dpi = params$plot_dpi,
        bg = "white",
        limitsize = FALSE
      )
      
      message("    Density plot saved for group: ", my_group, " (", group_filename, ")")
    }
  }
  
  message("\n  All ridge plots saved to: ", density_by_group_and_time_dir)
  message("  (Not printed inline — open files directly to view)")
  
} else {
  message("  Warning: No data available for ridge plots (missing group/time metadata)")
}


message("\nDistribution plots saved to: ", plots_dir)
```

---

# Group controls

Define control repeat lengths for each group and analysis range.

Group controls provide baseline reference values for calculating instability metrics.

**Sample selection methods:**

- `flagged`: Use samples flagged with `group_control_sample = 1`
- `earliest`: Use first timepoint (time == min)
- `all`: Average all samples in group

**Setpoint metrics:**

- `modal_length`: Most frequent repeat length (robust, default)
- `mean_length`: Arithmetic mean (sensitive to outliers)
- `median_length`: Median repeat length (robust middle value)

```{r t_controls_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r group_controls, echo=FALSE, message=FALSE}
if (params$group_control) {
  
  # Map control_setpoint_metric to column name
  metric_column_map <- c(
    "modal_length" = "modal_length",
    "mean_length" = "mean_length",
    "median_length" = "median_length"
  )
  
  # Get the column name to use
  if (!params$control_setpoint_metric %in% names(metric_column_map)) {
    stop("Invalid control_setpoint_metric: ", params$control_setpoint_metric,
         ". Must be one of: ", paste(names(metric_column_map), collapse = ", "))
  }
  
  metric_column <- metric_column_map[[params$control_setpoint_metric]]
  
  # Aggregator across control samples (mirrors calculate_control_setpoint() in 06_range_analysis.R)
  valid_agg_methods <- c("mean", "median", "trimmed_mean")
  if (!params$control_aggregation_method %in% valid_agg_methods) {
    stop("Invalid control_aggregation_method: ", params$control_aggregation_method,
         ". Must be one of: ", paste(valid_agg_methods, collapse = ", "))
  }
  agg_fun <- switch(params$control_aggregation_method,
    "mean"         = function(x) mean(x, na.rm = TRUE),
    "median"       = function(x) median(x, na.rm = TRUE),
    "trimmed_mean" = function(x) mean(x, trim = 0.1, na.rm = TRUE)
  )
  
  # Get unique analysis ranges
  analysis_range_names <- unique(distribution_summary$analysis_range_name)
  
  # Store control sample data for plotting
  control_samples_list <- list()
  
  # Calculate group controls for each range
  group_control_list <- lapply(analysis_range_names, function(my_range_name) {
    
    # Filter to this range
    control_data <- distribution_summary %>%
      filter(
        analysis_range_name == my_range_name,
        is.na(exclude),
        !is.na(!!sym(metric_column)),
        !is.na(group)
      )
    
    if (nrow(control_data) == 0) {
      return(NULL)
    }
    
    # Select control samples based on method
    if (params$control_sample_selection == "earliest") {
      # Use earliest timepoint
      control_data <- control_data %>%
        group_by(group) %>%
        filter(time == min(time, na.rm = TRUE)) %>%
        ungroup()
    } else if (params$control_sample_selection == "flagged") {
      # Use flagged controls
      control_data <- control_data %>%
        filter(!is.na(group_control_sample))
    }
    # If "all", use all samples (no filtering)
    
    # Store control samples for this range
    control_samples_list[[my_range_name]] <<- control_data
    
    # Calculate group control values for all three metrics
    control_summary <- control_data %>%
      group_by(group) %>%
      summarise(
        control_modal_length = agg_fun(modal_length),
        control_mean_length = agg_fun(mean_length),
        control_median_length = agg_fun(median_length),
        control_count = n(),
        control_sd = sd(!!sym(metric_column), na.rm = TRUE),
        control_sem = control_sd / sqrt(control_count),
        .groups = "drop"
      ) %>%
      mutate(
        analysis_range_name = my_range_name,
        # Set control setpoint based on user's choice
        control_setpoint = !!sym(paste0("control_", params$control_setpoint_metric)),
        control_metric_used = params$control_setpoint_metric,
        control_aggregation_used = params$control_aggregation_method
      )
    
    return(control_summary)
  })
  
  # Combine results
  group_controls <- bind_rows(group_control_list) %>%
    relocate(group, analysis_range_name)
  
  message("\nGroup controls calculated:")
  message("  Groups: ", length(unique(group_controls$group)))
  message("  Ranges: ", length(unique(group_controls$analysis_range_name)))
  message("  Sample selection: ", params$control_sample_selection)
  message("  Setpoint metric: ", params$control_setpoint_metric)
  
  # Display group controls
  caption_text <- paste0("Group control setpoints per analysis range (using ", 
                        params$control_setpoint_metric, ")")
  
  DT::datatable(
    group_controls,
    caption = caption_text,
    options = list(
      pageLength = 25,
      scrollX = TRUE,
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel')
    ),
    rownames = FALSE,
    filter = 'top',
    class = 'cell-border stripe compact'
  ) %>%
    DT::formatRound(columns = colnames(group_controls)[sapply(group_controls, is.numeric)], digits = 2) %>%
    DT::formatStyle(columns = colnames(group_controls), fontSize = '10px')
  
} else {
  message("\nGroup controls disabled (group_control = FALSE)")
  group_controls <- NULL
}
```

```{r t_controls_end, echo=FALSE}
timings["Group controls"] <- proc.time()["elapsed"] - .t
```

```{r add_controls_to_comprehensive_summary, echo=FALSE, message=FALSE}
# Add control information to comprehensive summary
if (params$group_control && exists("group_controls") && !is.null(group_controls)) {
  
  message("\n=== Adding Group Controls to Comprehensive Summary ===")
  
  # Get control setpoints per group and range
  control_setpoints <- group_controls %>%
    select(group, analysis_range_name, control_setpoint, control_metric_used,
           control_aggregation_used)
  
  # Join to comprehensive summary
  comprehensive_range_summary <- comprehensive_range_summary %>%
    left_join(control_setpoints, by = c("group", "analysis_range_name"))
  
  message("  Control setpoints added successfully")
  
  # Dynamically identify peak columns for formatting
  peak_cols <- grep("^peak[0-9]+_(length|frequency)$", 
                    colnames(comprehensive_range_summary), 
                    value = TRUE)
  peak_length_cols <- grep("^peak[0-9]+_length$", peak_cols, value = TRUE)
  peak_freq_cols <- grep("^peak[0-9]+_frequency$", peak_cols, value = TRUE)
  
  # Display final comprehensive table
  dt_final <- DT::datatable(
    comprehensive_range_summary,
    caption = "Range summary - configuration, peaks, and controls",
    extensions = c('Buttons', 'FixedColumns'),
    options = list(
      pageLength = 25,
      scrollX = TRUE,
      scrollY = "700px",
      dom = 'Bfrtip',
      buttons = list(
        'copy',
        list(extend = 'csv', filename = 'comprehensive_range_summary_complete'),
        list(extend = 'excel', filename = 'comprehensive_range_summary_complete'),
        'colvis',
        list(
          extend = 'pdfHtml5',
          orientation = 'landscape',
          pageSize = 'A3'
        )
      ),
      fixedColumns = list(leftColumns = 2),
      columnDefs = list(
        list(width = '110px', targets = 0:1)
      )
    ),
    rownames = FALSE,
    filter = 'top',
    class = 'cell-border stripe compact hover',
    style = 'bootstrap'
  )
  
  # Format numeric columns
  numeric_cols <- c('settings_min', 'settings_max', 'analysis_min', 'analysis_max',
                    'observed_min', 'observed_max', 
                    'manual_control_repeat_length', 'control_setpoint')
  numeric_cols <- c(numeric_cols, peak_length_cols)
  numeric_cols <- numeric_cols[numeric_cols %in% colnames(comprehensive_range_summary)]
  
  dt_final <- dt_final %>%
    DT::formatRound(columns = numeric_cols, digits = 1)
  
  # Highlight peak1 (most important) if it exists
  if ("peak1_length" %in% colnames(comprehensive_range_summary)) {
    dt_final <- dt_final %>%
      DT::formatStyle(
        c('peak1_length', 'peak1_frequency'),
        fontWeight = 'bold',
        backgroundColor = '#e3f2fd'
      )
  }
  
  # Highlight peak2 if it exists
  if ("peak2_length" %in% colnames(comprehensive_range_summary)) {
    dt_final <- dt_final %>%
      DT::formatStyle(
        c('peak2_length', 'peak2_frequency'),
        backgroundColor = '#f0f0f0'
      )
  }
  
  # Highlight additional peaks (peak3, peak4, etc.) with progressively lighter grey
  for (i in 3:length(peak_length_cols)) {
    peak_len_col <- paste0("peak", i, "_length")
    peak_freq_col <- paste0("peak", i, "_frequency")
    if (peak_len_col %in% colnames(comprehensive_range_summary)) {
      # Progressively lighter grey: #f5f5f5, #f8f8f8, #fafafa
      grey_value <- sprintf("#f%df%df%d", 2+i, 2+i, 2+i)
      dt_final <- dt_final %>%
        DT::formatStyle(
          c(peak_len_col, peak_freq_col),
          backgroundColor = grey_value
        )
    }
  }
  
  # Highlight manual controls and control setpoints
  dt_final <- dt_final %>%
    DT::formatStyle(
      'manual_control_repeat_length',
      backgroundColor = DT::styleEqual(NA, 'white', default = '#fff3cd')
    ) %>%
    DT::formatStyle(
      'control_setpoint',
      backgroundColor = DT::styleEqual(NA, 'white', default = '#d4edda'),
      fontWeight = 'bold'
    ) %>%
    DT::formatStyle(
      columns = colnames(comprehensive_range_summary),
      fontSize = '10px'
    )
  
  dt_final
}
```

```{r plot_group_controls, echo=FALSE, fig.width=10, fig.height=5, message=FALSE}
if (params$group_control && !is.null(group_controls)) {

  # ── Plot 09: Control setpoints — one point per group per range ────────────────
  # Simple dot plot: x = range, y = setpoint, colour = group.
  # Interpretable regardless of cohort size.
  p_controls <- ggplot(group_controls,
                       aes(x = analysis_range_name, y = control_setpoint,
                           colour = group)) +
    geom_point(size = 4, alpha = 0.85) +
    labs(title = "Group control setpoints by analysis range",
         subtitle = paste0("One point per group per range, using ", params$control_setpoint_metric,
                           ". All values also exported in Excel."),
         x = "Analysis range",
         y = paste0(tools::toTitleCase(gsub("_", " ", params$control_setpoint_metric)),
                    " (repeats)"),
         colour = "Group") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "right")

  print(p_controls)
  ggsave(file.path(plots_dir, "11_control_setpoints.png"),
         p_controls, width = 10, height = 5, dpi = params$plot_dpi)

  # ── Control sample counts — inline table, no plot ─────────────────────────────
  control_samples <- bind_rows(control_samples_list)
  sample_counts <- control_samples %>%
    group_by(group, analysis_range_name) %>%
    summarise(n_control_samples = n(), .groups = "drop")

  DT::datatable(
    sample_counts,
    caption = "Control sample counts per group and range",
    options = list(pageLength = 25, scrollX = TRUE, dom = 't'),
    rownames = FALSE,
    class = 'cell-border stripe compact'
  ) %>%
    DT::formatStyle(columns = colnames(sample_counts), fontSize = '10px')
}
```

---

# Instability metrics

Calculate instability metrics relative to setpoints using the actual read distribution.

**Two setpoint types (both using user-selected metric):**

1. **"sample"** - Each sample's own value of the chosen metric
2. **"control"** - Group control (aggregated from control samples using the same metric)

**User-selected metric:** `r params$control_setpoint_metric`  
**Control aggregation:** `r params$control_aggregation_method` of control samples

**Example:** If user selects `modal_length` with `median` aggregation:

- Sample setpoint = Sample's modal_length (e.g., 19)
- Control setpoint = Median of control samples' modal_lengths (e.g., 17)

For each sample and setpoint type, we calculate:

- **Instability indices**: Mean deviations from setpoint (instability, expansion, contraction)
- **Read counts**: Number of reads at, below, and above setpoint
- **Tail ratios**: Relative sizes of expansion and contraction tails

```{r t_instability_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r instability_metrics, echo=FALSE, message=FALSE}
# ALWAYS calculate instability metrics (at minimum, relative to sample itself)
# If group_control = TRUE, also calculate relative to control setpoint

# Instability functions are in range_analysis.R (already sourced)

# Get unique analysis ranges
analysis_range_names <- unique(distribution_summary$analysis_range_name)

# Calculate instability for each range
instability_metrics_list <- lapply(analysis_range_names, function(my_range_name) {
  
  # Get distribution summaries for this range
  range_data <- distribution_summary %>%
    filter(analysis_range_name == my_range_name)
  
  # Get control setpoints for this range (if available)
  if (params$group_control && !is.null(group_controls)) {
    range_controls <- group_controls %>%
      filter(analysis_range_name == my_range_name)
  } else {
    range_controls <- NULL
  }
  
  # Calculate instability for each sample (parallelised)
  sample_file_stems <- unique(range_data$file_stem)
  
  sample_instability_list <- apply_fn(sample_file_stems, function(my_file_stem) {
    
    # Get sample's distribution summary
    sample_summary <- range_data %>%
      filter(file_stem == my_file_stem)
    
    if (nrow(sample_summary) == 0) return(NULL)
    
    # Get sample's group
    sample_group <- sample_summary$group[1]
    
    # Get raw read data for this sample/range
    sample_reads <- alignment_clustered %>%
      filter(
        file_stem == my_file_stem,
        !!sym(params$repeat_count_method) >= sample_summary$range_min[1],
        !!sym(params$repeat_count_method) <= sample_summary$range_max[1]
      ) %>%
      pull(!!sym(params$repeat_count_method))
    
    if (length(sample_reads) == 0) {
      warning(sprintf("No reads for sample '%s' in range '%s'", my_file_stem, my_range_name))
      return(NULL)
    }
    
    # Get floor for this range
    floor_val <- sample_summary$floor_threshold[1]
    
    # Get sample's own setpoint value
    sample_setpoint <- sample_summary[[params$control_setpoint_metric]][1]
    
    if (is.na(sample_setpoint)) {
      warning(sprintf("Sample '%s' has NA for metric '%s'", 
                     my_file_stem, params$control_setpoint_metric))
      return(NULL)
    }
    
    # ALWAYS calculate instability relative to SAMPLE's own setpoint
    instability_sample <- calculate_instability_metrics(
      numbers = sample_reads,
      setpoint = sample_setpoint,
      floor = floor_val
    ) %>%
      mutate(
        file_stem = my_file_stem,
        analysis_range_name = my_range_name,
        group = sample_group,
        setpoint_type = "sample_relative",
        setpoint_metric = params$control_setpoint_metric,
        setpoint_aggregation = NA_character_
      )
    
    # CONDITIONALLY calculate instability relative to CONTROL setpoint
    instability_control <- NULL
    
    if (params$group_control && !is.null(range_controls) && !is.na(sample_group)) {
      # Get control setpoint for this sample's group
      group_control <- range_controls %>%
        filter(group == sample_group)
      
      if (nrow(group_control) > 0) {
        control_setpoint_value <- group_control$control_setpoint[1]
        
        if (!is.na(control_setpoint_value)) {
          instability_control <- calculate_instability_metrics(
            numbers = sample_reads,
            setpoint = control_setpoint_value,
            floor = floor_val
          ) %>%
            mutate(
              file_stem = my_file_stem,
              analysis_range_name = my_range_name,
              group = sample_group,
              setpoint_type = "control_relative",
              setpoint_metric = params$control_setpoint_metric,
              setpoint_aggregation = params$control_aggregation_method,
              control_sd = group_control$control_sd[1],
              control_sem = group_control$control_sem[1],
              control_n = group_control$control_count[1]
            )
          
          # Add z-score for control comparison
          instability_control <- instability_control %>%
            mutate(
              z_score = if_else(
                !is.na(control_sd) & control_sd > 0,
                (sample_setpoint - setpoint_repeat_length) / control_sd,
                NA_real_
              )
            )
        }
      }
    }
    
    # Combine sample_relative and control_relative (if exists)
    bind_rows(instability_sample, instability_control)
  })
  
  # Extract results from apply_fn
  sample_instability_list <- extract_apply_fn_result(sample_instability_list)
  
  # Combine all samples for this range
  bind_rows(sample_instability_list)
})

# Combine all ranges
instability_metrics <- bind_rows(instability_metrics_list) %>%
  relocate(file_stem, analysis_range_name, group, setpoint_type,
           setpoint_metric, setpoint_aggregation)

# Add relative distribution statistics (modal, median, mean differences from setpoint)
instability_metrics <- instability_metrics %>%
  left_join(
    distribution_summary %>% 
      select(file_stem, analysis_range_name, 
             modal_length, median_length, mean_length),
    by = c("file_stem", "analysis_range_name")
  ) %>%
  mutate(
    # Calculate differences from setpoint
    modal_diff = modal_length - setpoint_repeat_length,
    median_diff = median_length - setpoint_repeat_length,
    mean_diff = mean_length - setpoint_repeat_length
  ) %>%
  # Reorder columns logically
  relocate(file_stem, analysis_range_name, group, setpoint_type,
           setpoint_metric, setpoint_aggregation,
           setpoint_repeat_length, 
           modal_length, median_length, mean_length,
           modal_diff, median_diff, mean_diff)

message("\nInstability metrics calculated:")
message("  Samples: ", length(unique(instability_metrics$file_stem)))
message("  Ranges: ", length(unique(instability_metrics$analysis_range_name)))
message("  Setpoint types: ", paste(unique(instability_metrics$setpoint_type), collapse = ", "))
message("  Setpoint metric: ", params$control_setpoint_metric)
if (params$group_control) {
  message("  Control aggregation: ", params$control_aggregation_method)
}
message("  Includes relative distribution statistics: modal_diff, median_diff, mean_diff")
```

```{r t_instability_end, echo=FALSE}
timings["Instability metrics"] <- proc.time()["elapsed"] - .t
```

```{r instability_display, echo=FALSE, message=FALSE}
if (nrow(instability_metrics) > 0) {
  
  # Display key columns including new relative distribution statistics
  display_cols <- c("file_stem", "analysis_range_name", "setpoint_type", 
                   "setpoint_repeat_length",
                   "modal_diff", "median_diff", "mean_diff",
                   "instability_index", "expansion_index", 
                   "contraction_index", "expansion_ratio", "z_score")
  
  caption_text <- if (params$group_control) {
    paste0("Instability metrics with relative distribution statistics (using ", 
           params$control_setpoint_metric, 
           " with ", params$control_aggregation_method, " of controls)")
  } else {
    paste0("Instability metrics with relative distribution statistics (using ", 
           params$control_setpoint_metric, ")")
  }
  
  datatable(instability_metrics[, display_cols],
            options = list(pageLength = 15, scrollX = TRUE),
            caption = caption_text,
            rownames = FALSE) %>%
    formatRound(columns = c('setpoint_repeat_length', 
                           'modal_diff', 'median_diff', 'mean_diff',
                           'instability_index', 'expansion_index',
                           'contraction_index', 'expansion_ratio', 'z_score'),
                digits = 2) %>%
    formatStyle(columns = display_cols, fontSize = '10px')
}
```

**Metrics explanation:**

**Relative distribution statistics** — measure the *shift* of the distribution from the setpoint:

- **modal_diff**: Difference between sample's modal_length and setpoint (modal - setpoint)
- **median_diff**: Difference between sample's median_length and setpoint (median - setpoint)
- **mean_diff**: Difference between sample's mean_length and setpoint (mean - setpoint)

**Instability metrics** — measure the *spread* of the distribution around the setpoint:

- **setpoint_type**: "sample_relative" (sample's own metric value) or "control_relative" (group control)
- **instability_index**: Mean absolute deviation from setpoint (weighted by read frequency)
- **expansion_index**: Mean deviation for reads above setpoint
- **contraction_index**: Mean deviation for reads below setpoint  
- **expansion_ratio**: (reads above) / (reads at setpoint)
- **z_score**: Standard deviations from control (control_relative only)

All metrics calculated from actual read distribution, not just summary statistics.

---

```{r plot_instability, echo=FALSE, fig.width=12, fig.height=7, message=FALSE}
# Cohort instability plots: x = analysis range, box+jitter per range, facet by group.
# All plots scale correctly regardless of sample count.

# Define setpoint type labels used across plots
setpoint_labels <- c(
  "sample_relative" = "Relative to Sample",
  "control_relative" = "Relative to Control"
)

# Helper: standard box+jitter+mean-diamond layer set for instability plots
# x aesthetic must be set in the parent ggplot() call
instability_box_layers <- function() {
  list(
    geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey50", width = 0.6),
    geom_jitter(colour = "grey40", width = 0.2, height = 0,
                alpha = 0.7, size = 2),
    stat_summary(fun = mean, geom = "point", shape = 18, size = 4,
                 colour = "red", alpha = 0.6)
  )
}

# ── Plot 12: Instability Index by range ──────────────────────────────────────────
p_instability <- ggplot(instability_metrics,
                        aes(x = analysis_range_name, y = instability_index)) +
  instability_box_layers() +
  facet_wrap(~ setpoint_type, ncol = 1,
             labeller = labeller(setpoint_type = setpoint_labels)) +
  labs(title = "Instability index by analysis range",
       subtitle = paste0("Mean absolute deviation from setpoint (using ",
                         params$control_setpoint_metric,
                         "). Box = IQR; points = samples; red diamond = mean."),
       x = "Analysis range", y = "Instability index",
) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_instability)
ggsave(file.path(plots_dir, "12_instability_index.png"),
       p_instability, width = 12, height = 8, dpi = params$plot_dpi)

# ── Plot 13: Expansion vs Contraction Indices by range ───────────────────────────
exp_cont_long <- instability_metrics %>%
  select(file_stem, analysis_range_name, group, setpoint_type,
         expansion_index, contraction_index) %>%
  pivot_longer(cols = c(expansion_index, contraction_index),
               names_to = "direction", values_to = "index") %>%
  mutate(direction = tools::toTitleCase(gsub("_index", "", direction)),
         direction = factor(direction, levels = c("Contraction", "Expansion")))

p_exp_cont <- ggplot(exp_cont_long,
                     aes(x = analysis_range_name, y = index, fill = direction)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.7,
               position = position_dodge(0.8)) +
  geom_jitter(aes(colour = direction),
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.6, size = 1.5) +
  facet_wrap(~ setpoint_type, ncol = 1,
             labeller = labeller(setpoint_type = setpoint_labels)) +
  scale_fill_manual(values = c("Expansion" = "coral", "Contraction" = "steelblue")) +
  scale_colour_manual(values = c("Expansion" = "coral", "Contraction" = "steelblue")) +
  labs(title = "Expansion vs contraction indices by range",
       subtitle = "Mean deviation above/below setpoint. Box = IQR; points = samples.",
       x = "Analysis range", y = "Index (repeats)",
       fill = "Direction", colour = "Direction") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

print(p_exp_cont)
ggsave(file.path(plots_dir, "13_expansion_contraction.png"),
       p_exp_cont, width = 12, height = 8, dpi = params$plot_dpi)

if (params$group_control && any(instability_metrics$setpoint_type == "control_relative")) {

  # ── Plot 14: Expansion Ratio by range (both setpoint types) ───────────────────
  p_ratio <- ggplot(instability_metrics,
                    aes(x = analysis_range_name, y = expansion_ratio)) +
    instability_box_layers() +
    facet_wrap(~ setpoint_type, ncol = 1,
               labeller = labeller(setpoint_type = setpoint_labels)) +
    labs(title = "Expansion ratio by range",
         subtitle = "(Reads above setpoint) / (Reads at setpoint). Box = IQR; points = samples; red diamond = mean.",
         x = "Analysis range", y = "Expansion ratio") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_ratio)
  ggsave(file.path(plots_dir, "14_expansion_ratio.png"),
         p_ratio, width = 12, height = 8, dpi = params$plot_dpi)

  # ── Plot 15: Contraction Ratio by range (both setpoint types) ─────────────────
  p_contraction_ratio <- ggplot(instability_metrics,
                                aes(x = analysis_range_name, y = contraction_ratio)) +
    instability_box_layers() +
    facet_wrap(~ setpoint_type, ncol = 1,
               labeller = labeller(setpoint_type = setpoint_labels)) +
    labs(title = "Contraction ratio by range",
         subtitle = "(Reads below setpoint) / (Reads at setpoint). Box = IQR; points = samples; red diamond = mean.",
         x = "Analysis range", y = "Contraction ratio") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  ggsave(file.path(plots_dir, "15_contraction_ratio.png"),
         p_contraction_ratio, width = 12, height = 8, dpi = params$plot_dpi)

  # ── Plot 17: Total Tail Ratio by range ─────────────────────────────────────────
  p_total_tail <- ggplot(instability_metrics,
                         aes(x = analysis_range_name, y = total_tail_ratio)) +
    instability_box_layers() +
    facet_wrap(~ setpoint_type, ncol = 1,
               labeller = labeller(setpoint_type = setpoint_labels)) +
    labs(title = "Total tail ratio by range",
         subtitle = "(Reads above + Reads below) / Reads at setpoint. Box = IQR; points = samples; red diamond = mean.",
         x = "Analysis range", y = "Total tail ratio") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  ggsave(file.path(plots_dir, "16_total_tail_ratio.png"),
         p_total_tail, width = 12, height = 8, dpi = params$plot_dpi)

  # ── Plot 18/19: Read proportions by region ─────────────────────────────────────
  # Absolute counts are exported in Excel (instability_metrics sheet).
  # Cohort plot shows proportion above/at/below per range, box+jitter.
  read_prop_data <- instability_metrics %>%
    mutate(
      total_reads = reads_below_setpoint + reads_at_setpoint + reads_above_setpoint,
      prop_below = reads_below_setpoint / total_reads,
      prop_at    = reads_at_setpoint    / total_reads,
      prop_above = reads_above_setpoint / total_reads
    ) %>%
    select(file_stem, analysis_range_name, group, setpoint_type,
           prop_below, prop_at, prop_above) %>%
    pivot_longer(cols = c(prop_below, prop_at, prop_above),
                 names_to = "region", values_to = "proportion") %>%
    mutate(region = case_when(
      region == "prop_below" ~ "Below setpoint",
      region == "prop_at"    ~ "At setpoint",
      region == "prop_above" ~ "Above setpoint"
    ),
    region = factor(region, levels = c("Below setpoint", "At setpoint", "Above setpoint")))

    p_read_prop <- ggplot(read_prop_data,
                        aes(x = analysis_range_name, y = proportion, colour = region)) +
    geom_boxplot(aes(fill = region), outlier.shape = NA, alpha = 0.3, width = 0.7,
                 position = position_dodge(0.8)) +
    geom_jitter(position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.8),
                size = 2.5, alpha = 0.9) +
    facet_wrap(~ setpoint_type, ncol = 1,
               labeller = labeller(setpoint_type = setpoint_labels)) +
    scale_fill_manual(
      values = c("Below setpoint" = "#08519C",
                 "At setpoint"    = "#525252",
                 "Above setpoint" = "#A50F15"),
      breaks = c("Below setpoint", "At setpoint", "Above setpoint")
    ) +
    scale_colour_manual(
      values = c("Below setpoint" = "#08519C",
                 "At setpoint"    = "#525252",
                 "Above setpoint" = "#A50F15"),
      breaks = c("Below setpoint", "At setpoint", "Above setpoint")
    ) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(title = "Read proportions by region and range",
         subtitle = "Proportion of reads below, at, and above setpoint. Absolute counts exported in Excel.",
         x = "Analysis range", y = "Proportion of reads",
         fill = "Region", colour = "Region") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")

  ggsave(file.path(plots_dir, "17_read_proportions_by_region.png"),
         p_read_prop, width = 12, height = 8, dpi = params$plot_dpi)

  # ── Plot 20: Expansion vs Contraction Balance (faceted by range) ───────────────
  # Saved to disk only — many range panels would be heavy inline.
  balance_data <- instability_metrics %>%
    filter(!is.na(contraction_ratio), !is.na(expansion_ratio))

  if (nrow(balance_data) > 0) {
    # Two separate plots — one per setpoint type, each facet_wrapped by range
    n_ranges <- length(unique(balance_data$analysis_range_name))
    plot_w <- max(14, n_ranges * 2.5)

    for (sp_type in unique(balance_data$setpoint_type)) {
      sp_data  <- balance_data %>% filter(setpoint_type == sp_type)
      sp_label <- setpoint_labels[sp_type]
      sp_slug  <- gsub("_", "-", sp_type)   # e.g. "control-relative"

      p_bal <- ggplot(sp_data,
                      aes(x = contraction_ratio, y = expansion_ratio)) +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
        geom_smooth(method = "lm", se = FALSE, linewidth = 0.7, colour = "#4393C3") +
        geom_point(colour = "grey30", size = 2.5, alpha = 0.75) +
        facet_wrap(~ analysis_range_name, scales = "free") +
        labs(title = paste0("Expansion vs contraction balance — ", sp_label),
             subtitle = "Each point = one sample. Dashed = balanced (expansion = contraction). Blue line = lm fit.",
             x = "Contraction ratio", y = "Expansion ratio") +
        theme_bw() +
        theme(strip.text = element_text(size = 8))

      fname <- paste0("18_expansion_contraction_balance-", sp_slug, ".png")
      n_cols  <- min(4, n_ranges)
      n_rows  <- ceiling(n_ranges / n_cols)
      ggsave(file.path(plots_dir, fname),
             p_bal,
             width  = n_cols * 4.5,
             height = n_rows * 4 + 2,
             dpi = params$plot_dpi, limitsize = FALSE)
      message("Balance plot saved: ", fname)
    }
  }

}

message("\nInstability plots saved to: ", plots_dir)
```

---

# Export to Excel

Export all results to Excel workbook.

```{r export_excel, echo=FALSE, message=FALSE}
# Prepare export list
# Note: comprehensive_range_summary contains all per-sample-range detail including
# peaks and controls; range_summary (3-column settings overview) is dropped as redundant.
excel_export <- list(
  range_summary = comprehensive_range_summary,
  modal_peaks = modal_peaks,
  distribution_summary = distribution_summary
)

if (!is.null(group_controls)) {
  excel_export$group_controls <- group_controls
}

if (nrow(instability_metrics) > 0) {
  excel_export$instability_metrics <- instability_metrics

  # Add read counts table (wide format: one row per sample/range/setpoint_type)
  read_counts_export <- instability_metrics %>%
    select(file_stem, analysis_range_name, group, setpoint_type, setpoint_repeat_length,
           reads_below_setpoint, reads_at_setpoint, reads_above_setpoint)
  excel_export$read_counts <- read_counts_export
}

# Export to Excel
excel_path <- file.path(range_dir, "range_analysis.xlsx")
openxlsx::write.xlsx(excel_export, file = excel_path)

message("\nResults exported to Excel: ", excel_path)
message("  Sheets: ", paste(names(excel_export), collapse = ", "))
```

**Excel file contains:**

- **range_summary**: Complete table — all analysis ranges per sample with detected peaks, control setpoints, and boundary details
- **modal_peaks**: Detected modal peaks per sample and range (long format)
- **distribution_summary**: ~30 distribution statistics per sample per range (includes `n_reads_total` per range)
- **group_controls**: Group control setpoints per range (if `group_control = TRUE`)
- **instability_metrics**: All instability metrics (instability index, expansion/contraction indices, ratios, z-scores) per sample, range, and setpoint type
- **read_counts**: Absolute read counts below, at, and above setpoint per sample/range/setpoint type

---

# Summary

**Range analysis complete:**

- **`r length(unique(modal_peaks$file_stem))`** samples processed
- **`r nrow(modal_peaks)`** total peaks detected
- **`r nrow(distribution_summary)`** distribution summaries calculated

---

# Save outputs

```{r save_outputs, echo=FALSE}
# Compile module output
module6_output <- list(
  settings = settings_parsed,
  comprehensive_range_summary = comprehensive_range_summary,
  modal_peaks = modal_peaks,
  distribution_summary = distribution_summary,
  group_controls = group_controls,
  instability_metrics = instability_metrics
)

# Save complete module output
results_dir <- file.path(params$dir_out, "module_data")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

output_path <- file.path(results_dir, "06_range_analysis.RData")
save(module6_output, file = output_path)

message("Module 6 outputs saved to: ", output_path)
```

**Module 6 Complete!**

Results saved to: `r output_path`

---

# Module diagnostics

## Timing

```{r timing_table, echo=FALSE}
timing_df <- build_timing_table(timings)
DT::datatable(
  timing_df,
  caption = "Section timing — Module 6",
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Section", "Seconds", "Elapsed (mm:ss)", "Cumulative (mm:ss)")
) %>%
  DT::formatStyle("section", target = "row",
                  fontWeight = DT::styleEqual("Total", "bold"))
```

## Output objects

```{r manifest_table, echo=FALSE}
manifest <- data.frame(
  object = names(module6_output),
  description = c(
    "Parsed settings: range definitions, floors, max peaks per sample",
    "Per-sample/range config, peaks, controls, and summary statistics in one joined table",
    "Detected modal peaks per sample and analysis range",
    "~30 distribution statistics per sample/range (mean, median, SD, skewness, tails, CV...)",
    "Aggregated control setpoints per group and range (NULL if group_control = FALSE)",
    "Instability indices, expansion/contraction metrics, and z-scores vs control setpoints"
  ),
  class = sapply(module6_output, function(x) if (is.null(x)) "NULL" else class(x)[1]),
  size = sapply(module6_output, function(x) {
    format_size(if (is.null(x)) 0 else as.numeric(object.size(x)))
  }),
  bytes = sapply(module6_output, function(x) {
    if (is.null(x)) 0L else as.numeric(object.size(x))
  }),
  stringsAsFactors = FALSE
)

DT::datatable(
  manifest[, c("object", "description", "class", "size")],
  caption = sprintf("Objects saved to module6_output — RData file: %s on disk (%s)",
                    format_size(file.size(output_path)), output_path),
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Object", "Description", "Class", "In-memory size")
)
```

```{r diagnostics_export, echo=FALSE}
write_excel_function(
  named_list = list("timing" = timing_df, "manifest" = manifest),
  path = excel_path,
  overwrite_sheet = TRUE
)
message(sprintf("Module 6 complete — total time: %s  |  RData: %s",
                format_elapsed(sum(timings)), format_size(file.size(output_path))))
```

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```


