---
title: "Duke Pipeline - Module 6: Range Analysis"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

```{r load_packages, echo=FALSE, message=FALSE}
# Required packages
required_packages <- c("dplyr", "tidyr", "ggplot2", "ggridges", "stringr", "readxl",
                      "knitr", "DT", "pbapply", "moments",
                      "openxlsx", "pracma", "scales", "pbmcapply")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r create_directories, echo=FALSE}
# Create output directory
range_dir <- file.path(params$dir_out, "06_range_analysis", "range_analysis")

if (!dir.exists(range_dir)) dir.create(range_dir, recursive = TRUE)
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}
```

---

# Module 6: Range Analysis & Instability Metrics

Analyze repeat length distributions within user-defined ranges and calculate instability metrics relative to control samples.

**Key features:**

- Settings-based range definitions
- Peak detection per range
- Distribution summaries (~30 metrics per range)
- Group control definition
- Instability metrics
- Excel exports

---

# Load data

```{r load_data, echo=FALSE}
# Load Module 4 output (clustered data with repeat counts)
results_dir <- file.path(params$dir_out, "module_data")
load(file.path(results_dir, "04_allele_calling_results.RData"))

# Extract clustered data
alignment_clustered <- module4_output$alignment_clustered

# Check data
message("Loaded clustered alignment data:")
message("  Samples: ", length(unique(alignment_clustered$file_stem)))
message("  Total reads: ", nrow(alignment_clustered))
message("  Clusters: ", length(unique(alignment_clustered$cluster_number)))
```

---

# Load settings

Settings Excel defines analysis ranges, group assignments, and control samples.

**Expected columns:**

- `file_name`: Sample filename
- `analysis_ranges`: Range definitions (e.g., `[0-35][36-1000]`)
- `floor`: Minimum frequency threshold per range (e.g., `[3][3]`)
- `max_peaks`: Maximum peaks to detect per range (e.g., `[2][1]`)
- `manual_control_repeat_length`: Optional manual control values (e.g., `[14][100]`)
- `group`: Group assignment for comparisons
- `group_control_sample`: Flag control samples (1 = control, NA = experimental)
- `time`: Timepoint or order
- `exclude`: Flag samples to exclude from analysis

```{r load_settings, echo=FALSE}
# Check if settings file exists
if (is.na(params$path_settings) || !file.exists(params$path_settings)) {
  stop("Settings file not found. Please provide settings_path parameter.")
}

# Load settings (supports CSV and Excel)
file_ext <- tolower(tools::file_ext(params$path_settings))

if (file_ext == "csv") {
  # Read CSV
  settings <- read.csv(params$path_settings, header = TRUE, stringsAsFactors = FALSE)
} else if (file_ext %in% c("xlsx", "xls")) {
  # Read Excel
  settings <- read_excel(params$path_settings, sheet = 1)
} else {
  stop("Unsupported settings file format: ", file_ext, 
       ". Please use .csv, .xlsx, or .xls")
}

# Convert file_name to file_stem (remove extensions)
settings <- settings %>%
  mutate(file_stem = str_remove(file_name, "\\.(fastq|fasta|fa|bam)(\\.gz)?$"))

message("Settings loaded from: ", params$path_settings)
message("  Format: ", toupper(file_ext))
message("  Samples: ", nrow(settings))

# Display settings table
datatable(settings,
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Settings file",
          rownames = FALSE) %>%
  formatStyle(columns = colnames(settings), fontSize = '10px')
```

---

# Parse analysis ranges

Parse range definitions from settings for each sample.

```{r parse_ranges, echo=FALSE}
# Parse ranges for all samples
settings_parsed <- settings %>%
  filter(!is.na(analysis_ranges), !is.na(floor), !is.na(max_peaks)) %>%
  rowwise() %>%
  mutate(
    # Parse ranges, floor, and max_peaks
    ranges_list = list(parse_analysis_ranges(analysis_ranges)),
    floor_list = list(parse_bracketed_param(floor)),
    max_peaks_list = list(parse_bracketed_param(max_peaks)),
    n_ranges = length(ranges_list)
  ) %>%
  ungroup()

# Show summary
range_summary <- settings_parsed %>%
  select(file_stem, analysis_ranges, n_ranges)

DT::datatable(
  range_summary,
  caption = "Analysis ranges per sample",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(range_summary), fontSize = '85%')

# Example for first sample
message("\nExample range parsing for: ", settings_parsed$file_stem[1])
for (rng in names(settings_parsed$ranges_list[[1]])) {
  r <- settings_parsed$ranges_list[[1]][[rng]]
  f <- settings_parsed$floor_list[[1]][[rng]]
  m <- settings_parsed$max_peaks_list[[1]][[rng]]
  message("  ", rng, ": [", r[1], "-", r[2], "] floor=", f, " max_peaks=", m)
}
```

---

# Comprehensive Range Summary

Build a comprehensive summary table showing all range parameters for each sample.

```{r comprehensive_range_summary, echo=FALSE}
message("\n=== Building Comprehensive Range Summary ===")

# Build comprehensive range summary for each sample and range
comprehensive_range_summary_list <- lapply(1:nrow(settings_parsed), function(i) {
  
  my_settings <- settings_parsed[i, ]
  my_file_stem <- my_settings$file_stem
  
  # Get ranges and parameters
  ranges <- my_settings$ranges_list[[1]]
  floor_vals <- my_settings$floor_list[[1]]
  max_peaks_vals <- my_settings$max_peaks_list[[1]]
  
  # Get manual controls if they exist
  manual_controls <- list()
  if (!is.null(my_settings$manual_control_list) && 
      length(my_settings$manual_control_list) > 0) {
    manual_controls <- my_settings$manual_control_list[[1]]
  }
  
  # Get sample data to calculate actual boundaries
  sample_data <- alignment_clustered %>%
    filter(file_stem == my_file_stem)
  
  if (nrow(sample_data) == 0) {
    warning("No data for sample: ", my_file_stem)
    return(NULL)
  }
  
  repeat_counts <- sample_data[[params$repeat_count_method]]
  
  # Build summary for each range
  range_summaries <- lapply(names(ranges), function(range_name) {
    
    my_range <- ranges[[range_name]]
    my_floor <- floor_vals[[range_name]]
    my_max_peaks <- max_peaks_vals[[range_name]]
    
    # Manual control if provided
    manual_control <- if (length(manual_controls) > 0 && 
                          range_name %in% names(manual_controls)) {
      manual_controls[[range_name]]
    } else {
      NA_real_
    }
    
    # Store original boundaries from settings
    settings_min <- my_range[1]
    settings_max <- my_range[2]
    
    # Replace NA with actual data boundaries for analysis
    if (is.na(my_range[1])) my_range[1] <- min(repeat_counts, na.rm = TRUE)
    if (is.na(my_range[2])) my_range[2] <- max(repeat_counts, na.rm = TRUE)
    
    # Filter to range to get observed boundaries
    counts_in_range <- repeat_counts[repeat_counts >= my_range[1] & 
                                     repeat_counts <= my_range[2]]
    
    # Observed boundaries (what's actually in the data)
    observed_min <- if (length(counts_in_range) > 0) {
      min(counts_in_range, na.rm = TRUE)
    } else {
      NA_real_
    }
    
    observed_max <- if (length(counts_in_range) > 0) {
      max(counts_in_range, na.rm = TRUE)
    } else {
      NA_real_
    }
    
    data.frame(
      file_stem = my_file_stem,
      analysis_range_name = range_name,
      settings_min = settings_min,
      settings_max = settings_max,
      analysis_min = my_range[1],
      analysis_max = my_range[2],
      observed_min = observed_min,
      observed_max = observed_max,
      floor_threshold = my_floor,
      max_peaks_allowed = my_max_peaks,
      manual_control_repeat_length = manual_control,
      stringsAsFactors = FALSE
    )
  })
  
  bind_rows(range_summaries)
})

# Combine all summaries
comprehensive_range_summary <- bind_rows(comprehensive_range_summary_list)

# Add group and time information
comprehensive_range_summary <- comprehensive_range_summary %>%
  left_join(
    select(settings, file_stem, group, group_control_sample, time),
    by = "file_stem"
  ) %>%
  relocate(file_stem, group, time, analysis_range_name)

message("  Rows created: ", nrow(comprehensive_range_summary))
message("  Samples: ", length(unique(comprehensive_range_summary$file_stem)))
message("  Ranges: ", length(unique(comprehensive_range_summary$analysis_range_name)))

# Display initial table (before peaks are added)
DT::datatable(
  comprehensive_range_summary,
  caption = "Comprehensive Range Summary - Configuration and Boundaries",
  extensions = 'Buttons',
  options = list(
    pageLength = 20,
    scrollX = TRUE,
    scrollY = "600px",
    dom = 'Bfrtip',
    buttons = list(
      'copy',
      list(extend = 'csv', filename = 'range_summary_configuration'),
      list(extend = 'excel', filename = 'range_summary_configuration'),
      'colvis'
    ),
    columnDefs = list(
      list(width = '120px', targets = 0)
    )
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact hover',
  style = 'bootstrap'
) %>%
  DT::formatRound(
    columns = c('settings_min', 'settings_max', 'analysis_min', 'analysis_max',
                'observed_min', 'observed_max', 'manual_control_repeat_length'),
    digits = 1
  ) %>%
  DT::formatStyle(
    'manual_control_repeat_length',
    backgroundColor = DT::styleEqual(NA, 'white', default = '#fff3cd'),
    fontWeight = 'bold'
  ) %>%
  DT::formatStyle(
    columns = colnames(comprehensive_range_summary),
    fontSize = '10px'
  )
```

**Column descriptions:**

- **settings_min/max**: Original range boundaries from settings file (may be NA for open ranges)
- **analysis_min/max**: Actual boundaries used for analysis (NA replaced with data min/max)
- **observed_min/max**: Minimum and maximum repeat lengths actually observed in the data within this range
- **floor_threshold**: Minimum read frequency threshold applied
- **max_peaks_allowed**: Maximum number of peaks allowed to detect
- **manual_control_repeat_length**: Manually specified control value (highlighted in yellow if present)

---

# Peak detection

Detect modal peaks within each analysis range for each sample.

```{r peak_detection, echo=FALSE}
# Get unique file stems that have settings
file_stems_with_settings <- settings_parsed$file_stem

# Peak detection per sample and range
modal_peaks_list <- apply_fn(file_stems_with_settings, function(my_file_stem) {
  
  # Get sample data
  sample_data <- alignment_clustered %>%
    filter(file_stem == my_file_stem)
  
  if (nrow(sample_data) == 0) {
    warning("No data for sample: ", my_file_stem)
    return(NULL)
  }
  
  # Get settings for this sample
  my_settings <- settings_parsed %>%
    filter(file_stem == my_file_stem) %>%
    slice(1)
  
  if (nrow(my_settings) == 0) {
    warning("No settings for sample: ", my_file_stem)
    return(NULL)
  }
  
  # Get repeat counts
  repeat_counts <- sample_data[[params$repeat_count_method]]
  
  # Detect peaks for each range
  ranges <- my_settings$ranges_list[[1]]
  floor_vals <- my_settings$floor_list[[1]]
  max_peaks_vals <- my_settings$max_peaks_list[[1]]
  
  range_results <- lapply(names(ranges), function(range_name) {
    
    my_range <- ranges[[range_name]]
    my_floor <- floor_vals[[range_name]]
    my_max_peaks <- max_peaks_vals[[range_name]]
    
    # Store original window (before NA replacement)
    original_window_min <- my_range[1]
    original_window_max <- my_range[2]
    
    # Replace NA in range with data min/max
    if (is.na(my_range[1])) {
      my_range[1] <- min(repeat_counts, na.rm = TRUE)
    }
    if (is.na(my_range[2])) {
      my_range[2] <- max(repeat_counts, na.rm = TRUE)
    }
    
    # Find peaks
    peaks <- find_modal_peaks(
      repeat_counts = repeat_counts,
      range_min = my_range[1],
      range_max = my_range[2],
      floor = my_floor,
      max_peaks = my_max_peaks,
      span = params$range_peak_span
    )
    
    # Add metadata
    peaks %>%
      mutate(
        file_stem = my_file_stem,
        analysis_range_name = range_name,
        window_min = original_window_min,  # Settings boundary (may be NA)
        window_max = original_window_max,  # Settings boundary (may be NA)
        range_min = my_range[1],           # Actual boundary used
        range_max = my_range[2]            # Actual boundary used
      )
  })
  
  bind_rows(range_results)
})

# Combine all results
modal_peaks_list <- extract_apply_fn_result(modal_peaks_list)

modal_peaks <- bind_rows(modal_peaks_list) %>%
  relocate(file_stem, analysis_range_name)

message("\nPeak detection complete:")
message("  Samples: ", length(unique(modal_peaks$file_stem)))
message("  Total peaks detected: ", sum(!is.na(modal_peaks$modal_length)))

# Display peak table
datatable(modal_peaks,
          options = list(pageLength = 15, scrollX = TRUE),
          caption = "Modal peaks detected per sample and range",
          rownames = FALSE) %>%
  formatRound(columns = c('modal_length'), digits = 1) %>%
  formatStyle(columns = colnames(modal_peaks), fontSize = '10px')
```

```{r add_peaks_to_comprehensive_summary, echo=FALSE}
# Add detected peaks to comprehensive summary
message("\n=== Adding Peaks to Comprehensive Summary ===")

# Get peaks in wide format for joining
message("  Converting peaks to wide format...")

# First, find the maximum number of peaks detected across all samples/ranges
max_peaks_found <- modal_peaks %>%
  filter(!is.na(modal_length)) %>%
  group_by(file_stem, analysis_range_name) %>%
  summarise(n = n(), .groups = "drop") %>%
  pull(n) %>%
  max()

message("  Maximum peaks detected in any sample/range: ", max_peaks_found)

# Dynamically create peak columns based on what exists
peaks_wide <- modal_peaks %>%
  filter(!is.na(modal_length)) %>%
  arrange(file_stem, analysis_range_name, peak_rank) %>%
  group_by(file_stem, analysis_range_name) %>%
  summarise(
    n_peaks_detected = n(),
    .groups = "drop"
  )

# Add peak columns dynamically
for (i in 1:max_peaks_found) {
  # Extract peaks for this rank
  peak_data <- modal_peaks %>%
    filter(!is.na(modal_length), peak_rank == i) %>%
    select(file_stem, analysis_range_name, modal_length, modal_frequency) %>%
    rename(!!paste0("peak", i, "_length") := modal_length,
           !!paste0("peak", i, "_frequency") := modal_frequency)
  
  # Join to peaks_wide
  peaks_wide <- peaks_wide %>%
    left_join(peak_data, by = c("file_stem", "analysis_range_name"))
}

message("  Created ", max_peaks_found, " peak column pairs (length + frequency)")

# Join peaks to comprehensive summary
comprehensive_range_summary <- comprehensive_range_summary %>%
  left_join(peaks_wide, by = c("file_stem", "analysis_range_name"))

message("  Peaks added successfully")
if ("peak1_length" %in% colnames(comprehensive_range_summary)) {
  message("  Total peak1 detected: ", sum(!is.na(comprehensive_range_summary$peak1_length)))
}

# Dynamically identify peak columns for formatting
peak_cols <- grep("^peak[0-9]+_(length|frequency)$", 
                  colnames(comprehensive_range_summary), 
                  value = TRUE)
peak_length_cols <- grep("^peak[0-9]+_length$", peak_cols, value = TRUE)
peak_freq_cols <- grep("^peak[0-9]+_frequency$", peak_cols, value = TRUE)

message("  Peak columns in table: ", paste(peak_cols, collapse = ", "))

# Display updated table with peaks
dt_table <- DT::datatable(
  comprehensive_range_summary,
  caption = "Comprehensive Range Summary - With Detected Peaks",
  extensions = c('Buttons', 'FixedColumns'),
  options = list(
    pageLength = 20,
    scrollX = TRUE,
    scrollY = "650px",
    dom = 'Bfrtip',
    buttons = list(
      'copy',
      list(extend = 'csv', filename = 'range_summary_with_peaks'),
      list(extend = 'excel', filename = 'range_summary_with_peaks'),
      'colvis'
    ),
    fixedColumns = list(leftColumns = 2),
    columnDefs = list(
      list(width = '110px', targets = 0:1)
    )
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact hover',
  style = 'bootstrap'
)

# Format numeric columns
numeric_cols <- c('settings_min', 'settings_max', 'analysis_min', 'analysis_max',
                  'observed_min', 'observed_max', 'manual_control_repeat_length')
numeric_cols <- c(numeric_cols, peak_length_cols)
numeric_cols <- numeric_cols[numeric_cols %in% colnames(comprehensive_range_summary)]

dt_table <- dt_table %>%
  DT::formatRound(columns = numeric_cols, digits = 1)

# Highlight peak1 (most important) if it exists
if ("peak1_length" %in% colnames(comprehensive_range_summary)) {
  dt_table <- dt_table %>%
    DT::formatStyle(
      c('peak1_length', 'peak1_frequency'),
      fontWeight = 'bold',
      backgroundColor = '#e3f2fd'
    )
}

# Highlight peak2 if it exists
if ("peak2_length" %in% colnames(comprehensive_range_summary)) {
  dt_table <- dt_table %>%
    DT::formatStyle(
      c('peak2_length', 'peak2_frequency'),
      backgroundColor = '#f0f0f0'
    )
}

# Highlight manual controls
dt_table <- dt_table %>%
  DT::formatStyle(
    'manual_control_repeat_length',
    backgroundColor = DT::styleEqual(NA, 'white', default = '#fff3cd')
  ) %>%
  DT::formatStyle(
    columns = colnames(comprehensive_range_summary),
    fontSize = '10px'
  )

dt_table
```

```{r plot_peaks, echo=FALSE, fig.width=12, fig.height=10}
# Create plots directory
plots_dir <- file.path(params$dir_out, file.path("06_range_analysis", "plots"))
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)

# Plot: Modal peaks with range windows shown as rectangles
# Prepare range boundaries for each sample
range_boxes <- modal_peaks %>%
  select(file_stem, analysis_range_name, range_min, range_max) %>%
  distinct() %>%
  # Create x-positions for rectangles
  mutate(sample_num = as.numeric(factor(file_stem, levels = unique(file_stem))))

# === RANGE BOUNDARY CACHING FOR EFFICIENT LOOKUPS ===
# Cache range boundaries to avoid repeated filtering in plotting loops
# Handles per-sample specific ranges correctly
message("\n=== CACHING RANGE BOUNDARIES ===")
range_boundaries_cache <- modal_peaks %>%
  select(file_stem, analysis_range_name, range_min, range_max) %>%
  distinct()

# Create fast lookup function
get_range_boundaries <- function(sample, range_name) {
  range_boundaries_cache %>%
    filter(file_stem == sample, analysis_range_name == range_name) %>%
    slice(1)
}

message(sprintf("Cached %d unique sample-range combinations", nrow(range_boundaries_cache)))
message("Example: Sample-specific ranges are preserved")
if (nrow(range_boundaries_cache) > 2) {
  message(sprintf("  %s / %s: [%d-%d]", 
                  range_boundaries_cache$file_stem[1],
                  range_boundaries_cache$analysis_range_name[1],
                  range_boundaries_cache$range_min[1],
                  range_boundaries_cache$range_max[1]))
  message(sprintf("  %s / %s: [%d-%d]", 
                  range_boundaries_cache$file_stem[2],
                  range_boundaries_cache$analysis_range_name[2],
                  range_boundaries_cache$range_min[2],
                  range_boundaries_cache$range_max[2]))
}
message("================================\n")


p_peaks <- ggplot() +
  # Add shaded rectangles showing the range windows
  geom_rect(data = range_boxes,
            aes(xmin = sample_num - 0.4, xmax = sample_num + 0.4,
                ymin = range_min, ymax = range_max,
                fill = analysis_range_name),
            alpha = 0.15) +
  # Add the actual peak points on top
  geom_point(data = modal_peaks %>% filter(!is.na(modal_length)),
             aes(x = as.numeric(factor(file_stem, levels = unique(file_stem))),
                 y = modal_length,
                 colour = analysis_range_name,
                 shape = factor(peak_rank)),
             size = 4, alpha = 0.9) +
  scale_x_continuous(breaks = 1:length(unique(modal_peaks$file_stem)),
                    labels = unique(modal_peaks$file_stem)) +
  labs(title = "Modal Peaks Detected",
       subtitle = "Shaded boxes show analysis range windows; points show detected peaks",
       x = "Sample",
       y = "Repeat Length",
       colour = "Analysis Range",
       fill = "Range Window",
       shape = "Peak Rank") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

print(p_peaks)
ggsave(file.path(plots_dir, "01_modal_peaks.png"), 
       p_peaks, width = 12, height = 7, dpi = 300)

# Generate individual peak plots by sample
message("\nGenerating individual peak plots by sample...")

modal_peaks_by_sample_dir <- file.path(plots_dir, "01_modal_peaks_by_sample")
if (!dir.exists(modal_peaks_by_sample_dir)) {
  dir.create(modal_peaks_by_sample_dir, recursive = TRUE)
}

for (my_sample in unique(modal_peaks$file_stem)) {
  sample_peaks <- modal_peaks %>%
    filter(file_stem == my_sample)
  
  if (nrow(sample_peaks) == 0) next
  
  # Get range boxes for this sample
  sample_boxes <- sample_peaks %>%
    select(analysis_range_name, window_min, window_max, range_min, range_max) %>%
    distinct() %>%
    mutate(y_position = seq_along(analysis_range_name))
  
  # Create peak plot for this sample
  p_sample_peaks <- ggplot() +
    # Range windows as shaded rectangles
    geom_rect(data = sample_boxes,
              aes(xmin = range_min, xmax = range_max,
                  ymin = y_position - 0.4, ymax = y_position + 0.4,
                  fill = analysis_range_name),
              alpha = 0.2) +
    # Detected peaks as points
    geom_point(data = sample_peaks %>% filter(!is.na(modal_length)),
               aes(x = modal_length, 
                   y = match(analysis_range_name, sample_boxes$analysis_range_name),
                   size = modal_frequency,
                   shape = factor(peak_rank)),
               colour = "black", alpha = 0.8) +
    # Labels for peaks
    geom_text(data = sample_peaks %>% filter(!is.na(modal_length)),
              aes(x = modal_length,
                  y = match(analysis_range_name, sample_boxes$analysis_range_name),
                  label = paste0(modal_length, " (", modal_frequency, ")")),
              vjust = -1.2, hjust = -0.1, angle = 45, size = 3) +
    scale_y_continuous(breaks = sample_boxes$y_position,
                      labels = sample_boxes$analysis_range_name) +
    scale_shape_manual(values = c("1" = 16, "2" = 17, "3" = 15),
                      name = "Peak Rank") +
    scale_size_continuous(range = c(3, 8), name = "Frequency") +
    labs(title = paste0("Modal Peaks - ", my_sample),
         subtitle = "Shaded regions = analysis ranges; Points = detected peaks; Numbers = (length, frequency)",
         x = "Repeat Length",
         y = "Analysis Range",
         fill = "Analysis Range") +
    theme_bw() +
    theme(legend.position = "right",
          panel.grid.major.y = element_blank())
  
  ggsave(file.path(modal_peaks_by_sample_dir, 
                  paste0(my_sample, "-modal_peaks.png")),
         p_sample_peaks, width = 12, height = max(4, nrow(sample_boxes) * 1.5), 
         dpi = 300)
}

message("  Individual peak plots saved to: ", modal_peaks_by_sample_dir)
```

---

# Distribution summaries

Calculate comprehensive distribution statistics for each range and sample.

**Range boundary columns:**

- **`window_min/max`**: Original boundaries from settings (e.g., `[0-35]` → 0, 35; may be NA for open-ended)
- **`range_min/max`**: Actual boundaries used (NA values replaced with data min/max)
- **`observed_min/max`**: Actual minimum and maximum repeat lengths observed in the data within this window

**Floor filtering:**

Low-frequency repeat lengths can be filtered before calculating statistics to reduce noise:

- **`floor_threshold`**: The minimum frequency threshold from settings (can be absolute count or proportion)
  - If 0-1: Proportion of modal frequency (e.g., 0.05 = 5% of peak height)
  - If ≥1: Absolute count (e.g., 3 = must appear at least 3 times)
- **`n_reads_total`**: Total reads in the range window (before filtering)
- **`n_reads_used`**: Reads remaining after floor filtering (used for all statistics)

**Example:** If `floor_threshold = 3`, any repeat length appearing <3 times is excluded. If you have 5000 reads total but 200 are low-frequency singletons/doubletons, then `n_reads_used = 4800`.

```{r distribution_summaries, echo=FALSE}
# Calculate distribution summaries
distribution_summary_list <- apply_fn(file_stems_with_settings, function(my_file_stem) {
  
  # Get sample data
  sample_data <- alignment_clustered %>%
    filter(file_stem == my_file_stem)
  
  if (nrow(sample_data) == 0) return(NULL)
  
  # Get settings
  my_settings <- settings_parsed %>%
    filter(file_stem == my_file_stem) %>%
    slice(1)
  
  if (nrow(my_settings) == 0) return(NULL)
  
  # Get repeat counts
  repeat_counts <- sample_data[[params$repeat_count_method]]
  
  # Get ranges
  ranges <- my_settings$ranges_list[[1]]
  floor_vals <- my_settings$floor_list[[1]]
  
  # Summary for each range
  range_summaries <- lapply(names(ranges), function(range_name) {
    
    my_range <- ranges[[range_name]]
    my_floor <- floor_vals[[range_name]]
    
    # Store original window (before NA replacement)
    original_window_min <- my_range[1]
    original_window_max <- my_range[2]
    
    # Replace NA with data min/max
    if (is.na(my_range[1])) my_range[1] <- min(repeat_counts, na.rm = TRUE)
    if (is.na(my_range[2])) my_range[2] <- max(repeat_counts, na.rm = TRUE)
    
    # Filter to range
    counts_in_range <- repeat_counts[repeat_counts >= my_range[1] & 
                                     repeat_counts <= my_range[2]]
    
    # Calculate observed min/max within this window
    observed_min <- if (length(counts_in_range) > 0) min(counts_in_range, na.rm = TRUE) else NA_real_
    observed_max <- if (length(counts_in_range) > 0) max(counts_in_range, na.rm = TRUE) else NA_real_
    
    # Calculate summary
    summary_stats <- distribution_summary_function(
      repeat_counts = counts_in_range,
      floor = my_floor
    )
    
    # Convert to data frame
    as.data.frame(summary_stats) %>%
      mutate(
        file_stem = my_file_stem,
        analysis_range_name = range_name,
        window_min = original_window_min,  # Settings boundary (may be NA)
        window_max = original_window_max,  # Settings boundary (may be NA)
        range_min = my_range[1],           # Actual boundary used (NA replaced)
        range_max = my_range[2],           # Actual boundary used (NA replaced)
        observed_min = observed_min,       # Observed minimum in data
        observed_max = observed_max        # Observed maximum in data
      )
  })
  
  bind_rows(range_summaries)
})

# Combine all summaries
distribution_summary_list <- extract_apply_fn_result(distribution_summary_list)

distribution_summary <- bind_rows(distribution_summary_list) %>%
  relocate(file_stem, analysis_range_name)

message("\nDistribution summaries calculated:")
message("  Total summaries: ", nrow(distribution_summary))

# Join with settings for group info
distribution_summary <- distribution_summary %>%
  left_join(
    select(settings, file_stem, group, group_control_sample, time, exclude),
    by = "file_stem"
  )

# Display summary table (key columns only)
key_cols <- c("file_stem", "analysis_range_name", 
              "window_min", "window_max", "range_min", "range_max",
              "observed_min", "observed_max",
              "modal_length", "modal_frequency", "mean_length", "median_length", 
              "sd", "floor_threshold", "n_reads_total", "n_reads_used")

datatable(distribution_summary[, key_cols],
          options = list(pageLength = 15, scrollX = TRUE),
          caption = "Distribution summary statistics (key metrics)",
          rownames = FALSE) %>%
  formatRound(columns = c('window_min', 'window_max', 'range_min', 'range_max',
                         'observed_min', 'observed_max',
                         'modal_length', 'mean_length', 'median_length', 'sd'), 
              digits = 2) %>%
  formatStyle(columns = key_cols, fontSize = '10px')
```

```{r plot_distributions, echo=FALSE, fig.width=10, fig.height=8}
# Plot 1: Modal vs mean lengths per sample
p_modal_mean <- ggplot(distribution_summary,
                       aes(x = modal_length, y = mean_length, colour = analysis_range_name)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Modal vs Mean Repeat Lengths",
       subtitle = "Points near diagonal indicate symmetric distributions",
       x = "Modal Length (repeats)",
       y = "Mean Length (repeats)",
       colour = "Analysis Range") +
  theme_bw() +
  theme(legend.position = "right")

print(p_modal_mean)
ggsave(file.path(plots_dir, "02_modal_vs_mean.png"),
       p_modal_mean, width = 10, height = 6, dpi = 300)

# Plot 2: Distribution spread (SD) per sample with range windows
p_spread <- ggplot(distribution_summary,
                   aes(x = file_stem, y = sd, fill = analysis_range_name)) +
  geom_col(position = "dodge", alpha = 0.8) +
  labs(title = "Distribution Spread by Sample",
       subtitle = "Standard deviation of repeat lengths",
       x = "Sample",
       y = "Standard Deviation",
       fill = "Analysis Range") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

print(p_spread)
ggsave(file.path(plots_dir, "03_spread_by_sample.png"),
       p_spread, width = 10, height = 6, dpi = 300)

# Plot 3: Filtering effect
filtering_data <- distribution_summary %>%
  mutate(pct_filtered = ((n_reads_total - n_reads_used) / n_reads_total) * 100)

p_filtering <- ggplot(filtering_data,
                      aes(x = file_stem, y = pct_filtered, fill = analysis_range_name)) +
  geom_col(position = "dodge", alpha = 0.8) +
  labs(title = "Floor Filtering Effect",
       subtitle = "Percentage of reads filtered out (floor threshold shown in table)",
       x = "Sample",
       y = "% Reads Filtered",
       fill = "Analysis Range") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

print(p_filtering)
ggsave(file.path(plots_dir, "04_filtering_effect.png"),
       p_filtering, width = 10, height = 6, dpi = 300)

# NEW DISTRIBUTION PLOTS - INSERT AFTER LINE 478 (after filtering plot)

# Plot 4: Tail Balance (Expansion vs Contraction Proportions)
# Calculate balanced proportion (reads at modal length)
tail_balance_data <- distribution_summary %>%
  mutate(balanced_proportion = 1 - contraction_proportion - expansion_proportion) %>%
  select(file_stem, analysis_range_name, group, 
         contraction_proportion, balanced_proportion, expansion_proportion) %>%
  pivot_longer(cols = c(contraction_proportion, balanced_proportion, expansion_proportion),
               names_to = "direction",
               values_to = "proportion") %>%
  mutate(direction = case_when(
    direction == "contraction_proportion" ~ "Contraction (below modal)",
    direction == "balanced_proportion" ~ "At modal",
    direction == "expansion_proportion" ~ "Expansion (above modal)"
  ),
  # Set factor levels in REVERSE order: expansion first stacks on TOP
  direction = factor(direction, levels = c("Expansion (above modal)", "At modal", "Contraction (below modal)")))

p_tail_balance <- ggplot(tail_balance_data,
                         aes(x = file_stem, y = proportion, fill = direction)) +
  geom_col(position = "stack", alpha = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "black") +
  facet_grid(analysis_range_name ~ group, scales = "free_x") +
  labs(title = "Tail Balance: Expansion vs Contraction Proportions",
       subtitle = "Proportion of reads above, at, and below modal length (dashed line = balanced)",
       x = "Sample",
       y = "Proportion of Reads",
       fill = "Direction") +
  scale_fill_manual(
    values = c("Contraction (below modal)" = "steelblue", 
               "At modal" = "grey60",
               "Expansion (above modal)" = "coral"),
    # Legend reads bottom to top, but stacking is reversed in code
    breaks = c("Contraction (below modal)", "At modal", "Expansion (above modal)")
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

print(p_tail_balance)
ggsave(file.path(plots_dir, "05_tail_balance.png"),
       p_tail_balance, width = 14, height = 8, dpi = 300)

# Plot 5: Coefficient of Variation
p_cv <- ggplot(distribution_summary,
               aes(x = file_stem, y = coefficient_of_variation, fill = analysis_range_name)) +
  geom_col(position = "dodge", alpha = 0.8) +
  labs(title = "Coefficient of Variation",
       subtitle = "Relative variability (SD / mean) - useful for comparing samples with different mean lengths",
       x = "Sample",
       y = "Coefficient of Variation (CV)",
       fill = "Analysis Range") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

print(p_cv)
ggsave(file.path(plots_dir, "06_coefficient_of_variation.png"),
       p_cv, width = 10, height = 6, dpi = 300)

# Plot 6: Percentile Ranges (5th-95th)
percentile_data <- distribution_summary %>%
  select(file_stem, analysis_range_name, group, 
         percentile_5th, percentile_95th, modal_length)

p_percentiles <- ggplot(percentile_data,
                        aes(x = file_stem, colour = analysis_range_name)) +
  geom_segment(aes(xend = file_stem, y = percentile_5th, yend = percentile_95th),
               linewidth = 3, alpha = 0.6) +
  geom_point(aes(y = modal_length), size = 3, shape = 23, fill = "white") +
  facet_wrap(~ group, scales = "free_x", nrow = 1) +
  labs(title = "Percentile Ranges (5th-95th) with Modal Length",
       subtitle = "Robust measure of spread (diamond = modal length)",
       x = "Sample",
       y = "Repeat Length",
       colour = "Analysis Range") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

print(p_percentiles)
ggsave(file.path(plots_dir, "07_percentile_ranges.png"),
       p_percentiles, width = 12, height = 6, dpi = 300)

# Plot 7: Distribution Shape (Skewness vs Kurtosis)
shape_data <- distribution_summary %>%
  filter(!is.na(skewness), !is.na(kurtosis))

if (nrow(shape_data) > 0) {
  p_shape <- ggplot(shape_data,
                    aes(x = skewness, y = kurtosis, 
                        colour = group, shape = analysis_range_name)) +
    geom_hline(yintercept = 3, linetype = "dashed", colour = "grey50", alpha = 0.5) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
    geom_point(size = 4, alpha = 0.7) +
    labs(title = "Distribution Shape: Skewness vs Kurtosis",
         subtitle = "Reference lines: skewness=0 (symmetric), kurtosis=3 (normal)",
         x = "Skewness (symmetry)",
         y = "Kurtosis (tail weight)",
         colour = "Group",
         shape = "Analysis Range") +
    theme_bw() +
    theme(legend.position = "right")
  
  print(p_shape)
  ggsave(file.path(plots_dir, "08_distribution_shape.png"),
         p_shape, width = 10, height = 7, dpi = 300)
}

# Plot 21: Density plots by group and time (full range + per analysis range)
message("\nGenerating density plots by group and time...")

# Create directory for density plots
density_by_group_and_time_dir <- file.path(plots_dir, "21_density_by_group_and_time")
if (!dir.exists(density_by_group_and_time_dir)) {
  dir.create(density_by_group_and_time_dir, recursive = TRUE)
}

# Prepare base data: join repeat counts with group/time metadata
density_data_full <- alignment_clustered %>%
  left_join(
    select(settings, file_stem, group, time),
    by = "file_stem"
  ) %>%
  filter(!is.na(group), !is.na(time), !is.na(repeat_count)) %>%
  arrange(group, time) %>%
  mutate(
    sample_label = paste0(file_stem, " (t=", time, ")"),
    # Order samples by time within each group (earliest = lowest number)
    time_order = as.numeric(time)
  )

if (nrow(density_data_full) > 0) {
  
  # Get unique analysis ranges from settings
  all_ranges <- unique(unlist(lapply(settings_parsed$ranges_list, names)))
  
  # Create a list of range definitions: "Full" + each analysis range
  range_list <- c(
    list(Full = c(-Inf, Inf)),  # Full range (all data)
    setNames(
      lapply(all_ranges, function(rng) {
        # Get the most common range boundaries for this range name
        range_bounds <- sapply(settings_parsed$ranges_list, function(x) {
          if (rng %in% names(x)) x[[rng]] else c(NA, NA)
        })
        # Take the first non-NA range
        valid_bounds <- range_bounds[, !is.na(range_bounds[1,]), drop = FALSE]
        if (ncol(valid_bounds) > 0) {
          bounds <- valid_bounds[, 1]
          # Replace NA with -Inf/Inf for filtering
          c(ifelse(is.na(bounds[1]), -Inf, bounds[1]),
            ifelse(is.na(bounds[2]), Inf, bounds[2]))
        } else {
          c(-Inf, Inf)
        }
      }),
      all_ranges
    )
  )
  
  # Generate plots for each range
  for (range_name in names(range_list)) {
    
    message("\n  Processing range: ", range_name)
    
    # Filter data by range
    range_bounds <- range_list[[range_name]]
    density_data_range <- density_data_full %>%
      filter(repeat_count >= range_bounds[1] & repeat_count <= range_bounds[2])
    
    if (nrow(density_data_range) == 0) {
      message("    No data in this range, skipping")
      next
    }
    
    # Create subdirectory for this range
    density_range_dir <- file.path(density_by_group_and_time_dir, range_name)
    if (!dir.exists(density_range_dir)) {
      dir.create(density_range_dir, recursive = TRUE)
    }
    
    # COMBINED PLOT: All groups in columns for this range
    p_density_combined <- ggplot(density_data_range, 
                              aes(x = repeat_count, 
                                  y = reorder(sample_label, -time_order),  # Earliest at top
                                  fill = group)) +
      geom_density_ridges(alpha = 0.7, scale = 2, rel_min_height = 0.01) +
      facet_grid(~ group, scales = "free_y") +  # Groups as COLUMNS
      scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
      labs(
        title = paste0("Repeat Length Distribution by Group and Time - ", range_name),
        subtitle = "Samples ordered by time within each group (earliest at top)",
        x = "Repeat Count",
        y = "Sample (Time)",
        fill = "Group"
      ) +
      theme_minimal() +
      theme(
        legend.position = "none",
        strip.text = element_text(face = "bold", size = 12),
        axis.text.y = element_text(size = 8)
      )
    
    print(p_density_combined)
    
    # Calculate dimensions
    n_samples_range <- length(unique(density_data_range$file_stem))
    n_groups <- length(unique(density_data_range$group))
    plot_height <- max(8, min(50, n_samples_range * 0.4))
    plot_width <- max(10, min(40, n_groups * 8))
    
    # Determine filename based on range
    combined_filename <- if (range_name == "Full") {
      "density.png"
    } else {
      paste0("density-", range_name, ".png")
    }
    
    ggsave(
      file.path(density_range_dir, combined_filename),
      p_density_combined,
      width = plot_width,
      height = plot_height,
      dpi = 300,
      bg = "white",
      limitsize = FALSE
    )
    
    message("    Combined plot saved: ", combined_filename)
    
    # INDIVIDUAL PLOTS: One per group for this range
    for (my_group in unique(density_data_range$group)) {
      group_data <- density_data_range %>%
        filter(group == my_group) %>%
        arrange(time)
      
      if (nrow(group_data) == 0) next
      
      # Create plot for this group
      p_density_group <- ggplot(group_data, 
                             aes(x = repeat_count, 
                                 y = reorder(sample_label, -time_order),  # Earliest at top
                                 fill = group)) +
        geom_density_ridges(alpha = 0.7, scale = 2, rel_min_height = 0.01) +
        scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
        labs(
          title = paste0("Repeat Length Distribution - ", my_group, " (", range_name, ")"),
          subtitle = "Samples ordered by time (earliest at top)",
          x = "Repeat Count",
          y = "Sample (Time)",
          fill = "Group"
        ) +
        theme_minimal() +
        theme(
          legend.position = "none",
          axis.text.y = element_text(size = 9)
        )
      
      # Calculate height for this group
      n_samples_group <- length(unique(group_data$file_stem))
      group_height <- max(6, min(40, n_samples_group * 0.5))
      
      # Determine filename based on range
      group_filename <- if (range_name == "Full") {
        paste0("density-", my_group, ".png")
      } else {
        paste0("density-", my_group, "-", range_name, ".png")
      }
      
      ggsave(
        file.path(density_range_dir, group_filename),
        p_density_group,
        width = 12,
        height = group_height,
        dpi = 300,
        bg = "white",
        limitsize = FALSE
      )
      
      message("    Density plot saved for group: ", my_group, " (", group_filename, ")")
    }
  }
  
  message("\n  All ridge plots saved to: ", density_by_group_and_time_dir)
  
} else {
  message("  Warning: No data available for ridge plots (missing group/time metadata)")
}


# Generate individual distribution plots by analysis range
message("\nGenerating individual distribution plots by analysis range...")

distribution_by_range_dir <- file.path(plots_dir, "distribution_by_range")
if (!dir.exists(distribution_by_range_dir)) {
  dir.create(distribution_by_range_dir, recursive = TRUE)
}

# Create subdirectories for specific plot types
modal_vs_mean_by_range_dir <- file.path(plots_dir, "02_modal_vs_mean_by_range")
if (!dir.exists(modal_vs_mean_by_range_dir)) {
  dir.create(modal_vs_mean_by_range_dir, recursive = TRUE)
}

tail_balance_by_range_dir <- file.path(plots_dir, "05_tail_balance_by_range")
if (!dir.exists(tail_balance_by_range_dir)) {
  dir.create(tail_balance_by_range_dir, recursive = TRUE)
}

for (my_range in unique(distribution_summary$analysis_range_name)) {
  range_data <- distribution_summary %>%
    filter(analysis_range_name == my_range)
  
  if (nrow(range_data) == 0) next
  
  # Plot 1: Modal vs Mean for this range
  p_modal_mean_range <- ggplot(range_data,
                               aes(x = modal_length, y = mean_length, colour = group)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(size = 4, alpha = 0.7) +
    labs(title = paste0("Modal vs Mean - ", my_range),
         subtitle = "Points near diagonal indicate symmetric distributions",
         x = "Modal Length",
         y = "Mean Length",
         colour = "Group") +
    theme_bw() +
    theme(legend.position = "right")
  
  ggsave(file.path(modal_vs_mean_by_range_dir, 
                  paste0("modal_vs_mean-", my_range, ".png")),
         p_modal_mean_range, width = 10, height = 6, dpi = 300)
  
  # Plot 2: Tail balance for this range
  tail_data_range <- range_data %>%
    mutate(balanced_proportion = 1 - contraction_proportion - expansion_proportion) %>%
    select(file_stem, group, contraction_proportion, balanced_proportion, expansion_proportion) %>%
    pivot_longer(cols = c(contraction_proportion, balanced_proportion, expansion_proportion),
                 names_to = "direction",
                 values_to = "proportion") %>%
    mutate(direction = case_when(
      direction == "contraction_proportion" ~ "Contraction",
      direction == "balanced_proportion" ~ "At modal",
      direction == "expansion_proportion" ~ "Expansion"
    ),
    # Set factor levels in REVERSE order: expansion first stacks on TOP
    direction = factor(direction, levels = c("Expansion", "At modal", "Contraction")))
  
  p_tail_range <- ggplot(tail_data_range,
                         aes(x = file_stem, y = proportion, fill = direction)) +
    geom_col(position = "stack", alpha = 0.8) +
    geom_hline(yintercept = 0.5, linetype = "dashed", colour = "black") +
    facet_wrap(~ group, scales = "free_x", nrow = 1) +
    scale_fill_manual(
      values = c("Contraction" = "steelblue", 
                 "At modal" = "grey60",
                 "Expansion" = "coral"),
      breaks = c("Contraction", "At modal", "Expansion")
    ) +
    labs(title = paste0("Tail Balance - ", my_range),
         subtitle = "Proportion of reads below, at, and above modal (dashed line = balanced)",
         x = "Sample",
         y = "Proportion",
         fill = "Direction") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  ggsave(file.path(tail_balance_by_range_dir, 
                  paste0("tail_balance-", my_range, ".png")),
         p_tail_range, width = 12, height = 6, dpi = 300)
  
  # Plot 3: Key metrics for this range
  p_metrics_range <- ggplot(range_data,
                            aes(x = file_stem)) +
    geom_col(aes(y = sd, fill = "SD"), alpha = 0.8, position = "dodge") +
    geom_point(aes(y = coefficient_of_variation * 10, colour = "CV×10"), 
               size = 3, shape = 17) +
    facet_wrap(~ group, scales = "free", nrow = 1) +
    scale_y_continuous(
      name = "Standard Deviation",
      sec.axis = sec_axis(~./10, name = "Coefficient of Variation")
    ) +
    labs(title = paste0("Distribution Metrics - ", my_range),
         x = "Sample",
         fill = "Metric",
         colour = "Metric") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  ggsave(file.path(distribution_by_range_dir, 
                  paste0("metrics_", my_range, ".png")),
         p_metrics_range, width = 12, height = 6, dpi = 300)
}

message("  Individual range plots saved to: ", distribution_by_range_dir)

# Generate individual distribution plots by sample
message("\nGenerating individual distribution plots by sample...")

distribution_by_sample_dir <- file.path(plots_dir, "distribution_by_sample")
if (!dir.exists(distribution_by_sample_dir)) {
  dir.create(distribution_by_sample_dir, recursive = TRUE)
}

for (my_sample in unique(distribution_summary$file_stem)) {
  sample_data <- distribution_summary %>%
    filter(file_stem == my_sample)
  
  if (nrow(sample_data) == 0) next
  
  # Create multi-panel plot for this sample
  # Panel 1: Central tendency
  central_data <- sample_data %>%
    select(analysis_range_name, modal_length, mean_length, median_length) %>%
    pivot_longer(cols = c(modal_length, mean_length, median_length),
                 names_to = "metric",
                 values_to = "value") %>%
    mutate(metric = gsub("_length", "", metric),
           metric = tools::toTitleCase(metric))
  
  p1 <- ggplot(central_data,
              aes(x = analysis_range_name, y = value, fill = metric)) +
    geom_col(position = "dodge", alpha = 0.8) +
    labs(title = "Central Tendency",
         x = "Analysis Range",
         y = "Repeat Length",
         fill = "Metric") +
    theme_bw() +
    theme(legend.position = "bottom")
  
  # Panel 2: Spread metrics
  spread_data <- sample_data %>%
    select(analysis_range_name, sd, interquartile_range, 
           percentile_5th, percentile_95th, modal_length)
  
  p2 <- ggplot(spread_data,
              aes(x = analysis_range_name)) +
    # Show percentile range as segments (absolute positions)
    geom_segment(aes(y = percentile_5th, yend = percentile_95th,
                     x = as.numeric(factor(analysis_range_name)) - 0.2,
                     xend = as.numeric(factor(analysis_range_name)) - 0.2),
                 colour = "coral", linewidth = 4, alpha = 0.6) +
    # Show SD as error bars around modal length
    geom_errorbar(aes(ymin = modal_length - sd, ymax = modal_length + sd,
                      x = as.numeric(factor(analysis_range_name)) + 0.2),
                  width = 0.2, colour = "steelblue", linewidth = 2, alpha = 0.8) +
    geom_point(aes(y = modal_length, 
                   x = as.numeric(factor(analysis_range_name)) + 0.2),
               colour = "steelblue", size = 3) +
    scale_x_continuous(breaks = 1:length(unique(spread_data$analysis_range_name)),
                      labels = unique(spread_data$analysis_range_name)) +
    labs(title = "Spread Metrics",
         subtitle = "Orange bars = 5th-95th percentile; Blue bars = SD around modal",
         x = "Analysis Range",
         y = "Repeat Length") +
    theme_bw()
  
  # Panel 3: Tail metrics
  tail_data <- sample_data %>%
    mutate(balanced_proportion = 1 - contraction_proportion - expansion_proportion) %>%
    select(analysis_range_name, contraction_proportion, balanced_proportion, expansion_proportion) %>%
    pivot_longer(cols = c(contraction_proportion, balanced_proportion, expansion_proportion),
                 names_to = "direction",
                 values_to = "proportion") %>%
    mutate(direction = case_when(
      direction == "contraction_proportion" ~ "Contraction",
      direction == "balanced_proportion" ~ "At modal",
      direction == "expansion_proportion" ~ "Expansion"
    ),
    # Set factor levels in REVERSE order: expansion first stacks on TOP
    direction = factor(direction, levels = c("Expansion", "At modal", "Contraction")))
  
  p3 <- ggplot(tail_data,
              aes(x = analysis_range_name, y = proportion, fill = direction)) +
    geom_col(position = "stack", alpha = 0.8) +
    geom_hline(yintercept = 0.5, linetype = "dashed", colour = "black") +
    scale_fill_manual(
      values = c("Contraction" = "steelblue", 
                 "At modal" = "grey60",
                 "Expansion" = "coral"),
      breaks = c("Contraction", "At modal", "Expansion")
    ) +
    labs(title = "Tail Balance",
         subtitle = "Proportion of reads below, at, and above modal (dashed line = balanced)",
         x = "Analysis Range",
         y = "Proportion of Reads",
         fill = "Direction") +
    theme_bw() +
    theme(legend.position = "bottom")
  
  # Panel 4: Shape metrics
  shape_data <- sample_data %>%
    filter(!is.na(skewness), !is.na(peak_sharpness))
  
  if (nrow(shape_data) > 0) {
    shape_long <- shape_data %>%
      select(analysis_range_name, skewness, peak_sharpness, entropy_normalised) %>%
      pivot_longer(cols = c(skewness, peak_sharpness, entropy_normalised),
                   names_to = "metric",
                   values_to = "value") %>%
      mutate(metric = gsub("_", " ", metric),
             metric = tools::toTitleCase(metric))
    
    p4 <- ggplot(shape_long,
                aes(x = analysis_range_name, y = value, fill = metric)) +
      geom_col(position = "dodge", alpha = 0.8) +
      labs(title = "Shape Metrics",
           x = "Analysis Range",
           y = "Value",
           fill = "Metric") +
      theme_bw() +
      theme(legend.position = "bottom")
  } else {
    p4 <- ggplot() + 
      labs(title = "Shape Metrics (not available)") + 
      theme_void()
  }
  
  # Combine all panels
  p_combined <- cowplot::plot_grid(p1, p2, p3, p4, ncol = 2,
                                   labels = c("A", "B", "C", "D"),
                                   label_size = 14)
  
  # Add title
  title_text <- cowplot::ggdraw() + 
    cowplot::draw_label(paste0("Distribution Summary - ", my_sample),
                       fontface = "bold", size = 16)
  
  p_final <- cowplot::plot_grid(title_text, p_combined, ncol = 1,
                                rel_heights = c(0.05, 1))
  
  ggsave(file.path(distribution_by_sample_dir, 
                  paste0(my_sample, "-distribution.png")),
         p_final, width = 14, height = 10, dpi = 300)
}

message("  Individual sample plots saved to: ", distribution_by_sample_dir)

message("\nDistribution plots saved to: ", plots_dir)
```

---

# Group controls

Define control repeat lengths for each group and analysis range.

Group controls provide baseline reference values for calculating instability metrics.

**Sample selection methods:**

- `flagged`: Use samples flagged with `group_control_sample = 1`
- `earliest`: Use first timepoint (time == min)
- `all`: Average all samples in group

**Setpoint metrics:**

- `modal_length`: Most frequent repeat length (robust, default)
- `mean_length`: Arithmetic mean (sensitive to outliers)
- `median_length`: Median repeat length (robust middle value)

```{r group_controls, echo=FALSE}
if (params$group_control) {
  
  # Map control_setpoint_metric to column name
  metric_column_map <- c(
    "modal_length" = "modal_length",
    "mean_length" = "mean_length",
    "median_length" = "median_length"
  )
  
  # Get the column name to use
  if (!params$control_setpoint_metric %in% names(metric_column_map)) {
    stop("Invalid control_setpoint_metric: ", params$control_setpoint_metric,
         ". Must be one of: ", paste(names(metric_column_map), collapse = ", "))
  }
  
  metric_column <- metric_column_map[[params$control_setpoint_metric]]
  
  # Get unique analysis ranges
  analysis_range_names <- unique(distribution_summary$analysis_range_name)
  
  # Store control sample data for plotting
  control_samples_list <- list()
  
  # Calculate group controls for each range
  group_control_list <- lapply(analysis_range_names, function(my_range_name) {
    
    # Filter to this range
    control_data <- distribution_summary %>%
      filter(
        analysis_range_name == my_range_name,
        is.na(exclude),
        !is.na(!!sym(metric_column)),
        !is.na(group)
      )
    
    if (nrow(control_data) == 0) {
      return(NULL)
    }
    
    # Select control samples based on method
    if (params$control_sample_selection == "earliest") {
      # Use earliest timepoint
      control_data <- control_data %>%
        group_by(group) %>%
        filter(time == min(time, na.rm = TRUE)) %>%
        ungroup()
    } else if (params$control_sample_selection == "flagged") {
      # Use flagged controls
      control_data <- control_data %>%
        filter(!is.na(group_control_sample))
    }
    # If "all", use all samples (no filtering)
    
    # Store control samples for this range
    control_samples_list[[my_range_name]] <<- control_data
    
    # Calculate group control values for all three metrics
    control_summary <- control_data %>%
      group_by(group) %>%
      summarise(
        control_modal_length = mean(modal_length, na.rm = TRUE),
        control_mean_length = mean(mean_length, na.rm = TRUE),
        control_median_length = mean(median_length, na.rm = TRUE),
        control_count = n(),
        control_sd = sd(!!sym(metric_column), na.rm = TRUE),
        control_sem = control_sd / sqrt(control_count),
        .groups = "drop"
      ) %>%
      mutate(
        analysis_range_name = my_range_name,
        # Set control setpoint based on user's choice
        control_setpoint = !!sym(paste0("control_", params$control_setpoint_metric)),
        control_metric_used = params$control_setpoint_metric
      )
    
    return(control_summary)
  })
  
  # Combine results
  group_controls <- bind_rows(group_control_list) %>%
    relocate(group, analysis_range_name)
  
  message("\nGroup controls calculated:")
  message("  Groups: ", length(unique(group_controls$group)))
  message("  Ranges: ", length(unique(group_controls$analysis_range_name)))
  message("  Sample selection: ", params$control_sample_selection)
  message("  Setpoint metric: ", params$control_setpoint_metric)
  
  # Display group controls
  caption_text <- paste0("Group control setpoints per analysis range (using ", 
                        params$control_setpoint_metric, ")")
  
  DT::datatable(
    group_controls,
    caption = caption_text,
    options = list(
      pageLength = 25,
      scrollX = TRUE,
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel')
    ),
    rownames = FALSE,
    filter = 'top',
    class = 'cell-border stripe compact'
  ) %>%
    DT::formatRound(columns = colnames(group_controls)[sapply(group_controls, is.numeric)], digits = 2)
  
} else {
  message("\nGroup controls disabled (group_control = FALSE)")
  group_controls <- NULL
}
```

```{r add_controls_to_comprehensive_summary, echo=FALSE}
# Add control information to comprehensive summary
if (params$group_control && exists("group_controls") && !is.null(group_controls)) {
  
  message("\n=== Adding Group Controls to Comprehensive Summary ===")
  
  # Get control setpoints per group and range
  control_setpoints <- group_controls %>%
    select(group, analysis_range_name, control_setpoint, control_metric_used)
  
  # Join to comprehensive summary
  comprehensive_range_summary <- comprehensive_range_summary %>%
    left_join(control_setpoints, by = c("group", "analysis_range_name"))
  
  message("  Control setpoints added successfully")
  
  # Dynamically identify peak columns for formatting
  peak_cols <- grep("^peak[0-9]+_(length|frequency)$", 
                    colnames(comprehensive_range_summary), 
                    value = TRUE)
  peak_length_cols <- grep("^peak[0-9]+_length$", peak_cols, value = TRUE)
  peak_freq_cols <- grep("^peak[0-9]+_frequency$", peak_cols, value = TRUE)
  
  # Display final comprehensive table
  dt_final <- DT::datatable(
    comprehensive_range_summary,
    caption = "COMPLETE: Comprehensive Range Summary - Configuration, Peaks, and Controls",
    extensions = c('Buttons', 'FixedColumns'),
    options = list(
      pageLength = 25,
      scrollX = TRUE,
      scrollY = "700px",
      dom = 'Bfrtip',
      buttons = list(
        'copy',
        list(extend = 'csv', filename = 'comprehensive_range_summary_complete'),
        list(extend = 'excel', filename = 'comprehensive_range_summary_complete'),
        'colvis',
        list(
          extend = 'pdfHtml5',
          orientation = 'landscape',
          pageSize = 'A3'
        )
      ),
      fixedColumns = list(leftColumns = 2),
      columnDefs = list(
        list(width = '110px', targets = 0:1)
      )
    ),
    rownames = FALSE,
    filter = 'top',
    class = 'cell-border stripe compact hover',
    style = 'bootstrap'
  )
  
  # Format numeric columns
  numeric_cols <- c('settings_min', 'settings_max', 'analysis_min', 'analysis_max',
                    'observed_min', 'observed_max', 
                    'manual_control_repeat_length', 'control_setpoint')
  numeric_cols <- c(numeric_cols, peak_length_cols)
  numeric_cols <- numeric_cols[numeric_cols %in% colnames(comprehensive_range_summary)]
  
  dt_final <- dt_final %>%
    DT::formatRound(columns = numeric_cols, digits = 1)
  
  # Highlight peak1 (most important) if it exists
  if ("peak1_length" %in% colnames(comprehensive_range_summary)) {
    dt_final <- dt_final %>%
      DT::formatStyle(
        c('peak1_length', 'peak1_frequency'),
        fontWeight = 'bold',
        backgroundColor = '#e3f2fd'
      )
  }
  
  # Highlight peak2 if it exists
  if ("peak2_length" %in% colnames(comprehensive_range_summary)) {
    dt_final <- dt_final %>%
      DT::formatStyle(
        c('peak2_length', 'peak2_frequency'),
        backgroundColor = '#f0f0f0'
      )
  }
  
  # Highlight additional peaks (peak3, peak4, etc.) with progressively lighter grey
  for (i in 3:length(peak_length_cols)) {
    peak_len_col <- paste0("peak", i, "_length")
    peak_freq_col <- paste0("peak", i, "_frequency")
    if (peak_len_col %in% colnames(comprehensive_range_summary)) {
      # Progressively lighter grey: #f5f5f5, #f8f8f8, #fafafa
      grey_value <- sprintf("#f%df%df%d", 2+i, 2+i, 2+i)
      dt_final <- dt_final %>%
        DT::formatStyle(
          c(peak_len_col, peak_freq_col),
          backgroundColor = grey_value
        )
    }
  }
  
  # Highlight manual controls and control setpoints
  dt_final <- dt_final %>%
    DT::formatStyle(
      'manual_control_repeat_length',
      backgroundColor = DT::styleEqual(NA, 'white', default = '#fff3cd')
    ) %>%
    DT::formatStyle(
      'control_setpoint',
      backgroundColor = DT::styleEqual(NA, 'white', default = '#d4edda'),
      fontWeight = 'bold'
    ) %>%
    DT::formatStyle(
      columns = colnames(comprehensive_range_summary),
      fontSize = '10px'
    )
  
  dt_final
}
```

```{r plot_group_controls, echo=FALSE, fig.width=12, fig.height=7}
if (params$group_control && !is.null(group_controls)) {
  
  # Combine all control samples for plotting
  control_samples <- bind_rows(control_samples_list)
  
  # Get the metric column name for plotting
  metric_col_name <- params$control_setpoint_metric
  
  # Plot 1: Control setpoints with violin + jitter showing actual samples
  p_controls <- ggplot() +
    # Violin plot showing distribution
    geom_violin(data = control_samples,
                aes(x = group, y = !!sym(metric_col_name), fill = analysis_range_name),
                position = position_dodge(0.9), alpha = 0.3, scale = "width") +
    # Jitter showing individual control samples (horizontal spread only, no vertical jitter)
    geom_jitter(data = control_samples,
                aes(x = group, y = !!sym(metric_col_name), colour = analysis_range_name),
                position = position_jitterdodge(jitter.width = 0.2, jitter.height = 0, dodge.width = 0.9),
                alpha = 0.6, size = 2) +
    # Mean setpoint as larger point
    geom_point(data = group_controls,
               aes(x = group, y = control_setpoint, fill = analysis_range_name),
               position = position_dodge(0.9),
               size = 4, shape = 23, colour = "black", stroke = 1) +
    labs(title = "Group Control Setpoints with Sample Distributions",
         subtitle = paste0("Using ", params$control_setpoint_metric, 
                          " | Violins show distribution, points show samples at actual values, diamonds show mean setpoint"),
         x = "Group",
         y = paste0(tools::toTitleCase(gsub("_", " ", metric_col_name)), " (repeats)"),
         fill = "Analysis Range",
         colour = "Analysis Range") +
    theme_bw() +
    theme(legend.position = "right")
  
  print(p_controls)
  ggsave(file.path(plots_dir, "09_control_setpoints.png"),
         p_controls, width = 12, height = 7, dpi = 300)
  
  # Plot 2: Show all three metrics with sample distributions
  # Prepare long-format data for all three metrics
  controls_samples_long <- control_samples %>%
    select(file_stem, group, analysis_range_name, 
           modal_length, mean_length, median_length) %>%
    pivot_longer(cols = c(modal_length, mean_length, median_length),
                names_to = "metric",
                values_to = "value") %>%
    mutate(metric = gsub("_length", "", metric),
           metric = tools::toTitleCase(metric))
  
  # Prepare setpoints
  controls_setpoints_long <- group_controls %>%
    select(group, analysis_range_name, 
           control_modal_length, control_mean_length, control_median_length) %>%
    pivot_longer(cols = starts_with("control_"),
                names_to = "metric",
                values_to = "setpoint") %>%
    mutate(metric = gsub("control_", "", metric),
           metric = gsub("_length", "", metric),
           metric = tools::toTitleCase(metric))
  
  p_metrics_compare <- ggplot() +
    # Violin plot
    geom_violin(data = controls_samples_long,
                aes(x = metric, y = value, fill = metric),
                alpha = 0.3, scale = "width") +
    # Individual samples (horizontal jitter only, no vertical jitter)
    geom_jitter(data = controls_samples_long,
                aes(x = metric, y = value, colour = metric),
                alpha = 0.5, size = 1.5, width = 0.2, height = 0) +
    # Mean setpoints
    geom_point(data = controls_setpoints_long,
               aes(x = metric, y = setpoint),
               size = 4, shape = 23, fill = "black", colour = "white", stroke = 1) +
    facet_grid(analysis_range_name ~ group, scales = "free_y") +
    labs(title = "Comparison of Control Metrics with Sample Distributions",
         subtitle = "Violins show distributions, points show samples at actual values, black diamonds show mean setpoints",
         x = "Metric",
         y = "Repeat Length",
         fill = "Metric",
         colour = "Metric") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  print(p_metrics_compare)
  ggsave(file.path(plots_dir, "10_control_metrics_comparison.png"),
         p_metrics_compare, width = 12, height = 8, dpi = 300)
  
  # Plot 3: Sample counts per group/range
  sample_counts <- control_samples %>%
    group_by(group, analysis_range_name) %>%
    summarise(n_samples = n(), .groups = "drop")
  
  p_counts <- ggplot(sample_counts,
                     aes(x = group, y = n_samples, fill = analysis_range_name, label = n_samples)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_text(position = position_dodge(0.9), vjust = -0.5, size = 4) +
    labs(title = "Number of Control Samples",
         subtitle = "Samples contributing to each group control",
         x = "Group",
         y = "Number of Samples",
         fill = "Analysis Range") +
    theme_bw() +
    theme(legend.position = "right")
  
  print(p_counts)
  ggsave(file.path(plots_dir, "11_control_sample_counts.png"),
         p_counts, width = 10, height = 6, dpi = 300)
}
```

---

# Instability metrics

Calculate instability metrics relative to setpoints using the actual read distribution.

**Two setpoint types (both using user-selected metric):**

1. **"sample"** - Each sample's own value of the chosen metric
2. **"control"** - Group control (aggregated from control samples using the same metric)

**User-selected metric:** `r params$control_setpoint_metric`  
**Control aggregation:** `r params$control_aggregation_method` of control samples

**Example:** If user selects `modal_length` with `median` aggregation:
- Sample setpoint = Sample's modal_length (e.g., 19)
- Control setpoint = Median of control samples' modal_lengths (e.g., 17)

For each sample and setpoint type, we calculate:

- **Instability indices**: Mean deviations from setpoint (instability, expansion, contraction)
- **Read counts**: Number of reads at, below, and above setpoint
- **Tail ratios**: Relative sizes of expansion and contraction tails

```{r instability_metrics, echo=FALSE}
# ALWAYS calculate instability metrics (at minimum, relative to sample itself)
# If group_control = TRUE, also calculate relative to control setpoint

# Instability functions are in range_analysis.R (already sourced)

# Get unique analysis ranges
analysis_range_names <- unique(distribution_summary$analysis_range_name)

# Calculate instability for each range
instability_metrics_list <- lapply(analysis_range_names, function(my_range_name) {
  
  # Get distribution summaries for this range
  range_data <- distribution_summary %>%
    filter(analysis_range_name == my_range_name)
  
  # Get control setpoints for this range (if available)
  if (params$group_control && !is.null(group_controls)) {
    range_controls <- group_controls %>%
      filter(analysis_range_name == my_range_name)
  } else {
    range_controls <- NULL
  }
  
  # Calculate instability for each sample (parallelised)
  sample_file_stems <- unique(range_data$file_stem)
  
  sample_instability_list <- apply_fn(sample_file_stems, function(my_file_stem) {
    
    # Get sample's distribution summary
    sample_summary <- range_data %>%
      filter(file_stem == my_file_stem)
    
    if (nrow(sample_summary) == 0) return(NULL)
    
    # Get sample's group
    sample_group <- sample_summary$group[1]
    
    # Get raw read data for this sample/range
    sample_reads <- alignment_clustered %>%
      filter(
        file_stem == my_file_stem,
        !!sym(params$repeat_count_method) >= sample_summary$range_min[1],
        !!sym(params$repeat_count_method) <= sample_summary$range_max[1]
      ) %>%
      pull(!!sym(params$repeat_count_method))
    
    if (length(sample_reads) == 0) {
      warning(sprintf("No reads for sample '%s' in range '%s'", my_file_stem, my_range_name))
      return(NULL)
    }
    
    # Get floor for this range
    floor_val <- sample_summary$floor_threshold[1]
    
    # Get sample's own setpoint value
    sample_setpoint <- sample_summary[[params$control_setpoint_metric]][1]
    
    if (is.na(sample_setpoint)) {
      warning(sprintf("Sample '%s' has NA for metric '%s'", 
                     my_file_stem, params$control_setpoint_metric))
      return(NULL)
    }
    
    # ALWAYS calculate instability relative to SAMPLE's own setpoint
    instability_sample <- calculate_instability_metrics(
      numbers = sample_reads,
      setpoint = sample_setpoint,
      floor = floor_val
    ) %>%
      mutate(
        file_stem = my_file_stem,
        analysis_range_name = my_range_name,
        group = sample_group,
        setpoint_type = "sample_relative",
        setpoint_metric = params$control_setpoint_metric
      )
    
    # CONDITIONALLY calculate instability relative to CONTROL setpoint
    instability_control <- NULL
    
    if (params$group_control && !is.null(range_controls) && !is.na(sample_group)) {
      # Get control setpoint for this sample's group
      group_control <- range_controls %>%
        filter(group == sample_group)
      
      if (nrow(group_control) > 0) {
        control_setpoint_value <- group_control$control_setpoint[1]
        
        if (!is.na(control_setpoint_value)) {
          instability_control <- calculate_instability_metrics(
            numbers = sample_reads,
            setpoint = control_setpoint_value,
            floor = floor_val
          ) %>%
            mutate(
              file_stem = my_file_stem,
              analysis_range_name = my_range_name,
              group = sample_group,
              setpoint_type = "control_relative",
              setpoint_metric = params$control_setpoint_metric,
              control_sd = group_control$control_sd[1],
              control_sem = group_control$control_sem[1],
              control_n = group_control$control_count[1]
            )
          
          # Add z-score for control comparison
          instability_control <- instability_control %>%
            mutate(
              z_score = if_else(
                !is.na(control_sd) & control_sd > 0,
                (sample_setpoint - setpoint_repeat_length) / control_sd,
                NA_real_
              )
            )
        }
      }
    }
    
    # Combine sample_relative and control_relative (if exists)
    bind_rows(instability_sample, instability_control)
  })
  
  # Extract results from apply_fn
  sample_instability_list <- extract_apply_fn_result(sample_instability_list)
  
  # Combine all samples for this range
  bind_rows(sample_instability_list)
})

# Combine all ranges
instability_metrics <- bind_rows(instability_metrics_list) %>%
  relocate(file_stem, analysis_range_name, group, setpoint_type, setpoint_metric)

# Add relative distribution statistics (modal, median, mean differences from setpoint)
instability_metrics <- instability_metrics %>%
  left_join(
    distribution_summary %>% 
      select(file_stem, analysis_range_name, 
             modal_length, median_length, mean_length),
    by = c("file_stem", "analysis_range_name")
  ) %>%
  mutate(
    # Calculate differences from setpoint
    modal_diff = modal_length - setpoint_repeat_length,
    median_diff = median_length - setpoint_repeat_length,
    mean_diff = mean_length - setpoint_repeat_length
  ) %>%
  # Reorder columns logically
  relocate(file_stem, analysis_range_name, group, setpoint_type, setpoint_metric,
           setpoint_repeat_length, 
           modal_length, median_length, mean_length,
           modal_diff, median_diff, mean_diff)

message("\nInstability metrics calculated:")
message("  Samples: ", length(unique(instability_metrics$file_stem)))
message("  Ranges: ", length(unique(instability_metrics$analysis_range_name)))
message("  Setpoint types: ", paste(unique(instability_metrics$setpoint_type), collapse = ", "))
message("  Setpoint metric: ", params$control_setpoint_metric)
if (params$group_control) {
  message("  Control aggregation: ", params$control_aggregation_method)
}
message("  Includes relative distribution statistics: modal_diff, median_diff, mean_diff")
```

```{r instability_display, echo=FALSE}
if (nrow(instability_metrics) > 0) {
  
  # Display key columns including new relative distribution statistics
  display_cols <- c("file_stem", "analysis_range_name", "setpoint_type", 
                   "setpoint_repeat_length",
                   "modal_diff", "median_diff", "mean_diff",
                   "instability_index", "expansion_index", 
                   "contraction_index", "expansion_ratio", "z_score")
  
  caption_text <- if (params$group_control) {
    paste0("Instability metrics with relative distribution statistics (using ", 
           params$control_setpoint_metric, 
           " with ", params$control_aggregation_method, " of controls)")
  } else {
    paste0("Instability metrics with relative distribution statistics (using ", 
           params$control_setpoint_metric, ")")
  }
  
  datatable(instability_metrics[, display_cols],
            options = list(pageLength = 15, scrollX = TRUE),
            caption = caption_text,
            rownames = FALSE) %>%
    formatRound(columns = c('setpoint_repeat_length', 
                           'modal_diff', 'median_diff', 'mean_diff',
                           'instability_index', 'expansion_index',
                           'contraction_index', 'expansion_ratio', 'z_score'),
                digits = 2) %>%
    formatStyle(columns = display_cols, fontSize = '10px')
}
```

**Metrics explanation:**

**Relative Distribution Statistics** (new):

- **modal_diff**: Difference between sample's modal_length and setpoint (modal - setpoint)
- **median_diff**: Difference between sample's median_length and setpoint (median - setpoint)
- **mean_diff**: Difference between sample's mean_length and setpoint (mean - setpoint)

Positive values indicate expansion relative to setpoint, negative values indicate contraction.

**Instability Metrics** (existing):

- **setpoint_type**: "sample_relative" (sample's own metric value) or "control_relative" (group control)
- **instability_index**: Mean absolute deviation from setpoint (weighted by read frequency)
- **expansion_index**: Mean deviation for reads above setpoint
- **contraction_index**: Mean deviation for reads below setpoint  
- **expansion_ratio**: (reads above) / (reads at setpoint)
- **z_score**: Standard deviations from control (control_relative only)

**Key distinction:**
- Relative distribution statistics measure the *shift* of the distribution from setpoint
- Instability metrics measure the *spread* of the distribution around setpoint

All metrics calculated from actual read distribution, not just summary statistics.

---

```{r plot_instability, echo=FALSE, fig.width=12, fig.height=10}
# Only show these plots if we have control comparisons
if (params$group_control && any(instability_metrics$setpoint_type == "control_relative")) {
  
  
  # Define facet labels for setpoint types
  setpoint_labels <- c(
    "sample_relative" = "Relative to Sample",
    "control_relative" = "Relative to Control"
  )
  # Plot 1: Instability Index (faceted by setpoint type)
  p_instability <- ggplot(instability_metrics,
                          aes(x = file_stem, y = instability_index, fill = analysis_range_name)) +
    geom_col(position = "dodge", alpha = 0.8) +
    facet_grid(setpoint_type ~ group, scales = "free_x",
           labeller = labeller(setpoint_type = setpoint_labels)) +
    labs(title = "Instability Index",
         subtitle = paste0("Mean absolute deviation from setpoint (using ", 
                          params$control_setpoint_metric, ")"),
         x = "Sample",
         y = "Instability Index",
         fill = "Analysis Range") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  print(p_instability)
  ggsave(file.path(plots_dir, "12_instability_index.png"),
         p_instability, width = 12, height = 7, dpi = 300)
  
  # Plot 2: Expansion vs Contraction
  expansion_contraction_data <- instability_metrics %>%
    select(file_stem, analysis_range_name, group, setpoint_type, 
           expansion_index, contraction_index) %>%
    pivot_longer(cols = c(expansion_index, contraction_index),
                names_to = "direction",
                values_to = "index") %>%
    mutate(direction = gsub("_index", "", direction),
           direction = tools::toTitleCase(direction))
  
  p_exp_cont <- ggplot(expansion_contraction_data,
                       aes(x = file_stem, y = index, fill = direction)) +
    geom_col(position = "dodge", alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    facet_grid(setpoint_type + analysis_range_name ~ group, scales = "free",
           labeller = labeller(setpoint_type = setpoint_labels)) +
    labs(title = "Expansion vs Contraction Indices",
         subtitle = "Mean deviations above and below setpoint",
         x = "Sample",
         y = "Index (repeats)",
         fill = "Direction") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  print(p_exp_cont)
  ggsave(file.path(plots_dir, "13_expansion_contraction.png"),
         p_exp_cont, width = 14, height = 10, dpi = 300)
  
  # Plot 3: Z-scores (control setpoint only)
  zscore_data <- instability_metrics %>%
    filter(setpoint_type == "control_relative", !is.na(z_score))
  
  if (nrow(zscore_data) > 0) {
    p_zscore <- ggplot(zscore_data,
                       aes(x = file_stem, y = z_score, fill = analysis_range_name)) +
      geom_col(position = "dodge", alpha = 0.8) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_hline(yintercept = c(-2, 2), linetype = "dotted", colour = "red", alpha = 0.5) +
      facet_wrap(~group, scales = "free_x", nrow = 1) +
      labs(title = "Z-Scores from Control Setpoint",
           subtitle = "Standard deviations from control (dotted lines at ±2 SD)",
           x = "Sample",
           y = "Z-Score",
           fill = "Analysis Range") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "bottom")
    
    print(p_zscore)
    ggsave(file.path(plots_dir, "14_zscore.png"),
           p_zscore, width = 12, height = 6, dpi = 300)
  }
  
  # Plot 4: Expansion Ratio (control setpoint)
  ratio_data <- instability_metrics %>%
    filter(setpoint_type == "control_relative")
  
  p_ratio <- ggplot(ratio_data,
                    aes(x = file_stem, y = expansion_ratio, fill = analysis_range_name)) +
    geom_col(position = "dodge", alpha = 0.8) +
    facet_wrap(~group, scales = "free_x", nrow = 1) +
    labs(title = "Expansion Ratio (Control Setpoint)",
         subtitle = "(Reads above control) / (Reads at control)",
         x = "Sample",
         y = "Expansion Ratio",
         fill = "Analysis Range") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  print(p_ratio)
  ggsave(file.path(plots_dir, "15_expansion_ratio.png"),
         p_ratio, width = 12, height = 6, dpi = 300)
  

# NEW INSTABILITY PLOTS - INSERT AFTER LINE 1047 (after expansion ratio plot)

  # Plot 5: Contraction Ratio (control setpoint)
  p_contraction_ratio <- ggplot(ratio_data,
                                aes(x = file_stem, y = contraction_ratio, fill = analysis_range_name)) +
    geom_col(position = "dodge", alpha = 0.8) +
    facet_wrap(~group, scales = "free_x", nrow = 1) +
    labs(title = "Contraction Ratio (Control Setpoint)",
         subtitle = "(Reads below control) / (Reads at control)",
         x = "Sample",
         y = "Contraction Ratio",
         fill = "Analysis Range") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  print(p_contraction_ratio)
  ggsave(file.path(plots_dir, "16_contraction_ratio.png"),
         p_contraction_ratio, width = 12, height = 6, dpi = 300)
  
  # Plot 6: Total Tail Ratio (both setpoint types)
  p_total_tail <- ggplot(instability_metrics,
                         aes(x = file_stem, y = total_tail_ratio, fill = analysis_range_name)) +
    geom_col(position = "dodge", alpha = 0.8) +
    facet_grid(setpoint_type ~ group, scales = "free_x",
           labeller = labeller(setpoint_type = setpoint_labels)) +
    labs(title = "Total Tail Ratio",
         subtitle = "(Reads above + Reads below) / Reads at setpoint",
         x = "Sample",
         y = "Total Tail Ratio",
         fill = "Analysis Range") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  print(p_total_tail)
  ggsave(file.path(plots_dir, "17_total_tail_ratio.png"),
         p_total_tail, width = 14, height = 8, dpi = 300)
  
  # Plot 7: Read Counts by Region (stacked bars)
  read_counts_data <- instability_metrics %>%
    select(file_stem, analysis_range_name, group, setpoint_type,
           reads_below_setpoint, reads_at_setpoint, reads_above_setpoint) %>%
    pivot_longer(cols = c(reads_below_setpoint, reads_at_setpoint, reads_above_setpoint),
                 names_to = "region",
                 values_to = "count") %>%
    mutate(region = case_when(
      region == "reads_below_setpoint" ~ "Below setpoint",
      region == "reads_at_setpoint" ~ "At setpoint",
      region == "reads_above_setpoint" ~ "Above setpoint"
    ),
    # Set factor levels: below first (bottom), at middle, above last (top)
    # This matches legend order
    region = factor(region, levels = c("Below setpoint", "At setpoint", "Above setpoint")))
  
  p_read_counts <- ggplot(read_counts_data,
                          aes(x = file_stem, y = count, fill = region)) +
    geom_col(position = "stack", alpha = 0.8) +
    facet_grid(setpoint_type + analysis_range_name ~ group, scales = "free",
           labeller = labeller(setpoint_type = setpoint_labels)) +
    labs(title = "Read Distribution by Region Relative to Setpoint",
         subtitle = "Absolute read counts below, at, and above setpoint",
         x = "Sample",
         y = "Number of Reads",
         fill = "Region") +
    scale_fill_manual(
      values = c("Below setpoint" = "steelblue", 
                 "At setpoint" = "grey70",
                 "Above setpoint" = "coral"),
      # Explicit breaks to ensure legend order
      breaks = c("Below setpoint", "At setpoint", "Above setpoint")
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  print(p_read_counts)
  ggsave(file.path(plots_dir, "18_read_counts_by_region.png"),
         p_read_counts, width = 14, height = 10, dpi = 300)
  
  # Plot 7b: Read Proportions by Region (stacked bars showing proportions)
  read_proportions_data <- instability_metrics %>%
    mutate(
      total_reads = reads_below_setpoint + reads_at_setpoint + reads_above_setpoint,
      prop_below = reads_below_setpoint / total_reads,
      prop_at = reads_at_setpoint / total_reads,
      prop_above = reads_above_setpoint / total_reads
    ) %>%
    select(file_stem, analysis_range_name, group, setpoint_type,
           prop_below, prop_at, prop_above) %>%
    pivot_longer(cols = c(prop_below, prop_at, prop_above),
                 names_to = "region",
                 values_to = "proportion") %>%
    mutate(region = case_when(
      region == "prop_below" ~ "Below setpoint",
      region == "prop_at" ~ "At setpoint",
      region == "prop_above" ~ "Above setpoint"
    ),
    # Set factor levels: below first (bottom), at middle, above last (top)
    # This matches legend order
    region = factor(region, levels = c("Below setpoint", "At setpoint", "Above setpoint")))
  
  p_read_proportions <- ggplot(read_proportions_data,
                               aes(x = file_stem, y = proportion, fill = region)) +
    geom_col(position = "stack", alpha = 0.8) +
    facet_grid(setpoint_type + analysis_range_name ~ group, scales = "free_x",
           labeller = labeller(setpoint_type = setpoint_labels)) +
    labs(title = "Read Distribution by Region Relative to Setpoint (Proportions)",
         subtitle = "Proportion of reads below, at, and above setpoint",
         x = "Sample",
         y = "Proportion of Reads",
         fill = "Region") +
    scale_fill_manual(
      values = c("Below setpoint" = "steelblue", 
                 "At setpoint" = "grey70",
                 "Above setpoint" = "coral"),
      # Explicit breaks to ensure legend order
      breaks = c("Below setpoint", "At setpoint", "Above setpoint")
    ) +
    scale_y_continuous(labels = scales::percent_format()) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
  
  print(p_read_proportions)
  ggsave(file.path(plots_dir, "19_read_proportions_by_region.png"),
         p_read_proportions, width = 14, height = 10, dpi = 300)
  
  # Plot 8: Expansion vs Contraction Balance (scatter)
  balance_data <- instability_metrics %>%
    filter(!is.na(contraction_ratio), !is.na(expansion_ratio))
  
  if (nrow(balance_data) > 0) {
    p_balance_scatter <- ggplot(balance_data,
                                aes(x = contraction_ratio, y = expansion_ratio, 
                                    colour = group, shape = analysis_range_name)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
      geom_point(size = 4, alpha = 0.7) +
      facet_wrap(~ setpoint_type, nrow = 1) +
      labs(title = "Expansion vs Contraction Balance",
           subtitle = "Dashed line = balanced (expansion = contraction). Points above = more expansion; below = more contraction",
           x = "Contraction Ratio (reads below / reads at setpoint)",
           y = "Expansion Ratio (reads above / reads at setpoint)",
           colour = "Group",
           shape = "Analysis Range") +
      theme_bw() +
      theme(legend.position = "right")
    
    print(p_balance_scatter)
    ggsave(file.path(plots_dir, "20_expansion_contraction_balance.png"),
           p_balance_scatter, width = 14, height = 7, dpi = 300)
  }
  

  # Generate individual plots by analysis range
  message("\nGenerating individual plots by analysis range...")
  
  instability_by_range_dir <- file.path(plots_dir, "instability_by_range")
  if (!dir.exists(instability_by_range_dir)) {
    dir.create(instability_by_range_dir, recursive = TRUE)
  }
  
  for (my_range in unique(instability_metrics$analysis_range_name)) {
    range_data <- instability_metrics %>%
      filter(analysis_range_name == my_range)
    
    if (nrow(range_data) == 0) next
    
    # Plot: All instability indices for this range
    p_range <- ggplot(range_data,
                      aes(x = file_stem, y = instability_index, fill = setpoint_type)) +
      geom_col(position = "dodge", alpha = 0.8) +
      facet_wrap(~ group, scales = "free_x", nrow = 1) +
      scale_fill_manual(values = c("sample_relative" = "steelblue", 
                                   "control_relative" = "coral"),
                       labels = c("sample_relative" = "Relative to Sample",
                                 "control_relative" = "Relative to Control")) +
      labs(title = paste0("Instability Index - ", my_range),
           x = "Sample",
           y = "Instability Index",
           fill = "Setpoint Type") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "bottom")
    
    ggsave(file.path(instability_by_range_dir, 
                    paste0("instability_index-", my_range, ".png")),
           p_range, width = 12, height = 6, dpi = 300)
    
    # Plot: Read counts by region for this range
    read_counts_range <- range_data %>%
      select(file_stem, group, setpoint_type,
             reads_below_setpoint, reads_at_setpoint, reads_above_setpoint) %>%
      pivot_longer(cols = c(reads_below_setpoint, reads_at_setpoint, reads_above_setpoint),
                   names_to = "region",
                   values_to = "count") %>%
      mutate(region = case_when(
        region == "reads_below_setpoint" ~ "Below setpoint",
        region == "reads_at_setpoint" ~ "At setpoint",
        region == "reads_above_setpoint" ~ "Above setpoint"
      ),
      # Set factor levels: above first so it stacks on TOP
      region = factor(region, levels = c("Above setpoint", "At setpoint", "Below setpoint")))
    
    p_counts_range <- ggplot(read_counts_range,
                             aes(x = file_stem, y = count, fill = region)) +
      geom_col(position = "stack", alpha = 0.8) +
      facet_grid(setpoint_type ~ group, scales = "free",
                 labeller = labeller(setpoint_type = setpoint_labels)) +
      scale_fill_manual(values = c("Below setpoint" = "steelblue", 
                                   "At setpoint" = "grey70",
                                   "Above setpoint" = "coral")) +
      labs(title = paste0("Read Counts by Region - ", my_range),
           x = "Sample",
           y = "Number of Reads",
           fill = "Region") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "bottom")
    
    ggsave(file.path(instability_by_range_dir, 
                    paste0("read_counts-", my_range, ".png")),
           p_counts_range, width = 12, height = 8, dpi = 300)
  }
  
  message("  Individual range plots saved to: ", instability_by_range_dir)
  
  # Generate individual plots by sample
  message("\nGenerating individual plots by sample...")
  
  instability_by_sample_dir <- file.path(plots_dir, "instability_by_sample")
  if (!dir.exists(instability_by_sample_dir)) {
    dir.create(instability_by_sample_dir, recursive = TRUE)
  }
  
  for (my_sample in unique(instability_metrics$file_stem)) {
    sample_data <- instability_metrics %>%
      filter(file_stem == my_sample)
    
    if (nrow(sample_data) == 0) next
    
    # Create a multi-panel plot for this sample
    # Panel 1: Instability indices
    indices_data <- sample_data %>%
      select(analysis_range_name, setpoint_type, 
             instability_index, expansion_index, contraction_index) %>%
      pivot_longer(cols = c(instability_index, expansion_index, contraction_index),
                   names_to = "metric",
                   values_to = "value") %>%
      mutate(metric = gsub("_", " ", metric),
             metric = tools::toTitleCase(metric))
    
    p1 <- ggplot(indices_data,
                aes(x = analysis_range_name, y = value, fill = metric)) +
      geom_col(position = "dodge", alpha = 0.8) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      facet_wrap(~ setpoint_type, nrow = 2,
                 labeller = labeller(setpoint_type = setpoint_labels)) +
      labs(title = "Instability Indices",
           x = "Analysis Range",
           y = "Index Value",
           fill = "Metric") +
      theme_bw() +
      theme(legend.position = "bottom")
    
    # Panel 2: Read counts
    counts_data <- sample_data %>%
      select(analysis_range_name, setpoint_type,
             reads_below_setpoint, reads_at_setpoint, reads_above_setpoint) %>%
      pivot_longer(cols = c(reads_below_setpoint, reads_at_setpoint, reads_above_setpoint),
                   names_to = "region",
                   values_to = "count") %>%
      mutate(region = case_when(
        region == "reads_below_setpoint" ~ "Below",
        region == "reads_at_setpoint" ~ "At",
        region == "reads_above_setpoint" ~ "Above"
      ),
      # Set factor levels: above first so it stacks on TOP
      region = factor(region, levels = c("Above", "At", "Below")))
    
    p2 <- ggplot(counts_data,
                aes(x = analysis_range_name, y = count, fill = region)) +
      geom_col(position = "stack", alpha = 0.8) +
      facet_wrap(~ setpoint_type, nrow = 2,
                 labeller = labeller(setpoint_type = setpoint_labels)) +
      scale_fill_manual(values = c("Below" = "steelblue", 
                                   "At" = "grey70",
                                   "Above" = "coral")) +
      labs(title = "Read Distribution",
           x = "Analysis Range",
           y = "Number of Reads",
           fill = "Region") +
      theme_bw() +
      theme(legend.position = "bottom")
    
    # Combine plots
    p_combined <- cowplot::plot_grid(p1, p2, ncol = 2,
                                     labels = c("A", "B"),
                                     label_size = 14)
    
    # Add title
    title_text <- cowplot::ggdraw() + 
      cowplot::draw_label(paste0("Instability Metrics - ", my_sample),
                         fontface = "bold", size = 16)
    
    p_final <- cowplot::plot_grid(title_text, p_combined, ncol = 1,
                                  rel_heights = c(0.05, 1))
    
    ggsave(file.path(instability_by_sample_dir, 
                    paste0(my_sample, "-instability.png")),
           p_final, width = 14, height = 10, dpi = 300)
  }
  
  message("  Individual sample plots saved to: ", instability_by_sample_dir)

  message("\nInstability plots saved to: ", plots_dir)
}
```

---

# Export to Excel

Export all results to Excel workbook.

```{r export_excel, echo=FALSE}
# Prepare export list - comprehensive summary first
excel_export <- list(
  comprehensive_range_summary = comprehensive_range_summary,
  range_summary = range_summary,
  modal_peaks = modal_peaks,
  distribution_summary = distribution_summary
)

if (!is.null(group_controls)) {
  excel_export$group_controls <- group_controls
}

if (nrow(instability_metrics) > 0) {
  excel_export$instability_metrics <- instability_metrics
}

# Export to Excel
excel_path <- file.path(range_dir, "range_analysis.xlsx")
openxlsx::write.xlsx(excel_export, file = excel_path)

message("\nResults exported to Excel: ", excel_path)
message("  Sheets: ", paste(names(excel_export), collapse = ", "))
```

**Excel file contains:**

- **comprehensive_range_summary**: Complete table with all parameters, peaks, and controls (NEW!)
- **range_summary**: Analysis range definitions per sample
- **modal_peaks**: Detected peaks per sample and range
- **distribution_summary**: Full distribution statistics
- **group_controls**: Control setpoints per group (if enabled)
- **instability_metrics**: Instability metrics (if enabled)

---

# Summary

**Range analysis complete:**

- **`r length(unique(modal_peaks$file_stem))`** samples processed
- **`r nrow(modal_peaks)`** total peaks detected
- **`r nrow(distribution_summary)`** distribution summaries calculated

---

# Save outputs

```{r save_outputs, echo=FALSE}
# Compile module output
module6_output <- list(
  settings = settings_parsed,
  comprehensive_range_summary = comprehensive_range_summary,
  modal_peaks = modal_peaks,
  distribution_summary = distribution_summary,
  group_controls = group_controls,
  instability_metrics = instability_metrics
)

# Save complete module output
results_dir <- file.path(params$dir_out, "module_data")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

output_path <- file.path(results_dir, "06_range_analysis_results.RData")
save(module6_output, file = output_path)

message("Module 6 outputs saved to: ", output_path)
message("  comprehensive_range_summary included in module6_output")
```

**Module 6 Complete!**

Results saved to: `r output_path`

---

**Session Info:**

```{r session_info, echo=FALSE}
sessionInfo()
```
