---
title: "Duke pipeline"
subtitle: "Module 7: Repeat visualisation"
author: "Michael Flower"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
pbapply::pboptions(type = "none")
```

```{r set_tab_title, echo=FALSE, results='asis'}
cat('<script>document.title = "Duke | Module 7: Repeat visualisation";</script>')
```

---

# Overview

**Module 7: Repeat visualisation**

Provides visualisation of repeat length distributions through histograms, scatter plots, and density plots. These complement the statistical analyses in Module 6 by showing individual reads and distribution shapes visually.

**Input:** Repeat counts and cluster assignments (Modules 3 and 4), modal peaks and distribution summaries (Module 6)

**Metric overlays** (modal, mean, median) are controlled by `params$repeat_distribution_metrics` in the pipeline settings.

**Plot types:**

- `repeat_histogram = TRUE` — frequency histograms, full range and per analysis range
- `repeat_scatter = TRUE` — cohort summary box+jitter and per-sample horizontal scatter/violin plots
- `repeat_density = TRUE` — per-sample ggridges density plots, full range and per analysis range

**Outputs:**

- Cohort repeat length summary plot (`cohort_repeat_summary.png`)
- Repeat histograms per sample — full range (`histograms_by_sample/full/`) and per analysis range (`histograms_by_sample/{range}/`)
- Per-sample horizontal scatter plots — full range (`scatter_by_sample/full/`) and per analysis range (`scatter_by_sample/{range}/`)
- Per-sample ggridges density plots — all reads as a single density (`density_by_sample/full/`) and one ridge per analysis range (`density_by_sample/ranges/`)

---

# Setup

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(scales)
library(DT)
library(ggnewscale)
library(data.table)
library(ggridges)
library(openxlsx)
```

```{r load_functions, echo=FALSE}
# Source shared functions - works for both interactive and pipeline execution
if (file.exists("lib/load_all.R")) {
  source("lib/load_all.R")
} else if (file.exists("../lib/load_all.R")) {
  source("../lib/load_all.R")
} else {
  stop("Cannot find lib/load_all.R")
}

# Create output directory
plots_dir <- file.path(params$dir_out, file.path("07_repeat_visualisation", "plots"))
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}

message("\n=== Module 7: Repeat Distribution Visualisation ===")
message("Output directory: ", plots_dir)
```

```{r timing_init, echo=FALSE}
module_start_time <- proc.time()["elapsed"]
timings <- c()
# Set default for repeat_density if not supplied by older runner scripts
if (is.null(params$repeat_density)) params$repeat_density <- TRUE
```

---

# Load data

Load repeat counts and cluster assignments from Modules 3 and 4, and modal peaks and distribution summaries from Module 6.

```{r load_data, echo=FALSE, message=FALSE}
results_dir <- file.path(params$dir_out, "module_data")

# Load Module 3 slim (repeat counts)
module3_path <- file.path(results_dir, "03_repeat_detection_results.RData")
if (!file.exists(module3_path)) stop("Module 3 output not found: ", module3_path)
load(module3_path)
counts_slim <- module3_output$alignment_counts_slim

# Load Module 4 slim (cluster assignments)
module4_path <- file.path(results_dir, "04_allele_calling_results.RData")
if (!file.exists(module4_path)) stop("Module 4 output not found: ", module4_path)
load(module4_path)
clusters_slim <- module4_output$alignment_clustered_slim

# Join: repeat counts + cluster assignments
alignment_clustered <- counts_slim %>%
  dplyr::left_join(clusters_slim, by = c("file_stem", "qname"))

# Load Module 6 output (modal peaks and distribution summary)
module6_path <- file.path(results_dir, "06_range_analysis_results.RData")
if (!file.exists(module6_path)) stop("Module 6 output not found: ", module6_path)
load(module6_path)
modal_peaks        <- module6_output$modal_peaks
distribution_summary <- module6_output$distribution_summary

repeat_method <- params$repeat_count_method

message("Data loaded:")
message("  Samples: ",  length(unique(alignment_clustered$file_stem)))
message("  Total reads: ", format(nrow(alignment_clustered), big.mark = ","))
message("  Repeat method: ", repeat_method)
message("  Modal peaks: ", nrow(modal_peaks))
message("  Distribution summaries: ", nrow(distribution_summary))
```

---

# Prepare data

Build summary metric points (modal, mean, median) for overlay on per-sample plots, controlled by `params$repeat_distribution_metrics`. Analysis range boundaries are extracted from Module 6 modal peaks.

```{r prepare_data, echo=FALSE, message=FALSE}
# alignment_clustered is already a single dataframe with all reads from all samples
alignment_data <- alignment_clustered

# Get analysis ranges from modal_peaks
analysis_ranges <- modal_peaks %>%
  select(file_stem, analysis_range_name, range_min, range_max) %>%
  distinct()

# Note: alignment_data is used as-is from Module 4 (no pre-filtering by ranges)
# Reads are filtered dynamically in plotting loops by repeat length
# This allows reads to contribute to multiple overlapping ranges as intended

# Prepare summary metric points based on user selection
summary_points_list <- list()

# Get selected metrics
selected_metrics <- params$repeat_distribution_metrics

# Add modal if requested
if ("modal_length" %in% selected_metrics) {
  modal_points <- modal_peaks %>%
    filter(!is.na(modal_length), peak_rank == 1) %>%
    select(file_stem, analysis_range_name, modal_length, modal_frequency) %>%
    mutate(name = "modal",
           x = modal_length,
           y = modal_frequency) %>%
    select(file_stem, analysis_range_name, name, x, y)
  
  summary_points_list[["modal"]] <- modal_points
}

# Add mean if requested
if ("mean_length" %in% selected_metrics) {
  mean_points <- distribution_summary %>%
    select(file_stem, analysis_range_name, mean_length) %>%
    mutate(name = "mean",
           x = mean_length,
           y = 0) %>%  # Will be at bottom of histogram
    select(file_stem, analysis_range_name, name, x, y)
  
  summary_points_list[["mean"]] <- mean_points
}

# Add median if requested
if ("median_length" %in% selected_metrics) {
  median_points <- distribution_summary %>%
    select(file_stem, analysis_range_name, median_length) %>%
    mutate(name = "median",
           x = median_length,
           y = 0) %>%  # Will be at bottom of histogram
    select(file_stem, analysis_range_name, name, x, y)
  
  summary_points_list[["median"]] <- median_points
}

# Combine all selected points
if (length(summary_points_list) > 0) {
  summary_points <- bind_rows(summary_points_list) %>%
    rename(range = analysis_range_name)  # Rename for plotting function
} else {
  # Create empty dataframe with correct structure if no metrics selected
  summary_points <- data.frame(
    file_stem = character(),
    analysis_range_name = character(),
    name = character(),
    x = numeric(),
    y = numeric()
  )
}

# Prepare range boxes
range_boxes <- analysis_ranges %>%
  mutate(xmin = range_min,
         xmax = range_max) %>%
  rename(name = analysis_range_name)

message("\nData prepared:")
message("  Total reads: ", nrow(alignment_data))
message("  Analysis ranges: ", length(unique(analysis_ranges$analysis_range_name)))
message("  Metrics to display: ", paste(selected_metrics, collapse = ", "))
```

---

# Repeat histograms

Frequency histograms of repeat lengths. All plots are saved to disk — none are printed inline as the number of samples × ranges makes inline display impractical.

**Full histograms** (`histograms_by_sample/full/`) — all reads for each sample across the full repeat length range, with no range window overlay.

**Per-range histograms** (`histograms_by_sample/{range}/`) — reads filtered to each analysis range, with a shaded yellow range window. Metric overlays (modal, mean, median) shown as coloured points: red = modal, blue = mean, purple = median.

```{r setup_range_colors, echo=FALSE, message=FALSE}
# Define consistent colors for analysis ranges
range_names <- unique(analysis_ranges$analysis_range_name)
range_colors <- setNames(
  scales::hue_pal()(length(range_names)),
  range_names
)
```

```{r t_hist_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r histograms_by_sample, echo=FALSE, message=FALSE}
if (params$repeat_histogram) {

  hist_by_sample_dir <- file.path(plots_dir, "histograms_by_sample")
  # Consistent metric colour/shape convention (used by per-range histograms and scatters)
  metric_colors <- c("modal" = "red",     "mean" = "#4393C3", "median" = "purple")
  metric_shapes <- c("modal" = 16,        "mean" = 17,        "median" = 15)
  metric_labels <- c("modal" = "Modal",   "mean" = "Mean",    "median" = "Median")

  add_metric_points_hist <- function(p, pts) {
    # pts: data frame with columns x, y, name
    # Colour only (no shape distinction) — consistent with scatter convention
    if (is.null(pts) || nrow(pts) == 0) return(p)
    present <- unique(pts$name)
    p +
      geom_point(data = pts,
                 aes(x = x, y = y, colour = name),
                 shape = 16, size = 4, inherit.aes = FALSE) +
      ggrepel::geom_text_repel(
        data = pts,
        aes(x = x, y = y,
            label = as.character(round(x, 1)),
            colour = name),
        size = 3.5, inherit.aes = FALSE, show.legend = FALSE) +
      scale_colour_manual(values = metric_colors[present],
                          labels = metric_labels[present],
                          name = "Metric")
  }

  # Per-range scatter metric helper — colour only, numeric labels
  add_metric_points_scatter <- function(p, pts) {
    if (is.null(pts) || nrow(pts) == 0) return(p)
    present <- unique(pts$name)
    p +
      ggnewscale::new_scale_colour() +
      geom_point(data = pts,
                 aes(x = x, y = 1, colour = name),
                 shape = 16, size = 3, inherit.aes = FALSE) +
      ggrepel::geom_text_repel(
        data = pts,
        aes(x = x, y = 1,
            label = as.character(round(x, 1)),
            colour = name),
        size = 3.5, inherit.aes = FALSE, show.legend = FALSE,
        direction = "y", nudge_y = 0.2) +
      scale_colour_manual(values = metric_colors[present],
                          labels = metric_labels[present],
                          name = "Metric")
  }

  # Dense x-axis breaks helper
  dense_x <- scale_x_continuous(breaks = scales::pretty_breaks(n = 15))



  # Empty single-colour box for per-range plots (grey, no range colour)
  grey_box_color <- c(grey = "grey70")

  for (my_sample in unique(alignment_clustered$file_stem)) {

    sample_data_full <- alignment_clustered %>%
      filter(file_stem == my_sample)
    if (nrow(sample_data_full) == 0) next

    sample_points <- summary_points %>%
      filter(file_stem == my_sample)

    # ── 1. Full-range histogram: no range boxes, save to full/ ────────────────
    full_dir <- file.path(hist_by_sample_dir, "full")
    if (!dir.exists(full_dir)) dir.create(full_dir, recursive = TRUE)

    p_full <- plot_frequency_histogram(
      data          = sample_data_full,
      repeat_col    = repeat_method,
      binwidth      = params$repeat_histogram_binwidth,
      manual_points = NULL,          # Too many overlapping points on full plot
      manual_boxes  = NULL,          # No range windows on full plot
      range_colors  = NULL,
      title         = my_sample,
      base_text_size = 14
    )

    p_full <- p_full + dense_x +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
    ggsave(file.path(full_dir, paste0(my_sample, "-histogram-full.png")),
           p_full, width = 10, height = 6, dpi = params$plot_dpi)

    # ── 2. Per-range histograms: grey box, save to {range}/ ──────────────────
    sample_ranges <- analysis_ranges %>%
      filter(file_stem == my_sample) %>%
      pull(analysis_range_name) %>%
      unique()

    for (my_range in sample_ranges) {

      range_info <- analysis_ranges %>%
        filter(file_stem == my_sample, analysis_range_name == my_range)
      if (nrow(range_info) == 0) next

      range_min <- range_info$range_min[1]
      range_max <- range_info$range_max[1]

      sample_range_data <- alignment_data %>%
        filter(
          file_stem == my_sample,
          .data[[repeat_method]] >= range_min,
          .data[[repeat_method]] <= range_max
        )
      if (nrow(sample_range_data) == 0) next

      range_points <- sample_points %>%
        filter(range == my_range)

      # Box named with range name so legend is correct
      grey_range_box <- data.frame(
        name = my_range, xmin = range_min, xmax = range_max
      )
      grey_named <- setNames("#D6EAF8", my_range)  # subtle blue tint

      # Pass manual_points = NULL; add with consistent metric colours below
      p_range <- plot_frequency_histogram(
        data          = sample_range_data,
        repeat_col    = repeat_method,
        binwidth      = params$repeat_histogram_binwidth,
        manual_points = NULL,
        manual_boxes  = grey_range_box,
        range_colors  = grey_named,
        title         = my_sample,
        base_text_size = 14
      ) +
        labs(subtitle = my_range) +
        # Shaded range window (added after function to control visibility)
        annotate("rect", xmin = range_min, xmax = range_max,
                 ymin = 0, ymax = Inf,
                 fill = "#FFD700", colour = NA, alpha = 0.1) +
        dense_x +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

      # Add metric points with consistent red/blue/purple colours
      p_range <- add_metric_points_hist(p_range, range_points)
      # Sentence case axis and legend titles (override function defaults)
      p_range <- p_range +
        labs(fill = "Analysis range", x = "Repeat length", y = "Number of reads")

      range_dir <- file.path(hist_by_sample_dir, my_range)
      if (!dir.exists(range_dir)) dir.create(range_dir, recursive = TRUE)

      ggsave(file.path(range_dir, paste0(my_sample, "-histogram-", my_range, ".png")),
             p_range, width = 10, height = 6, dpi = params$plot_dpi)
    }
  }

  message("Histograms saved:")
  message("  Full:      histograms_by_sample/full/{sample}-histogram-full.png")
  message("  Per-range: histograms_by_sample/{range}/{sample}-histogram-{range}.png")
}
```

```{r t_hist_end, echo=FALSE}
timings["Histograms"] <- proc.time()["elapsed"] - .t
```

---

# Repeat scatter plots

## Cohort repeat length summary

Box+jitter plot showing modal repeat length per analysis range — one point per sample. Scales to hundreds of samples. This complements Module 6 plot 03 (modal peaks by range) and is included here for convenience within the visualisation module.

```{r t_scatter_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r cohort_repeat_summary, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
if (params$repeat_scatter) {

  # Use distribution_summary from Module 6 — one row per sample × range
  # Shows modal (primary peak) repeat length per range
  cohort_data <- distribution_summary %>%
    select(file_stem, analysis_range_name, modal_length, mean_length) %>%
    filter(!is.na(modal_length))

  p_cohort <- ggplot(cohort_data,
                     aes(x = analysis_range_name, y = modal_length)) +
    geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey50",
                 width = 0.6) +
    geom_jitter(colour = "grey40", width = 0.2, height = 0,
                alpha = 0.7, size = 2) +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 4,
                 colour = "red", alpha = 0.6) +
    labs(title = "Cohort repeat length summary by analysis range",
         subtitle = "Modal repeat length per sample per range. Box = IQR; points = samples; red diamond = mean.",
         x = "Analysis range",
         y = "Modal repeat length") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_cohort)

  ggsave(file.path(plots_dir, "cohort_repeat_summary.png"),
         p_cohort, width = 12, height = 6, dpi = params$plot_dpi)

  message("Cohort repeat length summary saved: cohort_repeat_summary.png")
}
```

## Horizontal scatter by sample

Horizontal scatter/violin plots for each sample. All plots saved to disk.

**Full scatter** (`scatter_by_sample/full/`) — all reads with a steelblue violin overlay, no range window.

**Per-range scatter** (`scatter_by_sample/{range}/`) — reads filtered to each analysis range, with a shaded yellow range window and coloured metric overlays (red = modal, blue = mean, purple = median).

```{r scatter_by_sample, echo=FALSE, message=FALSE}
if (params$repeat_scatter) {

  scatter_by_sample_dir <- file.path(plots_dir, "scatter_by_sample")
  grey_box_color <- c(grey = "grey70")  # kept for full scatter (no box used)

  # Dense x-axis for scatter
  dense_x_scatter <- scale_x_continuous(breaks = scales::pretty_breaks(n = 15),
                                        guide = guide_axis(angle = 45))

  # Transparent violin overlay on top of points (outline only)
  violin_overlay <- geom_violin(trim = TRUE, scale = "width",
                                fill = "steelblue", colour = "steelblue4",
                                alpha = 0.3, linewidth = 0.8)

  for (my_sample in unique(alignment_data$file_stem)) {

    sample_data <- alignment_data %>%
      filter(file_stem == my_sample)
    if (nrow(sample_data) == 0) next

    sample_points <- summary_points %>%
      filter(file_stem == my_sample)

    # ── 1. Full scatter: no range windows ─────────────────────────────────────
    full_dir <- file.path(scatter_by_sample_dir, "full")
    if (!dir.exists(full_dir)) dir.create(full_dir, recursive = TRUE)

    p_full <- plot_scatter_horizontal(
      data       = sample_data,
      x_var      = repeat_method,
      points     = NULL,         # Too many overlapping points on full plot
      boxes      = NULL,
      range_colors = NULL,
      title      = my_sample,
      base_text_size = 14
    ) +
      violin_overlay +
      dense_x_scatter

    ggsave(file.path(full_dir, paste0(my_sample, "-scatter.png")),
           p_full, width = 12, height = 4, dpi = params$plot_dpi)

    # ── 2. Per-range scatter: grey box, range in subtitle ─────────────────────
    sample_ranges <- analysis_ranges %>%
      filter(file_stem == my_sample) %>%
      pull(analysis_range_name) %>%
      unique()

    for (my_range in sample_ranges) {

      range_info <- analysis_ranges %>%
        filter(file_stem == my_sample, analysis_range_name == my_range)
      if (nrow(range_info) == 0) next

      range_min <- range_info$range_min[1]
      range_max <- range_info$range_max[1]

      sample_range_data <- alignment_data %>%
        filter(
          file_stem == my_sample,
          .data[[repeat_method]] >= range_min,
          .data[[repeat_method]] <= range_max
        )
      if (nrow(sample_range_data) == 0) next

      range_points <- sample_points %>%
        filter(range == my_range)

      # Name box with range name so legend shows correctly
      grey_range_box <- data.frame(
        name = my_range, xmin = range_min, xmax = range_max
      )
      grey_named_scatter <- setNames("#D6EAF8", my_range)  # subtle blue tint

      p_range <- plot_scatter_horizontal(
        data       = sample_range_data,
        x_var      = repeat_method,
        points     = NULL,           # Added manually below for numeric labels
        boxes      = grey_range_box,
        range_colors = grey_named_scatter,
        title      = my_sample,
        base_text_size = 14
      ) +
        labs(subtitle = my_range,
             fill = "Analysis range", x = "Repeat length") +
        # Shaded range window
        annotate("rect", xmin = range_min, xmax = range_max,
                 ymin = -Inf, ymax = Inf,
                 fill = "#FFD700", colour = NA, alpha = 0.1) +
        violin_overlay +
        dense_x_scatter

      # Add metric points with numeric-only labels
      p_range <- add_metric_points_scatter(p_range, range_points)

      range_dir <- file.path(scatter_by_sample_dir, my_range)
      if (!dir.exists(range_dir)) dir.create(range_dir, recursive = TRUE)

      ggsave(file.path(range_dir, paste0(my_sample, "-scatter-", my_range, ".png")),
             p_range, width = 12, height = 4, dpi = params$plot_dpi)
    }
  }

  message("Scatter by sample saved:")
  message("  Full:      scatter_by_sample/full/{sample}-scatter.png")
  message("  Per-range: scatter_by_sample/{range}/{sample}-scatter-{range}.png")
}
```

```{r t_scatter_end, echo=FALSE}
timings["Scatter plots"] <- proc.time()["elapsed"] - .t
```

---

# Density plots by sample

ggridges density plots per sample. All plots saved to disk.

**Full density** (`density_by_sample/full/`) — a single density ridge of all reads for each sample, regardless of analysis ranges. Useful for seeing the overall repeat length distribution shape.

**Ranges density** (`density_by_sample/ranges/`) — one ridge per analysis range for each sample, showing how the distribution shape differs across ranges. Reads are filtered to each range window; ranges may overlap so some reads appear in multiple ridges.

```{r t_density_start, echo=FALSE}
.t <- proc.time()["elapsed"]
```

```{r density_by_sample, echo=FALSE, message=FALSE}
if (params$repeat_density) {

  density_by_sample_dir <- file.path(plots_dir, "density_by_sample")

  # Use same range_colors defined in setup_range_colors chunk
  # for ridge fills consistent with Module 6 style

  for (my_sample in unique(alignment_data$file_stem)) {

    sample_data_full <- alignment_data %>%
      filter(file_stem == my_sample)
    if (nrow(sample_data_full) == 0) next

    sample_ranges <- analysis_ranges %>%
      filter(file_stem == my_sample) %>%
      pull(analysis_range_name) %>%
      unique()

    # ── 1. Full density: ALL reads for this sample, single ridge ─────────────
    full_dir <- file.path(density_by_sample_dir, "full")
    if (!dir.exists(full_dir)) dir.create(full_dir, recursive = TRUE)

    # Add group for fill colouring (matching Module 6 format)
    sample_group <- distribution_summary %>%
      filter(file_stem == my_sample) %>%
      pull(group) %>%
      unique() %>%
      .[1]

    p_density_full <- ggplot(sample_data_full,
                             aes(x = .data[[repeat_method]], y = "")) +
      geom_density_ridges(fill = "steelblue", colour = "black",
                          alpha = 0.7, scale = 1.2, rel_min_height = 0.01,
                          jittered_points = TRUE,
                          position = position_points_jitter(width = 0.5, height = 0.1),
                          point_shape = 16, point_size = 0.2,
                          point_alpha = 0.2, point_colour = "black") +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
      coord_cartesian(xlim = c(0, NA)) +
      labs(title    = my_sample,
           subtitle = "Repeat length density — all reads",
           x = "Repeat length",
           y = NULL) +
      theme_bw() +
      theme(axis.text.x  = element_text(angle = 45, hjust = 1),
            axis.text.y  = element_blank(),
            axis.ticks.y = element_blank())

    ggsave(file.path(full_dir, paste0(my_sample, "-density-full.png")),
           p_density_full, width = 12, height = 4, dpi = params$plot_dpi)

    # ── 2. Ranges density: one ridge per analysis range, coloured by range ────
    # Build tall dataset — each read tagged with every range it falls in
    density_ranges_data <- lapply(sample_ranges, function(rng) {
      ri <- analysis_ranges %>%
        filter(file_stem == my_sample, analysis_range_name == rng)
      if (nrow(ri) == 0) return(NULL)
      sample_data_full %>%
        filter(.data[[repeat_method]] >= ri$range_min[1],
               .data[[repeat_method]] <= ri$range_max[1]) %>%
        mutate(analysis_range_name = rng)
    }) %>%
      bind_rows() %>%
      mutate(analysis_range_name = factor(analysis_range_name,
                                          levels = rev(sort(unique(analysis_range_name)))))

    if (nrow(density_ranges_data) > 0) {
      n_ranges_sample <- length(sample_ranges)

      p_density_ranges <- ggplot(density_ranges_data,
                                 aes(x = .data[[repeat_method]],
                                     y = analysis_range_name)) +
        geom_density_ridges(fill = "steelblue", colour = "black",
                            alpha = 0.7, scale = 1.5, rel_min_height = 0.01,
                            jittered_points = TRUE,
                            position = position_points_jitter(width = 0.5, height = 0.1),
                            point_shape = 16, point_size = 0.2,
                            point_alpha = 0.2, point_colour = "black") +
        scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
        labs(title    = my_sample,
             subtitle = "Repeat length density by analysis range",
             x = "Repeat length",
             y = "Analysis range") +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

      ranges_dir <- file.path(density_by_sample_dir, "ranges")
      if (!dir.exists(ranges_dir)) dir.create(ranges_dir, recursive = TRUE)

      ggsave(file.path(ranges_dir, paste0(my_sample, "-density-ranges.png")),
             p_density_ranges,
             width  = 12,
             height = max(5, n_ranges_sample * 0.6 + 2),
             dpi = params$plot_dpi)
    }
  }

  message("Density plots by sample saved:")
  message("  Full:   density_by_sample/full/{sample}-density-full.png")
  message("  Ranges: density_by_sample/ranges/{sample}-density-ranges.png")
}
```

```{r t_density_end, echo=FALSE}
timings["Density plots"] <- proc.time()["elapsed"] - .t
```

---

# Module diagnostics

## Timing

```{r timing_table, echo=FALSE}
timing_df <- build_timing_table(timings)
DT::datatable(
  timing_df,
  caption = "Section timing — Module 7",
  options = list(pageLength = 15, scrollX = TRUE, dom = 't'),
  rownames = FALSE,
  class = 'cell-border stripe compact',
  colnames = c("Section", "Seconds", "Elapsed (mm:ss)", "Cumulative (mm:ss)")
) %>%
  DT::formatStyle("section", target = "row",
                  fontWeight = DT::styleEqual("Total", "bold"))
```

```{r diagnostics_export, echo=FALSE}
vis_excel_path <- file.path(params$dir_out, "07_repeat_visualisation", "repeat_visualisation.xlsx")
dir.create(dirname(vis_excel_path), showWarnings = FALSE, recursive = TRUE)
write_excel_function(
  named_list = list("timing" = timing_df),
  path = vis_excel_path,
  overwrite_sheet = TRUE
)
message(sprintf("Module 7 complete — total time: %s", format_elapsed(sum(timings))))
```

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```
