---
title: "Duke Pipeline - Module 1: Import and QC"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
```

# Overview

**Module 1: Import and QC**

This module handles:

1. File discovery and import (FASTQ, FASTA, BAM)
2. Duplicate read name handling
3. Optional downsampling
4. Quality control metrics
5. Adapter trimming

**Outputs:**

- Trimmed sequences ready for alignment
- QC metrics and visualisations
- Adapter trimming statistics


```{r load_packages, message=FALSE, warning=FALSE, echo=FALSE}
# Set working directory to where this Rmd is located
# This needs to happen BEFORE sourcing lib/load_all.R
# if (interactive()) {
#   setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
# } else {
#   # When rendering via rmarkdown::render()
#   # The working directory should already be set by duke_run.R
#   # But we can also try to detect the Rmd location
#   rmd_path <- knitr::current_input(dir = TRUE)
#   if (!is.null(rmd_path) && rmd_path != "") {
#     setwd(rmd_path)
#   }
# }

# Required packages
required_packages <- c("ShortRead", "Biostrings", "Rsamtools", "pbapply", 
                      "dplyr", "tidyr", "data.table", "stringr", "ggplot2",
                      "scales", "kableExtra", "DT", "openxlsx", "parallel",
                      "pbmcapply", "tibble")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r create_directories, echo=FALSE}
# Create output directories
temp_dir <- file.path(params$dir_out, "temp")
qc_dir <- file.path(params$dir_out, file.path("01_import_qc", "qc"))
plots_dir <- file.path(params$dir_out, file.path("01_import_qc", "plots"))  # NEW: plots directory

if (!dir.exists(temp_dir)) dir.create(temp_dir, recursive = TRUE)
if (!dir.exists(qc_dir)) dir.create(qc_dir, recursive = TRUE)
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)  # NEW
```

```{r load_shared_functions, echo=FALSE}
# Source shared functions
source("lib/load_all.R")
```

---

# Sequencing data

Discover and import sequencing files from the data directory.

```{r file_paths, echo=FALSE}
# List files
file_paths <- list.files(params$dir_data,
                         full.names = TRUE,
                         pattern = paste(params$import_patterns, collapse = "|"),
                         recursive = params$import_recursive)

# Extract file names
file_names <- basename(file_paths)

# Extract file stems, first removing gz double extensions
file_stems <- gsub("\\.gz$", "", file_names)
file_stems <- tools::file_path_sans_ext(file_stems)

# Extract formats
file_formats <- ifelse(grepl("\\.gz$", file_names), 
                       paste0(tools::file_ext(gsub("\\.gz$", "", file_names)), ".gz"), 
                       tools::file_ext(file_names))

# Vector of file names and stems
file_name_vector <- setNames(file_names, file_names)
file_stem_vector <- setNames(file_stems, file_stems)

# Table of file details
file_details <- data.frame(file_name = file_names,
                           file_stem = file_stems,
                           file_format = file_formats,
                           file_path = file_paths) %>%
  dplyr::mutate(paired = replace_na(grepl(params$r1_pattern, file_name) | grepl(params$r2_pattern, file_name), FALSE),
                paired_read_direction = case_when(grepl(params$r1_pattern, file_name) ~ "R1",
                                                  grepl(params$r2_pattern, file_name) ~ "R2",
                                                  TRUE ~ NA),
                include = case_when(paired & paired_read_direction == params$select_one_of_pair ~ TRUE,
                                    paired ~ FALSE,
                                    TRUE ~ TRUE),
                sample_stem = ifelse(paired, gsub("_R[12]_.*", "", file_name), file_stem)) %>%
  dplyr::relocate(c(file_format, paired, paired_read_direction, include, sample_stem), 
                  .after = "file_name")
```

## File format summary

```{r file_format_table, echo=FALSE}
data.frame(table(file_formats)) %>%
  dplyr::rename(file_format = file_formats,
                count = Freq) %>%
  kbl(caption = "File format") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Paired-end detection

```{r paired_end_summary, echo=FALSE}
file_details %>%
  distinct(sample_stem, paired) %>%
  group_by(paired) %>%
  dplyr::summarise(count = n(), .groups = "drop") %>%
  kbl(caption = "Count single and paired end sequencing files") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## File details

```{r file_details_table, echo=FALSE}
DT::datatable(
  file_details,
  caption = "File summary",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(file_details), fontSize = '85%')


# Export table
write_excel_function(named_list = list("file_details" = file_details),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

```{r select_one_of_pair, echo=FALSE}
if (!is.na(params$select_one_of_pair)) {
  # Files to keep
  file_stem_keep <- file_details %>%
    dplyr::filter(include)
  file_stem_keep <- file_stem_keep$file_stem
  
  # Filter file stem vector
  file_stem_vector <- file_stem_vector[file_stem_vector %in% file_stem_keep]
}
```

```{r sequencing_import_functions, echo=FALSE}
# Function to import fastq including quality data
fastq_import <- function(path) {
  readfastq_file <- readFastq(path)
  read_names <- as.character(ShortRead::id(readfastq_file))
  # Strip tags from read names by keeping only the first field before the first space
  clean_read_names <- sapply(strsplit(read_names, " "), `[`, 1)
  quality_scaled_dnastringset <- 
      QualityScaledDNAStringSet(sread(readfastq_file), as(quality(readfastq_file), "PhredQuality"))
  names(quality_scaled_dnastringset) <- clean_read_names
  return(quality_scaled_dnastringset)
}

# Function to import fasta
fasta_import <- function(path) {
  readDNAStringSet(path, format = "fasta")
}

# Function to import bam
bam_import <- function(path) {
  bam_file <- BamFile(path)
  bam_data <- scanBam(bam_file)
  bam_sequences <- bam_data[[1]]$seq
  bam_qualities <- bam_data[[1]]$qual
  bam_read_names <- bam_data[[1]]$qname
  quality_scaled_dnastringset <-
    QualityScaledDNAStringSet(bam_sequences, bam_qualities)
  names(quality_scaled_dnastringset) <- bam_read_names
  return(quality_scaled_dnastringset)
}

# Function to import sequencing files
import_sequencing <- function(path, format) {
  if (format == "fastq" || format == "fastq.gz") {
    return(fastq_import(path))
  } else if (format == "fasta" || format == "fa") {
    return(fasta_import(path))
  } else if (format == "bam") {
    return(bam_import(path))
  } else {
    stop("Unsupported file format")
  }
}
```

```{r sequencing_import, echo=FALSE}
# Import sequencing files
seq <- apply_fn(file_stem_vector, function(file_stem, file_details) {
  file_path <- file_details$file_path[match(file_stem, file_details$file_stem)]
  file_format <- file_details$file_format[match(file_stem, file_details$file_stem)]
  import_sequencing(file_path, file_format)
}, file_details = file_details)

# Extract results if using parallel processing
seq <- extract_apply_fn_result(seq)
```

---

# Check for duplicate read names

Some sequencing technologies may contain duplicate read names. This section identifies duplicates and retains only the longest instance of each duplicated read.

```{r select_unique_readnames, echo=FALSE, message=FALSE, results='hide'}
# Find duplicate read names and select the longest
my_path <- file.path(temp_dir, "01_unique_readname_result.RData")
if (file.exists(my_path) && params$resume) {
  load(my_path)
} else {
  unique_readname_result <- apply_fn(seq, select_longest_duplicate_readname)
  unique_readname_result <- extract_apply_fn_result(unique_readname_result)
  save(unique_readname_result, file = my_path)
}

# Make new sequencing file
seq_unique <- lapply(unique_readname_result, "[[", "dna_stringset")
```

```{r count_read_names, echo=FALSE}
# Read name counts
read_name_count <- lapply(file_stem_vector, function(file_stem, unique_readname_result) {
  unique_readname_result[[file_stem]][["read_name_count"]] %>%
    dplyr::mutate(file_stem = file_stem) %>%
    relocate(file_stem)
}, unique_readname_result = unique_readname_result)
read_name_count <- rbindlist(read_name_count)

# Summarise read name count
read_name_count_summary <- read_name_count %>%
  group_by(file_stem, count) %>%
  dplyr::summarise(n_readnames = n(), .groups = "drop")

# Total number of read names
total_readnames <- read_name_count %>%
  group_by(file_stem) %>%
  dplyr::summarise(total_readnames = n(), .groups = "drop")

# Add total read name count
read_name_count_summary <- read_name_count_summary %>%
  left_join(total_readnames, by = join_by(file_stem)) %>%
  dplyr::mutate(proportion = n_readnames / total_readnames)
```

## Duplicate read names

```{r plot_read_name_count, fig.height=8, fig.width=12, echo=FALSE}
p_duplicates <- ggplot(read_name_count_summary, aes(x = file_stem, y = proportion)) +
  geom_bar(aes(fill = as.factor(count)),
           stat = "identity", position = "stack", colour = "black") +
  scale_x_discrete(labels = function(x) stringr::str_wrap(gsub("(.{50})", "\\1\n", x), width = 50),
                   guide = guide_axis(angle = 70)) +
  scale_y_continuous(breaks = pretty_breaks()) +
  labs(x = "File stem",
       y = "Proportion of reads",
       fill = "Read name\ncount") +
  theme_bw()

print(p_duplicates)

# Export plot
# Calculate width with cap at 50 inches (ggplot2 limit)
n_samples <- length(unique(read_name_count_summary$file_stem))
plot_width <- min(50, max(10, n_samples * 0.8))

ggsave(
  filename = file.path(plots_dir, "read_name_duplicates.png"),
  plot = p_duplicates,
  width = plot_width,
  height = 8,
  dpi = 300,
  bg = "white",
  limitsize = FALSE
)
```

---

# Read count comparison

```{r count_original_vs_unique_readnames, echo=FALSE}
# Compare read count
readcount_unique_readnames <- 
  compare_read_count(dna_stringset1 = seq,
                     dna_stringset2 = seq_unique,
                     name1 = "original",
                     name2 = "unique_read_name")

# Remove previous sequencing file
if (params$remove_intermediate) rm(seq)
```

## Read count comparison

```{r plot_original_vs_unique_readnames, echo=FALSE, fig.height=8, fig.width=12}
# Long format
readcount_unique_readnames_long <- readcount_unique_readnames %>%
  pivot_longer(!file_stem, names_to = "dataset", values_to = "read_count")

# Plot
p_unique_reads <- ggplot(readcount_unique_readnames_long, aes(x = file_stem, y = read_count)) +
  facet_grid(rows = vars(dataset),
             labeller = as_labeller(c("original" = "Original",
                                      "unique_read_name" = "Unique read\nnames"))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = scales::comma(read_count)),
            angle = 70, vjust = -0.25, hjust = -0.25, size = 2) +
  scale_x_discrete(labels = function(x) stringr::str_wrap(gsub("(.{50})", "\\1\n", x), width = 50),
                   guide = guide_axis(angle = 70)) +
  scale_y_continuous(breaks = pretty_breaks(),
                     expand = expansion(mult = c(NA, 0.3))) +
  labs(x = "File stem",
       y = "Read count") +
  theme_bw()

print(p_unique_reads)

# Export plot
# Calculate width with cap at 50 inches
n_samples <- length(unique(readcount_unique_readnames_long$file_stem))
plot_width <- min(50, max(10, n_samples * 0.8))

ggsave(
  filename = file.path(plots_dir, "unique_read_names.png"),
  plot = p_unique_reads,
  width = plot_width,
  height = 8,
  dpi = 300,
  bg = "white",
  limitsize = FALSE
)
```

```{r table_original_vs_unique_readnames, echo=FALSE}
# Display table
DT::datatable(
  readcount_unique_readnames,
  caption = "Count unique read names",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(readcount_unique_readnames), fontSize = '85%')

# Export table
write_excel_function(named_list = list("readcount_unique_readnames" = readcount_unique_readnames),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

---

# Sample reads

Optionally downsample reads for faster processing and visualisation.

```{r downsample, echo=FALSE}
# Downsample
if (!is.na(params$downsample)) {
  seq_sample <- apply_fn(seq_unique, 
                         downsample, 
                         sample_size = params$downsample)
  seq_sample <- extract_apply_fn_result(seq_sample)
} else {
  seq_sample <- seq_unique
}
```

```{r downsample_read_counts, echo=FALSE}
# Compare read count
read_count_sample <- 
  compare_read_count(dna_stringset1 = seq_unique,
                     dna_stringset2 = seq_sample,
                     name1 = "original",
                     name2 = "sample")
```

## Downsampling reads

```{r plot_downsample_read_counts, echo=FALSE, fig.height=8, fig.width=12}
# Long format
read_count_sample_long <- read_count_sample %>%
  pivot_longer(!file_stem, names_to = "dataset", values_to = "read_count")

# Plot
p_filtered_counts <- ggplot(read_count_sample_long, aes(x = file_stem, y = read_count)) +
  facet_grid(rows = vars(dataset),
             labeller = as_labeller(c("original" = "Original",
                                      "sample" = "Sampled reads")),
             scales = "free_y") +
  geom_bar(stat = "identity") +
  geom_text(aes(label = scales::comma(read_count)),
            angle = 70, vjust = -0.25, hjust = -0.25, size = 2) +
  scale_x_discrete(labels = function(x) stringr::str_wrap(gsub("(.{50})", "\\1\n", x), width = 50),
                   guide = guide_axis(angle = 70)) +
  scale_y_continuous(breaks = pretty_breaks(),
                     expand = expansion(mult = c(NA, 0.3))) +
  labs(x = "File stem",
       y = "Read count") +
  theme_bw()

print(p_filtered_counts)

# Export plot
# Calculate width with cap at 50 inches
n_samples <- length(unique(read_count_sample_long$file_stem))
plot_width <- min(50, max(10, n_samples * 0.8))

ggsave(
  filename = file.path(plots_dir, "read_counts_filtered.png"),
  plot = p_filtered_counts,
  width = plot_width,
  height = 8,
  dpi = 300,
  bg = "white",
  limitsize = FALSE
)
```

```{r table_downsample_read_counts, echo=FALSE}
# Display table
DT::datatable(
  read_count_sample,
  caption = "Sampled read count",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(read_count_sample), fontSize = '85%')

# Export table
write_excel_function(named_list = list("read_count_sample" = read_count_sample),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)

# Remove previous sequencing file
if (params$remove_intermediate) rm(seq_unique)
```

---

# Quality

Calculate read quality metrics and length distributions.

```{r qc, echo=FALSE}
# Quality and length of each read
sequencing_qc <- apply_fn(file_stem_vector, function(my_file_stem, seq_sample) {
  
  # Extract sequencing file
  sequencing_file <- seq_sample[[my_file_stem]]
  
  # Check if quality scores exist
  if(inherits(sequencing_file, "QualityScaledDNAStringSet")) {
  
    # Extract qualities
    quals <- quality(sequencing_file)
    
    # Sum -log10 pvalues for each read
    read_scores <- alphabetScore(quals)
    
  } else {
    
    # NA if no quality scores
    read_scores <- rep(NA, length(sequencing_file))
    
  }
  
  # Read lengths
  read_lengths <- width(sequencing_file)
  
  # Table
  qc_table <- data.frame(file_stem = rep(my_file_stem, length(sequencing_file)),
                         read_name = names(sequencing_file),
                         phred_sum = read_scores,
                         length = read_lengths) %>%
    dplyr::mutate(read_phred = phred_sum / length)
  
  # Output
  return(qc_table)
  
}, seq_sample = seq_sample)

# Extract results if using parallel processing
sequencing_qc <- extract_apply_fn_result(sequencing_qc)
  
# Summarise QC
sequencing_qc_summary <- 
  rbindlist(sequencing_qc) %>%
    group_by(file_stem) %>%
    dplyr::summarise(read_count = n(),
                     mean_length = mean(length, na.rm = TRUE),
                     median_length = median(length, na.rm = TRUE),
                     sd_length = sd(length, na.rm = TRUE),
                     mean_phred = mean(read_phred, na.rm = TRUE),
                     median_phred = median(read_phred, na.rm = TRUE),
                     sd_phred = sd(read_phred, na.rm = TRUE),
                     .groups = "drop")
```

## QC metrics

```{r qc_plot, echo=FALSE, message=FALSE}
# Prepare data
filtered_read_summary <- sequencing_qc_summary %>%
  dplyr::mutate(mean_phred_plot = ifelse(is.na(mean_phred), 0, mean_phred),
                phred_available = ifelse(!is.na(mean_phred), "Phred available", "Phred unavailable"))

# Combined plot
p_qc_scatter <- filtered_read_summary %>%
  ggplot(aes(x = mean_length, y = mean_phred_plot)) +
  facet_wrap(vars(phred_available), scales = "free") +
  geom_point(aes(size = read_count), 
             fill = "skyblue", colour = "black", shape = 21, alpha = 0.75) +
  geom_smooth() +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(breaks = pretty_breaks()) +
  scale_fill_viridis_c(option = "D", direction = 1) +
  labs(x = "Mean read length (bp)",
       y = "Mean phred score",
       size = "Read count") +
  theme_minimal()

print(p_qc_scatter)

# Export combined plot
ggsave(
  filename = file.path(plots_dir, "quality_vs_length.png"),
  plot = p_qc_scatter,
  width = 12,
  height = 6,
  dpi = 300,
  bg = "white"
)

# Create subdirectory for per-sample plots
quality_by_sample_dir <- file.path(plots_dir, "quality_vs_length_by_sample")
dir.create(quality_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

# Individual per-sample plots
for (sample in unique(filtered_read_summary$file_stem)) {
  sample_data <- filtered_read_summary %>% filter(file_stem == sample)
  
  p_sample <- ggplot(sample_data, aes(x = mean_length, y = mean_phred_plot)) +
    geom_point(aes(size = read_count), 
               fill = "skyblue", colour = "black", shape = 21, alpha = 0.75) +
    geom_smooth(se = FALSE) +
    scale_x_continuous(breaks = pretty_breaks()) +
    scale_y_continuous(breaks = pretty_breaks()) +
    labs(title = paste("Read quality vs length -", sample),
         x = "Mean read length (bp)",
         y = "Mean phred score",
         size = "Read count") +
    theme_minimal()
  
  ggsave(
    filename = file.path(quality_by_sample_dir, paste0(sample, "-quality_vs_length.png")),
    plot = p_sample,
    width = 8,
    height = 6,
    dpi = 300,
    bg = "white"
  )
}
```

```{r qc_table, echo=FALSE}
# Display table
DT::datatable(
  sequencing_qc_summary,
  caption = "Sequencing QC",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(sequencing_qc_summary), fontSize = '85%')

# Export table
write_excel_function(named_list = list("sequencing_qc_summary" = sequencing_qc_summary),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

---

# Trim

Trim adapter sequences from reads.

```{r trim_function, echo=FALSE}
trim_reads <- function(dna_stringset, trim_patterns, max_mismatch, with_indels) {
  
  # Vector of trim sequence names
  trim_pattern_names <- setNames(names(trim_patterns), 
                                 names(trim_patterns))
  
  # Log locations of trim sequences
  trim_locations <- lapply(trim_pattern_names, function(trim_pattern_name,
                                                        trim_patterns,
                                                        dna_stringset,
                                                        max_mismatch,
                                                        with_indels) {
    # Extract trim sequence
    trim_pattern <- trim_patterns[[trim_pattern_name]]
    
    # Trim coordinates
    trim_location <- trimLRPatterns(Lpattern = trim_pattern,
                                    Rpattern = trim_pattern,
                                    subject = dna_stringset,
                                    max.Lmismatch = max_mismatch,
                                    max.Rmismatch = max_mismatch,
                                    with.Lindels = with_indels,
                                    with.Rindels = with_indels,
                                    ranges = TRUE)
    
    # Convert to dataframe
    trim_location <- as.data.frame(trim_location) %>%
      dplyr::mutate(trim_pattern_name = trim_pattern_name,
                    read_name = names(dna_stringset),
                    read_length = width(dna_stringset)) %>%
      relocate(c(trim_pattern_name, read_name, read_length))
    
    # Output
    return(trim_location)
    
  },
  trim_patterns = trim_patterns,
  dna_stringset = dna_stringset,
  max_mismatch = max_mismatch,
  with_indels = with_indels)
  
  # Rbind
  trim_locations <- rbindlist(trim_locations) %>%
    dplyr::mutate(trim_length = read_length - width,
                  trim = width < read_length)
  
  # Trim
  dna_stringset_trimmed <- dna_stringset
  for (trim_pattern in trim_patterns) {
    dna_stringset_trimmed <- trimLRPatterns(Lpattern = trim_pattern,
                                            Rpattern = trim_pattern,
                                            subject = dna_stringset_trimmed,
                                            max.Lmismatch = max_mismatch,
                                            max.Rmismatch = max_mismatch,
                                            with.Lindels = with_indels,
                                            with.Rindels = with_indels)
  }
  
  # Output
  return(list(trim_locations = trim_locations,
              dna_stringset_trimmed = dna_stringset_trimmed))
}
```

```{r trim_sequences, echo=FALSE}
# Import trim sequences
trim_sequences <- read.csv(params$path_trim_patterns, fileEncoding = "UTF-8-BOM")
trim_sequences <- setNames(trim_sequences$adapter_sequence,
                           trim_sequences$adapter_name)

# Reverse complement the trim sequences
trim_sequences_rc <- as.character(reverseComplement(DNAStringSet(trim_sequences)))
names(trim_sequences_rc) <- paste0(names(trim_sequences_rc), "_rc")

# Combine sense and reverse complement trim sequences
combined_trim_sequences <- c(trim_sequences, trim_sequences_rc)

# Display table
trim_df <- enframe(combined_trim_sequences, name = "adapter_name", value = "adapter_sequence")
kbl(trim_df, caption = "Trim sequences") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# Export table
write_excel_function(named_list = list("trim_sequences" = trim_df),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

```{r trim, echo=FALSE, message=FALSE, results='hide'}
# Run trim function
my_path <- file.path(temp_dir, "02_trim_result.RData")
if (file.exists(my_path) && params$resume) {
  load(my_path)
} else {
  trim_result <- apply_fn(seq_sample, trim_reads,
                          trim_patterns = combined_trim_sequences,
                          max_mismatch = params$trim_max_mismatch,
                          with_indels = params$trim_with_indels)
  trim_result <- extract_apply_fn_result(trim_result)
  save(trim_result, file = my_path)
}

# Extract trimmed sequencing files
seq_trim <- lapply(trim_result, "[[", "dna_stringset_trimmed")

# Remove previous sequencing file
if (params$remove_intermediate) rm(seq_sample)
```

## Proportion of reads trimmed

```{r trim_proportion, echo=FALSE}
# Extract trim locations
trim_locations <- lapply(file_stem_vector, function(my_file_stem, trim_result) {
  trim_result[[my_file_stem]]$trim_locations %>%
    dplyr::mutate(file_stem = my_file_stem) %>%
    relocate(file_stem)
}, trim_result = trim_result)

# Rbind
trim_locations <- rbindlist(trim_locations)

# Count trimmed reads by trim sequence for each file
count_trimmed <- trim_locations %>%
  group_by(file_stem, trim_pattern_name) %>%
  dplyr::summarise(read_count = sum(trim), .groups = "drop")

# Count untrimmed reads
count_untrimmed <- trim_locations %>%
  group_by(file_stem, read_name) %>%
  dplyr::summarise(all_untrimmed = all(!trim), .groups = "drop") %>%
  group_by(file_stem) %>%
  dplyr::summarise(read_count = sum(all_untrimmed), .groups = "drop")

# Merge trimmed and untrimmed read counts
count_trimmed <- plyr::rbind.fill(count_trimmed, count_untrimmed) %>%
  arrange(file_stem, trim_pattern_name) %>%
  dplyr::filter(read_count > 0) %>%
  dplyr::mutate(trim_pattern_name = ifelse(is.na(trim_pattern_name), "untrimmed", trim_pattern_name))

# Count total reads for each file
total_reads <- trim_locations %>%
  distinct(file_stem, read_name) %>%
  group_by(file_stem) %>%
  dplyr::summarise(total_reads = n(), .groups = "drop")

# Calculate proportion of reads trimmed
count_trimmed <- count_trimmed %>%
  left_join(total_reads, by = join_by(file_stem)) %>%
  dplyr::mutate(proportion = read_count / total_reads)

# Calculate mean proportion for each trim pattern to reorder the x-axis
mean_trim <- count_trimmed %>%
  group_by(trim_pattern_name) %>%
  dplyr::summarise(mean_proportion = mean(proportion, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(desc(mean_proportion))

# Reorder trim_pattern_name by mean count
count_trimmed <- count_trimmed %>%
  dplyr::mutate(trim_pattern_name = factor(trim_pattern_name, levels = mean_trim$trim_pattern_name))
```

```{r trim_proportion_plot, echo=FALSE}
# Calculate appropriate number of decimals based on data range
proportion_range <- max(count_trimmed$proportion) - min(count_trimmed$proportion)
n_decimals <- if (proportion_range < 0.01) {
  4  # Very small differences
} else if (proportion_range < 0.1) {
  3  # Small differences
} else if (proportion_range < 1) {
  2  # Moderate differences
} else {
  1  # Large differences
}

p_trim_prop <- ggplot(count_trimmed, aes(x = trim_pattern_name, y = proportion)) +
  stat_summary(fun = mean, geom = "bar", fill = "lightblue", colour = "black") +
  geom_jitter(height = 0, width = 0.1, alpha = 0.5, size = 3) +
  stat_summary(fun = mean, geom = "text", show.legend = FALSE, colour = "black", size = 4,
               vjust = -0.5,
               aes(label = after_stat(sprintf(paste0("%.", n_decimals, "f"), y)))) +
  scale_x_discrete(guide = guide_axis(angle = 75)) +
  scale_y_continuous(breaks = pretty_breaks(),
                     expand = expansion(mult = c(0.2))) +
  labs(x = "Trim sequence",
       y = "Proportion of reads") +
  theme_minimal()

print(p_trim_prop)

# Export plot
ggsave(
  filename = file.path(plots_dir, "adapter_trimming_proportion.png"),
  plot = p_trim_prop,
  width = 10,
  height = 6,
  dpi = 300,
  bg = "white"
)
```

## Trimming quality summary

Trimming occurs from sequence **ends** using `Biostrings::trimLRPatterns`:

- **left** = 5' end (start of read)
- **right** = 3' end (end of read)  
- **both** = adapters found at both ends

Compare trim length statistics to `adapter_length`:

- **Similar to adapter length** = good match (adapter fully present)
- **Shorter than adapter** = partial adapter or early termination
- **Longer than adapter** = adapter + extra sequence removed

### Summary by adapter

```{r trim_summary_by_adapter, echo=FALSE}
# Comprehensive summary by adapter
trim_by_adapter <- trim_locations %>%
  filter(trim == TRUE) %>%
  mutate(
    trim_side = case_when(
      start == 1 & end < read_length ~ "left",
      start > 1 & end == read_length ~ "right",
      start == 1 & end == read_length ~ "both",
      TRUE ~ "internal"
    )
  ) %>%
  group_by(trim_pattern_name, trim_side) %>%
  summarise(
    n_trimmed = n(),
    median_trim_length = median(trim_length, na.rm = TRUE),
    mean_trim_length = mean(trim_length, na.rm = TRUE),
    sd_trim_length = sd(trim_length, na.rm = TRUE),
    min_trim_length = min(trim_length, na.rm = TRUE),
    max_trim_length = max(trim_length, na.rm = TRUE),
    .groups = "drop"
  )

# Add adapter info and calculate proportion
total_reads <- length(unique(trim_locations$read_name))

trim_by_adapter <- trim_by_adapter %>%
  left_join(
    data.frame(
      trim_pattern_name = names(combined_trim_sequences),
      adapter_sequence = as.character(combined_trim_sequences),
      adapter_length = nchar(combined_trim_sequences)
    ),
    by = "trim_pattern_name"
  ) %>%
  mutate(
    proportion = n_trimmed / total_reads,
    total_reads = total_reads
  ) %>%
  relocate(adapter_sequence, adapter_length, trim_side, .after = trim_pattern_name) %>%
  relocate(n_trimmed, total_reads, proportion, .after = trim_side) %>%
  arrange(desc(n_trimmed))

# Display table
kbl(trim_by_adapter, 
    caption = "Trimming summary by adapter (across all samples)",
    digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                font_size = 10)
```

### Summary by sample

```{r trim_summary_by_sample, echo=FALSE}
# Comprehensive summary by sample
trim_by_sample <- trim_locations %>%
  filter(trim == TRUE) %>%
  mutate(
    trim_side = case_when(
      start == 1 & end < read_length ~ "left",
      start > 1 & end == read_length ~ "right",
      start == 1 & end == read_length ~ "both",
      TRUE ~ "internal"
    )
  ) %>%
  group_by(file_stem, trim_pattern_name, trim_side) %>%
  summarise(
    n_trimmed = n(),
    median_trim_length = median(trim_length, na.rm = TRUE),
    mean_trim_length = mean(trim_length, na.rm = TRUE),
    sd_trim_length = sd(trim_length, na.rm = TRUE),
    min_trim_length = min(trim_length, na.rm = TRUE),
    max_trim_length = max(trim_length, na.rm = TRUE),
    .groups = "drop"
  )

# Add adapter info and calculate proportion per sample
sample_totals <- trim_locations %>%
  group_by(file_stem) %>%
  summarise(total_reads = n_distinct(read_name), .groups = "drop")

trim_by_sample <- trim_by_sample %>%
  left_join(
    data.frame(
      trim_pattern_name = names(combined_trim_sequences),
      adapter_sequence = as.character(combined_trim_sequences),
      adapter_length = nchar(combined_trim_sequences)
    ),
    by = "trim_pattern_name"
  ) %>%
  left_join(sample_totals, by = "file_stem") %>%
  mutate(proportion = n_trimmed / total_reads) %>%
  relocate(adapter_sequence, adapter_length, trim_side, .after = trim_pattern_name) %>%
  relocate(n_trimmed, total_reads, proportion, .after = trim_side) %>%
  arrange(file_stem, desc(n_trimmed))

# Display table
DT::datatable(
  trim_by_sample,
  caption = "Trimming summary by sample (per adapter)",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(trim_by_sample), fontSize = '85%') %>%
  DT::formatRound(columns = c('median_trim_length', 'mean_trim_length', 
                               'sd_trim_length', 'proportion'), digits = 2)

# Export both tables
write_excel_function(
  named_list = list(
    "trim_by_adapter" = trim_by_adapter,
    "trim_by_sample" = trim_by_sample
  ),
  path = file.path(qc_dir, "qc.xlsx"),
  overwrite_sheet = TRUE
)
```

### Trim length distribution

```{r trim_length_plot, echo=FALSE}
# Use trim_by_sample for plotting
if (nrow(trim_by_sample) > 0) {
  # Calculate mean trim length for each adapter (for ordering x-axis)
  adapter_order <- trim_by_sample %>%
    group_by(trim_pattern_name) %>%
    summarise(mean_overall = mean(mean_trim_length, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_overall))
  
  # Prepare data for plotting
  plot_data <- trim_by_sample %>%
    mutate(trim_pattern_name = factor(trim_pattern_name, levels = adapter_order$trim_pattern_name))
  
  # Plot
  p_trim_length <- ggplot(plot_data, aes(x = trim_pattern_name, y = mean_trim_length)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.3) +
    geom_jitter(height = 0, width = 0.1, alpha = 0.5, size = 3) +
    stat_summary(fun = mean, geom = "point", shape = 18, colour = "red", size = 4) +
    stat_summary(fun = mean, geom = "text", colour = "red", size = 3.5,
                 angle = 45, vjust = -0.5, hjust = -0.25,
                 aes(label = after_stat(round(y, 1)))) +
    geom_point(aes(y = adapter_length), colour = "blue", shape = 3, size = 3) +
    scale_x_discrete(guide = guide_axis(angle = 75)) +
    scale_y_continuous(breaks = pretty_breaks(),
                       expand = expansion(mult = c(0.05, 0.2))) +
    labs(title = "Trim length distribution by adapter",
         subtitle = "Red diamond = mean, blue cross = expected adapter length, points = samples",
         x = "Adapter",
         y = "Trim length (bp)") +
    theme_minimal()
  
  print(p_trim_length)
  
  # Export plot
  ggsave(
    filename = file.path(plots_dir, "trim_length_distribution.png"),
    plot = p_trim_length,
    width = 10,
    height = 6,
    dpi = 300,
    bg = "white"
  )
}

message("All QC plots exported to: ", plots_dir)
```

---

# Summary

## Final read counts

```{r final_summary, echo=FALSE}
final_summary <- data.frame(
  sample = names(seq_trim),
  n_reads = sapply(seq_trim, length),
  mean_length = sapply(seq_trim, function(x) mean(width(x))),
  stringsAsFactors = FALSE
)

DT::datatable(
  final_summary,
  caption = "Final trimmed sequences",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(final_summary), fontSize = '85%') %>%
  DT::formatRound(columns = 'mean_length', digits = 1)

write_excel_function(named_list = list("final_summary" = final_summary),
                     path = file.path(qc_dir, "qc.xlsx"),
                     overwrite_sheet = TRUE)
```

---

# Save output

```{r save_output, echo=FALSE}
# Compile module output
module1_output <- list(
  sequences = seq_trim,
  file_details = file_details,
  qc_summary = sequencing_qc_summary,
  trim_summary = count_trimmed,
  trim_by_adapter = trim_by_adapter,
  trim_by_sample = trim_by_sample,
  final_summary = final_summary
)

# Create results directory
results_dir <- file.path(params$dir_out, "module_data")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)

# Save complete module output
output_path <- file.path(results_dir, "01_import_qc_results.RData")
save(module1_output, file = output_path)
```

**Module 1 Complete!**

Results saved to: `r output_path`

- **`r length(seq_trim)`** samples processed
- **`r format(sum(final_summary$n_reads), big.mark = ",")`** total reads
- Mean reads per sample: **`r format(round(mean(final_summary$n_reads)), big.mark = ",")`**

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```

