---
title: "Duke Pipeline - Module 3: Repeat Detection"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

```{r load_packages, message=FALSE, warning=FALSE, echo=FALSE}
# Required packages
required_packages <- c("Biostrings", "pbapply", "pbmcapply", "dplyr", "tidyr", 
                       "data.table", "stringr", "ggplot2", "kableExtra", "DT", 
                       "openxlsx", "ggridges", "scales")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r load_functions, echo=FALSE}
# Source all shared functions
source("lib/load_all.R")
```

```{r create_directories, echo=FALSE}
# Create output directories (derived from params$dir_out)
temp_dir <- file.path(params$dir_out, "temp")
results_dir <- file.path(params$dir_out, "module_data")
repeat_dir <- file.path(params$dir_out, "03_repeat_detection", "repeat_detection")
plots_dir <- file.path(params$dir_out, file.path("03_repeat_detection", "plots"))  # NEW: plots directory

if (!dir.exists(temp_dir)) dir.create(temp_dir, recursive = TRUE)
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
if (!dir.exists(repeat_dir)) dir.create(repeat_dir, recursive = TRUE)
if (!dir.exists(plots_dir)) dir.create(plots_dir, recursive = TRUE)  # NEW
```

```{r set_param_defaults, echo=FALSE}
# Set default values for optional parameters
if (is.null(params$repeat_count_round_digits)) params$repeat_count_round_digits <- 0
if (is.null(params$histogram_binwidth)) params$histogram_binwidth <- 1
if (is.null(params$plot_detection_rate)) params$plot_detection_rate <- TRUE
if (is.null(params$plot_histograms)) params$plot_histograms <- TRUE
if (is.null(params$repeat_count_method)) params$repeat_count_method <- "repeat_count_full"
if (is.null(params$na_repeat_handling)) params$na_repeat_handling <- "convert_to_zero"
```

# Module 3: Repeat Detection

This module identifies and counts repeat tracts in the mid sequence.

**Purpose:**

1. Find repeat tracts matching the target pattern (e.g., CAG)
2. Validate repeat detection rates
3. Count repeats using robust length-based method
4. Quality control and basic visualisation

**Input:** `02_alignment_results.RData` from Module 2  
**Output:** `03_repeat_detection_results.RData`

---

# Load previous results

```{r load_results, echo=FALSE}
# Load Module 2 results
module2_path <- file.path(results_dir, "02_alignment_results.RData")

if (!file.exists(module2_path)) {
  stop("Module 2 results not found. Please run Module 2 first.")
}

load(module2_path)

# Extract alignment_split
alignment_split <- module2_output$alignment_split

message("Loaded Module 2 results:")
message("  Reads: ", format(nrow(alignment_split), big.mark = ","))
message("  Samples: ", n_distinct(alignment_split$file_stem))
```

---

# Find repeat tracts

Search for repeat tracts in the mid sequence (between left and right flanks).

**Parameters:**

| Parameter | Value |
|-----------|-------|
| Pattern | `r params$rpt_pattern` |
| Min repeats | `r params$rpt_min_repeats` |
| Max mismatch | `r params$rpt_max_mismatch` |
| Perfect at start | `r params$rpt_start_perfect_repeats` |
| Perfect at end | `r params$rpt_end_perfect_repeats` |
| Max gap between repeats | `r params$rpt_max_gap` |
| Return option | `r params$rpt_return_option` |

```{r find_repeat_tracts, echo=FALSE}
message("Finding repeat tracts in mid sequences...")

# Check for overall cached results first
tract_cache_path <- file.path(temp_dir, "03_repeat_tracts.RData")

if (params$resume && file.exists(tract_cache_path)) {
  message("Loading cached repeat tract results...")
  load(tract_cache_path)
  
} else {
  # No overall cache - process with apply_fn (parallelisation + per-sample caching)
  message("Processing repeat tracts with parallelisation...")
  
  # Get sample names
  samples <- unique(alignment_split$file_stem)
  file_stem_vector <- setNames(samples, samples)
  
  # Apply find_repeat_tracts_sample to each sample using apply_fn
  repeat_tracts_list <- apply_fn(
    file_stem_vector,
    find_repeat_tracts_sample,
    alignment_split = alignment_split,
    sequence_column = "mid",
    pattern = params$rpt_pattern,
    min_repeats = params$rpt_min_repeats,
    max_mismatch = params$rpt_max_mismatch,
    start_perfect_repeats = params$rpt_start_perfect_repeats,
    end_perfect_repeats = params$rpt_end_perfect_repeats,
    max_gap = params$rpt_max_gap,
    max_tract_gap = params$rpt_max_tract_gap,
    return_option = params$rpt_return_option,
    out_dir = temp_dir,
    resume = params$resume
  )
  
  # Extract results
  repeat_tracts_list <- extract_apply_fn_result(repeat_tracts_list)
  
  # Combine all samples
  alignment_repeats <- bind_rows(repeat_tracts_list)
  
  # Save overall cache
  save(alignment_repeats, file = tract_cache_path)
  message("Saved overall cache: ", tract_cache_path)
}

message("Repeat tract detection complete!")
message("  Total reads processed: ", format(nrow(alignment_repeats), big.mark = ","))
```

## Detection rates

```{r repeat_tract_summary, echo=FALSE}
# Summarise repeat tract detection
tract_summary <- alignment_repeats %>%
  group_by(file_stem) %>%
  dplyr::summarise(
    total_reads = n(),
    reads_with_tract = sum(!is.na(mid_pattern_start)),
    reads_without_tract = sum(is.na(mid_pattern_start)),
    detection_rate = mean(!is.na(mid_pattern_start)),
    .groups = "drop"
  )

DT::datatable(
  tract_summary,
  caption = "Repeat tract detection by sample",
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(tract_summary), fontSize = '85%') %>%
  DT::formatRound(columns = 'detection_rate', digits = 3)
```

```{r plot_detection_rate, echo=FALSE, fig.height=6, fig.width=10, eval=params$plot_detection_rate}
if (params$plot_detection_rate) {
  p_detection <- ggplot(tract_summary, aes(x = file_stem, y = detection_rate)) +
    geom_bar(stat = "identity", fill = "#619CFF") +
    geom_text(aes(label = scales::percent(detection_rate, accuracy = 0.1)),
              vjust = -0.5, size = 3) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1.05)) +
    scale_x_discrete(guide = guide_axis(angle = 45)) +
    labs(title = "Repeat tract detection rate per sample",
         subtitle = paste0("Pattern: ", params$rpt_pattern),
         x = "Sample",
         y = "Detection rate") +
    theme_minimal()
  
  print(p_detection)
  
  # Export plot
  # Calculate width (capped at 50 inches)
  n_samples <- length(unique(tract_summary$file_stem))
  plot_width <- min(50, max(8, n_samples * 0.8))
  
  ggsave(
    filename = file.path(plots_dir, "repeat_detection_rate.png"),
    plot = p_detection,
    width = plot_width,
    height = 6,
    dpi = 300,
    bg = "white",
    limitsize = FALSE
  )
}
```

---

# Reposition segments

After finding the repeat tract, reposition segments around it:

- **left** = left + mid.1 (sequence before tract)
- **mid** = mid.2 (the tract itself)  
- **right** = mid.3 + right (sequence after tract)

This isolates the repeat-containing region for accurate counting.

```{r reposition_segments, echo=FALSE}
# Reposition segments around the identified repeat tract
# Following original duke.Rmd logic
alignment_repositioned <- alignment_repeats %>%
  dplyr::mutate(
    # Append mid.1 (pre-tract sequence) to left flank
    left = paste0(na_to_blank(left), na_to_blank(mid.1)),
    
    # Prepend mid.3 (post-tract sequence) to right flank
    right = paste0(na_to_blank(mid.3), na_to_blank(right))
  ) %>%
  # Remove mid.1 and mid.3 (now incorporated into flanks)
  dplyr::select(-mid.1, -mid.3) %>%
  # Rename mid.2 to mid (this is now just the repeat tract)
  dplyr::rename(mid = mid.2) %>%
  # Remove pattern position columns (keep in final output if needed for QC)
  dplyr::select(-mid_pattern_start, -mid_pattern_end, -mid_pattern_count)

message("Segments repositioned around repeat tracts")
```

---

# Count repeats

Calculate repeat counts using multiple methods. The pipeline provides three counting approaches, with the primary method selected via the `repeat_count_method` parameter.

## Counting Methods

**1. repeat_count_full** (Default - Most robust)
- Formula: Tract length ÷ pattern length (rounded to nearest integer)
- Best for: Oxford Nanopore data with sequencing errors
- Example: 63 bp CAG tract ÷ 3 bp = 21 repeats

**2. repeat_count_match** (Stringent)
- Formula: Count only perfect pattern matches
- Best for: High-accuracy data (PacBio HiFi, Illumina)
- Example: Requires exact CAG CAG CAG... (no interruptions)

**3. repeat_count_tracts** (Structural analysis)
- Formula: Reports tract structure as "3,2,2" (sizes separated by gaps)
- Use: QC visualization and interruption analysis
- Note: Returns string, not used as primary count

## Current Method

This analysis uses: **`r params$repeat_count_method`**

All three methods are calculated and available in the output data for comparison and quality control.

```{r count_repeats, echo=FALSE}
message("Counting repeats...")

# Apply repeat counting (vectorised for speed)
alignment_counts <- apply_count_repeats(
  alignment_info = alignment_repositioned,
  string_column = "mid",
  pattern = params$rpt_pattern,
  round_digits = params$repeat_count_round_digits
)

# Use the selected counting method as primary
# (Options: "repeat_count_full", "repeat_count_match")
alignment_counts <- alignment_counts %>%
  dplyr::mutate(repeat_count = !!sym(params$repeat_count_method))

message("Repeat counting complete! Using method: ", params$repeat_count_method)

# Handle reads where no repeat tract was found (repeat_count = NA)
# These could be: sequencing artifacts, genuine deletions (CRISPR), or off-target reads
n_na <- sum(is.na(alignment_counts$repeat_count))

if (n_na > 0) {
  na_rate <- n_na / nrow(alignment_counts)
  message("Found ", format(n_na, big.mark = ","), 
          " reads where no repeat tract was detected (", 
          round(100 * na_rate, 1), "%)")
  
  # Add QC flag for tracking (always useful)
  alignment_counts <- alignment_counts %>%
    dplyr::mutate(no_tract_found = is.na(repeat_count))
  
  # Handle based on user parameter
  if (params$na_repeat_handling == "convert_to_zero") {
    # OPTION 1: Convert NA to 0 (for CRISPR deletion detection)
    alignment_counts <- alignment_counts %>%
      dplyr::mutate(
        repeat_count_full = tidyr::replace_na(repeat_count_full, 0),
        repeat_count_match = tidyr::replace_na(repeat_count_match, 0),
        repeat_count = tidyr::replace_na(repeat_count, 0)
      )
    message("→ Converted NA to 0 (treating as: no repeat tract present)")
    message("  Note: Includes sequencing artifacts AND genuine deletions")
    
  } else if (params$na_repeat_handling == "filter") {
    # OPTION 2: Remove reads with NA (conservative, removes artifacts)
    n_before <- nrow(alignment_counts)
    alignment_counts <- alignment_counts %>%
      dplyr::filter(!no_tract_found)
    n_after <- nrow(alignment_counts)
    message("→ Filtered out ", format(n_before - n_after, big.mark = ","), 
            " reads with NA repeat counts")
    message("  Note: Removes artifacts but also genuine deletions")
    
  } else if (params$na_repeat_handling == "flag_only") {
    # OPTION 3: Keep NA as-is, just flag for awareness
    message("→ Keeping NA repeat counts as-is (flagged with 'no_tract_found' column)")
    message("  Note: Downstream analyses will need to handle NAs")
    message("  You can filter using: alignment_counts %>% filter(!no_tract_found)")
  }
} else {
  # No NA values found - still add flag column for consistency
  alignment_counts <- alignment_counts %>%
    dplyr::mutate(no_tract_found = FALSE)
  message("All reads have valid repeat tracts detected")
}

message("Final read count: ", format(nrow(alignment_counts), big.mark = ","))
```

---

# Visualisation

Basic visualisations of repeat count distributions.

## Repeat count histograms

```{r plot_histograms, echo=FALSE, fig.height=6, fig.width=12, eval=params$plot_histograms}
if (params$plot_histograms) {
  # Combined faceted histogram
  p_histogram_facet <- ggplot(alignment_counts, aes(x = repeat_count)) +
    geom_histogram(binwidth = params$histogram_binwidth, 
                   fill = "#619CFF", color = "black", alpha = 0.7) +
    facet_wrap(~ file_stem, scales = "free_y") +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
    labs(title = "Repeat count distribution per sample",
         subtitle = "Length-based counting method (tract length ÷ pattern length)",
         x = "Repeat count",
         y = "Number of reads") +
    theme_minimal() +
    theme(strip.text = element_text(face = "bold"))
  
  print(p_histogram_facet)
  
  # Save faceted plot
  n_samples <- length(unique(alignment_counts$file_stem))
  ncol_facet <- ceiling(sqrt(n_samples))
  nrow_facet <- ceiling(n_samples / ncol_facet)
  
  ggsave(
    filename = file.path(plots_dir, "repeat_histogram_all_samples.png"),
    plot = p_histogram_facet,
    width = max(12, ncol_facet * 4),
    height = max(6, nrow_facet * 3),
    dpi = 300,
    bg = "white"
  )
  
  # Individual histograms per sample
  for (sample in unique(alignment_counts$file_stem)) {
    sample_data <- alignment_counts %>% filter(file_stem == sample)
    
    p_histogram_sample <- ggplot(sample_data, aes(x = repeat_count)) +
      geom_histogram(binwidth = params$histogram_binwidth,
                     fill = "#619CFF", color = "black", alpha = 0.7) +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
      labs(title = paste("Repeat count distribution -", sample),
           x = "Repeat count",
           y = "Number of reads") +
      theme_minimal()
    
    ggsave(
      filename = file.path(plots_dir, paste0("repeat_histogram_", sample, ".png")),
      plot = p_histogram_sample,
      width = 10,
      height = 6,
      dpi = 300,
      bg = "white"
    )
  }
}
```

## Violin plot

## Distribution visualisation

**Important:** Samples are typically **bimodal** (two alleles with different repeat counts). 
Mean ± SD is misleading for bimodal distributions - it falls between the peaks and doesn't represent either allele!

```{r plot_distributions, echo=FALSE, fig.height=8, fig.width=12}
# Violin plot shows the actual distribution shape (bimodal structure visible)
p_violin <- ggplot(alignment_counts, aes(x = file_stem, y = repeat_count)) +
  geom_violin(fill = "#619CFF", alpha = 0.6, scale = "width") +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.8) +
  stat_summary(fun = median, geom = "point", size = 3, color = "red") +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 15)) +
  labs(title = "Repeat count distribution per sample",
       subtitle = "Violin plot shows distribution shape (bimodal for diploid samples)\nBoxplot shows quartiles, red dot = median",
       x = "Sample",
       y = "Repeat count") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10))

print(p_violin)

# Calculate width (capped at 50 inches)
n_samples <- length(unique(alignment_counts$file_stem))
plot_width <- min(50, max(10, n_samples * 0.8))

ggsave(
  filename = file.path(plots_dir, "repeat_violin_all_samples.png"),
  plot = p_violin,
  width = plot_width,
  height = 8,
  dpi = 300,
  bg = "white",
  limitsize = FALSE
)

# Individual violin plots per sample
for (sample in unique(alignment_counts$file_stem)) {
  sample_data <- alignment_counts %>% filter(file_stem == sample)
  
  p_violin_sample <- ggplot(sample_data, aes(x = 1, y = repeat_count)) +
    geom_violin(fill = "#619CFF", alpha = 0.6, scale = "width", width = 0.8) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8, fill = "white") +
    stat_summary(fun = median, geom = "point", size = 4, color = "red") +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 15)) +
    labs(title = paste("Repeat count distribution -", sample),
         subtitle = "Violin shows distribution shape (bimodal for diploid), boxplot shows quartiles, red dot = median",
         x = "",
         y = "Repeat count") +
    theme_minimal() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())
  
  ggsave(
    filename = file.path(plots_dir, paste0("repeat_violin_", sample, ".png")),
    plot = p_violin_sample,
    width = 6,
    height = 8,
    dpi = 300,
    bg = "white"
  )
}
```

```{r plot_ridges, echo=FALSE, fig.height=10, fig.width=12, message=FALSE, warning=FALSE}
# Ridge plot - excellent for comparing distributions across samples
if (requireNamespace("ggridges", quietly = TRUE)) {
  library(ggridges)
  
  p_ridge_all <- ggplot(alignment_counts, aes(x = repeat_count, y = file_stem, fill = file_stem)) +
    geom_density_ridges(alpha = 0.7, scale = 2) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
    labs(title = "Repeat count distributions across samples",
         subtitle = "Each ridge shows the full distribution - peaks indicate alleles",
         x = "Repeat count",
         y = "Sample") +
    theme_minimal() +
    theme(legend.position = "none")
  
  print(p_ridge_all)
  
  # Save combined ridge plot
  # Calculate height (capped at 50 inches)
  n_samples <- length(unique(alignment_counts$file_stem))
  plot_height <- min(50, max(8, n_samples * 0.6))
  
  ggsave(
    filename = file.path(plots_dir, "repeat_ridge_all_samples.png"),
    plot = p_ridge_all,
    width = 12,
    height = plot_height,
    dpi = 300,
    bg = "white",
    limitsize = FALSE
  )
  
  # Individual density plots per sample
  for (sample in unique(alignment_counts$file_stem)) {
    sample_data <- alignment_counts %>% filter(file_stem == sample)
    
    p_density_sample <- ggplot(sample_data, aes(x = repeat_count)) +
      geom_density(fill = "#619CFF", alpha = 0.7, color = "black", linewidth = 1) +
      scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
      labs(title = paste("Repeat count density -", sample),
           subtitle = "Peaks indicate alleles",
           x = "Repeat count",
           y = "Density") +
      theme_minimal()
    
    ggsave(
      filename = file.path(plots_dir, paste0("repeat_density_", sample, ".png")),
      plot = p_density_sample,
      width = 10,
      height = 6,
      dpi = 300,
      bg = "white"
    )
  }
  
  message("Repeat detection plots exported to: ", plots_dir)
  
} else {
  message("Install 'ggridges' package for ridge plot: install.packages('ggridges')")
}
```

```{r summary_table_clean, echo=FALSE}
# Simple summary statistics (leave allele detection for Module 4 clustering)
clean_summary <- alignment_counts %>%
  dplyr::group_by(file_stem) %>%
  dplyr::summarise(
    n_reads = n(),
    median_repeats = median(repeat_count, na.rm = TRUE),
    mean_repeats = mean(repeat_count, na.rm = TRUE),
    mode_repeats = as.numeric(names(sort(table(repeat_count), decreasing = TRUE)[1])),
    min_repeats = min(repeat_count, na.rm = TRUE),
    max_repeats = max(repeat_count, na.rm = TRUE),
    range_repeats = max_repeats - min_repeats,
    sd_repeats = sd(repeat_count, na.rm = TRUE),
    .groups = "drop"
  )

DT::datatable(
  clean_summary,
  caption = "Repeat count summary by sample",
  colnames = c("Sample", "N reads", "Median", "Mean", "Mode", "Min", "Max", "Range", "SD"),
  options = list(
    pageLength = 25,
    scrollX = TRUE,
    scrollY = "500px",
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ),
  rownames = FALSE,
  filter = 'top',
  class = 'cell-border stripe compact',
  style = 'bootstrap',
  fillContainer = FALSE
) %>%
  DT::formatStyle(columns = colnames(clean_summary), fontSize = '85%') %>%
  DT::formatRound(columns = c('median_repeats', 'mean_repeats', 'mode_repeats', 
                               'min_repeats', 'max_repeats', 'range_repeats', 'sd_repeats'), 
                  digits = 1)
```

**Note:** For bimodal samples (visible in plots above), mean and median fall between peaks.
Use the **violin/ridge plots** to visually identify alleles. Formal allele calling is performed in Module 4 (clustering).

---

# Save outputs

Save processed data for downstream modules.

```{r save_outputs, echo=FALSE}
# Compile module output
module3_output <- list(
  alignment_counts = alignment_counts,
  count_summary = clean_summary,
  tract_summary = tract_summary,
  parameters = params
)

# Export summaries to Excel
repeat_excel_path <- file.path(repeat_dir, "repeat_summaries.xlsx")
write_excel_function(
  named_list = list(
    "Count Summary" = clean_summary,
    "Tract Detection" = tract_summary
  ),
  path = repeat_excel_path
)
message("Repeat summaries exported to: ", repeat_excel_path)

# Save complete module output
output_path <- file.path(results_dir, "03_repeat_detection_results.RData")
save(module3_output, file = output_path)

message("Module 3 outputs saved to: ", output_path)
```

```{r cleanup, echo=FALSE}
# Clean up if requested
if (params$remove_intermediate) {
  rm(alignment_split, alignment_repeats, alignment_repositioned)
  gc()
}
```

**Module 3 Complete!**

Results saved to: `r output_path`

- **`r format(nrow(alignment_counts), big.mark = ",")`** reads with repeat counts
- **`r n_distinct(alignment_counts$file_stem)`** samples analysed
- Mean repeat count across samples: **`r round(mean(clean_summary$mean_repeats), 1)`**

---

**Next Steps:**

- **Module 4:** Cluster reads by repeat length to identify alleles
- **Module 5:** Analyze distributions and calculate instability metrics
- **Module 6 (Optional):** Generate waterfall visualisations

---

# Session info

```{r session_info, echo=FALSE}
sessionInfo()
```
