---
title: "Duke Pipeline - Module 4: Allele Calling"
author: "Michael Flower"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

```{r load_packages, message=FALSE, warning=FALSE, echo=FALSE}
# Required packages
required_packages <- c("Biostrings", "DECIPHER", "mclust", "cluster", 
                      "pbapply", "dplyr", "tidyr", "data.table", 
                      "stringr", "ggplot2", "kableExtra", "openxlsx", "DT",
                      "pbmcapply")

# Load packages
invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r load_functions, echo=FALSE}
# Source all shared functions
source("lib/load_all.R")
```

# Module 4: Allele Calling

This module performs **clustering** to assign reads to distinct alleles, generates **consensus sequences**, and **calls variants** against the reference.

## Overview

**Input:** Repeat count data from Module 3

**Processing:**

1. Cluster reads by repeat length, haplotype, or both
2. Assign cluster numbers ordered by mean repeat length
3. Generate consensus sequences for each cluster
4. Call variants in consensus sequences (if reference provided)
5. Calculate cluster-specific summary statistics

**Output:** Read-level cluster assignments, consensus sequences, and variants

**Key Parameters:**

- `cluster_by`: Clustering strategy - "repeat", "haplotype", or "both"
- `haplotype_cluster_max`: Maximum haplotype clusters to detect (auto-detect)
- `repeat_cluster_max`: Maximum repeat-length clusters to detect (auto-detect)
- `cluster_consensus`: Whether to generate consensus sequences
- `call_variants`: Whether to call variants against reference

---

```{r setup_directories, echo=FALSE}
# Derive directory paths from params$dir_out
results_dir <- file.path(params$dir_out, "module_data")
temp_dir <- file.path(params$dir_out, "temp")

# Module 4 output directories
consensus_dir <- file.path(params$dir_out, file.path("04_allele_calling", "consensus"))
variants_dir <- file.path(params$dir_out, file.path("04_allele_calling", "variants"))

# Create directories if needed
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(temp_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(consensus_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(variants_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r load_data, echo=FALSE, results='asis'}
message("Loading Module 3 output...")

# Load Module 3 results
module3_path <- file.path(results_dir, "03_repeat_detection_results.RData")
if (!file.exists(module3_path)) {
  stop("Module 3 output not found: ", module3_path, 
       "\nPlease run Module 3 first!")
}

load(module3_path)
alignment_counts <- module3_output$alignment_counts

message("Loaded ", format(nrow(alignment_counts), big.mark = ","), " reads from Module 3")

# Format sample list for HTML
samples <- unique(alignment_counts$file_stem)
n_samples <- length(samples)
if (n_samples <= 10) {
  cat("\n\n**", n_samples, " samples:**\n\n", 
      paste(samples, collapse = ", "), "\n\n", sep="")
} else {
  cat("\n\n**", n_samples, " samples:**\n\n", 
      paste(head(samples, 10), collapse = ", "), 
      ", ... and ", n_samples - 10, " more\n\n", sep="")
}

# Load reference sequence from Module 2 if variant calling is enabled
reference_seq <- NULL
if (params$call_variants) {
  module2_path <- file.path(results_dir, "02_alignment_results.RData")
  if (file.exists(module2_path)) {
    load(module2_path)
    if (!is.null(module2_output$ref_full_sequence)) {
      reference_seq <- module2_output$ref_full_sequence
      message("Loaded reference sequence from Module 2: ", nchar(reference_seq), " bp")
    } else {
      message("Warning: Module 2 output does not contain full reference sequence")
      message("Variant calling will be skipped")
    }
  } else {
    message("Warning: Module 2 output not found")
    message("Variant calling will be skipped")
  }
}
```

---

# Clustering

Assign reads to clusters (alleles) using the strategy: **`r params$cluster_by`**

```{r clustering, echo=FALSE}
# Check resume
cache_path <- file.path(temp_dir, "04_clustering.RData")

if (file.exists(cache_path) && params$resume) {
  message("Loading cached clustering results...")
  load(cache_path)
  message("Clustering loaded from cache")
  
} else {
  
  if (params$cluster) {
    
    message("Performing clustering...")
    message("Strategy: ", params$cluster_by)
    if (params$cluster_by %in% c("haplotype", "both")) {
      message("  Haplotype region: ", params$haplotype_region)
      message("  Haplotype method: ", params$haplotype_method)
    }
    
    # Split by sample for per-sample clustering
    alignment_list <- split(alignment_counts, alignment_counts$file_stem)
    n_samples <- length(alignment_list)
    
    message("\nClustering ", n_samples, " samples...")
    
    # Apply clustering to each sample using apply_fn (parallelised if threads > 1)
    alignment_clustered_list <- apply_fn(seq_along(alignment_list), function(i) {
      sample_data <- alignment_list[[i]]
      sample_name <- unique(sample_data$file_stem)
      
      # Check for cached result
      sample_cache_file <- file.path(temp_dir, paste0("cluster_", sample_name, ".RData"))
      if (params$resume && file.exists(sample_cache_file)) {
        load(sample_cache_file)  # Loads 'sample_clustered'
        return(sample_clustered)
      }
      
      # Apply clustering using the wrapper function
      cluster_result <- apply_clustering(
        alignment_data = sample_data,
        cluster_by = params$cluster_by,
        repeat_column = params$repeat_count_method,
        sequence_column = "query_sequence",
        left_column = "left",
        right_column = "right",
        haplotype_cluster_max = params$haplotype_cluster_max,
        repeat_cluster_max = params$repeat_cluster_max,
        haplotype_method = params$haplotype_method,
        haplotype_region = params$haplotype_region,
        haplotype_trim_length = params$haplotype_trim_length,
        haplotype_subsample = params$haplotype_subsample
      )
      
      # Handle result - can be vector or list with metadata
      if (is.list(cluster_result) && !is.null(cluster_result$cluster_number)) {
        # "both" strategy returns metadata
        sample_data$cluster_number <- cluster_result$cluster_number
        sample_data$repeat_cluster <- cluster_result$repeat_cluster
        sample_data$haplotype_cluster <- cluster_result$haplotype_cluster
      } else {
        # Simple strategies return vector
        sample_data$cluster_number <- cluster_result
        sample_data$repeat_cluster <- NA_character_
        sample_data$haplotype_cluster <- NA_character_
      }
      
      # Save per-sample cache
      sample_clustered <- sample_data
      save(sample_clustered, file = sample_cache_file)
      
      return(sample_clustered)
    })
    
    # Extract results (follows pattern from modules 1-3)
    alignment_clustered_list <- extract_apply_fn_result(alignment_clustered_list)
    
    # Recombine into single data frame
    alignment_clustered <- bind_rows(alignment_clustered_list)
    
    # Relocate cluster_number column
    alignment_clustered <- alignment_clustered %>%
      relocate(cluster_number, .after = all_of(params$repeat_count_method))
    
    # Summary statistics
    total_clusters <- alignment_clustered %>%
      group_by(file_stem) %>%
      summarise(n_clusters = length(unique(cluster_number[!is.na(cluster_number)])),
                .groups = "drop")
    
    message("Clustering complete!")
    message("  Total samples: ", n_samples)
    message("  Total clusters: ", sum(total_clusters$n_clusters))
    message("  Clusters per sample: ", 
            sprintf("%.1f ± %.1f (range: %d-%d)",
                    mean(total_clusters$n_clusters),
                    sd(total_clusters$n_clusters),
                    min(total_clusters$n_clusters),
                    max(total_clusters$n_clusters)))
    
  } else {
    
    # No clustering - assign all reads to cluster 1
    message("Clustering disabled - assigning all reads to cluster 1")
    alignment_clustered <- alignment_counts %>%
      mutate(cluster_number = 1) %>%
      relocate(cluster_number, .after = all_of(params$repeat_count_method))
    
  }
  
  # Cache results
  save(alignment_clustered, file = cache_path)
  message("Clustering results cached to: ", cache_path)
}
```

**Clustering Strategy Explanation:**

```{r clustering_explanation, echo=FALSE, results='asis'}
# Handle both vector and single value inputs
if (length(params$cluster_by) == 2) {
  # Two-step clustering
  cat("**Combined Clustering:**\n\n")
  cat("- Clustering order: ", paste(params$cluster_by, collapse = " → "), "\n")
  cat("- Step 1: Group by ", params$cluster_by[1], "\n")
  cat("- Step 2: Within each group, cluster by ", params$cluster_by[2], "\n")
  cat("- Captures both sequence variants AND repeat length differences\n")
  if (identical(params$cluster_by, c("repeat", "haplotype"))) {
    cat("- **Repeat-first**: More robust to sequencing errors (recommended for long reads)\n")
  } else {
    cat("- **Haplotype-first**: Traditional approach\n")
  }
} else if (length(params$cluster_by) == 1) {
  # Single method clustering
  if (params$cluster_by == "repeat") {
    cat("**Repeat Length Clustering:**\n\n")
    cat("- Reads grouped solely by repeat count\n")
    cat("- Standard approach for detecting alleles by size\n")
    cat("- Best when alleles differ primarily in repeat length\n")
  } else if (params$cluster_by == "haplotype") {
    cat("**Haplotype Clustering:**\n\n")
    cat("- Reads grouped by variants in flanking regions (", params$haplotype_region, ")\n")
    cat("- Uses ", params$haplotype_method, " distance to measure similarity\n")
    cat("- Best for detecting alleles that differ by SNPs/indels, not just repeat length\n")
  } else if (params$cluster_by == "none") {
    cat("**No Clustering:**\n\n")
    cat("- All reads combined into single consensus\n")
    cat("- Use for homozygous samples or when clustering not needed\n")
  }
}
```

---

## Cluster Summary

Statistics for each cluster per sample.

```{r cluster_summary, echo=FALSE}
# Calculate cluster summary statistics
cluster_summary <- alignment_clustered %>%
  group_by(file_stem, cluster_number) %>%
  summarise(
    read_count = n(),
    repeat_cluster = dplyr::first(repeat_cluster),      # Use dplyr::first() explicitly
    haplotype_cluster = dplyr::first(haplotype_cluster), # Use dplyr::first() explicitly
    repeat_count_mode = {
      # Mode: most common value (best for discrete integer data)
      counts <- !!sym(params$repeat_count_method)
      counts <- counts[!is.na(counts)]
      if (length(counts) == 0) {
        NA_real_
      } else {
        ux <- unique(counts)
        tab <- tabulate(match(counts, ux))
        ux[which.max(tab)]
      }
    },
    repeat_count_median = median(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_mean = mean(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_sd = sd(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_min = min(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_max = max(!!sym(params$repeat_count_method), na.rm = TRUE),
    repeat_count_iqr = IQR(!!sym(params$repeat_count_method), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(file_stem, cluster_number)

# Renumber clusters by read count within each sample (cluster 1 = most abundant)
cluster_summary <- cluster_summary %>%
  group_by(file_stem) %>%
  arrange(file_stem, desc(read_count)) %>%
  mutate(
    cluster_number_original = cluster_number,  # Keep original for reference
    cluster_number = row_number(),             # Renumber: 1 = most abundant
    read_proportion = read_count / sum(read_count)
  ) %>%
  ungroup() %>%
  relocate(cluster_number_original, .after = cluster_number) %>%
  relocate(read_proportion, .after = read_count)

# Also renumber in alignment_clustered to match
cluster_mapping <- cluster_summary %>%
  select(file_stem, cluster_number_original, cluster_number)

alignment_clustered <- alignment_clustered %>%
  left_join(
    cluster_mapping %>% rename(cluster_number_old = cluster_number_original, 
                                cluster_number_new = cluster_number),
    by = c("file_stem", "cluster_number" = "cluster_number_old")
  ) %>%
  select(-cluster_number) %>%
  rename(cluster_number = cluster_number_new) %>%
  relocate(cluster_number, .after = all_of(params$repeat_count_method))

message("Clusters renumbered by abundance (cluster 1 = most reads)")
message("Updated cluster numbers in both cluster_summary and alignment_clustered")

# Display table with DT::datatable for better handling of large tables
datatable(cluster_summary,
          options = list(
            pageLength = 15,
            autoWidth = TRUE,
            scrollX = TRUE,
            initComplete = JS("function(settings, json) {",
                              "$(this.api().table().header()).css({'font-size': '10px', 'font-family': 'Arial'});",
                              "}")
          ),
          rownames = FALSE,
          class = 'cell-border stripe',
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-size: 12px; font-weight: bold;',
            "Cluster Summary (cluster 1 = most abundant, R = repeat cluster, H = haplotype cluster)"
          ),
          colnames = c("Sample", "Cluster", "Original", "Reads", "Proportion",
                       "R", "H",
                       "Mode", "Median", "Mean", "SD", "Min", "Max", "IQR")
) %>%
  formatStyle(columns = names(cluster_summary), fontSize = '10px', fontFamily = 'Arial') %>%
  formatStyle(columns = 0, fontSize = '10px', fontFamily = 'Arial') %>%
  formatRound(columns = c('read_proportion', 'repeat_count_median', 'repeat_count_mean', 
                          'repeat_count_sd'), digits = 3) %>%
  formatRound(columns = c('repeat_count_mode', 'repeat_count_min', 'repeat_count_max', 
                          'repeat_count_iqr'), digits = 1)
```

**Interpretation:**

- **Cluster numbers renumbered** by abundance (cluster 1 = most reads)

- **Original cluster number** kept for reference

- **R (Repeat cluster)** and **H (Haplotype cluster)** show sub-clustering:
  - Same R value = same repeat length cluster
  - Same H value = same haplotype cluster
  - Example: R1-H1 and R1-H2 have same repeat but different haplotypes
  - Useful for diagnosing why many clusters were detected

- **Proportion** shows percentage of total reads per sample

- **Mode** is the most common repeat count (robust to outliers)

- For **heterozygous diploid** samples, expect 2 clusters with ~0.5 proportion each

- For **homozygous** samples, expect 1 dominant cluster with ~1.0 proportion

- **Read count imbalance** may indicate allele dropout or amplification bias

---

## Cluster Visualisation

```{r cluster_plots, echo=FALSE, fig.height=8, fig.width=12, warning=FALSE}
# Generate colors for any number of clusters
n_clusters <- length(unique(alignment_clustered$cluster_number[!is.na(alignment_clustered$cluster_number)]))
if (n_clusters <= 12) {
  cluster_colors <- RColorBrewer::brewer.pal(max(3, n_clusters), "Set3")
} else {
  # Use viridis for many clusters
  cluster_colors <- viridisLite::viridis(n_clusters, option = "turbo")
}

# Create plots directory
plots_dir <- file.path(params$dir_out, file.path("04_allele_calling", "plots"))
dir.create(plots_dir, showWarnings = FALSE, recursive = TRUE)

# Determine dynamic plot dimensions based on number of samples
n_samples <- length(unique(alignment_clustered$file_stem))
n_clusters_max <- max(table(cluster_summary$file_stem))

# Define plotting strategy based on sample count
# Small (≤30): All plots
# Medium (31-100): Combined plot limited, all individual plots
# Large (101-200): Skip combined plot, generate individual plots with message
# Very large (>200): Skip combined plot, generate only summary statistics
if (n_samples <= 30) {
  plot_strategy <- "all"
  message("Generating all plots (", n_samples, " samples)")
} else if (n_samples <= 100) {
  plot_strategy <- "limited_combined"
  message("Note: ", n_samples, " samples detected.")
  message("  Combined violin plot will be crowded but generated for overview.")
  message("  Individual plots per sample will be clearer for detailed analysis.")
} else if (n_samples <= 200) {
  plot_strategy <- "individual_only"
  message("Note: ", n_samples, " samples detected (>100).")
  message("  Skipping combined violin plot (would be unreadable).")
  message("  Generating individual plots per sample for detailed analysis.")
} else {
  plot_strategy <- "summary_only"
  message("WARNING: ", n_samples, " samples detected (>200).")
  message("  Skipping individual violin plots (would generate >200 files).")
  message("  Generating summary plots only (scatter, ridge).")
  message("  For detailed per-sample plots, filter to a subset of samples.")
}

# Calculate violin plot dimensions (3 columns of facets)
violin_ncol <- 3
violin_nrow <- ceiling(n_samples / violin_ncol)
# Cap height at 30 inches (enough for ~10 rows) to avoid graphics device limits
violin_height <- min(30, max(6, violin_nrow * 3))  # Between 6-30 inches
violin_width <- 12

# Calculate scatter plot dimensions
# Cap width at 40 inches to avoid graphics device limits
scatter_width <- min(40, max(8, n_samples * 0.8))  # Between 8-40 inches
scatter_height <- 6

# Violin plot colored by cluster - faceted by sample for clarity
# Add read count indicators (circle size) at bottom
cluster_counts <- alignment_clustered %>%
  group_by(file_stem, cluster_number) %>%
  summarise(n = n(), .groups = "drop")

# Get y-axis range for positioning
y_min <- min(alignment_clustered[[params$repeat_count_method]], na.rm = TRUE)
y_max <- max(alignment_clustered[[params$repeat_count_method]], na.rm = TRUE)
y_indicator <- y_min - (y_max - y_min) * 0.1  # Position below data

alignment_with_counts <- alignment_clustered %>%
  left_join(cluster_counts, by = c("file_stem", "cluster_number"))

# Only generate combined plot if strategy allows (≤200 samples)
if (plot_strategy %in% c("all", "limited_combined", "individual_only")) {
  
  # Create combined violin plot
  p_violin_combined <- ggplot(alignment_with_counts, aes(x = factor(cluster_number), 
                                y = !!sym(params$repeat_count_method), 
                                fill = factor(cluster_number))) +
  geom_violin(alpha = 0.7, scale = "width") +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9, fill = "white") +
  # Add read count indicators (circles scaled to n)
  geom_point(data = cluster_counts, 
             aes(x = factor(cluster_number), y = y_indicator, size = n),
             shape = 21, fill = "grey30", color = "black", alpha = 0.8, 
             inherit.aes = FALSE) +
  geom_text(data = cluster_counts,
            aes(x = factor(cluster_number), y = y_indicator, label = n),
            size = 2.5, color = "black", fontface = "bold",
            angle = 90, vjust = 0.5,
            inherit.aes = FALSE) +
  facet_wrap(~ file_stem, scales = "free_x", ncol = violin_ncol) +
  scale_fill_manual(values = cluster_colors, name = "Cluster") +
  scale_size_continuous(name = "Read count", range = c(3, 15)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 15)) +
  labs(title = "Repeat count distribution by cluster",
       subtitle = paste("Clustering strategy:", params$cluster_by, 
                       "- Each panel is one sample. Circle size at bottom = read count"),
       x = "Cluster",
       y = paste0("Repeat count (", params$repeat_count_method, ")")) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
        strip.text = element_text(face = "bold"),
        legend.position = "right")

# Display combined plot
print(p_violin_combined)

# Save combined violin plot
ggsave(
  filename = file.path(plots_dir, "violin.png"),
  plot = p_violin_combined,
  width = violin_width,
  height = violin_height,
  dpi = 300,
  bg = "white"
)

} else {
  message("Skipped combined violin plot (>200 samples)")
}

# Create subdirectory for per-sample plots
violin_by_sample_dir <- file.path(plots_dir, "violin_by_sample")
dir.create(violin_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

# Create and save individual violin plots per sample
# Skip if >200 samples (too many files)
if (plot_strategy != "summary_only") {
  
  for (sample in unique(alignment_clustered$file_stem)) {
    sample_data <- alignment_with_counts %>% filter(file_stem == sample)
    sample_counts <- cluster_counts %>% filter(file_stem == sample)
    
    p_violin_sample <- ggplot(sample_data, aes(x = factor(cluster_number), 
                                               y = !!sym(params$repeat_count_method), 
                                               fill = factor(cluster_number))) +
      geom_violin(alpha = 0.7, scale = "width") +
      geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.9, fill = "white") +
      geom_point(data = sample_counts, 
                 aes(x = factor(cluster_number), y = y_indicator, size = n),
                 shape = 21, fill = "grey30", color = "black", alpha = 0.8, 
                 inherit.aes = FALSE) +
      geom_text(data = sample_counts,
                aes(x = factor(cluster_number), y = y_indicator, label = n),
                size = 3, color = "black", fontface = "bold",
                angle = 90, vjust = 0.5,
                inherit.aes = FALSE) +
      scale_fill_manual(values = cluster_colors, name = "Cluster") +
      scale_size_continuous(name = "Read count", range = c(4, 15)) +
      labs(title = paste("Repeat count distribution -", sample),
           x = "Cluster",
           y = paste0("Repeat count (", params$repeat_count_method, ")")) +
      theme_minimal() +
      theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
            legend.position = "right")
    
    ggsave(
      filename = file.path(violin_by_sample_dir, paste0(sample, "-violin.png")),
      plot = p_violin_sample,
      width = 10,
      height = 6,
      dpi = 300,
      bg = "white"
    )
  }
  
  message("Violin plots exported to: ", violin_by_sample_dir)
  
} else {
  message("Skipped individual violin plots (>200 samples - would generate ", n_samples, " files)")
  message("  For per-sample plots, filter your data to a subset of samples before Module 4")
}
```

```{r cluster_scatter, echo=FALSE, fig.height=8, fig.width=12, warning=FALSE}
# Scatter plot with position dodge to separate overlapping error bars
# Color error bars by cluster to match points

# Add position dodge to separate overlapping points/error bars
dodge_width <- 0.6

p_scatter_combined <- ggplot(cluster_summary, aes(x = file_stem, y = repeat_count_mean, 
                             fill = factor(cluster_number), 
                             color = factor(cluster_number),  # Add color for error bars
                             size = read_count,
                             group = cluster_number)) +  # Group by cluster for dodging
  geom_errorbar(aes(ymin = repeat_count_mean - repeat_count_sd,
                    ymax = repeat_count_mean + repeat_count_sd),
                width = 0.3, alpha = 0.6, linewidth = 0.8,
                position = position_dodge(width = dodge_width)) +  # Dodge error bars
  geom_point(alpha = 0.8, shape = 21, stroke = 0.8,
             position = position_dodge(width = dodge_width)) +  # Dodge points
  geom_text(aes(label = read_count), size = 2.5, color = "black", fontface = "bold",
            position = position_dodge(width = dodge_width)) +  # Dodge labels
  scale_fill_manual(values = cluster_colors, name = "Cluster") +
  scale_color_manual(values = cluster_colors, guide = "none") +  # Match error bar colors, hide legend
  scale_size_continuous(name = "Read count", range = c(5, 20), 
                       breaks = c(100, 1000, 5000, 10000)) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 15)) +
  labs(title = "Cluster centres by sample",
       subtitle = "Error bars colored by cluster and dodged to reduce overlap. Circle size = read count, error bars = ± 1 SD",
       x = "Sample",
       y = paste0("Mean repeat count (", params$repeat_count_method, ")")) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10),
        legend.position = "right")

# Display combined plot
print(p_scatter_combined)

# Save combined scatter plot with dynamic width
ggsave(
  filename = file.path(plots_dir, "scatter.png"),
  plot = p_scatter_combined,
  width = scatter_width,
  height = scatter_height,
  dpi = 300,
  bg = "white"
)

# Create subdirectory for per-sample plots
scatter_by_sample_dir <- file.path(plots_dir, "scatter_by_sample")
dir.create(scatter_by_sample_dir, showWarnings = FALSE, recursive = TRUE)

# Create and save individual scatter plots per sample
for (sample in unique(cluster_summary$file_stem)) {
  sample_data <- cluster_summary %>% filter(file_stem == sample)
  
  p_scatter_sample <- ggplot(sample_data, aes(x = factor(cluster_number), y = repeat_count_mean, 
                               fill = factor(cluster_number), size = read_count)) +
    geom_errorbar(aes(ymin = repeat_count_mean - repeat_count_sd,
                      ymax = repeat_count_mean + repeat_count_sd),
                  width = 0.3, alpha = 0.6, linewidth = 0.8,
                  color = "grey40") +
    geom_point(alpha = 0.8, shape = 21, color = "black", stroke = 0.8) +
    geom_text(aes(label = read_count), size = 3, color = "black", fontface = "bold") +
    scale_fill_manual(values = cluster_colors, name = "Cluster") +
    scale_size_continuous(name = "Read count", range = c(6, 20)) +
    labs(title = paste("Cluster centers -", sample),
         x = "Cluster",
         y = paste0("Mean repeat count (", params$repeat_count_method, ")")) +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 11),
          legend.position = "right")
  
  ggsave(
    filename = file.path(scatter_by_sample_dir, paste0(sample, "-scatter.png")),
    plot = p_scatter_sample,
    width = max(8, nrow(sample_data) * 0.6),
    height = 6,
    dpi = 300,
    bg = "white"
  )
}

message("Scatter plots exported to: ", plots_dir)
```

---

# Consensus Generation

Generate consensus sequences for each cluster.

```{r consensus_generation, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Suppress DECIPHER progress bars
options(verbose = FALSE)
old_verbose <- getOption("verbose")
on.exit(options(verbose = old_verbose))

# Check resume
consensus_cache_path <- file.path(temp_dir, "04_consensus.RData")

if (file.exists(consensus_cache_path) && params$resume) {
  message("Loading cached consensus results...")
  load(consensus_cache_path)
  message("Consensus sequences loaded from cache")
  
} else {
  
  if (params$cluster_consensus) {
    
    message("Generating consensus sequences...")
    
    # Create composite key for splitting (more efficient than nest())
    alignment_clustered$cluster_key <- paste0(alignment_clustered$file_stem, 
                                              "_cluster", 
                                              alignment_clustered$cluster_number)
    
    # Split by cluster key (much faster than nest() for large datasets)
    cluster_list <- split(alignment_clustered, alignment_clustered$cluster_key)
    
    # Create lookup for sample names and cluster numbers
    cluster_info <- alignment_clustered %>%
      select(cluster_key, file_stem, cluster_number) %>%
      distinct()
    
    message("\nGenerating consensus sequences for ", length(cluster_list), " clusters...")
    
    # Generate consensus for each cluster using apply_fn
    consensus_results_list <- apply_fn(seq_along(cluster_list), function(i) {
      
      cluster_data <- cluster_list[[i]]
      cluster_key <- names(cluster_list)[i]
      
      # Extract sample name and cluster number from first row
      sample_name <- cluster_data$file_stem[1]
      cluster_num <- cluster_data$cluster_number[1]
      
      # Check for cached result
      cluster_cache_file <- file.path(temp_dir, 
                                      paste0("consensus_", sample_name, "_cluster", cluster_num, ".RData"))
      if (params$resume && file.exists(cluster_cache_file)) {
        load(cluster_cache_file)  # Loads 'cluster_consensus'
        return(cluster_consensus)
      }
      
      # Get sequences for this cluster
      sequences <- cluster_data$query_sequence
      
      # Skip if no sequences
      if (length(sequences) == 0 || all(is.na(sequences))) {
        cluster_consensus <- list(
          file_stem = sample_name,
          cluster_number = cluster_num,
          consensus_full = NA_character_,
          consensus_left = NA_character_,
          consensus_mid = NA_character_,
          consensus_right = NA_character_,
          read_count = 0
        )
        save(cluster_consensus, file = cluster_cache_file)
        return(cluster_consensus)
      }
      
      # Create DNAStringSet
      valid_seqs <- sequences[!is.na(sequences) & sequences != ""]
      if (length(valid_seqs) == 0) {
        cluster_consensus <- list(
          file_stem = sample_name,
          cluster_number = cluster_num,
          consensus_full = NA_character_,
          consensus_left = NA_character_,
          consensus_mid = NA_character_,
          consensus_right = NA_character_,
          read_count = 0
        )
        save(cluster_consensus, file = cluster_cache_file)
        return(cluster_consensus)
      }
      
      dna_stringset <- DNAStringSet(valid_seqs)
      
      # Generate consensus (suppress DECIPHER progress bars)
      sink(tempfile())  # Redirect output
      consensus_result <- tryCatch({
        consensus_function(
          dna_stringset = dna_stringset,
          threads = params$threads,
          consensus_threshold = params$consensus_threshold,
          sample_size = params$consensus_downsample
        )
      }, finally = {
        sink()  # Restore output
      })
      
      # Also generate consensus for segments (suppress DECIPHER output)
      sink(tempfile())
      consensus_left <- tryCatch({
        if (!all(is.na(cluster_data$left))) {
          left_seqs <- cluster_data$left[!is.na(cluster_data$left) & cluster_data$left != ""]
          if (length(left_seqs) > 0) {
            consensus_function(DNAStringSet(left_seqs), 
                             threads = params$threads,
                             sample_size = params$consensus_downsample)$consensus
          } else NA_character_
        } else NA_character_
      }, finally = {})
      
      consensus_mid <- tryCatch({
        if (!all(is.na(cluster_data$mid))) {
          mid_seqs <- cluster_data$mid[!is.na(cluster_data$mid) & cluster_data$mid != ""]
          if (length(mid_seqs) > 0) {
            consensus_function(DNAStringSet(mid_seqs),
                             threads = params$threads,
                             sample_size = params$consensus_downsample)$consensus
          } else NA_character_
        } else NA_character_
      }, finally = {})
      
      consensus_right <- tryCatch({
        if (!all(is.na(cluster_data$right))) {
          right_seqs <- cluster_data$right[!is.na(cluster_data$right) & cluster_data$right != ""]
          if (length(right_seqs) > 0) {
            consensus_function(DNAStringSet(right_seqs),
                             threads = params$threads,
                             sample_size = params$consensus_downsample)$consensus
          } else NA_character_
        } else NA_character_
      }, finally = {
        sink()  # Restore output after all segments done
      })
      
      # Save per-cluster cache
      cluster_consensus <- list(
        file_stem = sample_name,
        cluster_number = cluster_num,
        consensus_full = consensus_result$consensus,
        consensus_left = consensus_left,
        consensus_mid = consensus_mid,
        consensus_right = consensus_right,
        read_count = length(valid_seqs)
      )
      save(cluster_consensus, file = cluster_cache_file)
      
      return(cluster_consensus)
    })
    
    # Extract results (follows pattern from modules 1-3)
    consensus_results_list <- extract_apply_fn_result(consensus_results_list)
    
    # Convert to data frame
    consensus_summary <- bind_rows(consensus_results_list)
    
    # Merge with cluster_summary to get repeat count statistics
    consensus_summary <- consensus_summary %>%
      left_join(cluster_summary %>% select(file_stem, cluster_number, repeat_count_mode, repeat_count_median, repeat_count_mean),
                by = c("file_stem", "cluster_number"))
    
    message("Consensus generation complete!")
    message("Generated consensus sequences for ", nrow(consensus_summary), " clusters")
    
  } else {
    
    message("Consensus generation disabled")
    consensus_summary <- NULL
    
  }
  
  # Cache results
  save(consensus_summary, file = consensus_cache_path)
  message("Consensus results cached to: ", consensus_cache_path)
}
```

```{r consensus_table, echo=FALSE}
# Display consensus summary
if (exists("consensus_summary") && !is.null(consensus_summary)) {
  
  # Create display table with sequence lengths
  consensus_display <- consensus_summary %>%
    mutate(
      full_length = nchar(consensus_full),
      mid_length = nchar(consensus_mid),
      left_length = nchar(consensus_left),
      right_length = nchar(consensus_right)
    ) %>%
    select(file_stem, cluster_number, read_count, full_length, mid_length, left_length, right_length)
  
  # Export to Excel
  consensus_excel_path <- file.path(consensus_dir, "consensus_summary.xlsx")
  write_excel_function(
    named_list = list("Consensus Summary" = consensus_display),
    path = consensus_excel_path
  )
  message("Consensus summary exported to: ", consensus_excel_path)
  
  # Display consensus summary table
  datatable(consensus_display,
            options = list(
              pageLength = 15,
              autoWidth = TRUE,
              scrollX = TRUE,
              initComplete = JS("function(settings, json) {",
                                "$(this.api().table().header()).css({'font-size': '10px', 'font-family': 'Arial'});",
                                "}")
            ),
            rownames = FALSE,
            class = 'cell-border stripe',
            caption = htmltools::tags$caption(
              style = 'caption-side: top; text-align: left; font-size: 12px; font-weight: bold;',
              "Consensus Sequence Summary"
            ),
            colnames = c("Sample", "Cluster", "Reads Used", "Full Length", "Mid Length", "Left Length", "Right Length")
  ) %>%
    formatStyle(columns = names(consensus_display), fontSize = '10px', fontFamily = 'Arial') %>%
    formatStyle(columns = 0, fontSize = '10px', fontFamily = 'Arial')
}
```

---

# Variant Calling

```{r variant_calling, echo=FALSE}
# Cache variant calling results
variant_cache_path <- file.path(temp_dir, "04_variants.RData")

if (file.exists(variant_cache_path) && params$resume) {
  message("Loading cached variant calling results...")
  load(variant_cache_path)
  message("Variant results loaded from cache")
  
} else {
  
  if (params$call_variants && exists("reference_seq") && !is.null(reference_seq) && 
      exists("consensus_summary") && !is.null(consensus_summary)) {
    
    message("Calling variants in consensus sequences...")
    
    # Call variants for all consensus sequences
    variant_results <- call_variants_for_clusters(consensus_summary, reference_seq)
    
    if (!is.null(variant_results) && nrow(variant_results) > 0) {
      
      message("Found ", nrow(variant_results), " variants across all clusters")
      
      # Summary of variants
      variant_summary_table <- summarise_variants(variant_results)
      
    } else {
      message("No variants found (all consensus sequences match reference)")
      variant_results <- NULL
      variant_summary_table <- NULL
    }
    
  } else {
    if (!params$call_variants) {
      message("Variant calling disabled")
    } else if (is.null(reference_seq)) {
      message("Variant calling skipped (no reference sequence provided)")
    }
    variant_results <- NULL
    variant_summary_table <- NULL
  }
  
  # Cache results
  save(variant_results, variant_summary_table, file = variant_cache_path)
  message("Variant results cached to: ", variant_cache_path)
}
```

```{r variant_table, echo=FALSE}
if (exists("variant_results") && !is.null(variant_results) && nrow(variant_results) > 0) {
  
  # Export summaries to Excel
  variant_excel_path <- file.path(variants_dir, "variant_summary.xlsx")
    write_excel_function(
      named_list = list(
        "Summary" = variant_summary_table,
        "All Variants" = variant_results
      ),
      path = variant_excel_path
    )
    message("Variant summaries exported to: ", variant_excel_path)
    
    # Display variant summary
    if (!is.null(variant_summary_table)) {
      # Get actual column names
      col_names <- names(variant_summary_table)
      
      # Create display names (capitalize first letter)
      display_names <- sapply(col_names, function(x) {
        if (x == "file_stem") return("Sample")
        if (x == "cluster_number") return("Cluster")
        # Capitalize first letter for variant types
        paste0(toupper(substring(x, 1, 1)), substring(x, 2))
      }, USE.NAMES = FALSE)
      
      datatable(variant_summary_table,
                options = list(
                  pageLength = 15,
                  autoWidth = TRUE,
                  scrollX = TRUE,
                  initComplete = JS("function(settings, json) {",
                                    "$(this.api().table().header()).css({'font-size': '10px', 'font-family': 'Arial'});",
                                    "}")
                ),
                rownames = FALSE,
                class = 'cell-border stripe',
                caption = htmltools::tags$caption(
                  style = 'caption-side: top; text-align: left; font-size: 12px; font-weight: bold;',
                  "Variant Summary by Cluster"
                ),
                colnames = display_names
      ) %>%
        formatStyle(columns = col_names, fontSize = '10px', fontFamily = 'Arial') %>%
        formatStyle(columns = 0, fontSize = '10px', fontFamily = 'Arial')
    }
}
```

---

# Export Files

```{r export_consensus, echo=FALSE}
# Check if consensus was generated
if (exists("consensus_summary") && !is.null(consensus_summary) && nrow(consensus_summary) > 0) {
  
  # Export full consensus sequences
  consensus_fasta_path <- file.path(consensus_dir, "consensus_sequences.fasta")
  
  # Calculate proportion of reads per cluster (within each sample)
  consensus_with_props <- consensus_summary %>%
    group_by(file_stem) %>%
    mutate(
      total_reads = sum(read_count),
      proportion = read_count / total_reads
    ) %>%
    ungroup()
  
  # Calculate dynamic padding for cluster numbers
  # Ensures proper alphabetical sorting even with >999 clusters
  max_cluster <- max(consensus_with_props$cluster_number, na.rm = TRUE)
  n_digits <- nchar(as.character(max_cluster))
  format_string <- paste0("%0", n_digits, "d")
  
  # Create informative FASTA names with read counts, mode repeat, and proportions
  # Format: {sample}_cluster{N}_n{count}_mode{repeat}_prop{proportion}
  # Mode is best for discrete integer data (most common value, robust to outliers)
  # Cluster number padding adjusts to max clusters (e.g., 01-99, 001-999, 0001-9999)
  consensus_names <- paste0(
    consensus_with_props$file_stem, 
    "_cluster", sprintf(format_string, consensus_with_props$cluster_number),
    "_n", consensus_with_props$read_count,
    "_mode", round(consensus_with_props$repeat_count_mode, 0),
    "_prop", sprintf("%.3f", consensus_with_props$proportion)
  )
  consensus_seqs <- setNames(consensus_with_props$consensus_full, consensus_names)
  
  # Remove NA sequences
  consensus_seqs <- consensus_seqs[!is.na(consensus_seqs)]
  
  if (length(consensus_seqs) > 0) {
    write_fasta(consensus_seqs, sequence_type = "DNA", output_path = consensus_fasta_path)
    message("Consensus sequences exported to: ", consensus_fasta_path)
    message("  Cluster number format: ", n_digits, " digits (max cluster: ", max_cluster, ")")
  }
  
  # Also export mid sequences (repeat regions only)
  mid_fasta_path <- file.path(consensus_dir, "consensus_mid_sequences.fasta")
  mid_names <- paste0(
    consensus_with_props$file_stem,
    "_cluster", sprintf(format_string, consensus_with_props$cluster_number),
    "_n", consensus_with_props$read_count,
    "_mode", round(consensus_with_props$repeat_count_mode, 0),
    "_prop", sprintf("%.3f", consensus_with_props$proportion),
    "_mid"
  )
  mid_seqs <- setNames(consensus_with_props$consensus_mid, mid_names)
  mid_seqs <- mid_seqs[!is.na(mid_seqs)]
  
  if (length(mid_seqs) > 0) {
    write_fasta(mid_seqs, sequence_type = "DNA", output_path = mid_fasta_path)
    message("Mid consensus sequences exported to: ", mid_fasta_path)
  }
  
  # Export left flank sequences (independently generated consensus)
  left_fasta_path <- file.path(consensus_dir, "consensus_left_sequences.fasta")
  left_names <- paste0(
    consensus_with_props$file_stem, 
    "_cluster", sprintf(format_string, consensus_with_props$cluster_number),
    "_n", consensus_with_props$read_count,
    "_mode", round(consensus_with_props$repeat_count_mode, 0),
    "_prop", sprintf("%.3f", consensus_with_props$proportion),
    "_left"
  )
  left_seqs <- setNames(consensus_with_props$consensus_left, left_names)
  left_seqs <- left_seqs[!is.na(left_seqs)]
  
  if (length(left_seqs) > 0) {
    write_fasta(left_seqs, sequence_type = "DNA", output_path = left_fasta_path)
    message("Left consensus sequences exported to: ", left_fasta_path)
  }
  
  # Export right flank sequences (independently generated consensus)
  right_fasta_path <- file.path(consensus_dir, "consensus_right_sequences.fasta")
  right_names <- paste0(
    consensus_with_props$file_stem, 
    "_cluster", sprintf(format_string, consensus_with_props$cluster_number),
    "_n", consensus_with_props$read_count,
    "_mode", round(consensus_with_props$repeat_count_mode, 0),
    "_prop", sprintf("%.3f", consensus_with_props$proportion),
    "_right"
  )
  right_seqs <- setNames(consensus_with_props$consensus_right, right_names)
  right_seqs <- right_seqs[!is.na(right_seqs)]
  
  if (length(right_seqs) > 0) {
    write_fasta(right_seqs, sequence_type = "DNA", output_path = right_fasta_path)
    message("Right consensus sequences exported to: ", right_fasta_path)
  }
}
```

```{r export_variants, echo=FALSE}
if (exists("variant_results") && !is.null(variant_results) && nrow(variant_results) > 0) {
  
  # Prepare variant export with clear column names
  # Check if position column exists, if not it might already be named position_on_reference
  if ("position" %in% names(variant_results)) {
    variant_export <- variant_results %>%
      dplyr::rename(position_on_reference = position) %>%
      dplyr::select(file_stem, cluster_number, position_on_reference, ref, query, variant_type)
  } else {
    # Column might already be renamed or have different name
    variant_export <- variant_results %>%
      dplyr::select(file_stem, cluster_number, position, ref, query, variant_type) %>%
      dplyr::rename(position_on_reference = position)
  }
  
  # Export variants to CSV
  variants_csv_path <- file.path(variants_dir, "variants.csv")
  write.csv(variant_export, variants_csv_path, row.names = FALSE)
  message("Variants exported to: ", variants_csv_path)
  
  # Export variants to VCF format (for IGV, etc.)
  # Filter out variants with N bases (ambiguous) and variants at position 1 (no anchor)
  # Create one VCF per sample
  samples <- unique(variant_export$file_stem)
  
  # Read reference sequence for anchor bases
  ref_seq <- as.character(reference_seq)
  ref_name <- sub("^>", "", readLines(params$path_ref, n = 1))
  
  # Filter variants for VCF: remove N bases and position 1
  vcf_valid_variants <- variant_export %>%
    filter(position_on_reference > 1) %>%
    filter(ref != "N" & query != "N" & ref != "-" & query != "-")
  
  # Individual VCFs: one column per cluster
  message("Exporting individual VCF files...")
  
  # Create subdirectory for individual sample VCF files
  vcf_by_sample_dir <- file.path(variants_dir, "vcf_by_sample")
  dir.create(vcf_by_sample_dir, showWarnings = FALSE, recursive = TRUE)
  
  vcf_export_count <- 0
  vcf_total_variants <- 0
  
  for (sample in samples) {
    sample_variants <- vcf_valid_variants %>%
      filter(file_stem == sample)
    
    if (nrow(sample_variants) > 0) {
      vcf_path <- file.path(vcf_by_sample_dir, paste0(sample, "_variants.vcf"))
      
      # Get all clusters for this sample
      clusters <- sort(unique(sample_variants$cluster_number))
      cluster_names <- paste0("cluster", clusters)
      
      # VCF header
      vcf_header <- c(
        "##fileformat=VCFv4.2",
        paste0("##fileDate=", format(Sys.Date(), "%Y%m%d")),
        "##source=Duke_Pipeline_Module4",
        paste0("##reference=", basename(params$path_ref)),
        paste0("##contig=<ID=", ref_name, ",length=", nchar(ref_seq), ">"),
        "##INFO=<ID=TYPE,Number=1,Type=String,Description=\"Variant type: SNP, INS, or DEL\">",
        "##FORMAT=<ID=GT,Number=1,Type=String,Description=\"Genotype\">",
        paste0("#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO\tFORMAT\t", paste(cluster_names, collapse = "\t"))
      )
      
      # Group variants by position and create merged records
      position_groups <- sample_variants %>%
        group_by(position_on_reference) %>%
        group_split()
      
      vcf_records <- lapply(position_groups, function(pos_variants) {
        # Get position info from first variant at this position
        pos <- unique(pos_variants$position_on_reference)[1]
        
        # Determine VCF position and alleles
        # For each unique REF/ALT at this position, create one VCF line
        unique_variants <- pos_variants %>%
          distinct(ref, query, variant_type, .keep_all = TRUE)
        
        # Process each unique variant at this position
        lapply(1:nrow(unique_variants), function(i) {
          var <- unique_variants[i, ]
          ref_base <- as.character(var$ref)
          query_base <- as.character(var$query)
          var_type <- as.character(var$variant_type)
          
          # Convert to VCF format
          ref_vcf <- NULL
          alt_vcf <- NULL
          pos_vcf <- NULL
          
          if (var_type %in% c("SNP", "mismatch")) {
            ref_vcf <- ref_base
            alt_vcf <- query_base
            pos_vcf <- pos
          } else if (var_type == "insertion" && query_base != "-") {
            if (pos > 1) {
              anchor <- substr(ref_seq, pos - 1, pos - 1)
              ref_vcf <- anchor
              alt_vcf <- paste0(anchor, query_base)
              pos_vcf <- pos - 1
            }
          } else if (var_type == "deletion" && ref_base != "-") {
            if (pos > 1) {
              anchor <- substr(ref_seq, pos - 1, pos - 1)
              ref_vcf <- paste0(anchor, ref_base)
              alt_vcf <- anchor
              pos_vcf <- pos - 1
            }
          }
          
          if (is.null(ref_vcf) || is.null(alt_vcf) || is.null(pos_vcf)) {
            return(NULL)
          }
          
          # Which clusters have this exact variant?
          clusters_with_variant <- pos_variants %>%
            filter(ref == ref_base, query == query_base) %>%
            pull(cluster_number)
          
          # Generate genotypes for all clusters
          genotypes <- sapply(clusters, function(c) {
            if (c %in% clusters_with_variant) "0/1" else "0/0"
          })
          
          chrom <- ref_name
          id <- "."
          qual <- "."
          filter <- "PASS"
          info <- paste0("TYPE=", var_type)
          format <- "GT"
          
          paste(chrom, pos_vcf, id, ref_vcf, alt_vcf, qual, filter, info, format,
                paste(genotypes, collapse = "\t"), sep = "\t")
        })
      })
      
      # Flatten and remove NULLs
      vcf_records <- unlist(unlist(vcf_records, recursive = FALSE))
      vcf_records <- vcf_records[!sapply(vcf_records, is.null)]
      
      if (length(vcf_records) > 0) {
        writeLines(c(vcf_header, vcf_records), vcf_path)
        vcf_export_count <- vcf_export_count + 1
        vcf_total_variants <- vcf_total_variants + length(vcf_records)
      }
    }
  }
  
  # Summary message instead of per-sample messages
  message("VCF files exported: ", vcf_export_count, " samples")
  message("  Location: ", variants_dir)
  message("  Total variant positions: ", vcf_total_variants)
  
  # Create combined multi-sample VCF with sample_cluster columns
  combined_vcf_path <- file.path(variants_dir, "all_samples_variants.vcf")
  
  if (nrow(vcf_valid_variants) > 0) {
    # Get all sample_cluster combinations
    sample_clusters <- vcf_valid_variants %>%
      distinct(file_stem, cluster_number) %>%
      arrange(file_stem, cluster_number) %>%
      mutate(sample_cluster = paste0(file_stem, "_cluster", cluster_number))
    
    all_sample_cluster_names <- sample_clusters$sample_cluster
    
    # VCF header
    vcf_header_combined <- c(
      "##fileformat=VCFv4.2",
      paste0("##fileDate=", format(Sys.Date(), "%Y%m%d")),
      "##source=Duke_Pipeline_Module4",
      paste0("##reference=", basename(params$path_ref)),
      paste0("##contig=<ID=", ref_name, ",length=", nchar(ref_seq), ">"),
      "##INFO=<ID=TYPE,Number=1,Type=String,Description=\"Variant type: SNP, INS, or DEL\">",
      "##FORMAT=<ID=GT,Number=1,Type=String,Description=\"Genotype\">",
      paste0("#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO\tFORMAT\t", 
             paste(all_sample_cluster_names, collapse = "\t"))
    )
    
    # Group all variants by position
    position_groups <- vcf_valid_variants %>%
      group_by(position_on_reference) %>%
      group_split()
    
    vcf_records_combined <- lapply(position_groups, function(pos_variants) {
      pos <- unique(pos_variants$position_on_reference)[1]
      
      # Get unique variants at this position
      unique_variants <- pos_variants %>%
        distinct(ref, query, variant_type, .keep_all = TRUE)
      
      # Process each unique variant
      lapply(1:nrow(unique_variants), function(i) {
        var <- unique_variants[i, ]
        ref_base <- as.character(var$ref)
        query_base <- as.character(var$query)
        var_type <- as.character(var$variant_type)
        
        # Convert to VCF format
        ref_vcf <- NULL
        alt_vcf <- NULL
        pos_vcf <- NULL
        
        if (var_type %in% c("SNP", "mismatch")) {
          ref_vcf <- ref_base
          alt_vcf <- query_base
          pos_vcf <- pos
        } else if (var_type == "insertion" && query_base != "-") {
          if (pos > 1) {
            anchor <- substr(ref_seq, pos - 1, pos - 1)
            ref_vcf <- anchor
            alt_vcf <- paste0(anchor, query_base)
            pos_vcf <- pos - 1
          }
        } else if (var_type == "deletion" && ref_base != "-") {
          if (pos > 1) {
            anchor <- substr(ref_seq, pos - 1, pos - 1)
            ref_vcf <- paste0(anchor, ref_base)
            alt_vcf <- anchor
            pos_vcf <- pos - 1
          }
        }
        
        if (is.null(ref_vcf) || is.null(alt_vcf) || is.null(pos_vcf)) {
          return(NULL)
        }
        
        # Which sample_clusters have this variant?
        sample_clusters_with_variant <- pos_variants %>%
          filter(ref == ref_base, query == query_base) %>%
          mutate(sample_cluster = paste0(file_stem, "_cluster", cluster_number)) %>%
          pull(sample_cluster)
        
        # Generate genotypes for all sample_clusters
        genotypes <- sapply(all_sample_cluster_names, function(sc) {
          if (sc %in% sample_clusters_with_variant) "0/1" else "0/0"
        })
        
        chrom <- ref_name
        id <- "."
        qual <- "."
        filter <- "PASS"
        info <- paste0("TYPE=", var_type)
        format <- "GT"
        
        paste(chrom, pos_vcf, id, ref_vcf, alt_vcf, qual, filter, info, format,
              paste(genotypes, collapse = "\t"), sep = "\t")
      })
    })
    
    # Flatten and remove NULLs
    vcf_records_combined <- unlist(unlist(vcf_records_combined, recursive = FALSE))
    vcf_records_combined <- vcf_records_combined[!sapply(vcf_records_combined, is.null)]
    
    if (length(vcf_records_combined) > 0) {
      writeLines(c(vcf_header_combined, vcf_records_combined), combined_vcf_path)
      message("Combined VCF exported: ", combined_vcf_path,
              " (", length(vcf_records_combined), " variant positions, ",
              length(all_sample_cluster_names), " sample_cluster combinations)")
    }
  }
  
  message("Note: VCF files can be viewed in IGV. Load reference first, then VCF files.")
}
```

---

# Save Results

```{r save_outputs, echo=FALSE}
# Compile module output
module4_output <- list(
  alignment_clustered = alignment_clustered,
  cluster_summary = cluster_summary,
  consensus_summary = consensus_summary,
  variant_results = if(exists("variant_results")) variant_results else NULL,
  parameters = params
)

# Save complete module output
output_path <- file.path(results_dir, "04_allele_calling_results.RData")
save(module4_output, file = output_path)

message("Module 4 outputs saved to: ", output_path)
```

```{r cleanup, echo=FALSE}
# Clean up if requested
if (params$remove_intermediate) {
  rm(alignment_counts)
  gc()
}
```

**Module 4 Complete!**

Results saved to: `r output_path`

- **`r format(nrow(alignment_clustered), big.mark = ",")`** reads with cluster assignments
- **`r n_distinct(alignment_clustered$file_stem)`** samples analysed
- **`r if(!is.null(consensus_summary)) nrow(consensus_summary) else 0`** consensus sequences generated
- **`r if(exists("variant_results") && !is.null(variant_results)) nrow(variant_results) else 0`** variants called

---

**Session Info:**

```{r session_info, echo=FALSE}
sessionInfo()
```
