---
title: "Duke Pipeline - Module 7: Repeat Visualisation"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

---

# Overview

This module provides comprehensive visualisation of repeat length distributions through:

1. **Repeat Histograms**: Frequency distribution plots showing the count of reads at each repeat length
2. **Scatter/Violin Plots**: Individual read distributions with violin overlays

These plots complement the statistical analyses in Module 6 by showing individual reads and distribution shapes visually.

---

# Setup

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(scales)
library(DT)
library(ggnewscale)
library(data.table)

# Source visualization functions
source("lib/visualisation.R")

# Create output directory
plots_dir <- file.path(params$dir_out, file.path("07_repeat_visualisation", "plots"))
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}

message("\n=== Module 7: Repeat Distribution Visualisation ===")
message("Output directory: ", plots_dir)
```

---

# Load Data

```{r load_data}
# Load Module 4 output (clustered data with repeat counts)
results_dir <- file.path(params$dir_out, "module_data")
load(file.path(results_dir, "04_allele_calling_results.RData"))

# Load Module 6 output (modal peaks and distribution summary)
load(file.path(results_dir, "06_range_analysis_results.RData"))

# Extract data from Module 4
alignment_clustered <- module4_output$alignment_clustered

# Extract data from Module 6
modal_peaks <- module6_output$modal_peaks
distribution_summary <- module6_output$distribution_summary

# Get repeat count method
repeat_method <- params$repeat_count_method

message("\nData loaded:")
message("  Samples: ", length(unique(alignment_clustered$file_stem)))
message("  Total reads: ", nrow(alignment_clustered))
message("  Repeat method: ", repeat_method)
message("  Modal peaks: ", nrow(modal_peaks))
message("  Distribution summaries: ", nrow(distribution_summary))
```

---

# Prepare Data

```{r prepare_data}
# alignment_clustered is already a single dataframe with all reads from all samples
alignment_data <- alignment_clustered

# Get analysis ranges from modal_peaks
analysis_ranges <- modal_peaks %>%
  select(file_stem, analysis_range_name, range_min, range_max) %>%
  distinct()

# Note: alignment_data is used as-is from Module 4 (no pre-filtering by ranges)
# Reads are filtered dynamically in plotting loops by repeat length
# This allows reads to contribute to multiple overlapping ranges as intended

# Prepare summary metric points based on user selection
summary_points_list <- list()

# Get selected metrics
selected_metrics <- params$repeat_distribution_metrics

# Add modal if requested
if ("modal_length" %in% selected_metrics) {
  modal_points <- modal_peaks %>%
    filter(!is.na(modal_length), peak_rank == 1) %>%
    select(file_stem, analysis_range_name, modal_length, modal_frequency) %>%
    mutate(name = "modal",
           x = modal_length,
           y = modal_frequency) %>%
    select(file_stem, analysis_range_name, name, x, y)
  
  summary_points_list[["modal"]] <- modal_points
}

# Add mean if requested
if ("mean_length" %in% selected_metrics) {
  mean_points <- distribution_summary %>%
    select(file_stem, analysis_range_name, mean_length) %>%
    mutate(name = "mean",
           x = mean_length,
           y = 0) %>%  # Will be at bottom of histogram
    select(file_stem, analysis_range_name, name, x, y)
  
  summary_points_list[["mean"]] <- mean_points
}

# Add median if requested
if ("median_length" %in% selected_metrics) {
  median_points <- distribution_summary %>%
    select(file_stem, analysis_range_name, median_length) %>%
    mutate(name = "median",
           x = median_length,
           y = 0) %>%  # Will be at bottom of histogram
    select(file_stem, analysis_range_name, name, x, y)
  
  summary_points_list[["median"]] <- median_points
}

# Combine all selected points
if (length(summary_points_list) > 0) {
  summary_points <- bind_rows(summary_points_list) %>%
    rename(range = analysis_range_name)  # Rename for plotting function
} else {
  # Create empty dataframe with correct structure if no metrics selected
  summary_points <- data.frame(
    file_stem = character(),
    analysis_range_name = character(),
    name = character(),
    x = numeric(),
    y = numeric()
  )
}

# Prepare range boxes
range_boxes <- analysis_ranges %>%
  mutate(xmin = range_min,
         xmax = range_max) %>%
  rename(name = analysis_range_name)

message("\nData prepared:")
message("  Total reads: ", nrow(alignment_data))
message("  Analysis ranges: ", length(unique(analysis_ranges$analysis_range_name)))
message("  Metrics to display: ", paste(selected_metrics, collapse = ", "))
```

---

# Repeat Histograms

Histograms show the frequency distribution of repeat lengths with modal and mean markers.

```{r setup_range_colors}
# Define consistent colors for analysis ranges
range_names <- unique(analysis_ranges$analysis_range_name)
range_colors <- setNames(
  scales::hue_pal()(length(range_names)),
  range_names
)
```

## Histograms by Sample

```{r histograms_by_sample, fig.width=10, fig.height=6}
if (params$repeat_histogram) {
  
  # Create directory
  hist_by_sample_dir <- file.path(plots_dir, "histograms_by_sample")
  if (!dir.exists(hist_by_sample_dir)) {
    dir.create(hist_by_sample_dir, recursive = TRUE)
  }
  
  for (my_sample in unique(alignment_clustered$file_stem)) {
    
    # Get all data for this sample (full range)
    sample_data_full <- alignment_clustered %>%
      filter(file_stem == my_sample)
    
    if (nrow(sample_data_full) == 0) next
    
    # Get range boxes for this specific sample
    sample_boxes <- range_boxes %>%
      filter(file_stem == my_sample) %>%
      distinct(name, xmin, xmax)
    
    # Get points for this sample (all selected metrics already in summary_points)
    sample_points <- summary_points %>%
      filter(file_stem == my_sample)
    
    # 1. Create full range histogram (all data)
    p_full <- plot_frequency_histogram(
      data = sample_data_full,
      repeat_col = repeat_method,
      binwidth = params$repeat_histogram_binwidth,
      manual_points = sample_points,
      manual_boxes = sample_boxes,
      range_colors = range_colors,
      title = paste0("Repeat Distribution - ", my_sample, " (Full Range)"),
      base_text_size = 14
    )
    
    print(p_full)
    
    ggsave(file.path(hist_by_sample_dir, 
                    paste0(my_sample, "_histogram_full.png")),
           p_full, width = 10, height = 6, dpi = 300)
    
    # 2. Create per-range histograms (only for ranges this sample has)
    sample_ranges <- analysis_ranges %>%
      filter(file_stem == my_sample) %>%
      pull(analysis_range_name) %>%
      unique()
    
    for (my_range in sample_ranges) {
      
      # Get range boundaries for this sample and range
      range_info <- analysis_ranges %>%
        filter(file_stem == my_sample, analysis_range_name == my_range)
      
      if (nrow(range_info) == 0) next
      
      range_min <- range_info$range_min[1]
      range_max <- range_info$range_max[1]
      
      # Get data for this sample in this range
      # Filter by REPEAT LENGTH, not by pre-assigned column
      sample_range_data <- alignment_data %>%
        filter(
          file_stem == my_sample,
          .data[[repeat_method]] >= range_min,
          .data[[repeat_method]] <= range_max
        )
      
      if (nrow(sample_range_data) == 0) next
      
      # Get range box for this specific range and sample
      range_box <- range_boxes %>%
        filter(file_stem == my_sample, name == my_range) %>%
        distinct(name, xmin, xmax)
      
      # Get points for this sample and range
      # These come from Module 6 and already have analysis_range_name
      range_points <- sample_points %>%
        filter(range == my_range)
      
      # Create range-specific histogram
      p_range <- plot_frequency_histogram(
        data = sample_range_data,
        repeat_col = repeat_method,
        binwidth = params$repeat_histogram_binwidth,
        manual_points = range_points,
        manual_boxes = range_box,
        range_colors = range_colors,
        title = paste0("Repeat Distribution - ", my_sample, " - ", my_range),
        base_text_size = 14
      )
      
      print(p_range)
      
      ggsave(file.path(hist_by_sample_dir, 
                      paste0(my_sample, "_histogram_", my_range, ".png")),
             p_range, width = 10, height = 6, dpi = 300)
    }
  }
  
  message("Histograms by sample created: ", hist_by_sample_dir)
  message("  - Full range histograms: {sample}_histogram_full.png")
  message("  - Per-range histograms: {sample}_histogram_{range}.png")
}
```

---

# Repeat Scatter/Violin Plots

Scatter plots show individual reads as points with violin overlay showing distribution shape.

## Combined Vertical Scatter

```{r scatter_vertical_combined, fig.width=14, fig.height=10}
if (params$repeat_scatter) {
  
  # Use summary_points which already has the selected metrics
  # Reshape for scatter plot (needs repeat_method column name)
  combined_points <- summary_points %>%
    rename(!!sym(repeat_method) := x)
  
  # Create plot
  p_scatter_combined <- plot_scatter_vertical(
    data = alignment_data,
    x_var = "file_stem",
    y_var = repeat_method,
    points = combined_points,
    title = "Repeat Length Distribution - All Samples",
    base_text_size = 14
  )
  
  print(p_scatter_combined)
  
  ggsave(file.path(plots_dir, "scatter_combined.png"),
         p_scatter_combined, width = 14, height = 8, dpi = 300)
  
  message("\nCombined scatter plot created")
}
```

## Scatter Plots by Analysis Range

```{r scatter_by_range, fig.width=12, fig.height=10}
if (params$repeat_scatter) {
  
  # Create directory
  scatter_by_range_dir <- file.path(plots_dir, "scatter_by_range")
  if (!dir.exists(scatter_by_range_dir)) {
    dir.create(scatter_by_range_dir, recursive = TRUE)
  }
  
  for (my_range in unique(analysis_ranges$analysis_range_name)) {
    
    # Get range boundaries for this range
    range_info <- analysis_ranges %>%
      filter(analysis_range_name == my_range) %>%
      slice(1)  # Take first row if multiple samples have same range name
    
    if (nrow(range_info) == 0) next
    
    range_min <- range_info$range_min[1]
    range_max <- range_info$range_max[1]
    
    # Filter data by REPEAT LENGTH
    range_data <- alignment_data %>%
      filter(
        .data[[repeat_method]] >= range_min,
        .data[[repeat_method]] <= range_max
      )
    
    if (nrow(range_data) == 0) next
    
    # Get points for this range
    # These come from Module 6 and already have analysis_range_name
    range_points <- summary_points %>%
      filter(range == my_range) %>%
      rename(!!sym(repeat_method) := x)
    
    # Create plot
    p_range_scatter <- plot_scatter_vertical(
      data = range_data,
      x_var = "file_stem",
      y_var = repeat_method,
      points = range_points,
      title = paste0("Repeat Distribution - ", my_range),
      base_text_size = 14
    )
    
    print(p_range_scatter)
    
    ggsave(file.path(scatter_by_range_dir,
                    paste0("scatter_", my_range, ".png")),
           p_range_scatter, width = 12, height = 6, dpi = 300)
  }
  
  message("Scatter plots by range created: ", scatter_by_range_dir)
}
```

## Horizontal Scatter by Sample

```{r scatter_by_sample, fig.width=12, fig.height=4}
if (params$repeat_scatter) {
  
  # Create directory
  scatter_by_sample_dir <- file.path(plots_dir, "scatter_by_sample")
  if (!dir.exists(scatter_by_sample_dir)) {
    dir.create(scatter_by_sample_dir, recursive = TRUE)
  }
  
  for (my_sample in unique(alignment_data$file_stem)) {
    
    # Filter data
    sample_data <- alignment_data %>%
      filter(file_stem == my_sample)
    
    if (nrow(sample_data) == 0) next
    
    # Get range boxes for this specific sample
    sample_boxes <- range_boxes %>%
      filter(file_stem == my_sample) %>%
      distinct(name, xmin, xmax)
    
    # Get points for this sample (already filtered in summary_points)
    sample_points <- summary_points %>%
      filter(file_stem == my_sample)
    
    # Create plot
    p_sample_scatter <- plot_scatter_horizontal(
      data = sample_data,
      x_var = repeat_method,
      points = sample_points,
      boxes = sample_boxes,
      range_colors = range_colors,
      title = paste0("Repeat Distribution - ", my_sample),
      base_text_size = 14
    )
    
    print(p_sample_scatter)
    
    ggsave(file.path(scatter_by_sample_dir,
                    paste0(my_sample, "_scatter.png")),
           p_sample_scatter, width = 12, height = 4, dpi = 300)
  }
  
  message("Scatter plots by sample created: ", scatter_by_sample_dir)
}
```

---

# Summary

```{r summary}
message("\n=== Module 7 Complete ===")
message("Plots saved to: ", plots_dir)

if (params$repeat_histogram) {
  message("\nHistogram plots:")
  message("  - By sample: histograms_by_sample/")
  message("    - Full range: {sample}_histogram_full.png")
  message("    - Per range: {sample}_histogram_{range}.png")
}

if (params$repeat_scatter) {
  message("\nScatter plots:")
  message("  - Combined: scatter_combined.png")
  message("  - By range: scatter_by_range/")
  message("  - By sample: scatter_by_sample/")
}

cat("\n")
```

---

# Session Info

```{r session_info}
sessionInfo()
```
